### 2017-10-31 thesisProposal.Rmd
First commit

---
title: 'Thesis proposal: Longitudinal models of human mobility'
author: "Boaz Sobrado"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: tp2.tex

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
*Student Number:* 5891272
*Supervisor Name:* Peter Lugtig
*Word Count:* 676

## Research Question

How active people are and how they move around is correlated with health, income and social capital (Goodchild & Janelle, 2010). A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies. Yet studies on human mobility are largely based on pen-and-paper travel diaries. These surveys have known methodological flaws, such as the short period of data collection (due to costs and burden to respondents), the underreporting of short trips (Wolf et al., 2013) and the underestimation of the duration of commutes (Delclòs-Alió et al. 2017).

Longitudinal data on human movement has become available through smartphones. While there have been advances in combining and standardizing this information (Zheng, Zheng, Xie & Yang, 2010) extracting meaningful information remains a difficult task. As an illustration, *Figure 1* depicts a few hours in the life of the author with the raw measurements of an Android phone as well as smoothed measurements using a simple Kalman filter (commonly used by engineers for filtering noisy sensor measurements).

![Raw GPS measurements of a journey from de Uithof to Tuinwijk on February 17th 2017. The measurements are in red, the filtered path is in blue. The circles denote 67% confidence intervals of the given GPS measurement. Measurements and fitted points which follow each other in time are connected by lines. The inaccurate measurements lead to estimates of irregular movements. The filtered movement estimate is less volatile, but lags behind the unfiltered movement.](Map1.png)

Noisy data leads to inaccurate conclusions if not accounted for in analyses. The Kalman filter (and other state space models) improve these measurements somewhat, however they are limited by the fact that mobility is modeled on a Markov chain, which means that estimated location at timepoint $k$ is based only upon previous timepoint $k-1$. This assumption ignores common sense valuable information. For instance, most people spend the nights at home and the day at work.

This project seeks to develop a new model for understanding and quantifying human mobility over time. The main question of interest is: given what we know about an individual's travel history (as measured through his or her phone) how well can we estimate their precise location? In other words, how do we minimise measurement error and deal with missing data in smartphone GPS measurements?

## State of the Art

Research with respect to the analysis of GPS data collected from living creatures is wide ranging, highly interdisciplinary and often serves different purposes. For instance, there is a vast literature of state space models of animal movements used by biologists and ecologists (Patterson et al. 2007). These models are primarily Markovian and are used to extract information about how the animal is interacting with the environment. For instance, short bird flights over a certain location can be identified as foraging trips whilst more directed trips are movements between foraging habitats.

On the other hand, there is a body of literature from systems engineers which aim to identify how transport systems work as a whole. As of recently, these include machine learning methods such as feedback algorithms to solve the problem of travel mode classification , merging an individual's measured path with an actual path (known as *map matching*) and trip detection (e.g. Yang et al. 2016, Rassmussen et al. 2016). Due to the field's interest in the system rather than the individuals it often uninterested in individual level measurement error.

Perhaps the most illustrative example of the challenges researchers face when analysing smartphone GPS data comes from a recent paper which aims to detect bipolar depression from GPS data (Palmis et al. 2017). Despite sophisticated feature extraction measures the authors use sub-optimal filtering and imputation methods, such as mean imputation. In fact, most researchers using GPS data find the data pre-processing challenging (Harari et al. 2016).

## Approach

Using travel logs and diaries, I will evaluate currently used families of movement models used in contemporary research in their ability to impute missing data. These include:

1. State space models (such as the Kalman filter) which take measurement $k$, measurement accuracy, the previous measurement $k-1$ and time between measurements as parameters.

2. Long range-persistence models, such as cascading walks models (e.g. Han et al. 2015) and the FarOut model (Sadilek & Krumm, 2012) which rely on self-similarity and autoregressive characteristics.

Moreover, I will compare the performance of these models with that of traditional missing data imputation techniques, such as multiple imputation.

## Journals for Possible Publication

Journal of the Royal Statistical Society: Series A or C, Computers, Environment and Urban Systems,Geojournal

## References

Delclòs-Alió, X., Marquet, O., & Miralles-Guasch, C. (2017). Keeping track of time: A Smartphone-based analysis of travel time perception in a suburban environment. Travel Behaviour and Society, 9(Supplement C), 1–9. https://doi.org/10.1016/j.tbs.2017.07.001

D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, 5(1),144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf

Goodchild, M. F., & Janelle, D. G. (2010). Toward critical spatial thinking in the social sciences and humanities. GeoJournal, 75(1), 3-13.

Han, X.-P., Wang, X.-W., Yan, X.-Y., & Wang, B.-H. (2015). Cascading Walks Model for Human Mobility Patterns. PLOS ONE, 10(4), e0124800. https://doi.org/10.1371/journal.pone.0124800

Harari, G. M., Lane, N. D., Wang, R., Crosier, B. S., Campbell, A. T., & Gosling, S. D. (2016). Using Smartphones to Collect Behavioral Data in Psychological Science: Opportunities, Practical Considerations, and Challenges. Perspectives on Psychological Science, 11(6), 838–854. https://doi.org/10.1177/1745691616650285

Palmius, N., Tsanas, A., Saunders, K. E. A., Bilderbeck, A. C., Geddes, J. R., Goodwin, G. M., & Vos, M. D. (2017). Detecting Bipolar Depression From Geographic Location Data. IEEE Transactions on Biomedical Engineering, 64(8), 1761–1771. https://doi.org/10.1109/TBME.2016.2611862

Patterson, T. A., Thomas, L., Wilcox, C., Ovaskainen, O., & Matthiopoulos, J. (2008). State–space models of individual animal movement. Trends in Ecology & Evolution, 23(2), 87–94. https://doi.org/10.1016/j.tree.2007.10.009

Sadilek, A., & Krumm, J. (2016). Far Out: Predicting Long-Term Human Mobility. Microsoft Research. Retrieved from https://www.microsoft.com/en-us/research/publication/far-predicting-long-term-human-mobility/

Rasmussen, T. K., Ingvardson, J. B., Halldórsdóttir, K., & Nielsen, O. A. (2015). Improved methods to deduct trip legs and mode from travel surveys using wearable GPS devices: A case study from the Greater Copenhagen area. Computers, Environment and Urban Systems, 54(Supplement C), 301–313. https://doi.org/10.1016/j.compenvurbsys.2015.04.001

Wang, R., Harari, G., Hao, P., Zhou, X., & Campbell, A. T. (2015). SmartGPA: How Smartphones Can Assess and Predict Academic Performance of College Students. In Proceedings of the 2015 ACM International Joint Conference on Pervasive and Ubiquitous Computing (pp. 295–306). New York, NY, USA: ACM. https://doi.org/10.1145/2750858.2804251

Wolf, J., Oliveira, M., & Thompson, M. (2003). Impact of Underreporting on Mileage and Travel Time Estimates: Results from Global Positioning System-Enhanced Household Travel Survey. Transportation Research Record: Journal of the Transportation Research Board, 1854, 189–198. https://doi.org/10.3141/1854-21

Yang, F., Yao, Z., Cheng, Y., Ran, B., & Yang, D. (2016). Multimode trip information detection using personal trajectory data. Journal of Intelligent Transportation Systems, 20(5), 449–460. https://doi.org/10.1080/15472450.2016.1151791

Zheng, Y., Zhang, L., Xie, X., & Ma, W.-Y. (2009). Mining Interesting Locations and Travel Sequences from GPS Trajectories. In Proceedings of the 18th International Conference on World Wide Web (pp. 791–800). New York, NY, USA: ACM. https://doi.org/10.1145/1526709.1526816

### 2017-10-31 thesisReport.Rmd
First commit

---
title: 'Research Report: Longitudinal models of human mobility'
author: "Boaz Sobrado"
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography:
  - thesisGPS.bib
  - rReferences.bib
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: tp2.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggthemes)
```

# Introduction

How active people are and how they interact with their environment affects a wide range of measures including health , income and social capital [@goodchild_toward_2010]. A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies. Yet a large part of studies on human mobility are largely based on pen-and-paper travel diaries. These surveys have known methodological flaws, such as the short period of data collection (due to costs and burden to respondents), the underreporting of short trips [@wolf_impact_2003] and the underestimation of the duration of commutes [@delclos-alio_keeping_2017].

Longitudinal data on human temporospatial behaviour has become available through smartphones and other sensors. This data has been used to investigate topics such as the effects of the food environment on eating patterns *Zenk*, the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017]. While there have been advances in processing this data [@zheng_mining_2009] extracting meaningful information remains a difficult task. Two important challenges are dealing with measurement noise and missing data.

With regards to measurement accuracy, studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location. Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this happens to be where most people in the developed world tend to spend most of their time. In addition, location data collected by more ubiquitous (but less specialised) smartphones are an amalgamation of sensor data. For instance, Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006]. This adds another layer of complexity as each of these measures has its own characteristics in terms of measurement accuracy and bias.

As for missing data it is a pervasive issue as it can arise due to multiple factors, both technical and behavioural. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home, switching the phone off, switching location tracking off, etc. As a result, applied researchers are often left with wide temporal gaps with no measurements. For instance, different groups studying the effect of bipolar disorder on human movement have reported missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Similar trends are consistently reported in other fields [e.g. @harari_using_2016;@jankowska_framework_2015].

There is currently no golden standard in how to deal with missing temporospatial data [@barnett_inferring_2016]. @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it. Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017] can increase the severity of the missing data problem. On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.

In this paper we will compare methods used to deal with measurement error and missing data in location. Specifically, we are interested in location tracks as measured by commerically available smartphones.

## State of the art

Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and often serves different purposes. The following section briefly illustrates the different methods used to solve the measurement inaccuracy and missing data problem.

### State Space Models

There is a vast literature of using state space models (SSMs) to improve measurements accuracy and deal with missing data. Behavioural ecologists for instance, have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex, for example *Priesler* uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. *Gurram* *Kalman* *Chen* *Brown*. The extended Kalman filter is the de facto standard for GPS measurements.

The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time. For our pourposes the main limitation is that these models are based on the Markov property. Thus, the estimated location at timepoint $k$ is often based only upon measurements at $k$ and the previous timepoint $k-1$. This assumption ignores the periodic nature of human movement, whereby people generally spend the nights at home and the day at work. Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or infeasible [@sadilek_far_2016].

### Spatiotemporal Imputation Methods

For highly correlated in time.

### Alternative models

Alternatives to state space models include long range-persistence models, such as cascading walks models  and the FarOut model which rely on self-similarity and autoregressive characteristics [@han_cascading_2015; @sadilek_far_2016]. The latter uses Fourier analysis and PCA to extract cyclical patterns in an individual's behaviour and reduce the dimensionality of the extracted features and yields interpretable predictions for an individuals location months in advance.

## Methods

### Data & Analyses

The data used was collected between 2014 and 2017 on different Android devices. A total of X individuals contributed to the analysis, yielding a total of Y data points. Moreover we used diary information from **SOMEWHERE**.

Analyses were performed using R and a multitude of other packages (@ggplot2 @dplyr @leaflet @sp @ggthemes @base @sp1 @sp2).

*Plots: data points over time*

![Raw GPS measurements of a journey from de Uithof to Tuinwijk on February 17th 2017. The measurements are in red, the filtered path is in blue. The circles denote 67% confidence intervals of the given GPS measurement. Measurements and fitted points which follow each other in time are connected by lines. The inacurate measurements lead to estimates of irregular movements. The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](//img//Map1.png)

#### Filtering

*Carlson et al 2015* describe a personal activity measurement system (PALMS), which can filter data points (among others). Their system is validated using cameras.

Like others (e.g. @palmius_detecting_2017) they detect invalid points using extreme speed.

*Google* offers an out of the box activity inferral mechanism.

### Discussion on missing data

Our missing data patterns fit the broader pattern of missing data reported by others [@palmius_detecting_2017;@saeb_mobile_2015; @grunerbl_smartphone-based_2015;@harari_using_2016].

Missing data can  Importantly, one cannot say that the data is missing at random as there are a myriad of reasons why missingness can be related to location, such as turning off the phone on the plane before a long flight.

```{r missingDataPlot,echo = FALSE,fig.cap="Missing data over time for the author. The x-axis denotes time, the y-axis shows how many measurements are made on each day. The fill of the points shows the amount of hours of captured data each day.",fig.height=5}

ps<- readRDS("..//data//missingDataAnalysisBoaz.rds")

pal<- c("#4228A8", "#0D0101","#E04AC2")

ggplot(ps, aes(x = Day, y=count, fill = type, colour= type))+geom_point()+
  theme_tufte()+
  xlab("")+
  ylab("Measurements")+
  ggtitle("Missing data over time")+
  scale_colour_manual(values=pal)+
  theme(legend.position = c(0.8, 0.8))+
  theme(legend.title=element_blank())

rm(ps)
```

### Models

#### Spatiotemporal imputation methods

## Results

## Discussion

## References

### 2017-10-31 addRCitations.R

@@ -1,0 +1,6 @@
+# adding all R packages citations
+
+library(Ref)
+listOfPackages <- c("dplyr","ggplot2","base","ggthemes","sp1","sp2","pandoc","citr","rstudio")
+
+write.bib(listOfPackages, file='rReferences')

### 2017-10-31 thesisProposalPlot.R

@@ -1,0 +1,125 @@
+library(ggmap)
+
+day<-readRDS("C:\\Users\\user1\\Desktop\\Life\\Academic\\Utrecht\\year2\\thesis\\kalmanFilters\\implementation\\feb17.rds")
+library(leaflet)
+library(dplyr)
+library(sp)
+
+#initializing variables
+count <- nrow(day) # amount of data points in the day
+z <- cbind(day$lon,day$lat) #measurements
+
+#Allocate space:
+xhat <- matrix(rep(0,2*count),ncol =2) #a posteri estimate at each step
+P <- array(0,dim=c(2,2,count))  #a posteri error estimate
+xhatminus <- matrix(rep(0,2*count),ncol =2) #a priori estimate
+Pminus <- array(0,dim=c(2,2,count)) #a priori error estimate
+K <- array(0,dim=c(2,2,count)) #gain
+
+#Initializing matrices
+A <-diag(2)
+H<-diag(2)
+R<-function(k) diag(2)* day$accuracy[k]^2#estimate of measurement variance
+Q<-function(k) diag(2)* as.numeric(day$timeNextMeasure[k])^1.5# the process variance
+
+#initialise guesses:
+xhat[1,] <- z[1,]
+P[,,1] <- diag(2)
+
+for (k in 2:count){
+  #time update
+  #project state ahead
+  xhatminus[k,] <- A %*% xhat[k-1,] #+ B %*% u[k-1]
+  
+  #project error covariance ahead
+  Pminus[,,k] <- A %*% P[,,k-1] %*%  t(A) + (Q(k))
+  
+  #measurement update
+  # kalman gain
+  K[,,k] <- Pminus[,,k] %*% t(H)/ (H %*% Pminus[,,k] %*% t(H) + R(k))
+  
+  #what if NaaN?
+  K[,,k][which(is.nan(K[,,k]))]<-0
+  
+  # update estimate with measurement
+  xhat[k,] <-  xhatminus[k,] + K[,,k] %*% (z[k,] - H %*% xhatminus[k,])
+  #update error covariance
+  P[,,k] = (diag(2) - K[,,k]%*% H) %*% Pminus[,,k]
+}
+
+lns <- Line(na.omit(cbind(day$lon,day$lat)))
+lnsKF<-Line(na.omit(xhat))
+
+index<- 350:430
+lns <- Line(na.omit(cbind(day$lon[index],day$lat[index])))
+lnsKF<-Line(na.omit(xhat[index,]))
+
+
[…]

### 2017-11-07 thesisReport.Rmd
added a few details

@@
 library(ggthemes)
+library(ggplot2)
 ```
@@
 
+*An asside as to my conversation with Peter on 07.11*
+
+We can get some trackers on people to measure movement constantly and be a "golden standard" to evaluate the models.
+
+For December, it would be nice to have the Lit Rev section, plus an exploration of the problem.
+
 ### Data & Analyses

### 2017-11-10 tex2pdf.4476/input.tex
added graph plus methods and abstract bits

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Research Report: Longitudinal models of human mobility},
            pdfauthor={Boaz Sobrado},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Research Report: Longitudinal models of human mobility}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Boaz Sobrado}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{10 November 2017}

\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

How active people are and how they interact with their environment
affects a wide range of measures including health , income and social
capital (Goodchild and Janelle 2010). A better understanding of both
within-person and between-person variability in geospatial patterns
could be conducive to better social, health and urban-planning policies.
Yet a large part of studies on human mobility are largely based on
pen-and-paper travel diaries. These surveys have known methodological
flaws, such as the short period of data collection (due to costs and
burden to respondents), the underreporting of short trips (Wolf,
Oliveira, and Thompson 2003) and the underestimation of the duration of
commutes (Delclòs-Alió, Marquet, and Miralles-Guasch 2017).

Objective data on human temporospatial behaviour has become available
throgh the Global Positioning System (GPS) which uses the distance
between a device and a number of satellites to determine location.
Within behavioural science, this type of data has been used to
investigate topics such as the effects of the food environment on eating
patterns (Zenk, Schulz, and Odoms-Young 2009), the movement correlates
of personality and academic performance (G. M. Harari et al. 2016; Wang
et al. 2015) and detecting bipolar disorder (Palmius et al. 2017). In
most of these studies participants are given a specialised GPS devices
or a smartphone with a custom made app. However, Barnett and Onnela
(2016) point out that these studies are not scalable due to cost and
burden to participants, moreover may be biased because of the
introduction of a new device to the participant's life and usually span
a short amount of time. A solution would be to take advantage of the
location logs that smartphones already collect, such as Google Location
History which stores Android phone users' locations (Location History
2017). Yet, because GPS sensors consume a significant amounts of
battery, these logs can be sparse and inaccurate. Hence, two important
challenges are dealing with measurement noise and missing data.

Studies of professional grade GPS trackers suggest that less than 80\%
of measurements fall within 10 meters of the true location. Moreover,
GPS measures are reported to be most inaccurate in high density urban
locations and indoors (Schipperijn et al. 2014; Duncan et al. 2013).
Unfortunately for social scientists , this happens to be where most
people in the developed world tend to spend most of their time. In
addition, location data collected by more ubiquitous (but less
specialised) smartphones are an amalgamation of sensor data. For
instance, Android phones collect location information from WiFi access
points, cellphone triangluation, and GPS measurements due to
computational and battery constraints (LaMarca et al. 2005; M. Y. Chen
et al. 2006). This adds another layer of complexity as each of these
measures has its own characteristics in terms of measurement accuracy
and bias.

As for missing data it is a pervasive issue as it can arise due to
multiple factors, both technical and behavioural. Technical reasons
include signal loss, battery failure and device failure. Behavioural
reasons include leaving the phone at home, switching the phone off,
switching location tracking off, etc. As a result, applied researchers
are often left with wide temporal gaps with no measurements. For
instance, different groups studying the effect of bipolar disorder on
human movement have reported missing data rates between 30\% to 50\%
(Saeb et al. 2015; Grünerbl et al. 2015; Palmius et al. 2017). Similar
trends are consistently reported in other fields (e.g. G. M. Harari et
al. 2016; Jankowska, Schipperijn, and Kerr 2015).

There is currently no golden standard in how to deal with missing
temporospatial track data (Barnett and Onnela 2016). Jankowska,
Schipperijn, and Kerr (2015) have pointed out that there is often little
transparency regarding decisions of how to deal with it. Methods
frequently used by researchers to reduce noise, such as throwing out
inaccurate measurements (e.g. Palmius et al. 2017) can increase the
severity of the missing data problem. On the other hand, noisy data can
lead to inaccurate conclusions if it is not accounted for.

In this paper we will compare methods used to deal with measurement
error and missing data in location information. Specifically, we are
interested in establishing accurate mobility patterns from smartphone
GPS logs.

\section{Problem description and literature
review}\label{problem-description-and-literature-review}

\subsection{The structure of smartphone location
logs}\label{the-structure-of-smartphone-location-logs}

Smartphone log data looks like this:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"img/missingdayPeter.png"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=36.28in]{img/missingdayPeter} \caption{Missing data caption.}\label{fig:measurementsPerDay}
\end{figure}

Linear Interpolation vs actual movement vs kalman filtered.

\subsection{State of the art in temporospatial
models}\label{state-of-the-art-in-temporospatial-models}

Research with respect to the analysis of GPS data is wideranging, highly
interdisciplinary and often serves different purposes. Given that
Barnett and Onnela (2016) is to our knowledge the only paper explicitly
focusing on dealing with missingness for inferring missingness using GPS
data the following section briefly illustrates the some methods used to
solve measurement inaccuracy and missing data problems in the field, as
well as their applicability to our question.

\subsubsection{Spatiotemporal Imputation
Methods}\label{spatiotemporal-imputation-methods}

Given fixed measurement stations there are sevaral imputation methods
for spatiotemporal measurements. For instance, Feng et al. (2014)
illustrate their CUTOFF method, which relies on estimating missing
values using the nearest observed neighbours in time, using rainfall
data from dozens of gauging stations across Australia. Similarly, Z.
Zhang et al. (2017) use a variety of machine learning methods to present
their model based on underground water data in China.

While Feng et al. (2014) claim their model could be used to establish
mobility patterns, ostentibly by dividing the sample space into rasters
which would be measurement stations indicating a probability of the
individual being there, this seems to be computationally unfeasible to
do. To our knowledge such models have not been implemented.

\subsubsection{State Space Models}\label{state-space-models}

There is a vast literature of using state space models (SSMs) to improve
measurements accuracy and deal with missing data. Behavioural ecologists
for instance, have used SSMs to explain how animals interact with their
environment (Patterson et al. 2008). These models can be quite complex,
for example Preisler et al. (2004) uses Markovian movement processes to
characterise the effect of roads, food patches and streams on cyclical
elk movements. The most well studied SSM is the Kalman filter, which is
the optimal algorithm for inferring linear Gaussian systems. The
extended Kalman filter is the de facto standard for GPS measurements (Z.
Chen and Brown 2013).

The advantage of state space models is that they are flexible, deal with
measurement inaccuracy, include information from different sources and
can be used in real time. For our pourposes the main limitation is that
these models are based on the Markov property. Thus, the estimated
location at timepoint \(k\) is often based only upon measurements at
\(k\) and the previous timepoint \(k-1\). This assumption ignores the
periodic nature of human movement, whereby people generally spend the
nights at home and the day at work. Hierarchical structuring and
conditioning on a larger context have been suggested as ways to improve
their performance, but these are often computationally intractable or
infeasible (Sadilek and Krumm 2016).

\subsubsection{Alternative models}\label{alternative-models}

Alternatives to state space models include long range-persistence
models, such as cascading walks models and the FarOut model which rely
on self-similarity and autoregressive characteristics (Han et al. 2015;
Sadilek and Krumm 2016). The latter uses Fourier analysis and PCA to
extract cyclical patterns in an individual's behaviour and reduce the
dimensionality of the extracted features and yields interpretable
predictions for an individuals location months in advance.

\subsection{Methods}\label{methods}

\emph{An asside as to my conversation with Peter on 07.11}

We can get some trackers on people to measure movement constantly and be
a ``golden standard'' to evaluate the models.

For December, it would be nice to have the Lit Rev section, plus an
exploration of the problem.

\subsubsection{Data \& Analyses}\label{data-analyses}

The data used was collected between 2014 and 2017 on different Android
devices. A total of X individuals contributed to the analysis, yielding
a total of Y data points. Moreover we used diary information from
\textbf{SOMEWHERE}.

Analyses were performed using R and a multitude of other packages
(Wickham (2009) Wickham and Francois (2016) ({\textbf{???}})
({\textbf{???}}) Arnold (2013) R Core Team (2017) E. J. Pebesma and
Bivand (2005) Bivand, Pebesma, and Gomez-Rubio (2013)).

\emph{Plots: data points over time}

\emph{Raw GPS measurements of a journey from de Uithof to Tuinwijk on
February 17th 2017. The measurements are in red, the filtered path is in
blue. The circles denote 67\% confidence intervals of the given GPS
measurement. Measurements and fitted points which follow each other in
time are connected by lines. The inacurate measurements lead to
estimates of irregular movements. The filtered movement estimate seems
more accurate in terms of the path, but lags behind the unfiltered
movement.}

\paragraph{Filtering}\label{filtering}

\emph{Carlson et al 2015} describe a personal activity measurement
system (PALMS), which can filter data points (among others). Their
system is validated using cameras.

Like others (e.g. Palmius et al. (2017)) they detect invalid points
using extreme speed.

\emph{Google} offers an out of the box activity inferral mechanism.

\subsubsection{Discussion on missing
data}\label{discussion-on-missing-data}

Our missing data patterns fit the broader pattern of missing data
reported by others (Palmius et al. 2017; Saeb et al. 2015; Grünerbl et
al. 2015; G. M. Harari et al. 2016).

Missing data can Importantly, one cannot say that the data is missing at
random as there are a myriad of reasons why missingness can be related
to location, such as turning off the phone on the plane before a long
flight.

\begin{figure}
\centering
\includegraphics{thesisReport_files/figure-latex/missingDataPlot-1.pdf}
\caption{Missing data over time for the author. The x-axis denotes time,
the y-axis shows how many measurements are made on each day. The fill of
the points shows the amount of hours of captured data each day.}
\end{figure}

\subsubsection{Models}\label{models}

\paragraph{Spatiotemporal imputation
methods}\label{spatiotemporal-imputation-methods-1}

\subsection{Results}\label{results}

\subsection{Discussion}\label{discussion}

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-ggthemes}{}
Arnold, Jeffrey B. 2013. \emph{Ggthemes: Extra Themes, Scales and Geoms
for Ggplot}. \url{https://CRAN.R-project.org/package=ggthemes}.

\hypertarget{ref-barnett_inferring_2016}{}
Barnett, Ian, and Jukka-Pekka Onnela. 2016. ``Inferring Mobility
Measures from GPS Traces with Missing Data.'' \emph{arXiv:1606.06328
{[}Stat{]}}, June. \url{http://arxiv.org/abs/1606.06328}.

\hypertarget{ref-sp2}{}
Bivand, Roger S., Edzer Pebesma, and Virgilio Gomez-Rubio. 2013.
\emph{Applied Spatial Data Analysis with R, Second Edition}. Springer,
NY. \url{http://www.asdar-book.org/}.

\hypertarget{ref-chen_practical_2006}{}
Chen, Mike Y., Timothy Sohn, Dmitri Chmelev, Dirk Haehnel, Jeffrey
Hightower, Jeff Hughes, Anthony LaMarca, Fred Potter, Ian Smith, and
Alex Varshavsky. 2006. ``Practical Metropolitan-Scale Positioning for
GSM Phones.'' In \emph{UbiComp 2006: Ubiquitous Computing}, 225--42.
Lecture Notes in Computer Science. Springer, Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11853565_14}{10.1007/11853565\_14}.

\hypertarget{ref-chen_state_2013}{}
Chen, Zhe, and Emery N. Brown. 2013. ``State Space Model.''
\emph{Scholarpedia} 8 (3): 30868.
doi:\href{https://doi.org/10.4249/scholarpedia.30868}{10.4249/scholarpedia.30868}.

\hypertarget{ref-delclos-alio_keeping_2017}{}
Delclòs-Alió, Xavier, Oriol Marquet, and Carme Miralles-Guasch. 2017.
``Keeping Track of Time: A Smartphone-Based Analysis of Travel Time
Perception in a Suburban Environment.'' \emph{Travel Behaviour and
Society} 9 (Supplement C): 1--9.
doi:\href{https://doi.org/10.1016/j.tbs.2017.07.001}{10.1016/j.tbs.2017.07.001}.

\hypertarget{ref-duncan_portable_2013}{}
Duncan, Scott, Tom I. Stewart, Melody Oliver, Suzanne Mavoa, Deborah
MacRae, Hannah M. Badland, and Mitch J. Duncan. 2013. ``Portable Global
Positioning System Receivers: Static Validity and Environmental
Conditions.'' \emph{American Journal of Preventive Medicine} 44 (2):
e19--29.
doi:\href{https://doi.org/10.1016/j.amepre.2012.10.013}{10.1016/j.amepre.2012.10.013}.

\hypertarget{ref-feng_cutoff:_2014}{}
Feng, Lingbing, Gen Nowak, T.J. O'Neill, and A Welsh. 2014. ``CUTOFF: A
Spatio-Temporal Imputation Method.'' \emph{Journal of Hydrology} 519
(November): 3591--3605.
doi:\href{https://doi.org/10.1016/j.jhydrol.2014.11.012}{10.1016/j.jhydrol.2014.11.012}.

\hypertarget{ref-goodchild_toward_2010}{}
Goodchild, Michael F., and Donald G. Janelle. 2010. ``Toward Critical
Spatial Thinking in the Social Sciences and Humanities.''
\emph{GeoJournal} 75 (1): 3--13.
doi:\href{https://doi.org/10.1007/s10708-010-9340-3}{10.1007/s10708-010-9340-3}.

\hypertarget{ref-grunerbl_smartphone-based_2015}{}
Grünerbl, Agnes, Amir Muaremi, Venet Osmani, Gernot Bahle, Stefan Ohler,
Gerhard Tröster, Oscar Mayora, Christian Haring, and Paul Lukowicz.
2015. ``Smartphone-Based Recognition of States and State Changes in
Bipolar Disorder Patients.'' \emph{IEEE Journal of Biomedical and Health
Informatics} 19 (1): 140--48.
doi:\href{https://doi.org/10.1109/JBHI.2014.2343154}{10.1109/JBHI.2014.2343154}.

\hypertarget{ref-han_cascading_2015}{}
Han, Xiao-Pu, Xiang-Wen Wang, Xiao-Yong Yan, and Bing-Hong Wang. 2015.
``Cascading Walks Model for Human Mobility Patterns.'' \emph{PLOS ONE}
10 (4): e0124800.
doi:\href{https://doi.org/10.1371/journal.pone.0124800}{10.1371/journal.pone.0124800}.

\hypertarget{ref-harari_using_2016}{}
Harari, Gabriella M., Nicholas D. Lane, Rui Wang, Benjamin S. Crosier,
Andrew T. Campbell, and Samuel D. Gosling. 2016. ``Using Smartphones to
Collect Behavioral Data in Psychological Science: Opportunities,
Practical Considerations, and Challenges.'' \emph{Perspectives on
Psychological Science} 11 (6): 838--54.
doi:\href{https://doi.org/10.1177/1745691616650285}{10.1177/1745691616650285}.

\hypertarget{ref-jankowska_framework_2015}{}
Jankowska, Marta M., Jasper Schipperijn, and Jacqueline Kerr. 2015. ``A
Framework for Using GPS Data in Physical Activity and Sedentary Behavior
Studies.'' \emph{Exercise and Sport Sciences Reviews} 43 (1): 48--56.
doi:\href{https://doi.org/10.1249/JES.0000000000000035}{10.1249/JES.0000000000000035}.

\hypertarget{ref-lamarca_place_2005}{}
LaMarca, Anthony, Yatin Chawathe, Sunny Consolvo, Jeffrey Hightower, Ian
Smith, James Scott, Timothy Sohn, et al. 2005. ``Place Lab: Device
Positioning Using Radio Beacons in the Wild.'' In \emph{Pervasive
Computing}, 116--33. Lecture Notes in Computer Science. Springer,
Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11428572_8}{10.1007/11428572\_8}.

\hypertarget{ref-location_history_timeline_nodate}{}
Location History, Google. 2017. ``Timeline.'' Accessed November 9.
\url{https://www.google.com/maps/timeline?pb}.

\hypertarget{ref-palmius_detecting_2017}{}
Palmius, N., A. Tsanas, K. E. A. Saunders, A. C. Bilderbeck, J. R.
Geddes, G. M. Goodwin, and M. De Vos. 2017. ``Detecting Bipolar
Depression from Geographic Location Data.'' \emph{IEEE Transactions on
Biomedical Engineering} 64 (8): 1761--71.
doi:\href{https://doi.org/10.1109/TBME.2016.2611862}{10.1109/TBME.2016.2611862}.

\hypertarget{ref-patterson_statespace_2008}{}
Patterson, Toby A., Len Thomas, Chris Wilcox, Otso Ovaskainen, and Jason
Matthiopoulos. 2008. ``State--space Models of Individual Animal
Movement.'' \emph{Trends in Ecology \& Evolution} 23 (2): 87--94.
doi:\href{https://doi.org/10.1016/j.tree.2007.10.009}{10.1016/j.tree.2007.10.009}.

\hypertarget{ref-sp1}{}
Pebesma, Edzer J., and Roger S. Bivand. 2005. ``Classes and Methods for
Spatial Data in R.'' \emph{R News} 5 (2): 9--13.
\url{https://CRAN.R-project.org/doc/Rnews/}.

\hypertarget{ref-preisler_modeling_2004}{}
Preisler, Haiganoush K., Alan A. Ager, Bruce K. Johnson, and John G.
Kie. 2004. ``Modeling Animal Movements Using Stochastic Differential
Equations.'' \emph{Environmetrics 15: P. 643-657}.
\url{https://www.fs.usda.gov/treesearch/pubs/33038}.

\hypertarget{ref-base}{}
R Core Team. 2017. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\hypertarget{ref-sadilek_far_2016}{}
Sadilek, Adam, and John Krumm. 2016. ``Far Out: Predicting Long-Term
Human Mobility.'' \emph{Microsoft Research}, December.
\url{https://www.microsoft.com/en-us/research/publication/far-predicting-long-term-human-mobility/}.

\hypertarget{ref-saeb_mobile_2015}{}
Saeb, Sohrab, Mi Zhang, Christopher J. Karr, Stephen M. Schueller, Marya
E. Corden, Konrad P. Kording, and David C. Mohr. 2015. ``Mobile Phone
Sensor Correlates of Depressive Symptom Severity in Daily-Life Behavior:
An Exploratory Study.'' \emph{Journal of Medical Internet Research} 17
(7): e175.
doi:\href{https://doi.org/10.2196/jmir.4273}{10.2196/jmir.4273}.

\hypertarget{ref-schipperijn_dynamic_2014}{}
Schipperijn, Jasper, Jacqueline Kerr, Scott Duncan, Thomas Madsen,
Charlotte Demant Klinker, and Jens Troelsen. 2014. ``Dynamic Accuracy of
GPS Receivers for Use in Health Research: A Novel Method to Assess GPS
Accuracy in Real-World Settings.'' \emph{Frontiers in Public Health} 2:
21.
doi:\href{https://doi.org/10.3389/fpubh.2014.00021}{10.3389/fpubh.2014.00021}.

\hypertarget{ref-wang_smartgpa:_2015}{}
Wang, Rui, Gabriella Harari, Peilin Hao, Xia Zhou, and Andrew T.
Campbell. 2015. ``SmartGPA: How Smartphones Can Assess and Predict
Academic Performance of College Students.'' In \emph{Proceedings of the
2015 ACM International Joint Conference on Pervasive and Ubiquitous
Computing}, 295--306. UbiComp '15. New York, NY, USA: ACM.
doi:\href{https://doi.org/10.1145/2750858.2804251}{10.1145/2750858.2804251}.

\hypertarget{ref-ggplot2}{}
Wickham, Hadley. 2009. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. Springer-Verlag New York. \url{http://ggplot2.org}.

\hypertarget{ref-dplyr}{}
Wickham, Hadley, and Romain Francois. 2016. \emph{Dplyr: A Grammar of
Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.

\hypertarget{ref-wolf_impact_2003}{}
Wolf, Jean, Marcelo Oliveira, and Miriam Thompson. 2003. ``Impact of
Underreporting on Mileage and Travel Time Estimates: Results from Global
Positioning System-Enhanced Household Travel Survey.''
\emph{Transportation Research Record: Journal of the Transportation
Research Board} 1854 (January): 189--98.
doi:\href{https://doi.org/10.3141/1854-21}{10.3141/1854-21}.

\hypertarget{ref-zenk_how_2009}{}
Zenk, Shannon N., Amy J. Schulz, and Angela Odoms-Young. 2009. ``How
Neighborhood Environments Contribute to Obesity.'' \emph{The American
Journal of Nursing} 109 (7): 61--64.
doi:\href{https://doi.org/10.1097/01.NAJ.0000357175.86507.c8}{10.1097/01.NAJ.0000357175.86507.c8}.

\hypertarget{ref-zhang_application_2017}{}
Zhang, Zhongrong, Xuan Yang, Hao Li, Weide Li, Haowen Yan, and Fei Shi.
2017. ``Application of a Novel Hybrid Method for Spatiotemporal Data
Imputation: A Case Study of the Minqin County Groundwater Level.''
\emph{Journal of Hydrology} 553 (Supplement C): 384--97.
doi:\href{https://doi.org/10.1016/j.jhydrol.2017.07.053}{10.1016/j.jhydrol.2017.07.053}.

\end{document}

### 2017-11-10 thesisReport.Rmd
added graph plus methods and abstract bits

@@
 library(ggplot2)
+library(knitr)
 ```
+
+# Abstract
+*There is a great interest in using objective location data to quantify the mobility characteristics of individuals. However, *
 
@@
 
-Longitudinal data on human temporospatial behaviour has become available through smartphones and other sensors.
-This data has been used to investigate topics such as the effects of the food environment on eating patterns *Zenk*, the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017].
-While there have been advances in processing this data [@zheng_mining_2009] extracting meaningful information remains a difficult task.
-Two important challenges are dealing with measurement noise and missing data.
+Objective data on human temporospatial behaviour has become available throgh the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
+Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017].
+In most of these studies participants are given a specialised GPS devices or a smartphone with a custom made app.
+However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life and usually span a short amount of time.
+A solution would be to take advantage of the location logs that smartphones already collect, such as Google Location History which stores Android phone users' locations [@location_history_timeline_nodate].
+Yet, because GPS sensors consume a significant amounts of battery, these logs can be sparse and inaccurate.
+Hence, two important challenges are dealing with measurement noise and missing data.
 
-With regards to measurement accuracy, studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location.
+Studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location.
 Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
-Unfortunately for researchers, this happens to be where most people in the developed world tend to spend most of their time.
+Unfortunately for social scientists , this happens to be where most people in the developed world tend to spend most of their time.
 In addition, location data collected by more ubiquitous (but less specialised) smartphones are an amalgamation of sensor data.
@@
 
-There is currently no golden standard in how to deal with missing temporospatial data [@barnett_inferring_2016].
+There is currently no golden standard in how to deal with missing temporospatial track data [@barnett_inferring_2016].
 @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it.
@@
 
-In this paper we will compare methods used to deal with measurement error and missing data in location.
-Specifically, we are interested in location tracks as measured by commerically available smartphones.
+In this paper we will compare methods used to deal with measurement error and missing data in location information.
+Specifically, we are interested in establishing accurate mobility patterns from smartphone GPS logs.
 
-## State of the art
+# Problem description and literature review
+
+## The structure of smartphone location logs
+
+Android logged data
+
+```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data over time for the author.
+The x-axis denotes time, the y-axis shows how many measurements are made and each point is a day.
+The fill of the points shows the amount of hours of captured data each day.",out.width = '100%'}
+include_graphics("img/missingdayPeter.png")# from the playingaroundwiththicken.r file
+```
+
+Linear Interpolation vs actual movement vs kalman filtered.
+
+## State of the art in temporospatial models
 
 Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and often serves different purposes.
-The following section briefly illustrates the different methods used to solve the measurement inaccuracy and missing data problem.
+Given that @barnett_inferring_2016 is to our knowledge the only paper explicitly focusing on dealing with missingness for inferring missingness using GPS data the following section briefly illustrates the some methods used to solve measurement inaccuracy and missing data problems in the field, as well as their applicability to our question.
+
+### Spatiotemporal Imputation Methods
+
+Given fixed measurement stations there are sevaral imputation methods for spatiotemporal measurements.
+For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
+Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.
+
+While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostentibly by dividing the sample space into rasters which would be measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible to do.
+To our knowledge such models have not been implemented.
 
@@
 Behavioural ecologists for instance, have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008].
-These models can be quite complex, for example *Priesler* uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
+These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
 The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
-*Gurram* *Kalman* *Chen* *Brown*.
-The extended Kalman filter is the de facto standard for GPS measurements.
+The extended Kalman filter is the de facto standard for GPS measurements [@chen_state_2013].
 
@@
 Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or infeasible [@sadilek_far_2016].
-
-### Spatiotemporal Imputation Methods
-
-For highly correlated in time.
 
@@
 The inacurate measurements lead to estimates of irregular movements.
-The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](//img//Map1.png)
+The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](img/Map1.png)
 
@@
 
-ps<- readRDS("..//data//missingDataAnalysisBoaz.rds")
-
-pal<- c("#4228A8", "#0D0101","#E04AC2")
-
-ggplot(ps, aes(x = Day, y=count, fill = type, colour= type))+geom_point()+
-  theme_tufte()+
-  xlab("")+
-  ylab("Measurements")+
-  ggtitle("Missing data over time")+
-  scale_colour_manual(values=pal)+
-  theme(legend.position = c(0.8, 0.8))+
-  theme(legend.title=element_blank())
-
-rm(ps)
-```
-
 ### Models

### 2017-11-10 missingDays.R

@@ -1,0 +1,73 @@
+##Graphs for determining where missing data is
+library(ggthemes)
+library(dplyr)
+
+# would be a lot easier to do with padr
+
+# read data
+loc<- readRDS("..\\all.rds")
+
+
+pointsdaily<-loc %>% group_by(Day) %>% summarise(count = n())
+pointshourly<-loc %>% group_by(Day, lubridate::hour(time)) %>%
+  summarise(count=n())
+rm(loc)
+
+colnames(pointshourly) <- c("Day","Hour","count")
+
+#generate dates
+x<-as.Date(pointsdaily$Day[1]):as.Date(pointsdaily$Day[nrow(pointsdaily)])
+x<-as.Date.numeric(x = x,origin="1970-01-01")
+h<- sort(rep(0:23,length(x)))
+allDayHour <- data.frame(Day = rep(x,24), Hour = h, count = 0)
+
+#dates with zero measurements
+noMeasureDay<-as.Date(setdiff(x,pointsdaily$Day), origin = "1970-01-01")
+# alternative approach
+
+u<- unique(rbind(pointshourly[,-3],allDayHour[,-3]))
+allDayHourPasted<- paste0(allDayHour$Day,"$",allDayHour$Hour)
+measuredDayHourPasted <- paste0(pointshourly$Day,"$",pointshourly$Hour)
+noMeasureHour<- setdiff(allDayHourPasted,measuredDayHourPasted)
+noMeasureHour<-strsplit(noMeasureHour,split = "\\$")
+#hour with zero measurements
+
+noMeasureHour <- setNames(do.call(rbind.data.frame, noMeasureHour), c("Day", "Hour"))
+noMeasureHour$count <- 0
+noMeasureHour$Day <- as.Date(noMeasureHour$Day)
+noMeasureHour$Hour <- as.numeric(noMeasureHour$Hour)
+
+t<- data.frame(table(noMeasureHour$Day))
+nm<- filter(t,Freq != 24 & Freq>=6) %>% select(Var1)
+nm<- as.Date(as.character(nm$Var1))
+
+ps<-rbind(pointsdaily,data.frame(Day = noMeasureDay,count = 0))
+ps$type <- "More than 6 hours"
+ps[which(ps$Day%in%nm),"type"] <- "6 hours or less"
+ps[which(ps$Day%in%noMeasureDay),"type"] <- "No Measure"
[…]

### 2017-11-10 thesisReport.Rmd
finished abstract, problem section

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 library(ggthemes)
@@
 # Abstract
-*There is a great interest in using objective location data to quantify the mobility characteristics of individuals. However, *
+*There is a great interest in using objective location data to quantify the mobility characteristics of individuals.
+Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals.
+However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment.
+In this paper we present a model which leverages the periodicity of human movement in order to impute missing data values.
+Moreover, we will compare the performance of the model compared to currently used methods, such as linear interpolation.*
+
+**Keywords: GPS, missing data, human mobility,big data**
 
@@
 
-Objective data on human temporospatial behaviour has become available throgh the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
+Objective data on human mobility has become available throgh the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
 Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017].
+
 In most of these studies participants are given a specialised GPS devices or a smartphone with a custom made app.
 However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life and usually span a short amount of time.
-A solution would be to take advantage of the location logs that smartphones already collect, such as Google Location History which stores Android phone users' locations [@location_history_timeline_nodate].
+A solution is to take advantage of the already exisiting smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years[@location_history_timeline_nodate].
+These logs can be accessed and shared by users.
 Yet, because GPS sensors consume a significant amounts of battery, these logs can be sparse and inaccurate.
@@
 
-Studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location.
+There is currently no golden standard in how to deal with missing data in this context [@barnett_inferring_2016].
+@jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it.
+Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
+@palmius_detecting_2017] can exacerbate the severity of the missing data problem.
+On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.
+In this paper we will compare methods used to deal with measurement error and missing data in location information.
+Specifically, we are interested in establishing accurate mobility patterns from smartphone GPS logs.
+
+# Problem description & literature review
+
+Given that there is next to no literature on missing data in smartphone location logs it is worth illustrating the typical characteristics of this data using an example data set.
+Moreover, although we could find no published papers on dealing with missing data in this narrow domain, there are however methods which deal with similar problems.
+In this section we will describe the problems with the data and models which could be applied to it.
+
+## Missing data and noise in location logs
+
+This example data set spans 3 years from January 2013 to January 2017.
+It contains 814 941 measurements, with approximately 742 measurements per day (*SD*=868.15).
+Moreover, it also contains estimates of velocity, altitude, heading and activity (e.g.
+in a vehicle, still).
+For the purposes of this paper we will focus only on lattitude, longitude and time.
+
+### Accuracy
+
+Even in professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
 Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
-Unfortunately for social scientists , this happens to be where most people in the developed world tend to spend most of their time.
-In addition, location data collected by more ubiquitous (but less specialised) smartphones are an amalgamation of sensor data.
-For instance, Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006].
-This adds another layer of complexity as each of these measures has its own characteristics in terms of measurement accuracy and bias.
+Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
 
-As for missing data it is a pervasive issue as it can arise due to multiple factors, both technical and behavioural.
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement.
+The x-axis denotes the accuracy in meters such that the user's actual location is expected to be within that radius with 67 percent confidence.
+Measures were classified into high accuracy (under 800m), medium accuracy (under 5000m) and low accuracy (over 5000m).
+Measures with accuracy of over 20000m were omitted from the graph.
+The maximum accuracy measure was over 40000m.
+Low accuracy measures are often near in time to more high accuracy measures.",out.width = '100%'}
+
+include_graphics("img/accuracyPeter.png")
+
+```
+
+Given that Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
+Based on this data set, less than 40% of measures fall with a 67% confidence radius of 20 meters.
+
+### Missingness
+
+Missing data is a pervasive issue as it can arise due to multiple factors, both technical and behavioural.
 Technical reasons include signal loss, battery failure and device failure.
-Behavioural reasons include leaving the phone at home, switching the phone off, switching location tracking off, etc.
+Behavioural reasons include leaving the phone at home, switching the phone off, switching location measurements off, and so on.
 As a result, applied researchers are often left with wide temporal gaps with no measurements.
@@
 
-There is currently no golden standard in how to deal with missing temporospatial track data [@barnett_inferring_2016].
-@jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it.
-Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
-@palmius_detecting_2017] can increase the severity of the missing data problem.
-On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.
-
-In this paper we will compare methods used to deal with measurement error and missing data in location information.
-Specifically, we are interested in establishing accurate mobility patterns from smartphone GPS logs.
-
-# Problem description and literature review
-
-## The structure of smartphone location logs
-
-Android logged data
-
 ```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data over time for the author.
@@
 The fill of the points shows the amount of hours of captured data each day.",out.width = '100%'}
+
 include_graphics("img/missingdayPeter.png")# from the playingaroundwiththicken.r file
 ```
-
-Linear Interpolation vs actual movement vs kalman filtered.
 
@@
 
-```{r missingDataPlot,echo = FALSE,fig.cap="Missing data over time for the author.
-The x-axis denotes time, the y-axis shows how many measurements are made on each day.
-The fill of the points shows the amount of hours of captured data each day.",fig.height=5}
-
 ### Models

### 2017-11-10 playingAroundwithThicken.R

@@ -2,25 +2,16 @@
 library(dplyr)
 library(padr)
 library(ggplot2)
+library(tibbletime)
 # read data
-loc<- readRDS("..\\all.rds")
+all<-readRDS("../data/peter/myLocationHistory.rds")
+attr(all$time, "tzone") <- "Europe/Paris"
 
-mDaily<- loc %>%
-  filter(who =="peter")%>%
-  select(time) %>%
-  mutate(time =  lubridate::force_tz(time, "Europe/Budapest")) %>%
-  top_n(207952)%>% 
-  thicken('day') %>%
-  group_by(time_day) %>%
-  summarise(count = n()) %>%
-  pad() %>%
-  fill_by_value(value = 0)
 
-mHour<- loc %>%
-  filter(who =="peter")%>%
+mHour<- all %>%
   select(time) %>%
-  mutate(time =  lubridate::force_tz(time, "Europe/Budapest")) %>%
-  top_n(207952)%>% 
+  as_tbl_time(index = time) %>%
+  time_filter(2000-01 ~ 2017-06) %>% 
   thicken('hour') %>%
   group_by(time_hour) %>%
   summarise(measurements = n()) %>%
@@ -48,10 +39,54 @@ missingDay<- ggplot(mHour, aes(x = time_hour_day, y=measurements, colour= hoursM
 
 ggsave(missingDay,filename = "img/missingdayPeter.png",device = "png",height = 6.5, units = "cm")
 
-pointsdaily<-loc %>% group_by(Day) %>% summarise(count = n())
-pointshourly<-loc %>%
-  group_by(Day, lubridate::hour(time)) %>%
-  summarise(count=n())
+accuracyData<- all %>%
+  as_tbl_time(time) %>%
+  select(time, accuracy)
 
-pointsdaily
-pointshourly
+hist(log(accuracyData$accuracy))
+## thing about this later, now do a dayplot of accuracy
+
+
+accuracyData<- all %>%
+  as_tbl_time(time) %>%
+  select(time, accuracy)%>%
+  time_filter(2016-02-05~2016-02-05)
+
+ggplot(accuracyData, aes(x = time, y=accuracy))+
+  geom_point()+theme_tufte()+
+  xlab("")+
+  ylab("Accuracy in Meters")+
+  ggtitle("Accuracy of measures in a day")
+
+
+  scale_colour_gradient2(low = muted("green"), mid = "grey50",
+                         high = muted("red"), midpoint = 12, space = "Lab",
[…]

### 2017-11-10 thesisReport.Rmd
started methods section

@@
 ---
-title: 'Research Report: Longitudinal models of human mobility'
+title: 'Handling missing data in smartphone location logs'
 author: "Boaz Sobrado"
@@
 Moreover, we will compare the performance of the model compared to currently used methods, such as linear interpolation.*
-
-**Keywords: GPS, missing data, human mobility,big data**
 
@@
 
+Missing: discussion of Barnett model.
+
 ## Methods
@@
 
-For December, it would be nice to have the Lit Rev section, plus an exploration of the problem.
-
 ### Data & Analyses
 
-The data used was collected between 2014 and 2017 on different Android devices.
-A total of X individuals contributed to the analysis, yielding a total of Y data points.
-Moreover we used diary information from **SOMEWHERE**.
+The data used was collected between 2014 and 2017 on different Android devices from X individuals across X continents. The table below provides more details:
 
-Analyses were performed using R and a multitude of other packages (@ggplot2 @dplyr @leaflet @sp @ggthemes @base @sp1 @sp2).
+```{r datadetailsTable}
+datadetails <- readRDS("datadetails.rds")
+kable(datadetails,digits = 2,col.names = c("Log duration",
+                                           "Observations",
+                                           "Missing days",
+                                           "Mean Accuracy",
+                                           "SD Accuracy"))
+```
 
-*Plots: data points over time*
+Analyses were performed using R and a multitude of other packages [@ggplot2;@dplyr;@leaflet; @ggthemes;@base;@sp1;@sp2].
+
+### Model evaluation
+Models were evaluated using a test set and a train set.
+During a period of 7 days participants wore a professional GPS device (TRX-100) which reported location every minute.
+In addition, they continued using their phones as normal.
+Missing data was imputed for the week for which more accurate data was available.
+
+Measures for which the accuracy was estimated include *barnett's path paper*.
+
+Like others (e.g. @palmius_detecting_2017) they detect invalid points using extreme speed.
+
+## Results
 
@@
 
-#### Filtering
-
-*Carlson et al 2015* describe a personal activity measurement system (PALMS), which can filter data points (among others). Their system is validated using cameras.
-
-Like others (e.g. @palmius_detecting_2017) they detect invalid points using extreme speed.
-
-*Google* offers an out of the box activity inferral mechanism.
-
-### Discussion on missing data
-
-Our missing data patterns fit the broader pattern of missing data reported by others [@palmius_detecting_2017;@saeb_mobile_2015; @grunerbl_smartphone-based_2015;@harari_using_2016].
-
-Missing data can  Importantly, one cannot say that the data is missing at random as there are a myriad of reasons why missingness can be related to location, such as turning off the phone on the plane before a long flight.
-
-### Models
-
-#### Spatiotemporal imputation methods
-
-## Results
-
 ## Discussion

### 2017-11-10 auxDataDetails.R

@@ -1,0 +1,18 @@
+## auxiliary data details
+
+# time span
+# observations
+# missing days
+# mean accuracy
+# sd accuracy
+
+# variables constructed in the playing around with thicken.R file
+
+datadetails<- data.frame(span = paste0("From ", min(mHour$time_hour_day), " to ", max(mHour$time_hour_day)),
+           observations = sum(mHour$measurements),
+           days_missing = nrow(mHour[which(mHour$hoursMissing == 24),]),
+           mean_accuracy = mean(all$accuracy),
+           sd_accuracy   = sd(all$accuracy)
+           )
+
+saveRDS(datadetails, "datadetails.rds")

### 2017-11-11 thesisReport.Rmd
added Palmius section as well as data table

@@
 # Abstract
-*There is a great interest in using objective location data to quantify the mobility characteristics of individuals.
+*There is a great interest in using objective location data to infer the mobility measures of individuals.
 Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals.
 However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment.
-In this paper we present a model which leverages the periodicity of human movement in order to impute missing data values.
+In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values.
 Moreover, we will compare the performance of the model compared to currently used methods, such as linear interpolation.*
@@
 
-## State of the art in temporospatial models
+## State of the art in spatiotemporal models
 
-Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and often serves different purposes.
-Given that @barnett_inferring_2016 is to our knowledge the only paper explicitly focusing on dealing with missingness for inferring missingness using GPS data the following section briefly illustrates the some methods used to solve measurement inaccuracy and missing data problems in the field, as well as their applicability to our question.
+Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and serves vastly different purposes.
+The following section briefly illustrates some methods used to deal with measurement inaccuracy and missing data problems in spatiotemporal data, as well as their applicability to our question.
+Moreover, we discuss in detail two approaches which deal explicitly with missing data in smartphone measured human location.
+
+Let $D$ element of $R^4$ be the recorded data, $x$ element of $R^3$ our actual place in time space.
 
@@
 
-While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostentibly by dividing the sample space into rasters which would be measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible to do.
+While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostentibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
 To our knowledge such models have not been implemented.
@@
 There is a vast literature of using state space models (SSMs) to improve measurements accuracy and deal with missing data.
-Behavioural ecologists for instance, have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008].
+Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008].
 These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
 The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
-The extended Kalman filter is the de facto standard for GPS measurements [@chen_state_2013].
+The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
+
+**Should I show the Kalman filter here?**
+In the Kalman filter the recorded data is modelled by a measurement equation.
 
@@
 Thus, the estimated location at timepoint $k$ is often based only upon measurements at $k$ and the previous timepoint $k-1$.
-This assumption ignores the periodic nature of human movement, whereby people generally spend the nights at home and the day at work.
+This may be suitable for ships at sea, but it ignores the highly periodic nature of human movement.
 Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or infeasible [@sadilek_far_2016].
@@
 
-Missing: discussion of Barnett model.
+### Linear interpolation
+
+@palmius_detecting_2017 deal with measuremement inaccuracy of $D$ by removing from the data set all unique low-accuracy data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
+Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
+Morever:
+
+>If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch.
+
+Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm.
+
+### Barnett's model
 
@@
 
-Like others (e.g. @palmius_detecting_2017) they detect invalid points using extreme speed.
-
 ## Results

### 2017-11-11 thesisReport.Rmd
added a lot of methods, but also some animation scripts

@@
 
-In most of these studies participants are given a specialised GPS devices or a smartphone with a custom made app.
-However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life and usually span a short amount of time.
-A solution is to take advantage of the already exisiting smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years[@location_history_timeline_nodate].
+In most of these studies participants are given a specialised GPS devices, resulting in accurate mobility GPS data (*specialised logs*).
+However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life.
+Because of this, specialised logs usually span a short amount of time.
+@barnett_inferring_2016 advocate installing custom-made tracking app on user's phones (*custom logs*).
+Another solution is to take advantage of exisiting smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years[@location_history_timeline_nodate] (*secondary logs*).
 These logs can be accessed and shared by users.
-Yet, because GPS sensors consume a significant amounts of battery, these logs can be sparse and inaccurate.
+Yet, because they were created for non-academic purposes under engineering constraints, these logs can be sparse and inaccurate.
 Hence, two important challenges are dealing with measurement noise and missing data.
 
-There is currently no golden standard in how to deal with missing data in this context [@barnett_inferring_2016].
-@jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it.
+There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016].
+@jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.
 Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
@@
 
-Given that there is next to no literature on missing data in smartphone location logs it is worth illustrating the typical characteristics of this data using an example data set.
-Moreover, although we could find no published papers on dealing with missing data in this narrow domain, there are however methods which deal with similar problems.
+Given that there is next to no literature on missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
+Moreover, although we could find no published papers on dealing with missing data in secondary logs, there are however methods which deal with similar problems.
 In this section we will describe the problems with the data and models which could be applied to it.
@@
 
-Even in professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
-Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
+In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
+GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
 Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
+
+Given that Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
+Based on this data set, less than 40% of measures fall with a 67% confidence radius of 20 meters.
+@palmius_detecting_2017 note that in their custom logs that inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+Similar patterns have been observed in secondary logs.
 
@@
 ```
-
-Given that Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
-Based on this data set, less than 40% of measures fall with a 67% confidence radius of 20 meters.
 
@@
 
-### Linear interpolation
+### Filtering & Mean imputation
 
@@
 ### Barnett's model
+
+@barnett_inferring_2016 deal with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. They handle missing data by:
+
+> simulat[ing] flights and pauses over the period of missingness where the direction,
+duration, and spatial length of each flight, the fraction of flights versus the fraction of
+pauses, and the duration of pauses are sampled from observed data
+
+This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses.
+In other words, for a given missing period, the individual's mobility can be estimated based on  measured movements in that area, at that point in time or movements in the last 24 hours.
 

### 2017-11-11 scripts/accuracyPlots.R

@@ -1,0 +1,105 @@
+## plots on accuracy
+# missing days with padr
+library(dplyr)
+library(padr)
+library(ggplot2)
+library(tibbletime)
+# read data
+all<-readRDS("../../data/peter/myLocationHistory.rds")
+all<-readRDS("../../data/boaz/myLocationHistory.rds")
+
+all$time<- lubridate::with_tz(all$time, "Europe/Budapest")
+
+accData <- all %>% select(accuracy) %>% filter (accuracy < 20000)
+accuracy <- data.frame(accuracy = accData$accuracy,
+                       group = ifelse(accData$accuracy < 800,"High",
+                                      ifelse(accData$accuracy < 5000, "Middle",
+                                             "Low")))
+
+accuracy$group <- factor(accuracy$group, levels = c("High", "Middle", "Low"))
+
+accuracyPlot<- ggplot(accuracy, aes(x = accuracy, fill = group)) + 
+  geom_histogram() + 
+  facet_grid(group ~ ., scales="free") + 
+  theme_tufte() +
+  theme(
+    legend.position = "none",
+    strip.placement = "outside",
+    strip.background = element_blank(),
+    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5)
+  ) +
+  labs(
+    x = "Accuracy in metres",
+    y = "Count",
+    title = "Location Log Accuracy"
+  )+scale_fill_manual(values=c(muted("green"),"grey50",muted("red")))
+
+
+ggsave(accuracyPlot,filename = "img/accuracyPeter.png",device = "png",height = 6.5, units = "cm")
+
+#similar to the plot with distance from home and accuracy from palmius
+library(leaflet)
+
+all %>%
+  select(time,lat,lon,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-02-15 ~ 2017-02-15) %>%
+  filter(accuracy>0) %>%
+  leaflet() %>%
+  addTiles() %>%
+  addMarkers(~lon, ~lat)
+
+# extract home
+#52.10421 5.113919 
+home<- c(5.113919,52.10421)
+#calculate distance from home
+
+boazday0218<- all %>%
+  select(time,lon,lat,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-02-15 ~ 2017-02-15)%>%
+  mutate( distanceHome = sp::spDistsN1(pts=matrix(c(lon,lat),ncol = 2),
[…]

### 2017-11-11 scripts/js/Rplot.js

@@ -1,0 +1,13 @@
+(function($) {
+    $(document).ready(function() {
+	
+	$('#Rplot').scianimator({
+	    'images': ['images/Rplot1.png', 'images/Rplot2.png', 'images/Rplot3.png', 'images/Rplot4.png', 'images/Rplot5.png', 'images/Rplot6.png', 'images/Rplot7.png', 'images/Rplot8.png', 'images/Rplot9.png', 'images/Rplot10.png', 'images/Rplot11.png', 'images/Rplot12.png', 'images/Rplot13.png', 'images/Rplot14.png', 'images/Rplot15.png', 'images/Rplot16.png', 'images/Rplot17.png', 'images/Rplot18.png', 'images/Rplot19.png', 'images/Rplot20.png', 'images/Rplot21.png', 'images/Rplot22.png', 'images/Rplot23.png', 'images/Rplot24.png', 'images/Rplot25.png', 'images/Rplot26.png', 'images/Rplot27.png', 'images/Rplot28.png', 'images/Rplot29.png', 'images/Rplot30.png', 'images/Rplot31.png', 'images/Rplot32.png', 'images/Rplot33.png', 'images/Rplot34.png', 'images/Rplot35.png', 'images/Rplot36.png', 'images/Rplot37.png', 'images/Rplot38.png', 'images/Rplot39.png', 'images/Rplot40.png', 'images/Rplot41.png', 'images/Rplot42.png', 'images/Rplot43.png', 'images/Rplot44.png', 'images/Rplot45.png', 'images/Rplot46.png', 'images/Rplot47.png', 'images/Rplot48.png', 'images/Rplot49.png', 'images/Rplot50.png', 'images/Rplot51.png', 'images/Rplot52.png', 'images/Rplot53.png', 'images/Rplot54.png', 'images/Rplot55.png', 'images/Rplot56.png', 'images/Rplot57.png', 'images/Rplot58.png', 'images/Rplot59.png', 'images/Rplot60.png', 'images/Rplot61.png', 'images/Rplot62.png', 'images/Rplot63.png', 'images/Rplot64.png', 'images/Rplot65.png', 'images/Rplot66.png', 'images/Rplot67.png', 'images/Rplot68.png', 'images/Rplot69.png', 'images/Rplot70.png', 'images/Rplot71.png', 'images/Rplot72.png', 'images/Rplot73.png', 'images/Rplot74.png', 'images/Rplot75.png', 'images/Rplo
[…]

### 2017-11-13 thesisReport.Rmd
added Kalman filter equations and resolved Peter's comments

@@
   rmarkdown::pdf_document:
+    citation_package: natbib
     fig_caption: yes        
@@
 # Abstract
-*There is a great interest in using objective location data to infer the mobility measures of individuals.
+*Using objective location data to infer the mobility measures of individuals is highly desirable, but methodologically difficult.
 Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals.
@@
 
-Objective data on human mobility has become available throgh the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
+Objective data on human mobility has become available through the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
 Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017].
 
-In most of these studies participants are given a specialised GPS devices, resulting in accurate mobility GPS data (*specialised logs*).
+In most of these studies participants are given a specialised devices, resulting in accurate mobility GPS data (*specialised logs*).
 However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life.
@@
 @barnett_inferring_2016 advocate installing custom-made tracking app on user's phones (*custom logs*).
-Another solution is to take advantage of exisiting smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years[@location_history_timeline_nodate] (*secondary logs*).
+Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_nodate] (*secondary logs*).
 These logs can be accessed and shared by users.
-Yet, because they were created for non-academic purposes under engineering constraints, these logs can be sparse and inaccurate.
+Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate.
 Hence, two important challenges are dealing with measurement noise and missing data.
@@
 @palmius_detecting_2017] can exacerbate the severity of the missing data problem.
-On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.
+Traditional missing data methods, such as multiple imputation, cannot be used easily due to the geospatial nature of the data.
+On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as underestimating the movement of individuals.
 In this paper we will compare methods used to deal with measurement error and missing data in location information.
@@
 Given that there is next to no literature on missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
-Moreover, although we could find no published papers on dealing with missing data in secondary logs, there are however methods which deal with similar problems.
-In this section we will describe the problems with the data and models which could be applied to it.
+Moreover, although we could find no published papers on dealing with missing data in secondary logs, there are methods which deal with similar problems.
+In this section we will describe the problems with the data and why they arise as well as the models which could be applied to it.
 
@@
 
-This example data set spans 3 years from January 2013 to January 2017.
+This example data set comes from Google Location History and spans 3 years from January 2013 to January 2017 with multiple different Android devices.
 It contains 814 941 measurements, with approximately 742 measurements per day (*SD*=868.15).
-Moreover, it also contains estimates of velocity, altitude, heading and activity (e.g.
-in a vehicle, still).
-For the purposes of this paper we will focus only on lattitude, longitude and time.
+The data set contains a wide range of variables.
+For the purposes of this paper we will focus only on latitude, longitude, accuracy and time.
+Accuracy is defined as the size in meters of the radius of a circle such that 67% of the measurements fall within it.
 
-### Accuracy
+### Accuracy in location logs
 
@@
 
-Given that Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
+Given that Android phones collect location information from WiFi access points, cellphone triangulation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
 Based on this data set, less than 40% of measures fall with a 67% confidence radius of 20 meters.
-@palmius_detecting_2017 note that in their custom logs that inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
-Similar patterns have been observed in secondary logs.
+@palmius_detecting_2017 note that in their custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+I observe similar patterns in secondary logs.
 
@@
 include_graphics("img/missingdayPeter.png")# from the playingaroundwiththicken.r file
+# to do: make into 5 minute slots
+
 ```
@@
 
-Let $D$ element of $R^4$ be the recorded data, $x$ element of $R^3$ our actual place in time space.
+Let $d \in \mathbb{R}^3$ be the recorded data containing lattitude, longitude, accuracy and $l \in \mathbb{R}^3$ be actual location in space.
+$d_k$ and $l_k$ denote the measured location and the true location at time $k$.
 
@@
 
-Given fixed measurement stations there are sevaral imputation methods for spatiotemporal measurements.
+Given fixed measurement stations there are several imputation methods for spatiotemporal measurements.
 For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
@@
 
-While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostentibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
+While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
 To our knowledge such models have not been implemented.
@@
 
-**Should I show the Kalman filter here?**
-In the Kalman filter the recorded data is modelled by a measurement equation.
+The state equation of the Kalman filter models the current location:
+
+$$
+l_k = Al_{k-1} + Bu_{k-1}+w_{k-1}
+$$
+Here $w_k$ is known as the process noise and is conceptually equivalent to the error term in a linear regression.
+$A$ is a square matrix of order $n$ which relates our current state to the previous state.
+$B$ is an $n \times x$ matrix which relates the optional control input $u \in \mathbb{R}^x$ to the state $l$.
+
+The measurement equation models the measurements at any given time:
+
+$$
+d_k = H l_k +v_k
+$$
+
+$v_k$ is known as the measurement noise and represents the errors in measurement and $H$ is an $m \times n$  matrix relating the state to the measurement.
+
+The noise in both equations assumed to be:
+
+$$p(w) \sim N(0,Q)$$
+
+$$p(v) \sim N(0,R)$$
+Where $Q$ is the process noise covariance matrix and $R$ is the measurement noise covariance.
+
+Then, let $\widehat{l}^{-}_k \in \mathbb{R}^n$ be the a priori state estimate given our knowledge of the process and $\widehat{l}_k \in \mathbb{R}^n$ be the a posteriori state estimate at step $k$ given measurement $d_k$.
+These two state estimates have error covariances denoted $P^{-}_k$ and $P_k$ respectively.
+Algebraically the Kalman filter can then be written as:
+
+$$
+\widehat{x}_k = \widehat{x}^{-}_k + K (z_k - H \widehat{x}^{-}_k)
+$$
+
+Where $K$ (the weight or Kalman gain) is a square matrix order $n \times m$ matrix which minimizes the posteriori error covariance.
 
 The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
-For our pourposes the main limitation is that these models are based on the Markov property.
+For our purposes the main limitation is that these models are based on the Markov property.
 Thus, the estimated location at timepoint $k$ is often based only upon measurements at $k$ and the previous timepoint $k-1$.
 This may be suitable for ships at sea, but it ignores the highly periodic nature of human movement.
-Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or infeasible [@sadilek_far_2016].
+Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or unfeasible [@sadilek_far_2016].
 
@@
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
-Morever:
+Moreover:
 
@@
 
+This paper is to the best of our knowledge the only attempt at establishing a principled approach to this problem
+
 ## Methods
 
-*An asside as to my conversation with Peter on 07.11*
-
-We can get some trackers on people to measure movement constantly and be a "golden standard" to evaluate the models.
-
-### Data & Analyses
+### Datasets & Analyses
 
@@
 
-```{r datadetailsTable}
-datadetails <- readRDS("datadetails.rds")
+```{r datadetailsTable, echo = F}
+datadetails <- readRDS("tempdata/datadetails.rds")
 kable(datadetails,digits = 2,col.names = c("Log duration",
@@
 
+In addition to the secondary logs, participants also volunteered to carry with them a specialised GPS tracker for week. Therefore we also used a week of parallel secondary and specialised logs.
+
 Analyses were performed using R and a multitude of other packages [@ggplot2;@dplyr;@leaflet; @ggthemes;@base;@sp1;@sp2].
 
-### Model evaluation
-Models were evaluated using a test set and a train set.
-During a period of 7 days participants wore a professional GPS device (TRX-100) which reported location every minute.
-In addition, they continued using their phones as normal.
-Missing data was imputed for the week for which more accurate data was available.
+### Data pre-processing
 
-Measures for which the accuracy was estimated include *barnett's path paper*.
+The data was discretised into 5 minute windows following @palmius_detecting_2017 (as described above, for more detail please consult the original paper).
+
+### Imputation methods
+
+Four imputation methods were selected in order to cover techniques applied in the literature.
+Briefly, the filtered mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016].
+Simple Kalman filtered linear interpolation was used as a benchmark models.
+
+### Evaluation criteria
+
+The entire length of the secondary logs were used as a training set.
+The specialised logs were used as a test set.
+The missing data imputation models were evaluated both directly, and on on two computed measures: amount of trips made and distance traveled.
+
+Measures of performance used were root mean square error (RMSE) and mean absolute error (MAE).
 

### 2017-11-13 scripts/accuracyPlots.R

@@ -87,13 +87,13 @@ box<- c(left = min(boazday0218$lon)-0.01,
 library(ggmap)
 map<- get_stamenmap(bbox=box,zoom = 15, maptype = "toner-lite")
 q<- ggmap(map)+
-  geom_point(aes(lon,lat,frame = time2,colour = accuracy),data = boazday0218)
+  geom_point(aes(lon,lat,frame = time2,colour = accuracy),data = boazday0218)+xlab("")+ylab("")
 
 gganimate(q,"output.gif",interval = .2,ani.width=1000, ani.height=800)
 library(magick)
 
-library(animation)
-gganimate(q, ani.options(convert = 'path/to/convert.exe'))
+gganimate(q,"output.mp4",interval = .1,ani.width=500, ani.height=400, ani.res = 300)
+
 
 install.packages("installr")
 library(installr); install.ImageMagick(URL = "http://www.imagemagick.org/script/download.php")

### 2017-11-14 scripts/missingDays.R
updated missingness plot to 5 minute intervals

@@ -1,73 +1,0 @@
-##Graphs for determining where missing data is
-library(ggthemes)
-library(dplyr)
-
-# would be a lot easier to do with padr
-
-# read data
-loc<- readRDS("..\\all.rds")
-
-
-pointsdaily<-loc %>% group_by(Day) %>% summarise(count = n())
-pointshourly<-loc %>% group_by(Day, lubridate::hour(time)) %>%
-  summarise(count=n())
-rm(loc)
-
-colnames(pointshourly) <- c("Day","Hour","count")
-
-#generate dates
-x<-as.Date(pointsdaily$Day[1]):as.Date(pointsdaily$Day[nrow(pointsdaily)])
-x<-as.Date.numeric(x = x,origin="1970-01-01")
-h<- sort(rep(0:23,length(x)))
-allDayHour <- data.frame(Day = rep(x,24), Hour = h, count = 0)
-
-#dates with zero measurements
-noMeasureDay<-as.Date(setdiff(x,pointsdaily$Day), origin = "1970-01-01")
-# alternative approach
-
-u<- unique(rbind(pointshourly[,-3],allDayHour[,-3]))
-allDayHourPasted<- paste0(allDayHour$Day,"$",allDayHour$Hour)
-measuredDayHourPasted <- paste0(pointshourly$Day,"$",pointshourly$Hour)
-noMeasureHour<- setdiff(allDayHourPasted,measuredDayHourPasted)
-noMeasureHour<-strsplit(noMeasureHour,split = "\\$")
-#hour with zero measurements
-
-noMeasureHour <- setNames(do.call(rbind.data.frame, noMeasureHour), c("Day", "Hour"))
-noMeasureHour$count <- 0
-noMeasureHour$Day <- as.Date(noMeasureHour$Day)
-noMeasureHour$Hour <- as.numeric(noMeasureHour$Hour)
-
-t<- data.frame(table(noMeasureHour$Day))
-nm<- filter(t,Freq != 24 & Freq>=6) %>% select(Var1)
-nm<- as.Date(as.character(nm$Var1))
-
-ps<-rbind(pointsdaily,data.frame(Day = noMeasureDay,count = 0))
-ps$type <- "More than 6 hours"
-ps[which(ps$Day%in%nm),"type"] <- "6 hours or less"
-ps[which(ps$Day%in%noMeasureDay),"type"] <- "No Measure"
[…]

### 2017-11-14 scripts/playingAroundwithThicken.R

@@ -1,92 +1,0 @@
-# missing days with padr
-library(dplyr)
-library(padr)
-library(ggplot2)
-library(tibbletime)
-# read data
-all<-readRDS("../data/peter/myLocationHistory.rds")
-attr(all$time, "tzone") <- "Europe/Paris"
-
-
-mHour<- all %>%
-  select(time) %>%
-  as_tbl_time(index = time) %>%
-  time_filter(2000-01 ~ 2017-06) %>% 
-  thicken('hour') %>%
-  group_by(time_hour) %>%
-  summarise(measurements = n()) %>%
-  pad() %>%
-  fill_by_value(value = 0) %>%
-  thicken("day") %>%
-  mutate(
-    type = case_when(
-      measurements== 0 ~ 1,
-      measurements > 0 ~ 0)) %>%
-  group_by(time_hour_day) %>%
-  summarise(hoursMissing = sum(type),
-            measurements = sum(measurements))
-
-library(scales)
-missingDay<- ggplot(mHour, aes(x = time_hour_day, y=measurements, colour= hoursMissing))+
-  geom_point()+theme_tufte()+
-  xlab("")+
-  ylab("Measurements")+
-  ggtitle("Amount of daily measurements over time")+
-  scale_colour_gradient2(low = muted("green"), mid = "grey50",
-                         high = muted("red"), midpoint = 12, space = "Lab",
-                         na.value = "grey50", guide = "colourbar")+
-  labs(x = NULL, colour = "Missing Hours")
-
-ggsave(missingDay,filename = "img/missingdayPeter.png",device = "png",height = 6.5, units = "cm")
-
-accuracyData<- all %>%
-  as_tbl_time(time) %>%
-  select(time, accuracy)
-
-hist(log(accuracyData$accuracy))
-## thing about this later, now do a dayplot of accuracy
-
-
-accuracyData<- all %>%
-  as_tbl_time(time) %>%
-  select(time, accuracy)%>%
-  time_filter(2016-02-05~2016-02-05)
-
-ggplot(accuracyData, aes(x = time, y=accuracy))+
-  geom_point()+theme_tufte()+
-  xlab("")+
-  ylab("Accuracy in Meters")+
-  ggtitle("Accuracy of measures in a day")
-
-
-  scale_colour_gradient2(low = muted("green"), mid = "grey50",
-                         high = muted("red"), midpoint = 12, space = "Lab",
-                         na.value = "grey50", guide = "colourbar")+
-  labs(x = NULL, colour = "Missing Hours")
-
[…]

### 2017-11-16 thesisReport.Rmd
updated missingness plot to 5 minute intervals
added two new plots plus an unfinished plot

@@
 Low accuracy measures are often near in time to more high accuracy measures.",out.width = '100%'}
-
-include_graphics("img/accuracyPeter.png")
+#
+include_graphics("img/journeyTillMiddayBoaz.png")
 
@@
 The fill of the points shows the amount of hours of captured data each day.",out.width = '100%'}
-
-include_graphics("img/missingdayPeter.png")# from the playingaroundwiththicken.r file
-# to do: make into 5 minute slots
+#fix width
+include_graphics("img/missingBoaz5minExample.png")# from the missingadataplot.R script
 

### 2017-11-16 scripts/missingnessPlot.R

@@ -4,6 +4,7 @@ library(padr)
 library(ggplot2)
 library(tibbletime)
 library(ggthemes)
+library(scales)
 # read data
 all<-readRDS("../../data/boaz/myLocationHistory.rds")
 attr(all$time, "tzone") <- "Europe/Paris"
@@ -75,12 +76,13 @@ ggsave(missingDay5min,filename = "../img/missingdayPeter5min.png",device = "png"
 
 # by 5 minute version 2
 m5minv2<- all %>%
-  select(time) %>%
+  select(time,accuracy) %>%
   as_tbl_time(index = time) %>%
   time_filter(2000-01 ~ 2017-06) %>% 
   thicken('5 min') %>%
   group_by(time_5_min) %>%
-  summarise(measurements = n()) %>%
+  summarise(measurements = n(),
+            accuracy = mean(accuracy)) %>%
   pad() %>%
   fill_by_value(value = 0)%>%
   mutate(
@@ -93,13 +95,26 @@ m5minv2<- all %>%
       measurements > 0 ~ 0)
       )
 
-m5minv2 %>%
+exampleMiss<- m5minv2 %>%
   as_tbl_time(index = time2)%>%
   time_filter(2017-02-15 ~ 2017-02-15) %>%
-  ggplot( aes(x = hour, y=measurements, colour = missing))+
+  ggplot( aes(x = hour, y=measurements, colour = factor(missing)))+
   geom_point()+theme_tufte()+
   xlab("")+
   ylab("Measurements")+
-  ggtitle("Missingness in daily measurements over time")
+  ggtitle("Measurements per 5 minute window")+
+  scale_x_datetime(breaks = date_breaks("4 hour"),
+                   minor_breaks=date_breaks("2 hour"),
+                   labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
+  scale_colour_manual(values=c("black","#DF2935"))+
+  theme(legend.position="none")
+
+ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, units = "cm")
+
+
+time1_p1 <- strptime(paste("2017-02-14", "00:00:00"), "%Y-%m-%d %H:%M:%S")
+time2_p1 <- strptime(paste("2017-02-16", "24:00:00"), "%Y-%m-%d %H:%M:%S")
[…]

### 2017-11-16 thesisReport.Rmd
added notation to the beginning, removed SSM equations

@@
 
-## Missing data and noise in location logs
+## Location logs and notation
 
-This example data set comes from Google Location History and spans 3 years from January 2013 to January 2017 with multiple different Android devices.
-It contains 814 941 measurements, with approximately 742 measurements per day (*SD*=868.15).
-The data set contains a wide range of variables.
+This example dataset comes from Google Location History and spans 3 years from January 2013 to January 2017 with multiple different Android devices.
+It contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15).
+The dataset contains a wide range of variables.
 For the purposes of this paper we will focus only on latitude, longitude, accuracy and time.
-Accuracy is defined as the size in meters of the radius of a circle such that 67% of the measurements fall within it.
+
+Borrowing from @barnett_inferring_2016 let a persons' true Cartesian location be $G(t) = [G_x(t) G_y(t)]$ and $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+In addition, let $a$ denote the estimated accuracy of the recorded data.
+accuracy.
+$G(t)$, $D$ and $a$ are indexed by the set $t_1 < ...
+< t_{n+1}$.
+For simplicity, let each entry in the discrete index set represent a 5 minute window.
+The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
+If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
 
@@
 
-In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
+In professional grade GPS trackers less than 80% of measurements $D$ fall within 10 meters of the true location $G(t)$.
 GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
@@
 
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement in a morning journey.
+The red circles denote the accuracy of all logged measurement points (the raw data).
+The points connected in time are connected by a line.
+The blue line shows the path without the most innaccurate (accurace > 400 meters) points filtered out.
+The red line shows the path with all measurements included.
+",out.width = '100%'}
+include_graphics("img/journeyTillMiddayBoaz.png")
+```
+
 Given that Android phones collect location information from WiFi access points, cellphone triangulation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
-Based on this data set, less than 40% of measures fall with a 67% confidence radius of 20 meters.
+Based on this data set, less than 40% of measures $D$ are paired with $a \leq 20$.
 @palmius_detecting_2017 note that in their custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
 I observe similar patterns in secondary logs.
-
-```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement.
-The x-axis denotes the accuracy in meters such that the user's actual location is expected to be within that radius with 67 percent confidence.
-Measures were classified into high accuracy (under 800m), medium accuracy (under 5000m) and low accuracy (over 5000m).
-Measures with accuracy of over 20000m were omitted from the graph.
-The maximum accuracy measure was over 40000m.
-Low accuracy measures are often near in time to more high accuracy measures.",out.width = '100%'}
-#
-include_graphics("img/journeyTillMiddayBoaz.png")
-
-```
 
@@
 
-```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data over time for the author.
-The x-axis denotes time, the y-axis shows how many measurements are made and each point is a day.
-The fill of the points shows the amount of hours of captured data each day.",out.width = '100%'}
+```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day.
+The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
+For this day there were several periods with no information.",out.width = '100%'}
 #fix width
-include_graphics("img/missingBoaz5minExample.png")# from the missingadataplot.R script
+include_graphics("img/missingBoaz5minExample.png")
 
@@
 Moreover, we discuss in detail two approaches which deal explicitly with missing data in smartphone measured human location.
-
-Let $d \in \mathbb{R}^3$ be the recorded data containing lattitude, longitude, accuracy and $l \in \mathbb{R}^3$ be actual location in space.
-$d_k$ and $l_k$ denote the measured location and the true location at time $k$.
 
@@
 
-The state equation of the Kalman filter models the current location:
-
-$$
-l_k = Al_{k-1} + Bu_{k-1}+w_{k-1}
-$$
-Here $w_k$ is known as the process noise and is conceptually equivalent to the error term in a linear regression.
-$A$ is a square matrix of order $n$ which relates our current state to the previous state.
-$B$ is an $n \times x$ matrix which relates the optional control input $u \in \mathbb{R}^x$ to the state $l$.
-
-The measurement equation models the measurements at any given time:
-
-$$
-d_k = H l_k +v_k
-$$
-
-$v_k$ is known as the measurement noise and represents the errors in measurement and $H$ is an $m \times n$  matrix relating the state to the measurement.
-
-The noise in both equations assumed to be:
-
-$$p(w) \sim N(0,Q)$$
-
-$$p(v) \sim N(0,R)$$
-Where $Q$ is the process noise covariance matrix and $R$ is the measurement noise covariance.
-
-Then, let $\widehat{l}^{-}_k \in \mathbb{R}^n$ be the a priori state estimate given our knowledge of the process and $\widehat{l}_k \in \mathbb{R}^n$ be the a posteriori state estimate at step $k$ given measurement $d_k$.
-These two state estimates have error covariances denoted $P^{-}_k$ and $P_k$ respectively.
-Algebraically the Kalman filter can then be written as:
-
-$$
-\widehat{x}_k = \widehat{x}^{-}_k + K (z_k - H \widehat{x}^{-}_k)
-$$
-
-Where $K$ (the weight or Kalman gain) is a square matrix order $n \times m$ matrix which minimizes the posteriori error covariance.
-
 The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
 For our purposes the main limitation is that these models are based on the Markov property.
-Thus, the estimated location at timepoint $k$ is often based only upon measurements at $k$ and the previous timepoint $k-1$.
+Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
 This may be suitable for ships at sea, but it ignores the highly periodic nature of human movement.
@@
 
-@palmius_detecting_2017 deal with measuremement inaccuracy of $D$ by removing from the data set all unique low-accuracy data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
+@palmius_detecting_2017 deal with measuremement inaccuracy of $D$ by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.

### 2017-11-16 scripts/missingnessPlot.R

@@ -109,7 +109,7 @@ exampleMiss<- m5minv2 %>%
   scale_colour_manual(values=c("black","#DF2935"))+
   theme(legend.position="none")
 
-ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, units = "cm")
+ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")
 
 
 time1_p1 <- strptime(paste("2017-02-14", "00:00:00"), "%Y-%m-%d %H:%M:%S")

### 2017-11-17 thesisReport.Rmd
added plots and table

@@
 
-Borrowing from @barnett_inferring_2016 let a persons' true Cartesian location be $G(t) = [G_x(t) G_y(t)]$ and $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+Borrowing from @barnett_inferring_2016 let a persons' true Cartesian location on an error minimising projection be $G(t) = [G_x(t) G_y(t)]$ and $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
 In addition, let $a$ denote the estimated accuracy of the recorded data.
@@
 Given that Android phones collect location information from WiFi access points, cellphone triangulation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
-Based on this data set, less than 40% of measures $D$ are paired with $a \leq 20$.
+Based on this data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
 @palmius_detecting_2017 note that in their custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
 I observe similar patterns in secondary logs.
+
+```{r, accuracyPlot2, echo=FALSE, fig.cap="The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
+The colour scale shows the distance between the measurement and the previous point.
+To ease interpretability all points with a distance between points of less than 5 meters have been filtered out.",out.width = '100%'}
+include_graphics("img/accuracyLocShift.png")
+
+```
 
@@
 @harari_using_2016;@jankowska_framework_2015].
+
+In our example data set, over 54% of the data is missing for the entire duration of the log.
+However,this is misleading as there are several long periods with no measurements whatsoever.
+The typical strucutre of missingness of a day with measurements is shown below.
 
@@
 
-```{r datadetailsTable, echo = F}
-datadetails <- readRDS("tempdata/datadetails.rds")
+```{r datadetailsTable, echo = F, fig.cap="caption here?"}
+datadetails <- readRDS("tempdata/datadescriptives.rds")
+datadetails <- datadetails[,-7]
 kable(datadetails,digits = 2,col.names = c("Log duration",
+                                           "Days of log",
                                            "Observations",
                                            "Missing days",
-                                           "Mean Accuracy",
-                                           "SD Accuracy"))
+                                           "Missing data",
+                                           "Mean Accuracy"))
 ```

### 2017-11-17 scripts/accuracyPlots.R

@@ -10,6 +10,7 @@ all<-readRDS("../../data/boaz/myLocationHistory.rds")
 
 all$time<- lubridate::with_tz(all$time, "Europe/Budapest")# don't forget to run
 
+#discarded histogram like plot
 
 accData <- all %>% select(accuracy) %>% filter (accuracy < 20000)
 accuracy <- data.frame(accuracy = accData$accuracy,
@@ -73,24 +74,34 @@ all %>%
 #52.10421 5.113919 
 home<- c(5.113919,52.10421)
 #calculate distance from home
-
-all %>%
+filter(distancePrev >.005) %>%
+  
+locShift<- all %>%
   select(time,lon,lat,accuracy) %>%
   as_tbl_time(index = time) %>%
   time_filter(2017-02-15 ~ 2017-02-15)%>%
   mutate( distanceHome = sp::spDistsN1(pts=matrix(c(lon,lat),ncol = 2),
-          pt = home,
-          longlat = T),
+                                       pt = home,
+                                       longlat = T),
           euclon = lon-home[1],
           euclat = lat-home[2],
-          time2 = lubridate::force_tz(time, "Asia/Singapore")) %>% 
-  ggplot( aes(x = time2, y=accuracy))+
+          time2 = lubridate::force_tz(time, "Asia/Singapore"),
+          distancePrev = c(0,
+                           sp::spDists(x=matrix(c(lon,lat),ncol = 2), segments = T))) %>%
+  ggplot( aes(x = time2, y=accuracy, colour = distancePrev*1000))+
   geom_point()+theme_tufte()+
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
                    labels=date_format("%H:%M:%S", tz = "Asia/Singapore"),
-                   limits = xlim_p1)
+                   limits = xlim_p1)+
+  xlab("")+ylab("Accuracy")+
+  scale_colour_gradient2(low = "#DF2935", mid = "grey50",
+                         high = "#18206F", space = "Lab",
+                         na.value = "grey50", guide = "colourbar")+
+  theme(legend.position = c(0.17, 0.85), legend.direction = "horizontal") +
+  labs(x = NULL, colour = "Distance from\nprevious point")
 
[…]

### 2017-11-17 scripts/auxDataDetails.R

@@ -8,11 +8,85 @@
 
 # variables constructed in the playing around with thicken.R file
 
-datadetails<- data.frame(span = paste0("From ", min(mHour$time_hour_day), " to ", max(mHour$time_hour_day)),
-           observations = sum(mHour$measurements),
-           days_missing = nrow(mHour[which(mHour$hoursMissing == 24),]),
-           mean_accuracy = mean(all$accuracy),
-           sd_accuracy   = sd(all$accuracy)
-           )
+# function
 
-saveRDS(datadetails, "datadetails.rds")
+extractDescriptives<- function(PATH="../../data/boaz/myLocationHistory.rds"){
+  all<-readRDS(PATH)
+  
+  all <- all %>% select(time,accuracy,lat,lon)
+  
+  # begin descriptive construction
+  
+  # n observations
+  observations <- all %>% nrow()
+  
+  # span
+  spanDf<- all %>%
+    as_tbl_time(index = time) %>%
+    time_filter( 2000  ~ 2018) %>% 
+    summarise(max = as.Date(max(time)),
+              min = as.Date(min(time)))
+  
+  span <- paste0("From ", spanDf[["min"]], " to ", spanDf[["max"]])
+  
+  # range in days
+  range <- spanDf[["max"]]- spanDf[["min"]]
+  
+  # accuracy
+  meanAc <- mean(all$accuracy)
+  sdAc <- sd(all$accuracy)
+  
+  # missing periods of 5 min
+  miss5min<- all %>%
+    select(time) %>%
+    as_tbl_time(index = time) %>%
+    time_filter(2000-01 ~ 2017-06) %>% 
+    thicken('5 min') %>%
+    group_by(time_5_min) %>%
+    summarise(measurements = n()) %>%
+    pad() %>%
+    fill_by_value(value = 0) %>%
+    thicken("day") %>%
+    mutate(
+      missing = case_when(
+        measurements== 0 ~ 1,
+        measurements > 0 ~ 0)) %>%
+    group_by(time_5_min_day) %>%
+    summarise(missing = sum(missing)/288)%>%
+    filter(missing != 1) %>% 
+    summarise(missing = mean(missing))
+  
+  missDay<- all %>%
+    select(time) %>%
+    as_tbl_time(index = time) %>%
+    time_filter(2000-01 ~ 2017-06) %>% 
+    thicken('day') %>%
+    group_by(time_day) %>%
+    summarise(measurements = n()) %>%
[…]

### 2017-11-17 thesisReport.Rmd
added a bit of meat to the methods section

@@
 
-The data used was collected between 2014 and 2017 on different Android devices from X individuals across X continents. The table below provides more details:
+The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals. The table below provides more details:
 
-```{r datadetailsTable, echo = F, fig.cap="caption here?"}
+```{r datadetailsTable, echo = F}
 datadetails <- readRDS("tempdata/datadescriptives.rds")
-datadetails <- datadetails[,-7]
+datadetails <- datadetails[,-7] #remove last column for now
 kable(datadetails,digits = 2,col.names = c("Log duration",
-                                           "Days of log",
+                                           "Logged days",
                                            "Observations",
@@
                                            "Missing data",
-                                           "Mean Accuracy"))
+                                           "Mean Accuracy"),
+      caption = "Table with descriptives about the data sets used to build the imputation methods. ")
 ```
 
-In addition to the secondary logs, participants also volunteered to carry with them a specialised GPS tracker for week. Therefore we also used a week of parallel secondary and specialised logs.
+In addition to the secondary logs, participants also volunteered to carry with them a specialised GPS tracker for a week. This specialised log was used to evaluate the models.
 
@@
 
-The data was discretised into 5 minute windows following @palmius_detecting_2017 (as described above, for more detail please consult the original paper).
+Three different filtering methods were tested.
+First, the filtered rolling-median downsampling method described by @palmius_detecting_2017.
+The second method is a weighted mean approach taking $f(a)$ as a weight.
+Finally, an extended Kalman filter.
+The output of all of these methods was taken as the input of the imputation methods.
 
@@
 Four imputation methods were selected in order to cover techniques applied in the literature.
-Briefly, the filtered mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016].
-Simple Kalman filtered linear interpolation was used as a benchmark models.
+Briefly, the mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016].
+Simple linear interpolation was used as a benchmark model.
 
@@
 
-Measures of performance used were root mean square error (RMSE) and mean absolute error (MAE).
+The direct evaluation involved calculating the error of each $D_t$ compared to $G(t)$ approximated by the specialised log.
+The error measures used were root mean square error (RMSE) and mean absolute error (MAE).
+
+The evaluation on computed measures involved calculating a mobility trace following the rectangual method of @rhee_human_2007 for each imputed dataset.
+Like @barnett_inferring_2016 I calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data.
+For simulation-based imputation approaches a mean value over 100 samples was taken.
 
 ## Results
-
-![Raw GPS measurements of a journey from de Uithof to Tuinwijk on February 17th 2017.
-The measurements are in red, the filtered path is in blue.
-The circles denote 67% confidence intervals of the given GPS measurement.
-Measurements and fitted points which follow each other in time are connected by lines.
-The inacurate measurements lead to estimates of irregular movements.
-The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](img/Map1.png)
 

### 2017-11-24 thesisReport.Rmd
implemented a little feedback

@@
 In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values.
-Moreover, we will compare the performance of the model compared to currently used methods, such as linear interpolation.*
+Moreover, we will assess the performance of the model relative to currently used methods, such as linear interpolation.*
 
@@
 
-How active people are and how they interact with their environment affects a wide range of measures including health , income and social capital [@goodchild_toward_2010].
+How active people are and how they interact with their environment affects a wide range of outcomes including health, income and social capital [@goodchild_toward_2010].
 A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies.
@@
 
-Objective data on human mobility has become available through the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location.
-Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017].
+These obstacles can be overcome by using objective data on human mobility.
+The Global Positioning System (GPS), which uses the distance between a device and a number of satellites to determine location, provides such data.
+Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality [@harari_using_2016], academic performance [@wang_smartgpa:_2015] and bipolar disorder [@palmius_detecting_2017].
 
-In most of these studies participants are given a specialised devices, resulting in accurate mobility GPS data (*specialised logs*).
-However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life.
-Because of this, specialised logs usually span a short amount of time.
-@barnett_inferring_2016 advocate installing custom-made tracking app on user's phones (*custom logs*).
+In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*).
+However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants.
+Moreover, they may be biased because of the introduction of a new device to the participant's life.
+Because of these drawbacks, specialised logs usually span a short amount of time.
+@barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
 Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_nodate] (*secondary logs*).
-These logs can be accessed and shared by users.
+By law, logs can be accessed and shared by users for free [*European Comission Reference*].
 Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate.
-Hence, two important challenges are dealing with measurement noise and missing data.
+Hence, two important challenges are dealing with measurement noise and missing data (*messy data*).
 
-There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016].
-@jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.
-Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
-@palmius_detecting_2017] can exacerbate the severity of the missing data problem.
-Traditional missing data methods, such as multiple imputation, cannot be used easily due to the geospatial nature of the data.
-On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as underestimating the movement of individuals.
-In this paper we will compare methods used to deal with measurement error and missing data in location information.
-Specifically, we are interested in establishing accurate mobility patterns from smartphone GPS logs.
-
-# Problem description & literature review
-
-Given that there is next to no literature on missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
-Moreover, although we could find no published papers on dealing with missing data in secondary logs, there are methods which deal with similar problems.
-In this section we will describe the problems with the data and why they arise as well as the models which could be applied to it.
-
-## Location logs and notation
-
-This example dataset comes from Google Location History and spans 3 years from January 2013 to January 2017 with multiple different Android devices.
-It contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15).
-The dataset contains a wide range of variables.
-For the purposes of this paper we will focus only on latitude, longitude, accuracy and time.
-
-Borrowing from @barnett_inferring_2016 let a persons' true Cartesian location on an error minimising projection be $G(t) = [G_x(t) G_y(t)]$ and $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
-In addition, let $a$ denote the estimated accuracy of the recorded data.
-accuracy.
-$G(t)$, $D$ and $a$ are indexed by the set $t_1 < ...
-< t_{n+1}$.
-For simplicity, let each entry in the discrete index set represent a 5 minute window.
-The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
-If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
-
-### Accuracy in location logs
-
-In professional grade GPS trackers less than 80% of measurements $D$ fall within 10 meters of the true location $G(t)$.
-GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
-Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
-
-```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement in a morning journey.
-The red circles denote the accuracy of all logged measurement points (the raw data).
-The points connected in time are connected by a line.
-The blue line shows the path without the most innaccurate (accurace > 400 meters) points filtered out.
-The red line shows the path with all measurements included.
-",out.width = '100%'}
-include_graphics("img/journeyTillMiddayBoaz.png")
-```
-
-Given that Android phones collect location information from WiFi access points, cellphone triangulation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006], the accuracy is substantially lower than in professional grade GPS trackers.
-Based on this data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
-@palmius_detecting_2017 note that in their custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
-I observe similar patterns in secondary logs.
-
-```{r, accuracyPlot2, echo=FALSE, fig.cap="The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
-The colour scale shows the distance between the measurement and the previous point.
-To ease interpretability all points with a distance between points of less than 5 meters have been filtered out.",out.width = '100%'}
-include_graphics("img/accuracyLocShift.png")
-
-```
-
-### Missingness
-
-Missing data is a pervasive issue as it can arise due to multiple factors, both technical and behavioural.
+Missing data is a pervasive issue as it can arise due to multiple reasons.
 Technical reasons include signal loss, battery failure and device failure.
@@
 
-In our example data set, over 54% of the data is missing for the entire duration of the log.
+There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016].
+Traditional missing data methods, such as mean imputation, cannot be used easily in spatiotemporal data because the measurements are correlated in time and space.
+For example, assume the simplistic case that an individual spends almost half her time at work, half her time at home and a small amount of time commuting between the two along a angled path.
+Using mean imputation would result in imputed values of her being at the geometric midpoint between the home and work for all missing values, even though she has never been there and never will be.
+Worringly, @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.
+
+The accuracy of GPS measurements in smartphones is substantially lower than in professional grade GPS trackers.
+For example, Android phones collect location information through a variety of methods, such as from WiFi access points, cellphone triangulation, and GPS measurements.
+They use different methods due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
+GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
+Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
+
+On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as overestimating the movement of individuals.
+For instance, suppose that a naive researcher calculates the distance travelled by an individual by drawing a line between each measured point and calculating the sum of the length of all of these lines.
+If there is noise, the measurements will vary even though the individual is not moving.
+If the measurements are frequent, then the researcher will end up with a lot of movement, even though the individual did not move at all.
+The problem is further complicated by the fact that missing data and noisy measurements are related to each other.
+For instance, methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
+@palmius_detecting_2017] can exacerbate the severity of the missing data problem.
+
+In this paper we will compare methods used to deal with measurement error and missing data in mobility patterns from secondary GPS logs.
+
+# A concrete example
+
+Given that there is little literature on dealing with missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
+The example dataset comes from the Google Location History of a single individual and spans from January 2013 to January 2017.
+It was recorded with multiple different Android devices and contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15).
+The dataset contains a wide range of variables including inferred activity and velocity.
+For the purposes of this paper we will focus only on latitude, longitude, accuracy (defined below) and a timestamp.
+
+## Location logs and notation
+
+GPS measurements report our location on a three dimensional planet, yet we are interested in placing these measurements on a two dimensional map.
+Projecting three dimensional measurements onto a two dimensional plane results in errors, in order to minimise these errors we borrow an error minimising projection method from @barnett_inferring_2016.
+
+Let a persons' true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
+Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+In addition, let $a$ denote the estimated accuracy of the recorded data.
+accuracy.
+$G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ...
+< t_{n+1}$.
+For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window.
+The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
+If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+
+### Accuracy in location logs
+
+In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
+@palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+We observe similar patterns in secondary logs.
+Figure 1 shows how accuracy tends to vary as a function of user behaviour, time and location.
+There are several recurring low-accuracy points, most likely cell-phone triangulation towers.
+
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement in a morning journey.
+The red circles denote the accuracy of all logged measurement points (the raw data).
+The points connected in time are connected by a line.
+The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out.
+The red line shows the path with all measurements included.
+",out.width = '100%'}
+include_graphics("img/journeyTillMiddayBoaz.png")
+```
+
+```{r, accuracyPlot2, echo=FALSE, fig.cap="The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
+The colour scale shows the distance between the measurement and the previous point.",out.width = '100%'}
+include_graphics("img/accuracyLocShift.png")
+
+```
+
+### Missingness
+
+Over 54% of the data is missing for the entire duration of the log.
 However,this is misleading as there are several long periods with no measurements whatsoever.
-The typical strucutre of missingness of a day with measurements is shown below.
+The structure of missingness of a day with measurements is shown below.
 
@@
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
-For this day there were several periods with no information.",out.width = '100%'}
+For this day there were several periods with no information.
+These points are filled with red and lie on the x-axis.",out.width = '100%'}
 #fix width
@@
 
-## State of the art in spatiotemporal models
+## Current methods for imputing & filtering spatiotemporal data
 
-Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and serves vastly different purposes.
-The following section briefly illustrates some methods used to deal with measurement inaccuracy and missing data problems in spatiotemporal data, as well as their applicability to our question.
-Moreover, we discuss in detail two approaches which deal explicitly with missing data in smartphone measured human location.
+As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs.
+We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions.
+For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data.
+In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with messy human mobility patterns.
+Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.
 
-### Spatiotemporal Imputation Methods
+### General Spatiotemporal Methods
 
-Given fixed measurement stations there are several imputation methods for spatiotemporal measurements.
+There is a vast literature of using SSMs to improve measurements accuracy and deal with missing data.
+Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008].
+These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
+The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
+In fact, the extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
+The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
+
+However, the main limitation of SSMs is that they ignore the fact that humans have regular movement routines, such as going to work or shopping for groceries on weekends.
+This limitation is due to the fact that SSMs are based on the Markov property.
+Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
+Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve the performance of Markovian models, but these solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
+For this reason we do not consider SSMs to be useful for imputing missing data, although they could be of use in filtering noise.
+
+Spatiotemporal imputation methods generally assume  fixed measurement stations there are several imputation methods for spatiotemporal measurements.
 For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
@@
 
-### State Space Models
-
-There is a vast literature of using state space models (SSMs) to improve measurements accuracy and deal with missing data.
-Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008].
-These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
-The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
-The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
-
-The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
-For our purposes the main limitation is that these models are based on the Markov property.
-Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
-This may be suitable for ships at sea, but it ignores the highly periodic nature of human movement.
-Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve their performance, but these are often computationally intractable or unfeasible [@sadilek_far_2016].
-
-### Alternative models
-
-Alternatives to state space models include long range-persistence models, such as cascading walks models  and the FarOut model which rely on self-similarity and autoregressive characteristics [@han_cascading_2015; @sadilek_far_2016].
-The latter uses Fourier analysis and PCA to extract cyclical patterns in an individual's behaviour and reduce the dimensionality of the extracted features and yields interpretable predictions for an individuals location months in advance.
-
 ### Filtering & Mean imputation
@@
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
-Moreover:
+Moreover, @palmius_detecting_2017 explain:
 
->If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch.
+>"If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".
 
@@
 
-@barnett_inferring_2016 deal with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. They handle missing data by:
+@barnett_inferring_2016 deal with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. @barnett_inferring_2016 handle missing data by:
 
-> simulat[ing] flights and pauses over the period of missingness where the direction,
+> "simulat[ing] flights and pauses over the period of missingness where the direction,
 duration, and spatial length of each flight, the fraction of flights versus the fraction of
-pauses, and the duration of pauses are sampled from observed data
+pauses, and the duration of pauses are sampled from observed data."
 
@@
 
-This paper is to the best of our knowledge the only attempt at establishing a principled approach to this problem
+This paper is to the best of our knowledge the only attempt at establishing a principled approach to this problem.
+
+*perhaps remove the section on alternative for the december deadline*
+### Alternative models
+
+Alternatives to state space models include long range-persistence models, such as cascading walks models  and the FarOut model which rely on self-similarity and autoregressive characteristics [@han_cascading_2015; @sadilek_far_2016].
+The latter uses Fourier analysis and PCA to extract cyclical patterns in an individual's behaviour and reduce the dimensionality of the extracted features and yields interpretable predictions for an individuals location months in advance.
 
@@
 The evaluation on computed measures involved calculating a mobility trace following the rectangual method of @rhee_human_2007 for each imputed dataset.
-Like @barnett_inferring_2016 I calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data.
+Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data.
 For simulation-based imputation approaches a mean value over 100 samples was taken.

### 2017-11-25 thesisReportApa.Rmd
edited a bit and added new APA format

--- 2017-11-24 thesisReport.Rmd
@@
 ---
-title: 'Handling missing data in smartphone location logs'
-author: "Boaz Sobrado"
-date: "`r format(Sys.time(), '%d %B %Y')`"
+title             : "Handling missing data in smartphone location logs"
+shorttitle        : "Missing Data in Smartphone Location Logs"
+
+author: 
+  - name          : "Boaz Sobrado"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : "Postal address"
+    email         : "boaz@boazsobrado.com"
+
+affiliation:
+  - id            : "1"
+    institution   : "Utrecht University"
+
+author_note: |
+  Department of Methodology & Statistics
+
+  Submitted as a research report.
+
+abstract: |
+  Using objective location data to infer the mobility measures of individuals is highly desirable, but methodologically difficult.
+Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals.
+However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment.
+In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values.
+Moreover, we will assess the performance of the model relative to currently used methods, such as linear interpolation.
+
+keywords          : "GPS, Human Mobility"
+wordcount         : "X"
+
 bibliography:
@@
   - rReferences.bib
-output:
-  rmarkdown::pdf_document:
+
+figsintext        : no
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : yes
+
+lang              : "english"
+class             : "man"
+output            : papaja::apa6_pdf
 citation_package: natbib
-    fig_caption: yes        
-    includes:  
-      in_header: tp2.tex
-
 ---
 
-```{r setup, include=FALSE}
+```{r load_packages, include = FALSE}
+library("papaja")
 knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
@@
 ```
-
-# Abstract
-*Using objective location data to infer the mobility measures of individuals is highly desirable, but methodologically difficult.
-Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals.
-However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment.
-In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values.
-Moreover, we will assess the performance of the model relative to currently used methods, such as linear interpolation.*
-
-# Introduction
 
@@
 @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
-Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_nodate] (*secondary logs*).
-By law, logs can be accessed and shared by users for free [*European Comission Reference*].
+Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_2017] (*secondary logs*).
+By law, logs can be accessed and shared by users for free [@commission_protecting_2017].
 Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate.
-Hence, two important challenges are dealing with measurement noise and missing data (*messy data*).
+Hence, two important challenges are dealing with measurement noise and missing data.
 
@@
 
-## Current methods for imputing & filtering spatiotemporal data
+## Current methods for imputing missing spatiotemporal data
 
@@
 For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data.
-In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with messy human mobility patterns.
+In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns.
 Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.
-
-### General Spatiotemporal Methods
 
@@
 
-Spatiotemporal imputation methods generally assume  fixed measurement stations there are several imputation methods for spatiotemporal measurements.
+In addition to SSMs there are spatiotemporal imputation methods often used in climate or geological research for estimating missing data.
 For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
@@
 
+The difficulty of applying this class of spatiotemporal imputation models to human mobility tracks lies in the fact that these models generally assume fixed measurement stations (such as rainfall gauging stations).
 While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
-To our knowledge such models have not been implemented.
+To our knowledge such models have not been implemented for mobility traces.
 
-### Filtering & Mean imputation
-
-@palmius_detecting_2017 deal with measuremement inaccuracy of $D$ by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
+On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns.
+@palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
@@
 
-### Barnett's model
-
-@barnett_inferring_2016 deal with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. @barnett_inferring_2016 handle missing data by:
+@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data.
+@barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes.
+In the words of the authors, @barnett_inferring_2016 handle missing data by:
 
@@
 This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses.
-In other words, for a given missing period, the individual's mobility can be estimated based on  measured movements in that area, at that point in time or movements in the last 24 hours.
-
-This paper is to the best of our knowledge the only attempt at establishing a principled approach to this problem.
-
-*perhaps remove the section on alternative for the december deadline*
-### Alternative models
-
-Alternatives to state space models include long range-persistence models, such as cascading walks models  and the FarOut model which rely on self-similarity and autoregressive characteristics [@han_cascading_2015; @sadilek_far_2016].
-The latter uses Fourier analysis and PCA to extract cyclical patterns in an individual's behaviour and reduce the dimensionality of the extracted features and yields interpretable predictions for an individuals location months in advance.
+In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours (*circadian proximity*).
 
@@
 
-The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals. The table below provides more details:
+The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (table 1).
 
@@
 
-Analyses were performed using R and a multitude of other packages [@ggplot2;@dplyr;@leaflet; @ggthemes;@base;@sp1;@sp2].
+Analyses were performed using R and a multitude of other statistical packages [@ggplot2;@dplyr;@leaflet;@base;@sp1;@sp2].
 
-### Data pre-processing
+### Data pre-processing & filtering
 
-Three different filtering methods were tested.
-First, the filtered rolling-median downsampling method described by @palmius_detecting_2017.
-The second method is a weighted mean approach taking $f(a)$ as a weight.
-Finally, an extended Kalman filter.
+The goal of filtering was to remove noise from the measurements and to aggregate multiple measurements into 12 per hour. Three different filtering methods were tested:
+
+1. The filtered rolling-median downsampling method described by @palmius_detecting_2017.
+2. A weighted mean approach taking $f(a)$ as a weight.
+3. A Kalman filter commonly used for GPS measurements [@doust_smoothing_2013].
+
 The output of all of these methods was taken as the input of the imputation methods.
@@
 
-Four imputation methods were selected in order to cover techniques applied in the literature.
-Briefly, the mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016].
-Simple linear interpolation was used as a benchmark model.
+Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:
+
+1. Mean imputation as described by @palmius_detecting_2017.
+2. The model developed by @barnett_inferring_2016 using spatial, temporal and circadian proximity.
+3. Simple linear interpolation was used as a benchmark model.
 
@@
 The specialised logs were used as a test set.
-The missing data imputation models were evaluated both directly, and on on two computed measures: amount of trips made and distance traveled.
+The missing data imputation models were evaluated both directly, and on two computed measures: amount of trips made and distance traveled.
 
@@
 
-The evaluation on computed measures involved calculating a mobility trace following the rectangual method of @rhee_human_2007 for each imputed dataset.
+The evaluation on computed measures involved calculating a mobility trace following the rectangular method of @rhee_human_2007 for each imputed dataset.
 Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data.
@@
 
-## Results
+\newpage
 
-## Discussion
+# References
 
-## References
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}

### 2017-11-25 thesisReportApa.tex
edited a bit and added new APA format

\documentclass[english,man]{apa6}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

% Table formatting
\usepackage{longtable, booktabs}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}

\newenvironment{lltable}
  {\begin{landscape}\begin{center}\begin{ThreePartTable}}
  {\end{ThreePartTable}\end{center}\end{landscape}}

  \usepackage{ifthen} % Only add declarations when endfloat package is loaded
  \ifthenelse{\equal{\string man}{\string man}}{%
   \DeclareDelayedFloatFlavor{ThreePartTable}{table} % Make endfloat play with longtable
   % \DeclareDelayedFloatFlavor{ltable}{table} % Make endfloat play with lscape
   \DeclareDelayedFloatFlavor{lltable}{table} % Make endfloat play with lscape & longtable
  }{}%

% The following enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand\getlongtablewidth{%
 \begingroup
  \ifcsname LT@\roman{LT@tables}\endcsname
  \global\longtablewidth=0pt
  \renewcommand\LT@entry[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}%
  \@nameuse{LT@\roman{LT@tables}}%
  \fi
\endgroup}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            pdfauthor={},
            pdftitle={Handling missing data in smartphone location logs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setlength{\parindent}{0pt}
%\setlength{\parskip}{0pt plus 0pt minus 0pt}

\setlength{\emergencystretch}{3em}  % prevent overfull lines

\ifxetex
  \usepackage{polyglossia}
  \setmainlanguage{}
\else
  \usepackage[english]{babel}
\fi

% Manuscript styling
\captionsetup{font=singlespacing,justification=justified}
\usepackage{csquotes}
\usepackage{upgreek}

 % Line numbering
  \usepackage{lineno}
  \linenumbers

\usepackage{tikz} % Variable definition to generate author note

% fix for \tightlist problem in pandoc 1.14
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Essential manuscript parts
  \title{Handling missing data in smartphone location logs}

  \shorttitle{Missing Data in Smartphone Location Logs}

  \author{Boaz Sobrado\textsuperscript{1}}

  \def\affdep{{""}}%
  \def\affcity{{""}}%

  \affiliation{
    \vspace{0.5cm}
          \textsuperscript{1} Utrecht University  }

  \authornote{
    \newcounter{author}
    Department of Methodology \& Statistics

    Submitted as a research report.

                      Correspondence concerning this article should be addressed to Boaz Sobrado, Postal address. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
                }

  \abstract{Using objective location data to infer the mobility measures of
individuals is highly desirable, but methodologically difficult. Using
commercially gathered location logs from smartphones holds great
promise, as they have already been gathered, often span years and can be
associated to individuals. However, due to technical constraints this
data is more sparse and inaccurate than that produced by specialised
equipment. In this paper we present a model which leverages the
periodicity of human mobility in order to impute missing data values.
Moreover, we will assess the performance of the model relative to
currently used methods, such as linear interpolation.}
  \keywords{GPS, Human Mobility \\

    \indent Word count: X
  }

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

\maketitle

\setcounter{secnumdepth}{0}

How active people are and how they interact with their environment
affects a wide range of outcomes including health, income and social
capital (Goodchild \& Janelle, 2010). A better understanding of both
within-person and between-person variability in geospatial patterns
could be conducive to better social, health and urban-planning policies.
Yet a large part of studies on human mobility are largely based on
pen-and-paper travel diaries. These surveys have known methodological
flaws, such as the short period of data collection (due to costs and
burden to respondents), the underreporting of short trips (Wolf,
Oliveira, \& Thompson, 2003) and the underestimation of the duration of
commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).

These obstacles can be overcome by using objective data on human
mobility. The Global Positioning System (GPS), which uses the distance
between a device and a number of satellites to determine location,
provides such data. Within behavioural science, this type of data has
been used to investigate topics such as the effects of the food
environment on eating patterns (Zenk, Schulz, \& Odoms-Young, 2009), the
movement correlates of personality (G. M. Harari et al., 2016), academic
performance (Wang, Harari, Hao, Zhou, \& Campbell, 2015) and bipolar
disorder (Palmius et al., 2017).

In most of these studies participants are given a specialised device,
resulting in accurate mobility GPS data (\emph{specialised logs}).
However, Barnett and Onnela (2016) point out that these studies are not
scalable due to cost and burden to participants. Moreover, they may be
biased because of the introduction of a new device to the participant's
life. Because of these drawbacks, specialised logs usually span a short
amount of time. Barnett and Onnela (2016) advocate installing a
custom-made tracking app on user's phones (\emph{custom logs}). Another
solution is to take advantage of existing smartphone location logs, such
as Google Location History, which store location information of millions
of users spanning several years (Location History, 2017)
(\emph{secondary logs}). By law, logs can be accessed and shared by
users for free (Commission, 2017). Yet, because they were created for
non-academic purposes under engineering constraints, the sensors do not
monitor continuously and the resulting logs can be sparse and
inaccurate. Hence, two important challenges are dealing with measurement
noise and missing data.

Missing data is a pervasive issue as it can arise due to multiple
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home,
switching the phone off, switching location measurements off, and so on.
As a result, applied researchers are often left with wide temporal gaps
with no measurements. For instance, different groups studying the effect
of bipolar disorder on human movement have reported missing data rates
between 30\% to 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb
et al., 2015). Similar trends are consistently reported in other fields
(e.g. G. M. Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

There is currently no golden standard in how to deal with missing data
in custom or secondary logs (Barnett \& Onnela, 2016). Traditional
missing data methods, such as mean imputation, cannot be used easily in
spatiotemporal data because the measurements are correlated in time and
space. For example, assume the simplistic case that an individual spends
almost half her time at work, half her time at home and a small amount
of time commuting between the two along a angled path. Using mean
imputation would result in imputed values of her being at the geometric
midpoint between the home and work for all missing values, even though
she has never been there and never will be. Worringly, Jankowska et al.
(2015) have pointed out that there is often little transparency
regarding decisions of how to deal with missing data.

The accuracy of GPS measurements in smartphones is substantially lower
than in professional grade GPS trackers. For example, Android phones
collect location information through a variety of methods, such as from
WiFi access points, cellphone triangulation, and GPS measurements. They
use different methods due to computational and battery constraints (M.
Y. Chen et al., 2006; LaMarca et al., 2005).In professional grade GPS
trackers less than 80\% of measurements fall within 10 meters of the
true location. GPS measures are reported to be most inaccurate in high
density urban locations and indoors (S. Duncan et al., 2013; Schipperijn
et al., 2014). Unfortunately for social scientists, this happens to be
where most people in the developed world tend to spend most of their
time.

On the other hand, noisy data can lead to inaccurate conclusions if it
is not accounted for, such as overestimating the movement of
individuals. For instance, suppose that a naive researcher calculates
the distance travelled by an individual by drawing a line between each
measured point and calculating the sum of the length of all of these
lines. If there is noise, the measurements will vary even though the
individual is not moving. If the measurements are frequent, then the
researcher will end up with a lot of movement, even though the
individual did not move at all. The problem is further complicated by
the fact that missing data and noisy measurements are related to each
other. For instance, methods used by researchers to reduce noise, such
as throwing out inaccurate measurements (e.g. Palmius et al., 2017) can
exacerbate the severity of the missing data problem.

In this paper we will compare methods used to deal with measurement
error and missing data in mobility patterns from secondary GPS logs.

\section{A concrete example}\label{a-concrete-example}

Given that there is little literature on dealing with missing data in
custom or secondary logs it is worth illustrating the typical
characteristics of this data using an example data set. The example
dataset comes from the Google Location History of a single individual
and spans from January 2013 to January 2017. It was recorded with
multiple different Android devices and contains 814 941 measurements,
with approximately 742 measurements per day
(\(\widehat{\sigma}\)=868.15). The dataset contains a wide range of
variables including inferred activity and velocity. For the purposes of
this paper we will focus only on latitude, longitude, accuracy (defined
below) and a timestamp.

\subsection{Location logs and
notation}\label{location-logs-and-notation}

GPS measurements report our location on a three dimensional planet, yet
we are interested in placing these measurements on a two dimensional
map. Projecting three dimensional measurements onto a two dimensional
plane results in errors, in order to minimise these errors we borrow an
error minimising projection method from Barnett and Onnela (2016).

Let a persons' true location on this two-dimensional plane be
\(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and \(G_y(t)\) denote the
location of the individual at time \(t\) on the x-axis and y-axis
respectively. Moreover, let \(D \in \mathbb{R}^2\) be the recorded data
containing lattitude and longitude. In addition, let \(a\) denote the
estimated accuracy of the recorded data. accuracy. \(G(t)\), \(D\) and
\(a\) are indexed by time labled by the set \(T = t_1 < ... < t_{n+1}\).
For simplicity, let each entry in the discrete index set \(T\) represent
a 5 minute window. The measure of accuracy \(a_t\) is given in meters
such that it represents the radius of a 67\% confidence circle. If
\(D_t = \emptyset\) it is considered \emph{missing} and it is not
missing otherwise.

\subsubsection{Accuracy in location
logs}\label{accuracy-in-location-logs}

In the example data set the distribution of \(a\) is highly right
skewed, with a median of 28, \(\mu = 127\) and the maximum value at 26
km. Palmius et al. (2017) note that in their Android based custom logs
inaccurate location values are interspersed between more accurate
location values at higher sample rates per hour. We observe similar
patterns in secondary logs. Figure 1 shows how accuracy tends to vary as
a function of user behaviour, time and location. There are several
recurring low-accuracy points, most likely cell-phone triangulation
towers.

\begin{figure}
\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement in a morning journey. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. }\label{fig:accuracyPlot}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{The accuracy of raw measures with time on the x-axis and accuracy on the y-axis. The colour scale shows the distance between the measurement and the previous point.}\label{fig:accuracyPlot2}
\end{figure}

\subsubsection{Missingness}\label{missingness}

Over 54\% of the data is missing for the entire duration of the log.
However,this is misleading as there are several long periods with no
measurements whatsoever. The structure of missingness of a day with
measurements is shown below.

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingBoaz5minExample} \caption{Example of missing data over a day. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

\subsection{Current methods for imputing missing spatiotemporal
data}\label{current-methods-for-imputing-missing-spatiotemporal-data}

As we have mentioned before, Jankowska et al. (2015) point out that
there is little transparency regarding decisions of how to deal with
missing data in GPS mobility logs. We suspect this is because the highly
interdisciplinary nature of the problem means researchers are unaware of
potential solutions. For this reason, we consider it important to
briefly discuss a few methods one could consider in order to deal with
measurement inaccuracy and missing data problems in spatiotemporal data.
In doing so, we will argue that state space models (SSMs) and
spatiotemporal imputation methods, which are extensively used to model
missing and noisy spatiotemporal data, are not well suited to deal with
human mobility patterns. Moreover, we discuss in detail two approaches
by Palmius et al. (2017) and Barnett and Onnela (2016) which deal
explicitly with missing data in mobility patterns from smartphone GPS
logs.

There is a vast literature of using SSMs to improve measurements
accuracy and deal with missing data. Behavioural ecologists for
instance, have used SSMs extensively to explain how animals interact
with their environment (Patterson, Thomas, Wilcox, Ovaskainen, \&
Matthiopoulos, 2008). These models can be quite complex, for example
Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. In fact, the extended Kalman filter is the de facto standard
for GPS navigation (Z. Chen \& Brown, 2013). The advantage of state
space models is that they are flexible, deal with measurement
inaccuracy, include information from different sources and can be used
in real time.

However, the main limitation of SSMs is that they ignore the fact that
humans have regular movement routines, such as going to work or shopping
for groceries on weekends. This limitation is due to the fact that SSMs
are based on the Markov property. Thus, the estimated location \(G(t)\)
at timepoint \(t\) is often based only upon measurements \(D_t\),
\(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\). Hierarchical structuring
and conditioning on a larger context have been suggested as ways to
improve the performance of Markovian models, but these solutions are
often computationally intractable or unfeasible (Sadilek \& Krumm,
2016). For this reason we do not consider SSMs to be useful for imputing
missing data, although they could be of use in filtering noise.

In addition to SSMs there are spatiotemporal imputation methods often
used in climate or geological research for estimating missing data. For
instance, Feng, Nowak, O'Neill, and Welsh (2014) illustrate their CUTOFF
method, which relies on estimating missing values using the nearest
observed neighbours in time, using rainfall data from dozens of gauging
stations across Australia. Similarly, Z. Zhang et al. (2017) use a
variety of machine learning methods to present their model based on
underground water data in China.

The difficulty of applying this class of spatiotemporal imputation
models to human mobility tracks lies in the fact that these models
generally assume fixed measurement stations (such as rainfall gauging
stations). While Feng et al. (2014) claim their model could be used to
establish mobility patterns, ostensibly by dividing the sample space
into rasters analogous to measurement stations indicating a probability
of the individual being there, this seems to be computationally
unfeasible. To our knowledge such models have not been implemented for
mobility traces.

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns. Palmius et al. (2017) deal
with the measuremement inaccuracy of \(D\) in custom logs by removing
from the data set all unique low-accuracy \(a\) data points that had
\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
sample the data to a sample rate of 12 per hour using a median filter.
Moreover, Palmius et al. (2017) explain:

\begin{quote}
\enquote{If the standard deviation of {[}\(D\){]} in both latitude and
longitude within a 1 h epoch was less than 0.01 km, then all samples
within the hour were set to the mean value of the recorded data,
otherwise a 5 min median filter window was applied to the recorded
latitude and longitude in the epoch}.
\end{quote}

Missing data was imputed using the mean of measurements close in time if
the participant was recorded within 500m of either end of a missing
section and the missing section had a length of \(\leq 2h\) or
\(\leq 12h\) after 9pm.

Barnett and Onnela (2016) follow a different approach which is, to the
best of our knowledge, the only pricipled approach to dealing with
missing data in human mobility data. Barnett and Onnela (2016) work with
custom logs where location is measured for 2 minutes and subsequently
not measured for 10 minutes. In the words of the authors, Barnett and
Onnela (2016) handle missing data by:

\begin{quote}
\enquote{simulat{[}ing{]} flights and pauses over the period of
missingness where the direction, duration, and spatial length of each
flight, the fraction of flights versus the fraction of pauses, and the
duration of pauses are sampled from observed data.}
\end{quote}

This method can be extended to imputing the data based on temporally,
spatially or periodically close flights and pauses. In other words, for
a given missing period, the individual's mobility can be estimated based
on measured movements in that area, at that point in time or movements
in the last 24 hours (\emph{circadian proximity}).

\subsection{Methods}\label{methods}

\subsubsection{Datasets \& Analyses}\label{datasets-analyses}

The data used to train the imputation methods was collected between 2013
and 2017 on different Android devices from several individuals (table
1).

\begin{table}

\caption{\label{tab:datadetailsTable}Table with descriptives about the data sets used to build the imputation methods. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r}
\hline
Log duration & Logged days & Observations & Missing days & Missing data & Mean Accuracy\\
\hline
From 2013-02-06 to 2017-03-29 & 1512 & 646376 & 635 & 0.22 & 127.78\\
\hline
From 2016-07-14 to 2017-05-10 & 300 & 158382 & 3 & 0.41 & 1394.60\\
\hline
From 2014-01-22 to 2017-01-23 & 1097 & 814941 & 80 & 0.25 & 121.83\\
\hline
\end{tabular}
\end{table}

In addition to the secondary logs, participants also volunteered to
carry with them a specialised GPS tracker for a week. This specialised
log was used to evaluate the models.

Analyses were performed using R and a multitude of other statistical
packages ({\textbf{???}}; Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J.
Pebesma \& Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \&
Francois, 2016).

\subsubsection{Data pre-processing \&
filtering}\label{data-pre-processing-filtering}

The goal of filtering was to remove noise from the measurements and to
aggregate multiple measurements into 12 per hour. Three different
filtering methods were tested:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The filtered rolling-median downsampling method described by Palmius
  et al. (2017).
\item
  A weighted mean approach taking \(f(a)\) as a weight.
\item
  A Kalman filter commonly used for GPS measurements (Doust, 2013).
\end{enumerate}

The output of all of these methods was taken as the input of the
imputation methods.

\subsubsection{Imputation methods}\label{imputation-methods}

Three imputation methods were selected in order to cover a wide range of
techniques applied in the literature:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mean imputation as described by Palmius et al. (2017).
\item
  The model developed by Barnett and Onnela (2016) using spatial,
  temporal and circadian proximity.
\item
  Simple linear interpolation was used as a benchmark model.
\end{enumerate}

\subsubsection{Evaluation criteria}\label{evaluation-criteria}

The entire length of the secondary logs were used as a training set. The
specialised logs were used as a test set. The missing data imputation
models were evaluated both directly, and on two computed measures:
amount of trips made and distance traveled.

The direct evaluation involved calculating the error of each \(D_t\)
compared to \(G(t)\) approximated by the specialised log. The error
measures used were root mean square error (RMSE) and mean absolute error
(MAE).

The evaluation on computed measures involved calculating a mobility
trace following the rectangular method of Rhee, Shin, Hong, Lee, and
Chong (2007) for each imputed dataset. Like Barnett and Onnela (2016) we
calculate bias by substracting the estimated measure under each approach
for the same measure calculated on the full data. For simulation-based
imputation approaches a mean value over 100 samples was taken.

\newpage

\section{References}\label{references}

\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\hypertarget{ref-barnett_inferring_2016}{}
Barnett, I., \& Onnela, J.-P. (2016). Inferring Mobility Measures from
GPS Traces with Missing Data. \emph{arXiv:1606.06328 {[}Stat{]}}.
Retrieved from \url{http://arxiv.org/abs/1606.06328}

\hypertarget{ref-sp2}{}
Bivand, R. S., Pebesma, E., \& Gomez-Rubio, V. (2013). \emph{Applied
spatial data analysis with R, second edition}. Springer, NY. Retrieved
from \url{http://www.asdar-book.org/}

\hypertarget{ref-chen_practical_2006}{}
Chen, M. Y., Sohn, T., Chmelev, D., Haehnel, D., Hightower, J., Hughes,
J., \ldots{} Varshavsky, A. (2006). Practical Metropolitan-Scale
Positioning for GSM Phones. In \emph{UbiComp 2006: Ubiquitous Computing}
(pp. 225--242). Springer, Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11853565_14}{10.1007/11853565\_14}

\hypertarget{ref-chen_state_2013}{}
Chen, Z., \& Brown, E. N. (2013). State space model.
\emph{Scholarpedia}, \emph{8}(3), 30868.
doi:\href{https://doi.org/10.4249/scholarpedia.30868}{10.4249/scholarpedia.30868}

\hypertarget{ref-commission_protecting_2017}{}
Commission, E. (2017). Protecting your data: Your rights - European
Commission. Retrieved from
\url{http://ec.europa.eu/justice/data-protection/individuals/rights/index_en.htm}

\hypertarget{ref-delclos-alio_keeping_2017}{}
Delclòs-Alió, X., Marquet, O., \& Miralles-Guasch, C. (2017). Keeping
track of time: A Smartphone-based analysis of travel time perception in
a suburban environment. \emph{Travel Behaviour and Society},
\emph{9}(Supplement C), 1--9.
doi:\href{https://doi.org/10.1016/j.tbs.2017.07.001}{10.1016/j.tbs.2017.07.001}

\hypertarget{ref-doust_smoothing_2013}{}
Doust, P. (2013). \emph{Smoothing - Smooth GPS data - Stack Overflow}.
Retrieved from
\url{https://stackoverflow.com/questions/1134579/smooth-gps-data}

\hypertarget{ref-duncan_portable_2013}{}
Duncan, S., Stewart, T. I., Oliver, M., Mavoa, S., MacRae, D., Badland,
H. M., \& Duncan, M. J. (2013). Portable global positioning system
receivers: Static validity and environmental conditions. \emph{American
Journal of Preventive Medicine}, \emph{44}(2), e19--29.
doi:\href{https://doi.org/10.1016/j.amepre.2012.10.013}{10.1016/j.amepre.2012.10.013}

\hypertarget{ref-feng_cutoff:_2014}{}
Feng, L., Nowak, G., O'Neill, T., \& Welsh, A. (2014). CUTOFF: A
spatio-temporal imputation method. \emph{Journal of Hydrology},
\emph{519}, 3591--3605.
doi:\href{https://doi.org/10.1016/j.jhydrol.2014.11.012}{10.1016/j.jhydrol.2014.11.012}

\hypertarget{ref-goodchild_toward_2010}{}
Goodchild, M. F., \& Janelle, D. G. (2010). Toward critical spatial
thinking in the social sciences and humanities. \emph{GeoJournal},
\emph{75}(1), 3--13.
doi:\href{https://doi.org/10.1007/s10708-010-9340-3}{10.1007/s10708-010-9340-3}

\hypertarget{ref-grunerbl_smartphone-based_2015}{}
Grünerbl, A., Muaremi, A., Osmani, V., Bahle, G., Ohler, S., Tröster,
G., \ldots{} Lukowicz, P. (2015). Smartphone-based recognition of states
and state changes in bipolar disorder patients. \emph{IEEE Journal of
Biomedical and Health Informatics}, \emph{19}(1), 140--148.
doi:\href{https://doi.org/10.1109/JBHI.2014.2343154}{10.1109/JBHI.2014.2343154}

\hypertarget{ref-harari_using_2016}{}
Harari, G. M., Lane, N. D., Wang, R., Crosier, B. S., Campbell, A. T.,
\& Gosling, S. D. (2016). Using Smartphones to Collect Behavioral Data
in Psychological Science: Opportunities, Practical Considerations, and
Challenges. \emph{Perspectives on Psychological Science}, \emph{11}(6),
838--854.
doi:\href{https://doi.org/10.1177/1745691616650285}{10.1177/1745691616650285}

\hypertarget{ref-jankowska_framework_2015}{}
Jankowska, M. M., Schipperijn, J., \& Kerr, J. (2015). A Framework For
Using GPS Data In Physical Activity And Sedentary Behavior Studies.
\emph{Exercise and Sport Sciences Reviews}, \emph{43}(1), 48--56.
doi:\href{https://doi.org/10.1249/JES.0000000000000035}{10.1249/JES.0000000000000035}

\hypertarget{ref-lamarca_place_2005}{}
LaMarca, A., Chawathe, Y., Consolvo, S., Hightower, J., Smith, I.,
Scott, J., \ldots{} Schilit, B. (2005). Place Lab: Device Positioning
Using Radio Beacons in the Wild. In \emph{Pervasive Computing} (pp.
116--133). Springer, Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11428572_8}{10.1007/11428572\_8}

\hypertarget{ref-location_history_timeline_2017}{}
Location History, G. (2017). Timeline. Retrieved from
\url{https://www.google.com/maps/timeline?pb}

\hypertarget{ref-palmius_detecting_2017}{}
Palmius, N., Tsanas, A., Saunders, K. E. A., Bilderbeck, A. C., Geddes,
J. R., Goodwin, G. M., \& Vos, M. D. (2017). Detecting Bipolar
Depression From Geographic Location Data. \emph{IEEE Transactions on
Biomedical Engineering}, \emph{64}(8), 1761--1771.
doi:\href{https://doi.org/10.1109/TBME.2016.2611862}{10.1109/TBME.2016.2611862}

\hypertarget{ref-patterson_statespace_2008}{}
Patterson, T. A., Thomas, L., Wilcox, C., Ovaskainen, O., \&
Matthiopoulos, J. (2008). State--space models of individual animal
movement. \emph{Trends in Ecology \& Evolution}, \emph{23}(2), 87--94.
doi:\href{https://doi.org/10.1016/j.tree.2007.10.009}{10.1016/j.tree.2007.10.009}

\hypertarget{ref-sp1}{}
Pebesma, E. J., \& Bivand, R. S. (2005). Classes and methods for spatial
data in R. \emph{R News}, \emph{5}(2), 9--13. Retrieved from
\url{https://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-preisler_modeling_2004}{}
Preisler, H. K., Ager, A. A., Johnson, B. K., \& Kie, J. G. (2004).
Modeling animal movements using stochastic differential equations.
\emph{Environmetrics 15: P. 643-657}. Retrieved from
\url{https://www.fs.usda.gov/treesearch/pubs/33038}

\hypertarget{ref-base}{}
R Core Team. (2017). \emph{R: A language and environment for statistical
computing}. Vienna, Austria: R Foundation for Statistical Computing.
Retrieved from \url{https://www.R-project.org/}

\hypertarget{ref-rhee_human_2007}{}
Rhee, I., Shin, M., Hong, S., Lee, K., \& Chong, S. (2007). Human
Mobility Patterns and Their Impact on Routing in Human-Driven Mobile
Networks. \emph{ACM HotNets 2007}. Retrieved from
\url{http://koasas.kaist.ac.kr/handle/10203/160927}

\hypertarget{ref-sadilek_far_2016}{}
Sadilek, A., \& Krumm, J. (2016). Far Out: Predicting Long-Term Human
Mobility. \emph{Microsoft Research}. Retrieved from
\url{https://www.microsoft.com/en-us/research/publication/far-predicting-long-term-human-mobility/}

\hypertarget{ref-saeb_mobile_2015}{}
Saeb, S., Zhang, M., Karr, C. J., Schueller, S. M., Corden, M. E.,
Kording, K. P., \& Mohr, D. C. (2015). Mobile Phone Sensor Correlates of
Depressive Symptom Severity in Daily-Life Behavior: An Exploratory
Study. \emph{Journal of Medical Internet Research}, \emph{17}(7), e175.
doi:\href{https://doi.org/10.2196/jmir.4273}{10.2196/jmir.4273}

\hypertarget{ref-schipperijn_dynamic_2014}{}
Schipperijn, J., Kerr, J., Duncan, S., Madsen, T., Klinker, C. D., \&
Troelsen, J. (2014). Dynamic Accuracy of GPS Receivers for Use in Health
Research: A Novel Method to Assess GPS Accuracy in Real-World Settings.
\emph{Frontiers in Public Health}, \emph{2}, 21.
doi:\href{https://doi.org/10.3389/fpubh.2014.00021}{10.3389/fpubh.2014.00021}

\hypertarget{ref-wang_smartgpa:_2015}{}
Wang, R., Harari, G., Hao, P., Zhou, X., \& Campbell, A. T. (2015).
SmartGPA: How Smartphones Can Assess and Predict Academic Performance of
College Students. In \emph{Proceedings of the 2015 ACM International
Joint Conference on Pervasive and Ubiquitous Computing} (pp. 295--306).
New York, NY, USA: ACM.
doi:\href{https://doi.org/10.1145/2750858.2804251}{10.1145/2750858.2804251}

\hypertarget{ref-ggplot2}{}
Wickham, H. (2009). \emph{Ggplot2: Elegant graphics for data analysis}.
Springer-Verlag New York. Retrieved from \url{http://ggplot2.org}

\hypertarget{ref-dplyr}{}
Wickham, H., \& Francois, R. (2016). \emph{Dplyr: A grammar of data
manipulation}. Retrieved from
\url{https://CRAN.R-project.org/package=dplyr}

\hypertarget{ref-wolf_impact_2003}{}
Wolf, J., Oliveira, M., \& Thompson, M. (2003). Impact of Underreporting
on Mileage and Travel Time Estimates: Results from Global Positioning
System-Enhanced Household Travel Survey. \emph{Transportation Research
Record: Journal of the Transportation Research Board}, \emph{1854},
189--198. doi:\href{https://doi.org/10.3141/1854-21}{10.3141/1854-21}

\hypertarget{ref-zenk_how_2009}{}
Zenk, S. N., Schulz, A. J., \& Odoms-Young, A. (2009). How Neighborhood
Environments Contribute to Obesity. \emph{The American Journal of
Nursing}, \emph{109}(7), 61--64.
doi:\href{https://doi.org/10.1097/01.NAJ.0000357175.86507.c8}{10.1097/01.NAJ.0000357175.86507.c8}

\hypertarget{ref-zhang_application_2017}{}
Zhang, Z., Yang, X., Li, H., Li, W., Yan, H., \& Shi, F. (2017).
Application of a novel hybrid method for spatiotemporal data imputation:
A case study of the Minqin County groundwater level. \emph{Journal of
Hydrology}, \emph{553}(Supplement C), 384--397.
doi:\href{https://doi.org/10.1016/j.jhydrol.2017.07.053}{10.1016/j.jhydrol.2017.07.053}

\end{document}

### 2017-11-25 thesisReport.Rmd
problem description comments

@@
 We observe similar patterns in secondary logs.
-Figure 1 shows how accuracy tends to vary as a function of user behaviour, time and location.
+Figure 1 shows how accuracy can vary as a function of user behaviour, time and location.
+Most notably, accuracy tends to be lowest during movement.
+Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.
 There are several recurring low-accuracy points, most likely cell-phone triangulation towers.
@@
 
-Over 54% of the data is missing for the entire duration of the log.
+Over 54% of the data is missing for the entire duration of the log (as shown in Figure X a).
 However,this is misleading as there are several long periods with no measurements whatsoever.
-The structure of missingness of a day with measurements is shown below.
+The structure of missingness of a day with measurements is shown in Figure X b.
+As you can see, there are several long periods over the course of the log for which there are no measurements.
+Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.
+
+```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of the log.
+The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
+For this day there were several periods with no information.
+These points are filled with red and lie on the x-axis.",out.width = '100%'}
+
+include_graphics("img/missingdayBoaz5min.png")
+
+```
 
@@
 include_graphics("img/missingBoaz5minExample.png")
-
 ```
 
-## Current methods for imputing & filtering spatiotemporal data
+## Current methods for imputing missing spatiotemporal data
 
@@
 
-Four imputation methods were selected in order to cover techniques applied in the literature.
-Briefly, the mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016].
-Simple linear interpolation was used as a benchmark model.
+Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:
+
+1. Mean imputation as described by @palmius_detecting_2017.
+2. The model developed by @barnett_inferring_2016 using both spatial and temporal proximity.
+3. Simple linear interpolation was used as a benchmark model.
 
@@
 
-## Results
-
-## Discussion
+Each imputation method used each of the three filtering methods as an input.
+Thus in the end we end up with eight methods to evaluate, three for each filtering method as well as four for each imputation method.
 

### 2017-11-25 scripts/accuracyPlots.R

@@ -45,7 +45,7 @@ library(leaflet)
 lnsRaw<- all %>%
   select(time,lat,lon,accuracy) %>%
   as_tbl_time(index = time) %>%
-  time_filter(2017-02-15 + 00:00:00 ~ 2017-02-15+ 12:00:00)%>%
+  time_filter(2016-02-15 + 00:00:00 ~ 2017-02-15+ 12:00:00)%>%
   select(lon, lat) %>%
   sp::Line()
   
@@ -139,3 +139,38 @@ x<- bind_rows(day1,week1)
 
 ggplot(x, aes(x=unit, y=accuracy)) + geom_boxplot()
 
+
+# new scatterplot
+
+all %>%
+  select(time,lon,lat,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-03-02 ~ 2017-03-02)%>%
+  mutate( time2 = lubridate::force_tz(time, "Asia/Singapore"),
+          distancePrev = c(0,
+                           sp::spDists(x=matrix(c(lon,lat),ncol = 2), segments = T)),
+          hour = as.POSIXct(
+            paste0("2014-01-22 ",
+                   strftime(time,format = "%H:%M", tz = "Europe/Budapest")))) %>%
+  ggplot( aes(x = hour, y=accuracy))+
+  geom_point()+theme_tufte()
+
+#wtf happened 2nd of march
+
+all %>%
+  select(time,lat,lon,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-02-15 ~ 2017-02-17) %>%
+  filter(accuracy > 1000) %>% 
+  leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
+  addTiles() %>%
+  addCircles(lng = ~lon, lat = ~lat,
+             radius = ~accuracy, fillOpacity = 0.02,color = "#DF2935")%>%
+  addProviderTiles(providers$CartoDB.Positron)
+
+all %>%
+  select(time,lat,lon,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-03-01 + 23:30:00 ~ 2017-03-02+ 10:00:00) %>%
+  ggplot( aes(x = time, y=accuracy))+
+  geom_point()+theme_tufte()

### 2017-11-25 scripts/missingnessPlot.R

@@ -71,13 +71,12 @@ missingDay5min<- ggplot(m5min, aes(x = time_5_min_day, y=measurements, colour= m
   xlab("")+
   ylab("Measurements")+
   ggtitle("Missingness in daily measurements over time")+
-  scale_colour_gradient2(low = muted("green"), mid = "grey50",
+  scale_colour_gradient2(low = "black", mid = "grey50",
                          high = muted("red"), midpoint = 144, space = "Lab",
                          na.value = "grey50", guide = "colourbar")+
-  labs(x = NULL, colour = "Missing \n5 minute\n segments")+
-  theme(legend.position = c(0.3, 0.85), legend.direction = "horizontal")
+  labs(x = NULL, colour = "Missing \n5 minute\n segments")
 
-ggsave(missingDay5min,filename = "../img/missingdayPeter5min.png",device = "png",height = 6.5, units = "cm")
+ggsave(missingDay5min,filename = "../img/missingdayBoaz5min.png",device = "png",height = 6.5, , width = 18, units = "cm")
 
 # by 5 minute version 2
 m5minv2<- all %>%
@@ -111,7 +110,7 @@ exampleMiss<- m5minv2 %>%
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
                    labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
-  scale_colour_manual(values=c("black","#DF2935"))+
+  scale_colour_manual(values=c("black",muted("red")))+
   theme(legend.position="none")
 
 ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")

### 2017-11-26 thesisReport.Rmd
plot detail and imagemagick

@@
 Figure 1 shows how accuracy can vary as a function of user behaviour, time and location.
-Most notably, accuracy tends to be lowest during movement.
+Inaccurate measures are often followed by more accurate measures.
+Most notably, low accuracy tends to be associated with movement (Figure 2).
 Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.
-There are several recurring low-accuracy points, most likely cell-phone triangulation towers.
+There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.
 
-```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement in a morning journey.
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement on a morning journey.
 The red circles denote the accuracy of all logged measurement points (the raw data).
@@
 
-```{r, accuracyPlot2, echo=FALSE, fig.cap="The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
-The colour scale shows the distance between the measurement and the previous point.",out.width = '100%'}
+```{r, accuracyPlot2, echo=FALSE, fig.cap="The upper chart shows the rolling mean distance from the next point in meters over the course of the day.
+The lower chart shows the rolling mean accuracy over the course of the day.
+In both cases, the rolling window is 3.",out.width = '100%'}
 include_graphics("img/accuracyLocShift.png")

### 2017-11-26 scripts/imageMagick.R

@@ -1,0 +1,29 @@
+#image magick
+library(magick)
+inac <- image_read("../img/journeyTillMiddayBoaz.png")
+ac <- image_read("../img/accuracyLocShift.png")
+
+im2 <- c(inac,ac)
+x<- image_append(image_scale(im2,"x500"))
+image_write(x,"../img/accPlots.png")
+# doesnt; work
+
+
+image_mosaic(ac, inac)
+
+
+bigdata <- image_read('https://jeroen.github.io/images/bigdata.jpg')
+frink <- image_read("https://jeroen.github.io/images/frink.png")
+logo <- image_read("https://www.r-project.org/logo/Rlogo.png")
+img <- c(bigdata, logo, frink)
+img <- image_scale(img, "300x300")
+image_mosaic(img)
+image_flatten(img)
+
+image_append(image_scale(img,"x200"))
+
+left_to_right <- image_append(image_scale(img, "x200"))
+image_background(left_to_right, "white", flatten = TRUE)
+
+top_to_bottom <- image_append(image_scale(img, "100"), stack = TRUE)
+image_background(top_to_bottom, "white", flatten = TRUE)

### 2017-11-26 scripts/rollDistanceAccuracyPlot.R

@@ -2,7 +2,7 @@
 # plz load packages and data first
 #
 
-rolmean<- rollify(mean,window = 3,na_value = F)
+rolmean<- rollify(mean,window = 3)
 
 home<- c(5.113919,52.10421)
 
@@ -50,6 +50,10 @@ library(grid)
 
 locShift<- grid.arrange(rMeanD,rMeanAcc,ncol=1)
 
+#square
+ggsave(locShift,filename = "../img/accuracyLocShiftSquare.png",device = "png",height = 6.5,width = 6.5, units = "cm")
+
+#long
 ggsave(locShift,filename = "../img/accuracyLocShift.png",device = "png",height = 6.5,width = 20, units = "cm")
 
 

### 2017-11-26 thesisReport.Rmd
simplified introduction

@@
 
-How active people are and how they interact with their environment affects a wide range of outcomes including health, income and social capital [@goodchild_toward_2010].
-A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies.
-Yet a large part of studies on human mobility are largely based on pen-and-paper travel diaries.
-These surveys have known methodological flaws, such as the short period of data collection (due to costs and burden to respondents), the underreporting of short trips [@wolf_impact_2003] and the underestimation of the duration of commutes [@delclos-alio_keeping_2017].
+How people move about in their environment affects a wide range of outcomes.
+These include health, income and social capital [@goodchild_toward_2010].
+A better understanding of mobility could lead to better health and urban-planning policies.
+A large part of studies on human mobility are conducted with pen-and-paper travel diaries.
+These surveys have known methodological flaws.
+The high cost and burden to respondents limits the period of data collection.
+Short trips are often under reported  [@wolf_impact_2003] .
+The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].
 
 These obstacles can be overcome by using objective data on human mobility.
-The Global Positioning System (GPS), which uses the distance between a device and a number of satellites to determine location, provides such data.
-Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality [@harari_using_2016], academic performance [@wang_smartgpa:_2015] and bipolar disorder [@palmius_detecting_2017].
+Such data  can now be obtained using the Global Positioning System (GPS).
+GPS uses the distance between a device and several satellites to determine location.
+Within behavioural sciences, researchers have used GPS data to investigate wideranging topics.
+For example,  @zenk_how_2009 investigate the effects of the food environment on eating patterns.
+@harari_using_2016 look at the movement correlates of personality.
+@wang_smartgpa:_2015  does the same for academic performance.
+@palmius_detecting_2017 use mobility patterns to predict bipolar depression.
 
-In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*).
-However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants.
-Moreover, they may be biased because of the introduction of a new device to the participant's life.
-Because of these drawbacks, specialised logs usually span a short amount of time.
+In most studies participants receive a specialised GPS devices to track their mobility.
+We call resulting logs *specialised logs*.
+@barnett_inferring_2016 point out several methodological issues with these studies.
+High costs and a high burden to participants limits their scalability.
+Besides, introducing a new device to the participant's life may bias the data.
+Due to these drawbacks specialised logs usually span a short amount of time.
 @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
-Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_2017] (*secondary logs*).
-By law, logs can be accessed and shared by users for free [@commission_protecting_2017].
-Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate.
+Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) .
+For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017].
+Often, secondary logs span several years.
+By law, secondary logs are accessible to users for free [@commission_protecting_2017].
+Yet, secondary logs also present methodological challenges.
+They were created for non-academic purposes under engineering constraints.
+These constraints mean that sensors do not track users continuously.
+The resulting logs can be sparse and inaccurate.
 Hence, two important challenges are dealing with measurement noise and missing data.
 
-Missing data is a pervasive issue as it can arise due to multiple reasons.
+Missing data is a pervasive issue as it can arise due to several reasons.
 Technical reasons include signal loss, battery failure and device failure.
-Behavioural reasons include leaving the phone at home, switching the phone off, switching location measurements off, and so on.
-As a result, applied researchers are often left with wide temporal gaps with no measurements.
-For instance, different groups studying the effect of bipolar disorder on human movement have reported missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017].
-Similar trends are consistently reported in other fields [e.g.
+Behavioural reasons include leaving the phone at home or switching the device off.
+As a result, secondary logs often contain wide temporal gaps with no measurements.
+For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017].
+Other researchers report similar trends in other fields [e.g.
 @harari_using_2016;@jankowska_framework_2015].
 
-There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016].
-Traditional missing data methods, such as mean imputation, cannot be used easily in spatiotemporal data because the measurements are correlated in time and space.
-For example, assume the simplistic case that an individual spends almost half her time at work, half her time at home and a small amount of time commuting between the two along a angled path.
-Using mean imputation would result in imputed values of her being at the geometric midpoint between the home and work for all missing values, even though she has never been there and never will be.
-Worringly, @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.
+There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016].
+Traditional methods, such as mean imputation, are not suitable for spatiotemporal data.
+This is because measurements are often correlated in time and space.
+For example, imagine an individual who splits almost all her time between work or at home.
+Suppose she spends a small amount of time commuting between the two along a circular path.
+Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.
+She has never and will never be there!
+Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].
 
-The accuracy of GPS measurements in smartphones is substantially lower than in professional grade GPS trackers.
-For example, Android phones collect location information through a variety of methods, such as from WiFi access points, cellphone triangulation, and GPS measurements.
-They use different methods due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
-GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
-Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
+Smartphone location measurements are substantially less accurate than those of professional GPS trackers.
+Android phones collect location information through a variety of less accurate methods.
+Other than GPS measurements, Androids use WiFi access points and cellphone triangulation.
+Different methods are used because of computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].
+In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location.
+GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
+Unfortunately for researchers, this is where people in the developed world spend most of their time.
 
-On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as overestimating the movement of individuals.
-For instance, suppose that a naive researcher calculates the distance travelled by an individual by drawing a line between each measured point and calculating the sum of the length of all of these lines.
-If there is noise, the measurements will vary even though the individual is not moving.
-If the measurements are frequent, then the researcher will end up with a lot of movement, even though the individual did not move at all.
-The problem is further complicated by the fact that missing data and noisy measurements are related to each other.
-For instance, methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
-@palmius_detecting_2017] can exacerbate the severity of the missing data problem.
+Noisy data can lead to inaccurate conclusions if it is not accounted for.
+Suppose we wish to calculate an individual's movement in a day.
+A simple approach would be to calculate the sum of the distance between each measurement.
+But if there is noise, the coordinates will vary even though the individual is not moving.
+If the measurements are frequent and noisy, we will calculate a lot of movement.
+Even if the individual did not move at all!
+The problem is further complicated because missing data and noisy measurements are related.
+Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
+@palmius_detecting_2017],  can exacerbate the severity of the missing data problem.
 
@@
 Inaccurate measures are often followed by more accurate measures.
-Most notably, low accuracy tends to be associated with movement (Figure 2).
+Most notably, low accuracy often (but not always) is  associated with movement (Figure 2).
 Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.

### 2017-11-26 scripts/imageMagick.R

@@ -6,7 +6,6 @@ ac <- image_read("../img/accuracyLocShift.png")
 im2 <- c(inac,ac)
 x<- image_append(image_scale(im2,"x500"))
 image_write(x,"../img/accPlots.png")
-# doesnt; work
 
 
 image_mosaic(ac, inac)

### 2017-11-27 thesisReport.Rmd
new plots, scripts and added to the text

@@
 
-How people move about in their environment affects a wide range of outcomes.
-These include health, income and social capital [@goodchild_toward_2010].
+How people move about in their environment affects a wide range of outcomes, such as  health, income and social capital [@goodchild_toward_2010].
 A better understanding of mobility could lead to better health and urban-planning policies.
-A large part of studies on human mobility are conducted with pen-and-paper travel diaries.
-These surveys have known methodological flaws.
-The high cost and burden to respondents limits the period of data collection.
-Short trips are often under reported  [@wolf_impact_2003] .
+Yet a large part of studies on human mobility are conducted with pen-and-paper travel diaries, despite well known methodological flaws.
+The high cost and burden to respondents limits the span of data collection.
+Short trips are frequently under-reported  [@wolf_impact_2003] .
 The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].
@@
 
-In most studies participants receive a specialised GPS devices to track their mobility.
+In most studies participants receive a specialised GPS devices to track their movement.
 We call resulting logs *specialised logs*.
 @barnett_inferring_2016 point out several methodological issues with these studies.
-High costs and a high burden to participants limits their scalability.
-Besides, introducing a new device to the participant's life may bias the data.
+Like studies using pen-and-paper travel diaries, collecting specialised logs is costly and places a high burden on participants.
+Besides, introducing a new device to the participant's life may bias their behaviour.
 Due to these drawbacks specialised logs usually span a short amount of time.
@@
 Yet, secondary logs also present methodological challenges.
-They were created for non-academic purposes under engineering constraints.
-These constraints mean that sensors do not track users continuously.
-The resulting logs can be sparse and inaccurate.
+They were created for non-academic purposes under engineering constraints (detailed subsequently).
+These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate.
 Hence, two important challenges are dealing with measurement noise and missing data.
@@
 For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017].
-Other researchers report similar trends in other fields [e.g.
+Other researchers report similar trends in different fields [e.g.
 @harari_using_2016;@jankowska_framework_2015].
@@
 There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016].
-Traditional methods, such as mean imputation, are not suitable for spatiotemporal data.
-This is because measurements are often correlated in time and space.
-For example, imagine an individual who splits almost all her time between work or at home.
+Importantly, spatiotemporal data measurements are often correlated in time and space.
+This means that common methods, such as mean imputation, are unsuitable.
+For example, imagine an individual who splits almost all her time between work and home.
 Suppose she spends a small amount of time commuting between the two along a circular path.
@@
 
-Smartphone location measurements are substantially less accurate than those of professional GPS trackers.
-Android phones collect location information through a variety of less accurate methods.
-Other than GPS measurements, Androids use WiFi access points and cellphone triangulation.
-Different methods are used because of computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].
+The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers.
+Android phones collect location information through a variety of methods.
+Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation.
+Different methods are used because of computational and battery constraints.
+GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
 In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location.
@@
 But if there is noise, the coordinates will vary even though the individual is not moving.
-If the measurements are frequent and noisy, we will calculate a lot of movement.
-Even if the individual did not move at all!
+If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
 The problem is further complicated because missing data and noisy measurements are related.
@@
 
-In this paper we will compare methods used to deal with measurement error and missing data in mobility patterns from secondary GPS logs.
+In this paper we will explore in detail the problem of missing data and measurement error in secondary location logs. Moreover, we will  compare methods used to deal with these problems.
 
@@
 
-Given that there is little literature on dealing with missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
-The example dataset comes from the Google Location History of a single individual and spans from January 2013 to January 2017.
-It was recorded with multiple different Android devices and contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15).
-The dataset contains a wide range of variables including inferred activity and velocity.
-For the purposes of this paper we will focus only on latitude, longitude, accuracy (defined below) and a timestamp.
-
-## Location logs and notation
-
-GPS measurements report our location on a three dimensional planet, yet we are interested in placing these measurements on a two dimensional map.
-Projecting three dimensional measurements onto a two dimensional plane results in errors, in order to minimise these errors we borrow an error minimising projection method from @barnett_inferring_2016.
-
-Let a persons' true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
-Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
-In addition, let $a$ denote the estimated accuracy of the recorded data.
-accuracy.
-$G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ...
-< t_{n+1}$.
-For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window.
-The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
-If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+There is little literature on dealing with missing data in custom or secondary logs.
+Thus it is worth illustrating the typical characteristics using an example data set.
+The example data set comes from the Google Location History of a single individual.
+It spans from January 2013 to January 2017 and contains  814 941  measurements.
+The data set contains a multitude of variables, including inferred activity and velocity.
+We will focus on measurements of latitude, longitude, accuracy (defined below).
+All measurements are paired with a timestamp.
 
@@
 
-In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
+Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle.
+In the example data set the distribution of accuracy is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
 @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
@@
 
-```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement on a morning journey.
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017.
+This includes all measurements from midnight to midday.
 The red circles denote the accuracy of all logged measurement points (the raw data).
@@
 
-```{r, accuracyPlot2, echo=FALSE, fig.cap="The upper chart shows the rolling mean distance from the next point in meters over the course of the day.
+```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the rolling mean distance from the next point in meters over the course of the day.
+The first peak corresponds to the first journey from the users home a gym around 8am.
+The second, smaller peak before 10 reflects a journey from the gym to the lecture theatre.
+Both journies can be seen in Figure 1.
 The lower chart shows the rolling mean accuracy over the course of the day.
-In both cases, the rolling window is 3.",out.width = '100%'}
+In both cases, the rolling window is 3.
+The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
 include_graphics("img/accuracyLocShift.png")
@@
 
-Over 54% of the data is missing for the entire duration of the log (as shown in Figure X a).
-However,this is misleading as there are several long periods with no measurements whatsoever.
+Over 54% of the data is missing for the entire duration of the log.
+This may be misleading as there are several long periods with no measurements whatsoever (see Figure X a).
+For days which were not entirely missing, approximately 22% of all five minute segments were missing.
 The structure of missingness of a day with measurements is shown in Figure X b.
@@
 
-```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day.
+```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data on February 15th 2017.
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
@@
 
-## Current methods for imputing missing spatiotemporal data
+## Methods
 
-As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs.
-We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions.
-For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data.
-In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns.
-Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.
+## Two-dimensional projections and notation
 
-There is a vast literature of using SSMs to improve measurements accuracy and deal with missing data.
-Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008].
-These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
+GPS measurements provide us with coordinates on the surface of the earth.
+Because most mobility metrics are computed for data in $\mathbb{R}^2$,  we are interested in mapping these $\mathbb{R}^3$ measurements on a 2D Euclidean plane.
+Projecting three dimensional measurements onto a two dimensional plane results in distortion.
+To minimise errors we borrow an error minimising projection method from @barnett_inferring_2016.
+
+Having thus converted lattitude and longitude onto coordinates unique to each individual, let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
+Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+In addition, let $a$ denote the estimated accuracy of the recorded data.
+accuracy.
+$G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ...
+< t_{n+1}$.
+For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window.
+The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
+If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+
+## Selecting candidate models
+
+There is no golden standard or established practice in how to deal with missing data in GPS logs.
+Researchers are generally vague about what practices they follow [@jankowska_framework_2015].
+Ostensibly this is because they are unaware of possible solutions.
+In an attempt to elucidate the topic, we explore potential solutions.
+We will argue that extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns.
+We also discuss in detail two approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
+
+There is a vast literature on using SSMs in spatiotemporal statistics.
+For example, ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008].
+These models can be quite complex.
+@preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
 The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
-In fact, the extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
+The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
 The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
 
-However, the main limitation of SSMs is that they ignore the fact that humans have regular movement routines, such as going to work or shopping for groceries on weekends.
-This limitation is due to the fact that SSMs are based on the Markov property.
-Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
-Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve the performance of Markovian models, but these solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
-For this reason we do not consider SSMs to be useful for imputing missing data, although they could be of use in filtering noise.
+For us, the main limitation of SSMs is that they ignore regular movement routines.
+For instance, humans tend to go to work on weekdays and sleep at night.
+Because SSMs are based on the Markov property, they cannot incorporate this information.
+The estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
+Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models.
+These solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
+For this reason we do not consider SSMs to be useful for imputing missing data.
+Nonetheless, they could be of use in filtering noise.
 
-In addition to SSMs there are spatiotemporal imputation methods often used in climate or geological research for estimating missing data.
-For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
-Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.
+In climate or geological research spatiotemporal imputation methods are often used.
+For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ].
+The authors illustrate their example using rainfall data from gauging stations across Australia.
+Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values.
+The example provided relates to underground water data.
+Generally these models assume fixed measurement stations (such as rainfall gauging stations).
 
-The difficulty of applying this class of spatiotemporal imputation models to human mobility tracks lies in the fact that these models generally assume fixed measurement stations (such as rainfall gauging stations).
-While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
-To our knowledge such models have not been implemented for mobility traces.
+For this reason they cannot be easily applied to missing mobility tracks.
+@feng_cutoff:_2014 claim their model could be used to establish mobility patterns.
+This may be possible by dividing the sample space into rasters.
+Each raster would be analogous to a measurement station.
+These artificial stations could "measure" the probability of the individual being there.
+To our knowledge such models have not been implemented for mobility traces and seems computationally inefficient.
 
@@
 
-## Methods
-
 ### Datasets & Analyses
@@
                                            "Mean Accuracy"),
-      caption = "Table with descriptives about the data sets used to build the imputation methods. ")
+      caption = "Table with descriptives about the data sets used to build the imputation methods.
+Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.")
 ```
@@
 
-Analyses were performed using R and a multitude of other statistical packages [@ggplot2;@dplyr;@leaflet;@base;@sp1;@sp2].
+Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].
 
@@
 Thus in the end we end up with eight methods to evaluate, three for each filtering method as well as four for each imputation method.
-
-## References

### 2017-11-27 scripts/accuracyPlots.R

@@ -148,33 +148,8 @@ x<- bind_rows(day1,week1)
 ggplot(x, aes(x=unit, y=accuracy)) + geom_boxplot()
 
 
-# new scatterplot
-
-all %>%
-  select(time,lon,lat,accuracy) %>%
-  as_tbl_time(index = time) %>%
-  time_filter(2017-03-02 ~ 2017-03-02)%>%
-  mutate( time2 = lubridate::force_tz(time, "Asia/Singapore"),
-          distancePrev = c(0,
-                           sp::spDists(x=matrix(c(lon,lat),ncol = 2), segments = T)),
-          hour = as.POSIXct(
-            paste0("2014-01-22 ",
-                   strftime(time,format = "%H:%M", tz = "Europe/Budapest")))) %>%
-  ggplot( aes(x = hour, y=accuracy))+
-  geom_point()+theme_tufte()
-
 #wtf happened 2nd of march
 
-all %>%
-  select(time,lat,lon,accuracy) %>%
-  as_tbl_time(index = time) %>%
-  time_filter(2017-02-15 ~ 2017-02-17) %>%
-  filter(accuracy > 1000) %>% 
-  leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
-  addTiles() %>%
-  addCircles(lng = ~lon, lat = ~lat,
-             radius = ~accuracy, fillOpacity = 0.02,color = "#DF2935")%>%
-  addProviderTiles(providers$CartoDB.Positron)
 
 all %>%
   select(time,lat,lon,accuracy) %>%

### 2017-11-27 scripts/accuracyTimeOfDay.R

@@ -1,0 +1,16 @@
+# accuracy and time of day plot
+
+# new scatterplot
+
+all %>%
+  select(time,lon,lat,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2016-10-02 ~ 2016-10-07)%>%
+  mutate( time2 = lubridate::force_tz(time, "Asia/Singapore"),
+          distancePrev = c(0,
+                           sp::spDists(x=matrix(c(lon,lat),ncol = 2), segments = T)),
+          hour = as.POSIXct(
+            paste0("2014-01-22 ",
+                   strftime(time,format = "%H:%M", tz = "Europe/Budapest")))) %>%
+  ggplot( aes(x = hour, y=accuracy))+
+  geom_point()+theme_tufte()

### 2017-11-27 scripts/dailyMissPlot.R

@@ -1,0 +1,60 @@
+# daily missing plot 
+
+# by 5 minute version 2
+m5minv2<- all %>%
+  select(time,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2000-01 ~ 2017-06) %>% 
+  thicken('5 min') %>%
+  group_by(time_5_min) %>%
+  summarise(measurements = n(),
+            accuracy = mean(accuracy)) %>%
+  pad() %>%
+  fill_by_value(value = 0)%>%
+  mutate(
+    hour = as.POSIXct(
+      paste0("2014-01-22 ",
+             strftime(time_5_min,format = "%H:%M", tz = "Europe/Budapest"))),
+    time2 = lubridate::force_tz(time_5_min, "Asia/Singapore"),
+    missing = case_when(
+      measurements== 0 ~ 1,
+      measurements > 0 ~ 0)
+  )
+
+exampleMiss<- m5minv2 %>%
+  as_tbl_time(index = time2)%>%
+  time_filter(2017-02-15 ~ 2017-02-15) %>%
+  ggplot( aes(x = hour, y=measurements, colour = factor(missing)))+
+  geom_point()+theme_tufte()+
+  xlab("")+
+  ylab("Measurements")+
+  ggtitle("Measurements per 5 minute window")+
+  scale_x_datetime(breaks = date_breaks("4 hour"),
+                   minor_breaks=date_breaks("2 hour"),
+                   labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
+  scale_colour_manual(values=c("black",muted("red")),labels=c("Not Missing", "Missing"))+
+  labs(x = NULL, colour = "")
+
+ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")
+
+# randomly select two days
+sample(x = c(13,14,15,16,17),1)#YEAR
+sample(x = c(1:12),1)#YEAR::
+sample(c(1:31),1)#day
+
+exampleMiss<- m5minv2 %>%
+  as_tbl_time(index = time2)%>%
+  time_filter(2014-10-27 ~ 2014-10-27) %>%
+  ggplot( aes(x = hour, y=measurements, colour = factor(missing)))+
+  geom_point()+theme_tufte()+
+  xlab("")+
+  ylab("Measurements")+
+  ggtitle("Measurements per 5 minute window")+
+  scale_x_datetime(breaks = date_breaks("4 hour"),
[…]

### 2017-11-27 scripts/missingnessPlot.R

@@ -76,44 +76,8 @@ missingDay5min<- ggplot(m5min, aes(x = time_5_min_day, y=measurements, colour= m
                          na.value = "grey50", guide = "colourbar")+
   labs(x = NULL, colour = "Missing \n5 minute\n segments")
 
-ggsave(missingDay5min,filename = "../img/missingdayBoaz5min.png",device = "png",height = 6.5, , width = 18, units = "cm")
+ggsave(missingDay5min,filename = "../img/missingdayBoaz5min.png",device = "png",height = 6.5, width = 18, units = "cm")
 
-# by 5 minute version 2
-m5minv2<- all %>%
-  select(time,accuracy) %>%
-  as_tbl_time(index = time) %>%
-  time_filter(2000-01 ~ 2017-06) %>% 
-  thicken('5 min') %>%
-  group_by(time_5_min) %>%
-  summarise(measurements = n(),
-            accuracy = mean(accuracy)) %>%
-  pad() %>%
-  fill_by_value(value = 0)%>%
-  mutate(
-    hour = as.POSIXct(
-      paste0("2014-01-22 ",
-           strftime(time_5_min,format = "%H:%M", tz = "Europe/Budapest"))),
-    time2 = lubridate::force_tz(time_5_min, "Asia/Singapore"),
-    missing = case_when(
-      measurements== 0 ~ 1,
-      measurements > 0 ~ 0)
-      )
-
-exampleMiss<- m5minv2 %>%
-  as_tbl_time(index = time2)%>%
-  time_filter(2017-02-15 ~ 2017-02-15) %>%
-  ggplot( aes(x = hour, y=measurements, colour = factor(missing)))+
-  geom_point()+theme_tufte()+
-  xlab("")+
-  ylab("Measurements")+
-  ggtitle("Measurements per 5 minute window")+
-  scale_x_datetime(breaks = date_breaks("4 hour"),
-                   minor_breaks=date_breaks("2 hour"),
-                   labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
-  scale_colour_manual(values=c("black",muted("red")))+
-  theme(legend.position="none")
-
-ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")
 
 
 time1_p1 <- strptime(paste("2017-02-14", "00:00:00"), "%Y-%m-%d %H:%M:%S")

### 2017-11-27 scripts/rollAccOverTime.R

@@ -1,0 +1,33 @@
+# rolling accuracy over time
+
+
+r100mean<- rollify(mean,window = 100)
+
+r100mean(1:100000)
+# by 5 minute
+m5min<- all %>%
+  select(time, accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2000-01 ~ 2017-06) %>% 
+  thicken('day') %>%
+  group_by(time_day) %>%
+  summarise(accuracy = mean(accuracy)) %>%
+  pad() %>%
+  fill_by_value(value = NA) %>%
+  ggplot(aes(x = time_day, y=accuracy,))+
+  geom_point()+theme_tufte()+
+  xlab("")+
+  ylab("Measurements")+
+  ggtitle("Missingness in daily measurements over time")+
+  scale_colour_gradient2(low = "black", mid = "grey50",
+                         high = muted("red"), midpoint = 144, space = "Lab",
+                         na.value = "grey50", guide = "colourbar")+
+  labs(x = NULL, colour = "Missing \n5 minute\n segments")
+
+all %>%
+  select(time, accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2016-01 ~ 2017-06) %>%
+  mutate(rollingAcc = r100mean(accuracy))%>%
+  ggplot(aes(x = time, y=rollingAcc))+
+  geom_point()+theme_tufte()

### 2017-11-30 thesisReport.Rmd
added pdf
added lables to accuracy plot, changed Y variable

@@
 GPS uses the distance between a device and several satellites to determine location.
+GPS measurements can be used to infer a vast range of socioeconomic and behavioural measures, including where the individual lives, how much time he or she spends at home and where (and how) they travel.
+
 Within behavioural sciences, researchers have used GPS data to investigate wideranging topics.
 For example,  @zenk_how_2009 investigate the effects of the food environment on eating patterns.
-@harari_using_2016 look at the movement correlates of personality.
-@wang_smartgpa:_2015  does the same for academic performance.
+@harari_using_2016 look at the movement correlates of personality, finding that extrovers individuals spend less time at home than introverts.
+@wang_smartgpa:_2015  looks at how academic performance is affected by movement patterns.
 @palmius_detecting_2017 use mobility patterns to predict bipolar depression.
@@
 If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
+This issue is also visualised in Figure 1.
 The problem is further complicated because missing data and noisy measurements are related.
@@
 
-```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the rolling mean distance from the next point in meters over the course of the day.
-The first peak corresponds to the first journey from the users home a gym around 8am.
-The second, smaller peak before 10 reflects a journey from the gym to the lecture theatre.
-Both journies can be seen in Figure 1.
-The lower chart shows the rolling mean accuracy over the course of the day.
-In both cases, the rolling window is 3.
+```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
+All journeys are marked with a red line.
+The first peak corresponds to the first journey from the user's home to a gym around 8am.
+The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
+Both journeys can be seen in Figure 1.
+All other journeys are not shown in Figure 1.
+The large jump between journey 5 and 6 is measurement error.
+The lower chart shows the accuracy over the course of the day.
 The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}

### 2017-11-30 scripts/exploratoryRawDataViz.R

@@ -2,6 +2,7 @@
 Data_Start<- as.POSIXct("2015-09-07 01:55:00")
 Data_End  <- as.POSIXct("2015-09-10 01:59:00")
 
+library(leaflet)
 exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
   
   # helper function
@@ -35,7 +36,6 @@ exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
     select(time,lat,lon,accuracy) %>%
     as_tbl_time(index = time) %>%
     time_filter(time_formula = time_formula) %>%
-    filter(accuracy > 1000) %>% 
     leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
     addTiles() %>%
     addCircles(lng = ~lon, lat = ~lat,
@@ -44,5 +44,7 @@ exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
   
 }
 
-exploreDates("2014-10-27","2014-10-27")
+exploreDates("2017-02-15+20:55:00","2017-02-15+23:55:00")
 
++12:00:00
++18:00:00

### 2017-11-30 scripts/rollDistanceAccuracyPlot.R

@@ -3,10 +3,27 @@
 #
 
 rolmean<- rollify(mean,window = 3)
+rolmax<- rollify(max,window = 3)
 
 home<- c(5.113919,52.10421)
 
-rMeanD<- all %>%
+events <- data.frame(name = c("Journey 1", "Journey 2", "Journey 3","Journey 4","Journey 5","Journey 6"),
+                     time =c(as.POSIXct("2017-02-15 08:00:00",
+                                      tz= "Europe/Budapest"),
+                     as.POSIXct("2017-02-15 10:00:00",
+                                tz= "Europe/Budapest"),
+                     as.POSIXct("2017-02-15 14:30:00",
+                                tz= "Europe/Budapest"),
+                     as.POSIXct("2017-02-15 17:30:00",
+                                tz= "Europe/Budapest"),
+                     as.POSIXct("2017-02-15 19:00:00",
+                                tz= "Europe/Budapest"),
+                     as.POSIXct("2017-02-15 20:45:00",
+                                tz= "Europe/Budapest"))
+                     )
+
+  
+distanceTime<- all %>%
   select(time,lon,lat,accuracy) %>%
   as_tbl_time(index = time) %>%
   time_filter(2017-02-15 ~ 2017-02-15)%>%
@@ -17,16 +34,19 @@ rMeanD<- all %>%
           euclat = lat-home[2],
           distancePrev = c(0,
                            sp::spDists(x=matrix(c(lon,lat),ncol = 2), segments = T)),
-          rDistPrev = rolmean(distancePrev)) %>%
+          rDistPrev = distancePrev) %>%
   ggplot( aes(x = time, y=rDistPrev*1000))+
   geom_line()+theme_tufte()+
-  xlab("")+ylab("Rolling Mean\nDistance")+
+  xlab("")+ylab("Distance over time")+
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
                    labels=date_format("%H:%M:%S", tz = "Europe/Budapest"),
-                   limits = xlim_p1)
+                   limits = xlim_p1)+
+  geom_text(data=events,
+            aes(x=time,y=50,label=name),size=3,angle = 90, colour= muted("red"))+
+  ylim(c(-1,57))
 
-rMeanAcc <- all %>%
+accuracyTime <- all %>%
[…]

### 2017-11-30 thesisReport.Rmd
bookdownified the thesis

@@
   - rReferences.bib
+biblio-style: apalike
+link-citations: yes
 output:
-  rmarkdown::pdf_document:
+  bookdown::pdf_book:
     citation_package: natbib
@@
 
-# A concrete example
+## A concrete example
 
@@
 
-## Methods
+# Methods
 

### 2017-12-01 thesisReport.Rmd
added aggregation section with plot, figure table references

@@
 
+### Aggregate measures
+
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled.
+As we discussed earlier, aggregations without data processing can be highly biased.
+However, as an example we calculate time spent at home of the user for each day of the week in the month of February (\@ref(fig:aggrePlot)).
+Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].
+
+```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017.
+We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+Missing values are filled in by the previous observed value.
+Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
+Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
+include_graphics("img/timeUse.png")
+```
+
 ### Accuracy in location logs
@@
 We observe similar patterns in secondary logs.
-Figure 1 shows how accuracy can vary as a function of user behaviour, time and location.
+\@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location.
 Inaccurate measures are often followed by more accurate measures.
-Most notably, low accuracy often (but not always) is  associated with movement (Figure 2).
+Most notably, low accuracy often (but not always) is  associated with movement (\@ref(fig:accuracyPlot2)).
 Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.
@@
 Over 54% of the data is missing for the entire duration of the log.
-This may be misleading as there are several long periods with no measurements whatsoever (see Figure X a).
+This may be misleading as there are several long periods with no measurements whatsoever (see Figure \@ref(fig:longMeasurementsPerDay)).
 For days which were not entirely missing, approximately 22% of all five minute segments were missing.
-The structure of missingness of a day with measurements is shown in Figure X b.
+The structure of missingness of a day with measurements is shown in Figure \@ref(fig:measurementsPerDay).
 As you can see, there are several long periods over the course of the log for which there are no measurements.

### 2017-12-01 scripts/aggregation.R

@@ -1,0 +1,45 @@
+# aggregator
+
+library(ggthemes)
+
+x<- all %>%
+  select(time,lon,lat,accuracy) %>%
+  as_tbl_time(index = time) %>%
+  time_filter(2017-02 ~ 2017-02) %>% 
+  thicken('5 min', colname = 'time2') %>%
+  group_by(time2) %>%
+  summarise(lon = mean(lon),
+            lat = mean(lat),
+            accuracy = mean(accuracy)
+  )%>% 
+  pad()%>%
+  tidyr::fill(c(lon,lat)) %>%
+  mutate(distanceHome = sp::spDistsN1(pts=matrix(c(lon,lat),ncol = 2),
+                                      pt = home,
+                                      longlat = T))
+  
+colnames(x)[1] <- "time"
+exploreDates(x,"2017-02-15","2017-02-15")
+# time spent at home
+
+
+ggplot(x,aes(x = distanceHome, y = time2))+
+  geom_point()
+
+athome<- x %>% mutate(atHome = case_when(distanceHome <0.25 ~ 1,
+                                distanceHome >0.25 ~ 0)) %>%
+  thicken('day', colname = 'time') %>%
+  group_by(time) %>%
+  summarise(atHome2 = sum(atHome)/288) %>%
+  mutate(day = strftime(time,"%A"))
+
+athome$day <- factor(athome$day, levels= c("Monday", 
+                                         "Tuesday", "Wednesday",
+                                         "Thursday", "Friday", "Saturday","Sunday"))
+
+timeUse<- ggplot(athome,aes(x = day, y = atHome2))+
+  geom_point()+theme_tufte()+xlab("Day of the week")+
+  ylab("Proportion of time\nspent at home")+ggtitle("Time spent at home - February 2017")
+
+ggsave(timeUse,filename = "../img/timeUse.png",device = "png",height = 6.5,width = 20, units = "cm")
+  

### 2017-12-01 scripts/dailyMissPlot.R

@@ -49,12 +49,12 @@ exampleMiss<- m5minv2 %>%
   geom_point()+theme_tufte()+
   xlab("")+
   ylab("Measurements")+
-  ggtitle("Measurements per 5 minute window")+
+  ggtitle("Measurements per 5 minute window",subtitle = "2014-10-27")+
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
                    labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
   scale_colour_manual(values=c("black",muted("red")),labels=c("Not Missing", "Missing"))+
   labs(x = NULL, colour = "")
-
-ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")
+missingBoaz5minExample20141027.png
+ggsave(exampleMiss,filename = "../img/missingBoaz5minExample20141027.png",device = "png",height = 6.5, width = 18, units = "cm")
 

### 2017-12-01 scripts/exploratoryRawDataViz.R

@@ -3,7 +3,7 @@ Data_Start<- as.POSIXct("2015-09-07 01:55:00")
 Data_End  <- as.POSIXct("2015-09-10 01:59:00")
 
 library(leaflet)
-exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
+exploreDates <- function(df = all, Data_Start= "2017-02-15", Data_End = "2017-02-17"){
   
   # helper function
   create_time_formula <- function(lhs, rhs) {
@@ -32,7 +32,7 @@ exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
   
   time_formula <- create_time_formula(Data_Start, Data_End)
   # new scatterplot
-  all %>%
+  df %>%
     select(time,lat,lon,accuracy) %>%
     as_tbl_time(index = time) %>%
     time_filter(time_formula = time_formula) %>%
@@ -41,10 +41,9 @@ exploreDates <- function(Data_Start= "2017-02-15", Data_End = "2017-02-17"){
     addCircles(lng = ~lon, lat = ~lat,
                radius = ~accuracy, fillOpacity = 0.02,color = "#DF2935")%>%
     addProviderTiles(providers$CartoDB.Positron)
-  
 }
 
-exploreDates("2017-02-15+20:55:00","2017-02-15+23:55:00")
+exploreDates(all,"2017-02-15+20:55:00","2017-02-15+23:55:00")
 
 +12:00:00
 +18:00:00

### 2017-12-01 scripts/testingApa.Rmd
made it APAstyle

---
title             : "The title"
shorttitle        : "Title"

author: 
  - name          : "First Author"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"

affiliation:
  - id            : "1"
    institution   : "Wilhelm-Wundt-University"
  - id            : "2"
    institution   : "Konstanz Business School"

author_note: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Enter abstract here. Each new line herein must be indented, like this line.

keywords          : "keywords"
wordcount         : "X"

bibliography      : []

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
library("dplyr")

```

```{r analysis_preferences, results='asis'}
# Seed for random number generation
set.seed(42)
descriptives <-data.frame(hello = 1:5, you = 1:5)

apa_table(
  descriptives
  , caption = "Descriptive statistics of correct recall by dosage."
  , note = "This table was created with apa_table()"
)

```

# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

# Discussion

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

### 2017-12-01 thesisReportApa.Rmd
added aggregation section with plot, figure table references
made it APAstyle

@@
     corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
+    address       : ""
     email         : "boaz@boazsobrado.com"
@@
 
-keywords          : "GPS, Human Mobility"
-wordcount         : "X"
+keywords          : "Missing data, Measurement Bias,GPS, Human Mobility"
+wordcount         : "2205"
 
@@
   - rReferences.bib
+biblio-style: apalike
 
@@
 citation_package: natbib
+
 ---
@@
 library(knitr)
+library(kableExtra)
 ```
 
-How active people are and how they interact with their environment affects a wide range of outcomes including health, income and social capital [@goodchild_toward_2010].
-A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies.
-Yet a large part of studies on human mobility are largely based on pen-and-paper travel diaries.
-These surveys have known methodological flaws, such as the short period of data collection (due to costs and burden to respondents), the underreporting of short trips [@wolf_impact_2003] and the underestimation of the duration of commutes [@delclos-alio_keeping_2017].
+How people move about in their environment affects a wide range of outcomes, such as  health, income and social capital [@goodchild_toward_2010].
+A better understanding of mobility could lead to better health and urban-planning policies.
+Yet a large part of studies on human mobility are conducted with pen-and-paper travel diaries, despite well known methodological flaws.
+The high cost and burden to respondents limits the span of data collection.
+Short trips are frequently under-reported  [@wolf_impact_2003] .
+The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].
 
 These obstacles can be overcome by using objective data on human mobility.
-The Global Positioning System (GPS), which uses the distance between a device and a number of satellites to determine location, provides such data.
-Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality [@harari_using_2016], academic performance [@wang_smartgpa:_2015] and bipolar disorder [@palmius_detecting_2017].
+Such data  can now be obtained using the Global Positioning System (GPS).
+GPS uses the distance between a device and several satellites to determine location.
+GPS measurements can be used to infer a vast range of socioeconomic and behavioural measures, including where the individual lives, how much time he or she spends at home and where (and how) they travel.
 
-In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*).
-However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants.
-Moreover, they may be biased because of the introduction of a new device to the participant's life.
-Because of these drawbacks, specialised logs usually span a short amount of time.
+Within behavioural sciences, researchers have used GPS data to investigate wideranging topics.
+For example,  @zenk_how_2009 investigate the effects of the food environment on eating patterns.
+@harari_using_2016 look at the movement correlates of personality, finding that extrovers individuals spend less time at home than introverts.
+@wang_smartgpa:_2015  looks at how academic performance is affected by movement patterns.
+@palmius_detecting_2017 use mobility patterns to predict bipolar depression.
+
+In most studies participants receive a specialised GPS devices to track their movement.
+We call resulting logs *specialised logs*.
+@barnett_inferring_2016 point out several methodological issues with these studies.
+Like studies using pen-and-paper travel diaries, collecting specialised logs is costly and places a high burden on participants.
+Besides, introducing a new device to the participant's life may bias their behaviour.
+Due to these drawbacks specialised logs usually span a short amount of time.
 @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
-Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_2017] (*secondary logs*).
-By law, logs can be accessed and shared by users for free [@commission_protecting_2017].
-Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate.
+Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) .
+For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017].
+Often, secondary logs span several years.
+By law, secondary logs are accessible to users for free [@commission_protecting_2017].
+Yet, secondary logs also present methodological challenges.
+They were created for non-academic purposes under engineering constraints (detailed subsequently).
+These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate.
 Hence, two important challenges are dealing with measurement noise and missing data.
 
-Missing data is a pervasive issue as it can arise due to multiple reasons.
+Missing data is a pervasive issue as it can arise due to several reasons.
 Technical reasons include signal loss, battery failure and device failure.
-Behavioural reasons include leaving the phone at home, switching the phone off, switching location measurements off, and so on.
-As a result, applied researchers are often left with wide temporal gaps with no measurements.
-For instance, different groups studying the effect of bipolar disorder on human movement have reported missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017].
-Similar trends are consistently reported in other fields [e.g.
+Behavioural reasons include leaving the phone at home or switching the device off.
+As a result, secondary logs often contain wide temporal gaps with no measurements.
+For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017].
+Other researchers report similar trends in different fields [e.g.
 @harari_using_2016;@jankowska_framework_2015].
 
-There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016].
-Traditional missing data methods, such as mean imputation, cannot be used easily in spatiotemporal data because the measurements are correlated in time and space.
-For example, assume the simplistic case that an individual spends almost half her time at work, half her time at home and a small amount of time commuting between the two along a angled path.
-Using mean imputation would result in imputed values of her being at the geometric midpoint between the home and work for all missing values, even though she has never been there and never will be.
-Worringly, @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.
+There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016].
+Importantly, spatiotemporal data measurements are often correlated in time and space.
+This means that common methods, such as mean imputation, are unsuitable.
+For example, imagine an individual who splits almost all her time between work and home.
+Suppose she spends a small amount of time commuting between the two along a circular path.
+Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.
+She has never and will never be there!
+Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].
 
-The accuracy of GPS measurements in smartphones is substantially lower than in professional grade GPS trackers.
-For example, Android phones collect location information through a variety of methods, such as from WiFi access points, cellphone triangulation, and GPS measurements.
-They use different methods due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location.
-GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
-Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.
+The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers.
+Android phones collect location information through a variety of methods.
+Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation.
+Different methods are used because of computational and battery constraints.
+GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
+In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location.
+GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013].
+Unfortunately for researchers, this is where people in the developed world spend most of their time.
 
-On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as overestimating the movement of individuals.
-For instance, suppose that a naive researcher calculates the distance travelled by an individual by drawing a line between each measured point and calculating the sum of the length of all of these lines.
-If there is noise, the measurements will vary even though the individual is not moving.
-If the measurements are frequent, then the researcher will end up with a lot of movement, even though the individual did not move at all.
-The problem is further complicated by the fact that missing data and noisy measurements are related to each other.
-For instance, methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
-@palmius_detecting_2017] can exacerbate the severity of the missing data problem.
+Noisy data can lead to inaccurate conclusions if it is not accounted for.
+Suppose we wish to calculate an individual's movement in a day.
+A simple approach would be to calculate the sum of the distance between each measurement.
+But if there is noise, the coordinates will vary even though the individual is not moving.
+If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
+This issue is also visualised in Figure \@ref(fig:accuracyPlot).
+The problem is further complicated because missing data and noisy measurements are related.
+Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g.
+@palmius_detecting_2017],  can exacerbate the severity of the missing data problem.
 
-In this paper we will compare methods used to deal with measurement error and missing data in mobility patterns from secondary GPS logs.
+In this paper we will explore in detail the problem of missing data and measurement error in secondary location logs. Moreover, we will  compare methods used to deal with these problems.
 
-# A concrete example
+## A concrete example
 
-Given that there is little literature on dealing with missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set.
-The example dataset comes from the Google Location History of a single individual and spans from January 2013 to January 2017.
-It was recorded with multiple different Android devices and contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15).
-The dataset contains a wide range of variables including inferred activity and velocity.
-For the purposes of this paper we will focus only on latitude, longitude, accuracy (defined below) and a timestamp.
+There is little literature on dealing with missing data in custom or secondary logs.
+Thus it is worth illustrating the typical characteristics using an example data set.
+The example data set comes from the Google Location History of a single individual.
+It spans from January 2013 to January 2017 and contains  814 941  measurements.
+The data set contains a multitude of variables, including inferred activity and velocity.
+We will focus on measurements of latitude, longitude, accuracy (defined below).
+All measurements are paired with a timestamp.
 
-## Location logs and notation
+### Aggregate measures
 
-GPS measurements report our location on a three dimensional planet, yet we are interested in placing these measurements on a two dimensional map.
-Projecting three dimensional measurements onto a two dimensional plane results in errors, in order to minimise these errors we borrow an error minimising projection method from @barnett_inferring_2016.
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled.
+As we discussed earlier, aggregations without data processing can be highly biased.
+However, as an example we calculate time spent at home of the user for each day of the week in the month of February (Figure \@ref(fig:aggrePlot)).
+Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].
 
-Let a persons' true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
-Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
-In addition, let $a$ denote the estimated accuracy of the recorded data.
-accuracy.
-$G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ...
-< t_{n+1}$.
-For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window.
-The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
-If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017.
+We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+Missing values are filled in by the previous observed value.
+Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
+Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
+include_graphics("img/timeUse.png")
+```
 
@@
 
-In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
+Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle.
+In the example data set the distribution of accuracy is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km.
 @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
 We observe similar patterns in secondary logs.
-Figure 1 shows how accuracy tends to vary as a function of user behaviour, time and location.
-There are several recurring low-accuracy points, most likely cell-phone triangulation towers.
+Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location.
+Inaccurate measures are often followed by more accurate measures.
+Most notably, low accuracy often (but not always) is  associated with movement (Figure \@ref(fig:accuracyPlot2)).
+Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.
+There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.
 
-```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement in a morning journey.
+```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017.
+This includes all measurements from midnight to midday.
 The red circles denote the accuracy of all logged measurement points (the raw data).
@@
 
-```{r, accuracyPlot2, echo=FALSE, fig.cap="The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
-The colour scale shows the distance between the measurement and the previous point.",out.width = '100%'}
+```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
+All journeys are marked with a red line.
+The first peak corresponds to the first journey from the user's home to a gym around 8am.
+The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
+Both journeys can be seen in Figure 1.
+All other journeys are not shown in Figure 1.
+The large jump between journey 5 and 6 is measurement error.
+The lower chart shows the accuracy over the course of the day.
+The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
 include_graphics("img/accuracyLocShift.png")
@@
 Over 54% of the data is missing for the entire duration of the log.
-However,this is misleading as there are several long periods with no measurements whatsoever.
-The structure of missingness of a day with measurements is shown below.
+This may be misleading as there are several long periods with no measurements whatsoever (see Figure \@ref(fig:longMeasurementsPerDay)).
+For days which were not entirely missing, approximately 22% of all five minute segments were missing.
+The structure of missingness of a day with measurements is shown in Figure \@ref(fig:measurementsPerDay).
+As you can see, there are several long periods over the course of the log for which there are no measurements.
+Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.
 
-```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day.
+```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of the log.
+The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
+For this day there were several periods with no information.
+These points are filled with red and lie on the x-axis.",out.width = '100%'}
+
+include_graphics("img/missingdayBoaz5min.png")
+
+```
+
+```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data on February 15th 2017.
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
@@
 include_graphics("img/missingBoaz5minExample.png")
-
 ```
 
-## Current methods for imputing missing spatiotemporal data
+# Methods
 
-As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs.
-We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions.
-For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data.
-In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns.
-Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.
+## Two-dimensional projections and notation
 
-There is a vast literature of using SSMs to improve measurements accuracy and deal with missing data.
-Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008].
-These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
+GPS measurements provide us with coordinates on the surface of the earth.
+Because most mobility metrics are computed for data in $\mathbb{R}^2$,  we are interested in mapping these $\mathbb{R}^3$ measurements on a 2D Euclidean plane.
+Projecting three dimensional measurements onto a two dimensional plane results in distortion.
+To minimise errors we borrow an error minimising projection method from @barnett_inferring_2016.
+
+Having thus converted lattitude and longitude onto coordinates unique to each individual, let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
+Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+In addition, let $a$ denote the estimated accuracy of the recorded data.
+accuracy.
+$G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ...
+< t_{n+1}$.
+For simplicity, let each entry in the discrete index set $t$ represent a 5 minute window.
+The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
+If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+
+When several data sets are available from individuals living in overlapping areas we can construct a $t \times i$ matrix $M$ where the entry $M(t,i)$ contains $G(t)$ for the individual $i$.
+
+## Selecting candidate models
+
+There is no golden standard or established practice in how to deal with missing data in GPS logs.
+Researchers are generally vague about what practices they follow [@jankowska_framework_2015].
+Ostensibly this is because they are unaware of possible solutions.
+In an attempt to elucidate the topic, we explore potential solutions.
+We will argue that extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns.
+We also discuss in detail two approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
+
+There is a vast literature on using SSMs in spatiotemporal statistics.
+For example, ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008].
+These models can be quite complex.
+@preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
 The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems.
-In fact, the extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
+The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013].
 The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.
 
-However, the main limitation of SSMs is that they ignore the fact that humans have regular movement routines, such as going to work or shopping for groceries on weekends.
-This limitation is due to the fact that SSMs are based on the Markov property.
-Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
-Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve the performance of Markovian models, but these solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
-For this reason we do not consider SSMs to be useful for imputing missing data, although they could be of use in filtering noise.
+For us, the main limitation of SSMs is that they ignore regular movement routines.
+For instance, humans tend to go to work on weekdays and sleep at night.
+Because SSMs are based on the Markov property, they cannot incorporate this information.
+The estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
+Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models.
+These solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
+For this reason we do not consider SSMs to be useful for imputing missing data.
+Nonetheless, they could be of use in filtering noise.
 
-In addition to SSMs there are spatiotemporal imputation methods often used in climate or geological research for estimating missing data.
-For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia.
-Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.
+In climate or geological research spatiotemporal imputation methods are often used.
+For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ].
+The authors illustrate their example using rainfall data from gauging stations across Australia.
+Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values.
+The example provided relates to underground water data.
+Generally these models assume fixed measurement stations (such as rainfall gauging stations).
 
-The difficulty of applying this class of spatiotemporal imputation models to human mobility tracks lies in the fact that these models generally assume fixed measurement stations (such as rainfall gauging stations).
-While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible.
-To our knowledge such models have not been implemented for mobility traces.
+For this reason they cannot be easily applied to missing mobility tracks.
+@feng_cutoff:_2014 claim their model could be used to establish mobility patterns.
+This may be possible by dividing the sample space into rasters.
+Each raster would be analogous to a measurement station.
+These artificial stations could "measure" the probability of the individual being there.
+To our knowledge such models have not been implemented for mobility traces and seems computationally inefficient.
 
@@
 
-## Methods
-
 ### Datasets & Analyses
 
-The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (table 1).
+The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (Table \@ref(tab:datadetailsTable)).
 
-```{r datadetailsTable, echo = F}
+```{r datadetailsTable, echo = F, results ='asis'}
 datadetails <- readRDS("tempdata/datadescriptives.rds")
 datadetails <- datadetails[,-7] #remove last column for now
-kable(datadetails,digits = 2,col.names = c("Log duration",
+apa_table(datadetails,digits = 2,col.names = c("Log duration",
                                            "Logged days",
                                            "Observations",
-                                           "Missing days",
-                                           "Missing data",
+                                           "Missing\ndays",
+                                           "Missing\ndata",
                                            "Mean Accuracy"),
-      caption = "Table with descriptives about the data sets used to build the imputation methods. ")
+      caption = "Table with descriptives about the data sets used to build the imputation methods.
+Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.",landscape = T)
 ```
@@
 1. Mean imputation as described by @palmius_detecting_2017.
-2. The model developed by @barnett_inferring_2016 using spatial, temporal and circadian proximity.
+2. The model developed by @barnett_inferring_2016 using both spatial and temporal proximity.
 3. Simple linear interpolation was used as a benchmark model.
@@
 
+Each imputation method used each of the three filtering methods as an input.
+Thus in the end we end up with eight methods to evaluate, three for each filtering method as well as four for each imputation method.
+
 \newpage

### 2017-12-01 thesisReportApa.tex
made it APAstyle

@@
 
-                      Correspondence concerning this article should be addressed to Boaz Sobrado, Postal address. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
+                      Correspondence concerning this article should be addressed to Boaz Sobrado. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
                 }
@@
 currently used methods, such as linear interpolation.}
-  \keywords{GPS, Human Mobility \\
+  \keywords{Missing data, Measurement Bias,GPS, Human Mobility \\
 
-    \indent Word count: X
+    \indent Word count: 2205
   }
@@
 
-How active people are and how they interact with their environment
-affects a wide range of outcomes including health, income and social
-capital (Goodchild \& Janelle, 2010). A better understanding of both
-within-person and between-person variability in geospatial patterns
-could be conducive to better social, health and urban-planning policies.
-Yet a large part of studies on human mobility are largely based on
-pen-and-paper travel diaries. These surveys have known methodological
-flaws, such as the short period of data collection (due to costs and
-burden to respondents), the underreporting of short trips (Wolf,
-Oliveira, \& Thompson, 2003) and the underestimation of the duration of
-commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
+How people move about in their environment affects a wide range of
+outcomes, such as health, income and social capital (Goodchild \&
+Janelle, 2010). A better understanding of mobility could lead to better
+health and urban-planning policies. Yet a large part of studies on human
+mobility are conducted with pen-and-paper travel diaries, despite well
+known methodological flaws. The high cost and burden to respondents
+limits the span of data collection. Short trips are frequently
+under-reported (Wolf, Oliveira, \& Thompson, 2003) . The self-reported
+duration of commutes is often underestimated (Delclòs-Alió, Marquet, \&
+Miralles-Guasch, 2017).
 
 These obstacles can be overcome by using objective data on human
-mobility. The Global Positioning System (GPS), which uses the distance
-between a device and a number of satellites to determine location,
-provides such data. Within behavioural science, this type of data has
-been used to investigate topics such as the effects of the food
-environment on eating patterns (Zenk, Schulz, \& Odoms-Young, 2009), the
-movement correlates of personality (G. M. Harari et al., 2016), academic
-performance (Wang, Harari, Hao, Zhou, \& Campbell, 2015) and bipolar
-disorder (Palmius et al., 2017).
+mobility. Such data can now be obtained using the Global Positioning
+System (GPS). GPS uses the distance between a device and several
+satellites to determine location. GPS measurements can be used to infer
+a vast range of socioeconomic and behavioural measures, including where
+the individual lives, how much time he or she spends at home and where
+(and how) they travel.
 
-In most of these studies participants are given a specialised device,
-resulting in accurate mobility GPS data (\emph{specialised logs}).
-However, Barnett and Onnela (2016) point out that these studies are not
-scalable due to cost and burden to participants. Moreover, they may be
-biased because of the introduction of a new device to the participant's
-life. Because of these drawbacks, specialised logs usually span a short
-amount of time. Barnett and Onnela (2016) advocate installing a
+Within behavioural sciences, researchers have used GPS data to
+investigate wideranging topics. For example, Zenk, Schulz, and
+Odoms-Young (2009) investigate the effects of the food environment on
+eating patterns. G. M. Harari et al. (2016) look at the movement
+correlates of personality, finding that extrovers individuals spend less
+time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
+(2015) looks at how academic performance is affected by movement
+patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
+depression.
+
+In most studies participants receive a specialised GPS devices to track
+their movement. We call resulting logs \emph{specialised logs}. Barnett
+and Onnela (2016) point out several methodological issues with these
+studies. Like studies using pen-and-paper travel diaries, collecting
+specialised logs is costly and places a high burden on participants.
+Besides, introducing a new device to the participant's life may bias
+their behaviour. Due to these drawbacks specialised logs usually span a
+short amount of time. Barnett and Onnela (2016) advocate installing a
 custom-made tracking app on user's phones (\emph{custom logs}). Another
-solution is to take advantage of existing smartphone location logs, such
-as Google Location History, which store location information of millions
-of users spanning several years (Location History, 2017)
-(\emph{secondary logs}). By law, logs can be accessed and shared by
-users for free (Commission, 2017). Yet, because they were created for
-non-academic purposes under engineering constraints, the sensors do not
-monitor continuously and the resulting logs can be sparse and
-inaccurate. Hence, two important challenges are dealing with measurement
-noise and missing data.
+solution is to take advantage of existing smartphone location logs
+(\emph{secondary logs}) . For instance, Google Location History contains
+information on millions of users (Location History, 2017). Often,
+secondary logs span several years. By law, secondary logs are accessible
+to users for free (Commission, 2017). Yet, secondary logs also present
+methodological challenges. They were created for non-academic purposes
+under engineering constraints (detailed subsequently). These constraints
+mean that sensors do not track users continuously, meaning that the
+resulting logs can be sparse and inaccurate. Hence, two important
+challenges are dealing with measurement noise and missing data.
 
-Missing data is a pervasive issue as it can arise due to multiple
+Missing data is a pervasive issue as it can arise due to several
 reasons. Technical reasons include signal loss, battery failure and
-device failure. Behavioural reasons include leaving the phone at home,
-switching the phone off, switching location measurements off, and so on.
-As a result, applied researchers are often left with wide temporal gaps
-with no measurements. For instance, different groups studying the effect
-of bipolar disorder on human movement have reported missing data rates
-between 30\% to 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb
-et al., 2015). Similar trends are consistently reported in other fields
-(e.g. G. M. Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).
+device failure. Behavioural reasons include leaving the phone at home or
+switching the device off. As a result, secondary logs often contain wide
+temporal gaps with no measurements. For instance, several research
+groups studying mental health report missing data rates between 30\% to
+50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
+Other researchers report similar trends in different fields (e.g. G. M.
+Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).
 
-There is currently no golden standard in how to deal with missing data
-in custom or secondary logs (Barnett \& Onnela, 2016). Traditional
-missing data methods, such as mean imputation, cannot be used easily in
-spatiotemporal data because the measurements are correlated in time and
-space. For example, assume the simplistic case that an individual spends
-almost half her time at work, half her time at home and a small amount
-of time commuting between the two along a angled path. Using mean
-imputation would result in imputed values of her being at the geometric
-midpoint between the home and work for all missing values, even though
-she has never been there and never will be. Worringly, Jankowska et al.
-(2015) have pointed out that there is often little transparency
-regarding decisions of how to deal with missing data.
+There is no golden standard for dealing with missing data in GPS logs
+(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
+are often correlated in time and space. This means that common methods,
+such as mean imputation, are unsuitable. For example, imagine an
+individual who splits almost all her time between work and home. Suppose
+she spends a small amount of time commuting between the two along a
+circular path. Using mean imputation to estimate her missing
+coordinates, we impute her to be at the midpoint between home and work.
+She has never and will never be there! Worryingly, there is little
+transparency on how researchers deal with missing data (Jankowska et
+al., 2015).
 
-The accuracy of GPS measurements in smartphones is substantially lower
-than in professional grade GPS trackers. For example, Android phones
-collect location information through a variety of methods, such as from
-WiFi access points, cellphone triangulation, and GPS measurements. They
-use different methods due to computational and battery constraints (M.
-Y. Chen et al., 2006; LaMarca et al., 2005).In professional grade GPS
-trackers less than 80\% of measurements fall within 10 meters of the
-true location. GPS measures are reported to be most inaccurate in high
-density urban locations and indoors (S. Duncan et al., 2013; Schipperijn
-et al., 2014). Unfortunately for social scientists, this happens to be
-where most people in the developed world tend to spend most of their
-time.
+The accuracy of smartphone location measurements is substantially lower
+than that of professional GPS trackers. Android phones collect location
+information through a variety of methods. Other than GPS measurements,
+Androids use less-accurate heuristics such as WiFi access points and
+cellphone triangulation. Different methods are used because of
+computational and battery constraints. GPS is the most energy consuming
+sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
+2005). In professional GPS trackers less than 80\% of measurements fall
+within 10 meters of the true location. GPS measures are most inaccurate
+in dense urban locations and indoors (S. Duncan et al., 2013;
+Schipperijn et al., 2014). Unfortunately for researchers, this is where
+people in the developed world spend most of their time.
 
-On the other hand, noisy data can lead to inaccurate conclusions if it
-is not accounted for, such as overestimating the movement of
-individuals. For instance, suppose that a naive researcher calculates
-the distance travelled by an individual by drawing a line between each
-measured point and calculating the sum of the length of all of these
-lines. If there is noise, the measurements will vary even though the
-individual is not moving. If the measurements are frequent, then the
-researcher will end up with a lot of movement, even though the
-individual did not move at all. The problem is further complicated by
-the fact that missing data and noisy measurements are related to each
-other. For instance, methods used by researchers to reduce noise, such
-as throwing out inaccurate measurements (e.g. Palmius et al., 2017) can
-exacerbate the severity of the missing data problem.
+Noisy data can lead to inaccurate conclusions if it is not accounted
+for. Suppose we wish to calculate an individual's movement in a day. A
+simple approach would be to calculate the sum of the distance between
+each measurement. But if there is noise, the coordinates will vary even
+though the individual is not moving. If the measurements are frequent
+and noisy, we will calculate a lot of movement, even if the individual
+did not move at all! This issue is also visualised in Figure
+\ref{fig:accuracyPlot}. The problem is further complicated because
+missing data and noisy measurements are related. Methods used by
+researchers to reduce noise, such as throwing out inaccurate
+measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
+the missing data problem.
 
-In this paper we will compare methods used to deal with measurement
-error and missing data in mobility patterns from secondary GPS logs.
+In this paper we will explore in detail the problem of missing data and
+measurement error in secondary location logs. Moreover, we will compare
+methods used to deal with these problems.
 
-\section{A concrete example}\label{a-concrete-example}
+\subsection{A concrete example}\label{a-concrete-example}
 
-Given that there is little literature on dealing with missing data in
-custom or secondary logs it is worth illustrating the typical
-characteristics of this data using an example data set. The example
-dataset comes from the Google Location History of a single individual
-and spans from January 2013 to January 2017. It was recorded with
-multiple different Android devices and contains 814 941 measurements,
-with approximately 742 measurements per day
-(\(\widehat{\sigma}\)=868.15). The dataset contains a wide range of
-variables including inferred activity and velocity. For the purposes of
-this paper we will focus only on latitude, longitude, accuracy (defined
-below) and a timestamp.
+There is little literature on dealing with missing data in custom or
+secondary logs. Thus it is worth illustrating the typical
+characteristics using an example data set. The example data set comes
+from the Google Location History of a single individual. It spans from
+January 2013 to January 2017 and contains 814 941 measurements. The data
+set contains a multitude of variables, including inferred activity and
+velocity. We will focus on measurements of latitude, longitude, accuracy
+(defined below). All measurements are paired with a timestamp.
 
-\subsection{Location logs and
-notation}\label{location-logs-and-notation}
+\subsubsection{Aggregate measures}\label{aggregate-measures}
 
-GPS measurements report our location on a three dimensional planet, yet
-we are interested in placing these measurements on a two dimensional
-map. Projecting three dimensional measurements onto a two dimensional
-plane results in errors, in order to minimise these errors we borrow an
-error minimising projection method from Barnett and Onnela (2016).
+Social scientists are most interested in aggregating spatiotemporal data
+to more socially relevant information, such as distance travelled. As we
+discussed earlier, aggregations without data processing can be highly
+biased. However, as an example we calculate time spent at home of the
+user for each day of the week in the month of February (Figure
+\ref{fig:aggrePlot}). Between individuals, time spent at home has been
+found to be a reliable predictor of extraversion (G. M. Harari et al.,
+2016).
 
-Let a persons' true location on this two-dimensional plane be
-\(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and \(G_y(t)\) denote the
-location of the individual at time \(t\) on the x-axis and y-axis
-respectively. Moreover, let \(D \in \mathbb{R}^2\) be the recorded data
-containing lattitude and longitude. In addition, let \(a\) denote the
-estimated accuracy of the recorded data. accuracy. \(G(t)\), \(D\) and
-\(a\) are indexed by time labled by the set \(T = t_1 < ... < t_{n+1}\).
-For simplicity, let each entry in the discrete index set \(T\) represent
-a 5 minute window. The measure of accuracy \(a_t\) is given in meters
-such that it represents the radius of a 67\% confidence circle. If
-\(D_t = \emptyset\) it is considered \emph{missing} and it is not
-missing otherwise.
+\begin{figure}
+\includegraphics[width=1\linewidth]{img/timeUse} \caption{Proportion of time spent at home in February 2017.
+We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+Missing values are filled in by the previous observed value.
+Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
+Moreover, the variance is greater on weekends, due to travel.}\label{fig:aggrePlot}
+\end{figure}
 
@@
 
-In the example data set the distribution of \(a\) is highly right
+Google location history provides a measure of accuracy that is given in
+meters such that it represents the radius of a 67\% confidence circle.
+In the example data set the distribution of accuracy is highly right
 skewed, with a median of 28, \(\mu = 127\) and the maximum value at 26
@@
 location values at higher sample rates per hour. We observe similar
-patterns in secondary logs. Figure 1 shows how accuracy tends to vary as
-a function of user behaviour, time and location. There are several
-recurring low-accuracy points, most likely cell-phone triangulation
-towers.
+patterns in secondary logs. Figure \ref{fig:accuracyPlot} shows how
+accuracy can vary as a function of user behaviour, time and location.
+Inaccurate measures are often followed by more accurate measures. Most
+notably, low accuracy often (but not always) is associated with movement
+(Figure \ref{fig:accuracyPlot2}). Stationary accuracy varies depending
+on phone battery level, wifi connection and user phone use. There are
+several recurring low-accuracy points, possibly the result of cell-phone
+tower triangulation.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement in a morning journey.
+\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement of a morning journey on February 15th 2017.
+This includes all measurements from midnight to midday.
 The red circles denote the accuracy of all logged measurement points (the raw data).
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{The accuracy of raw measures with time on the x-axis and accuracy on the y-axis.
-The colour scale shows the distance between the measurement and the previous point.}\label{fig:accuracyPlot2}
+\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
+All journeys are marked with a red line.
+The first peak corresponds to the first journey from the user's home to a gym around 8am.
+The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
+Both journeys can be seen in Figure 1.
+All other journeys are not shown in Figure 1.
+The large jump between journey 5 and 6 is measurement error.
+The lower chart shows the accuracy over the course of the day.
+The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.}\label{fig:accuracyPlot2}
 \end{figure}
@@
 Over 54\% of the data is missing for the entire duration of the log.
-However,this is misleading as there are several long periods with no
-measurements whatsoever. The structure of missingness of a day with
-measurements is shown below.
+This may be misleading as there are several long periods with no
+measurements whatsoever (see Figure \ref{fig:longMeasurementsPerDay}).
+For days which were not entirely missing, approximately 22\% of all five
+minute segments were missing. The structure of missingness of a day with
+measurements is shown in Figure \ref{fig:measurementsPerDay}. As you can
+see, there are several long periods over the course of the log for which
+there are no measurements. Moreover, even during a single day there are
+continuous periods where there is missing data, mostly during the late
+hours of the night in this case.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/missingBoaz5minExample} \caption{Example of missing data over a day.
+\includegraphics[width=1\linewidth]{img/missingdayBoaz5min} \caption{Example of missing data over the entire duration of the log.
+The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
+For this day there were several periods with no information.
+These points are filled with red and lie on the x-axis.}\label{fig:longMeasurementsPerDay}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{img/missingBoaz5minExample} \caption{Example of missing data on February 15th 2017.
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
@@
 
-\subsection{Current methods for imputing missing spatiotemporal
-data}\label{current-methods-for-imputing-missing-spatiotemporal-data}
+\section{Methods}\label{methods}
 
-As we have mentioned before, Jankowska et al. (2015) point out that
-there is little transparency regarding decisions of how to deal with
-missing data in GPS mobility logs. We suspect this is because the highly
-interdisciplinary nature of the problem means researchers are unaware of
-potential solutions. For this reason, we consider it important to
-briefly discuss a few methods one could consider in order to deal with
-measurement inaccuracy and missing data problems in spatiotemporal data.
-In doing so, we will argue that state space models (SSMs) and
-spatiotemporal imputation methods, which are extensively used to model
-missing and noisy spatiotemporal data, are not well suited to deal with
-human mobility patterns. Moreover, we discuss in detail two approaches
-by Palmius et al. (2017) and Barnett and Onnela (2016) which deal
-explicitly with missing data in mobility patterns from smartphone GPS
-logs.
+\subsection{Two-dimensional projections and
+notation}\label{two-dimensional-projections-and-notation}
 
-There is a vast literature of using SSMs to improve measurements
-accuracy and deal with missing data. Behavioural ecologists for
-instance, have used SSMs extensively to explain how animals interact
+GPS measurements provide us with coordinates on the surface of the
+earth. Because most mobility metrics are computed for data in
+\(\mathbb{R}^2\), we are interested in mapping these \(\mathbb{R}^3\)
+measurements on a 2D Euclidean plane. Projecting three dimensional
+measurements onto a two dimensional plane results in distortion. To
+minimise errors we borrow an error minimising projection method from
+Barnett and Onnela (2016).
+
+Having thus converted lattitude and longitude onto coordinates unique to
+each individual, let a person's true location on this two-dimensional
+plane be \(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and \(G_y(t)\)
+denote the location of the individual at time \(t\) on the x-axis and
+y-axis respectively. Moreover, let \(D \in \mathbb{R}^2\) be the
+recorded data containing lattitude and longitude. In addition, let \(a\)
+denote the estimated accuracy of the recorded data. accuracy. \(G(t)\),
+\(D\) and \(a\) are indexed by time labled by the countable set
+\(t = t_1 < ... < t_{n+1}\). For simplicity, let each entry in the
+discrete index set \(t\) represent a 5 minute window. The measure of
+accuracy \(a_t\) is given in meters such that it represents the radius
+of a 67\% confidence circle. If \(D_t = \emptyset\) it is considered
+\emph{missing} and it is not missing otherwise.
+
+When several data sets are available from individuals living in
+overlapping areas we can construct a \(t \times i\) matrix \(M\) where
+the entry \(M(t,i)\) contains \(G(t)\) for the individual \(i\).
+
+\subsection{Selecting candidate
+models}\label{selecting-candidate-models}
+
+There is no golden standard or established practice in how to deal with
+missing data in GPS logs. Researchers are generally vague about what
+practices they follow (Jankowska et al., 2015). Ostensibly this is
+because they are unaware of possible solutions. In an attempt to
+elucidate the topic, we explore potential solutions. We will argue that
+extensively used spatiotemporal methods, such as state space models
+(SSMs), are not well suited to deal with human mobility patterns. We
+also discuss in detail two approaches which deal explicitly with
+mobility patterns from custom or secondary logs (Barnett \& Onnela,
+2016; Palmius et al., 2017 ).
+
+There is a vast literature on using SSMs in spatiotemporal statistics.
+For example, ecologists have used SSMs to explain how animals interact
 with their environment (Patterson, Thomas, Wilcox, Ovaskainen, \&
-Matthiopoulos, 2008). These models can be quite complex, for example
-Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
-processes to characterise the effect of roads, food patches and streams
-on cyclical elk movements. The most well studied SSM is the Kalman
-filter, which is the optimal algorithm for inferring linear Gaussian
-systems. In fact, the extended Kalman filter is the de facto standard
-for GPS navigation (Z. Chen \& Brown, 2013). The advantage of state
-space models is that they are flexible, deal with measurement
-inaccuracy, include information from different sources and can be used
-in real time.
+Matthiopoulos, 2008). These models can be quite complex. Preisler, Ager,
+Johnson, and Kie (2004) uses Markovian movement processes to
+characterise the effect of roads, food patches and streams on cyclical
+elk movements. The most well studied SSM is the Kalman filter, which is
+the optimal algorithm for inferring linear Gaussian systems. The
+extended Kalman filter is the de facto standard for GPS navigation (Z.
+Chen \& Brown, 2013). The advantage of state space models is that they
+are flexible, deal with measurement inaccuracy, include information from
+different sources and can be used in real time.
 
-However, the main limitation of SSMs is that they ignore the fact that
-humans have regular movement routines, such as going to work or shopping
-for groceries on weekends. This limitation is due to the fact that SSMs
-are based on the Markov property. Thus, the estimated location \(G(t)\)
-at timepoint \(t\) is often based only upon measurements \(D_t\),
+For us, the main limitation of SSMs is that they ignore regular movement
+routines. For instance, humans tend to go to work on weekdays and sleep
+at night. Because SSMs are based on the Markov property, they cannot
+incorporate this information. The estimated location \(G(t)\) at
+timepoint \(t\) is often based only upon measurements \(D_t\),
 \(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\). Hierarchical structuring
-and conditioning on a larger context have been suggested as ways to
-improve the performance of Markovian models, but these solutions are
-often computationally intractable or unfeasible (Sadilek \& Krumm,
-2016). For this reason we do not consider SSMs to be useful for imputing
-missing data, although they could be of use in filtering noise.
+and conditioning on a larger context have been suggested as ways to add
+periodicity to Markovian models. These solutions are often
+computationally intractable or unfeasible (Sadilek \& Krumm, 2016). For
+this reason we do not consider SSMs to be useful for imputing missing
+data. Nonetheless, they could be of use in filtering noise.
 
-In addition to SSMs there are spatiotemporal imputation methods often
-used in climate or geological research for estimating missing data. For
-instance, Feng, Nowak, O'Neill, and Welsh (2014) illustrate their CUTOFF
-method, which relies on estimating missing values using the nearest
-observed neighbours in time, using rainfall data from dozens of gauging
-stations across Australia. Similarly, Z. Zhang et al. (2017) use a
-variety of machine learning methods to present their model based on
-underground water data in China.
+In climate or geological research spatiotemporal imputation methods are
+often used. For instance, the CUTOFF method estimates missing values
+using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
+2014 ). The authors illustrate their example using rainfall data from
+gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
+a variety of machine learning methods to impute missing values. The
+example provided relates to underground water data. Generally these
+models assume fixed measurement stations (such as rainfall gauging
+stations).
 
-The difficulty of applying this class of spatiotemporal imputation
-models to human mobility tracks lies in the fact that these models
-generally assume fixed measurement stations (such as rainfall gauging
-stations). While Feng et al. (2014) claim their model could be used to
-establish mobility patterns, ostensibly by dividing the sample space
-into rasters analogous to measurement stations indicating a probability
-of the individual being there, this seems to be computationally
-unfeasible. To our knowledge such models have not been implemented for
-mobility traces.
+For this reason they cannot be easily applied to missing mobility
+tracks. Feng et al. (2014) claim their model could be used to establish
+mobility patterns. This may be possible by dividing the sample space
+into rasters. Each raster would be analogous to a measurement station.
+These artificial stations could \enquote{measure} the probability of the
+individual being there. To our knowledge such models have not been
+implemented for mobility traces and seems computationally inefficient.
 
@@
 
-\subsection{Methods}\label{methods}
-
 \subsubsection{Datasets \& Analyses}\label{datasets-analyses}
@@
 The data used to train the imputation methods was collected between 2013
-and 2017 on different Android devices from several individuals (table
-1).
+and 2017 on different Android devices from several individuals (Table
+\ref{tab:datadetailsTable}).
 
-\begin{table}
-
-\caption{\label{tab:datadetailsTable}Table with descriptives about the data sets used to build the imputation methods. }
-\centering
-\begin{tabular}[t]{l|r|r|r|r|r}
-\hline
-Log duration & Logged days & Observations & Missing days & Missing data & Mean Accuracy\\
-\hline
-From 2013-02-06 to 2017-03-29 & 1512 & 646376 & 635 & 0.22 & 127.78\\
-\hline
-From 2016-07-14 to 2017-05-10 & 300 & 158382 & 3 & 0.41 & 1394.60\\
-\hline
-From 2014-01-22 to 2017-01-23 & 1097 & 814941 & 80 & 0.25 & 121.83\\
-\hline
-\end{tabular}
-\end{table}
+\begin{lltable}
+\begin{longtable}{llllll}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:datadetailsTable}Table with descriptives about the data sets used to build the imputation methods.
+Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.}\\
+\toprule
+Log duration & Logged days & Observations & Missing
+days & Missing
+data & Mean Accuracy\\
+\midrule
+2013-02-06 to 2017-03-29 & 1,512.00 & 646,376.00 & 635.00 & 0.22 & 127.78\\
+2016-07-14 to 2017-05-10 & 300.00 & 158,382.00 & 3.00 & 0.41 & 1,394.60\\
+2014-01-22 to 2017-01-23 & 1,097.00 & 814,941.00 & 80.00 & 0.25 & 121.83\\
+\bottomrule
+\end{longtable}
+\end{lltable}
 
@@
 Analyses were performed using R and a multitude of other statistical
-packages ({\textbf{???}}; Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J.
-Pebesma \& Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \&
-Francois, 2016).
+packages (Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J. Pebesma \&
+Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \& Francois,
+2016).
 
@@
 \item
-  The model developed by Barnett and Onnela (2016) using spatial,
-  temporal and circadian proximity.
+  The model developed by Barnett and Onnela (2016) using both spatial
+  and temporal proximity.
 \item
@@
 imputation approaches a mean value over 100 samples was taken.
+
+Each imputation method used each of the three filtering methods as an
+input. Thus in the end we end up with eight methods to evaluate, three
+for each filtering method as well as four for each imputation method.
 

### 2017-12-02 thesisReportApa.Rmd
changing a few plots and captions

@@
 
-  Submitted as a research report.
+  Submitted as a research report conforming to APA manuscript guidelines (6th edition).
 
@@
 
-keywords          : "Missing data, Measurement Bias,GPS, Human Mobility"
+keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
 wordcount         : "2205"
@@
 We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+The user is coded as at home if within 250 meters of the home coordiantes.
 Missing values are filled in by the previous observed value.
 Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
+Moreover, the variance is greater on Saturday, due to travel.",out.width = '100%'}
 include_graphics("img/timeUse.png")
@@
 ```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
-All journeys are marked with a red line.
+All journeys are labled at the top of the chart.
 The first peak corresponds to the first journey from the user's home to a gym around 8am.
 The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
-Both journeys can be seen in Figure 1.
-All other journeys are not shown in Figure 1.
+Both journeys can be seen in Figure 2.
+All other journeys are not shown in Figure 2.
 The large jump between journey 5 and 6 is measurement error.
@@
 
+#WARNING: FIGURE REFERENCE NOT AUTOGENERATED. DOUBLE CHECK. 
 ```
@@
 For this day there were several periods with no information.
-These points are filled with red and lie on the x-axis.",out.width = '100%'}
+These points are filled with red and lie on the x-axis.
+Missingness over long time periods is related to the type of device the user has.",out.width = '100%'}
 #fix width

### 2017-12-02 thesisReportApa.tex
changing a few plots and captions

@@
 
-    Submitted as a research report.
+    Submitted as a research report conforming to APA manuscript guidelines
+    (6th edition).
 
@@
 currently used methods, such as linear interpolation.}
-  \keywords{Missing data, Measurement Bias,GPS, Human Mobility \\
+  \keywords{Missing Data, Measurement Bias, GPS, Human Mobility \\
 
@@
 We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+The user is coded as at home if within 250 meters of the home coordiantes.
 Missing values are filled in by the previous observed value.
 Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on weekends, due to travel.}\label{fig:aggrePlot}
+Moreover, the variance is greater on Saturday, due to travel.}\label{fig:aggrePlot}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
-All journeys are marked with a red line.
+All journeys are labled at the top of the chart.
 The first peak corresponds to the first journey from the user's home to a gym around 8am.
 The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
-Both journeys can be seen in Figure 1.
-All other journeys are not shown in Figure 1.
+Both journeys can be seen in Figure 2.
+All other journeys are not shown in Figure 2.
 The large jump between journey 5 and 6 is measurement error.
@@
 For this day there were several periods with no information.
-These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
+These points are filled with red and lie on the x-axis.
+Missingness over long time periods is related to the type of device the user has.}\label{fig:measurementsPerDay}
 \end{figure}

### 2017-12-02 scripts/auxDataDetails.R

@@ -95,4 +95,4 @@ saveRDS(allData, "datadescriptives.rds")
 
 d<- readRDS("../tempdata/datadescriptives.rds")
 d$span <- gsub(pattern = "From ",replacement = "",d$span)
-saveRDS(allData, "datadescriptives.rds")
+saveRDS(d, "../tempdata/datadescriptives.rds")

### 2017-12-02 scripts/dailyMissPlot.R

@@ -28,14 +28,14 @@ exampleMiss<- m5minv2 %>%
   geom_point()+theme_tufte()+
   xlab("")+
   ylab("Measurements")+
-  ggtitle("Measurements per 5 minute window")+
+  ggtitle("Measurements per 5 minute window",subtitle = "February 15 2017")+
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
                    labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
   scale_colour_manual(values=c("black",muted("red")),labels=c("Not Missing", "Missing"))+
-  labs(x = NULL, colour = "")
+  labs(x = NULL, colour = "")+xlab("Time")
 
-ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 6.5, width = 18, units = "cm")
+ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 10, width = 20, units = "cm")
 
 # randomly select two days
 sample(x = c(13,14,15,16,17),1)#YEAR

### 2017-12-15 thesisReportApa.Rmd
forgotten commit

@@
 keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
-wordcount         : "2205"
+wordcount         : "2297"
 
@@
 There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016].
-Importantly, spatiotemporal data measurements are often correlated in time and space.
+Importantly, spatiotemporal measurements are often correlated in time and space.
 This means that common methods, such as mean imputation, are unsuitable.
@@
 
-Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled.
-As we discussed earlier, aggregations without data processing can be highly biased.
-However, as an example we calculate time spent at home of the user for each day of the week in the month of February (Figure \@ref(fig:aggrePlot)).
-Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information.
+For instance, @palmius_detecting_2017 calculate several measures to predict bipolar depression, such as the  total distance travelled, the amount of time spent at home, travelling and the regularity of diurnal movement.
+As an example we calculate two of these measures: time spent at home and distance travelled for the month of February.
+Figure \@ref(fig:aggrePlot) shows how the aggregate measures are influenced by the noise in the measurements and the missingness.
+In the case of distance travelled, simply filtering out the most inaccurate measures leads to significantly lower estimates.
+This phenomenon is mapped out on Figure \@ref(fig:accuracyPlot) .
+On the other hand, the time spent at home would be greatly influenced at times by the imputation method, as several days with long periods of missing data.
 
-```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017.
-We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home  and distance travelled in February 2017.
+We estimate time spent at home by calculating the mean lattitude and longitude for every 5 minute time period in the month.
 The user is coded as at home if within 250 meters of the home coordiantes.
-Missing values are filled in by the previous observed value.
-Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on Saturday, due to travel.",out.width = '100%'}
-include_graphics("img/timeUse.png")
+One can see that several days have long periods with missing data.
+More accurate estimates can be reached by imputing the missing measurements.
+As for distance travelled, it is evident that although the behavioural trends remain the same, the absolute estimate of distance travelled varies depending on the filter.",out.width = '100%'}
+include_graphics("img/aggPlot.png")
 ```
@@
 
-Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].
+Analyses were performed using R and a multitude of other statistical packages [@dplyr;@ggplot2;@base;@ggthemes;@sp1;@sp2;@citr;@tidyr;@leaflet;@tibbletime;@padr;@bibtex;@gridExtra;@scales;@rstudio].
 
@@
 
-Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:
+Four imputation methods were selected in order to cover a wide range of techniques applied in the literature:
 

### 2017-12-15 thesisReportApa2.tex
forgotten commit

--- 2017-12-02 thesisReportApa.tex
@@
 
-    \indent Word count: 2205
+    \indent Word count: 2297
   }
@@
 Odoms-Young (2009) investigate the effects of the food environment on
-eating patterns. G. M. Harari et al. (2016) look at the movement
-correlates of personality, finding that extrovers individuals spend less
+eating patterns. Harari et al. (2016) look at the movement
+correlates of personality, finding that extroverted individuals spend less
 time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
@@
 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
-Other researchers report similar trends in different fields (e.g. G. M.
+Other researchers report similar trends in different fields (e.g.
 Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).
@@
 There is no golden standard for dealing with missing data in GPS logs
-(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
-are often correlated in time and space. This means that common methods,
-such as mean imputation, are unsuitable. For example, imagine an
-individual who splits almost all her time between work and home. Suppose
-she spends a small amount of time commuting between the two along a
-circular path. Using mean imputation to estimate her missing
-coordinates, we impute her to be at the midpoint between home and work.
-She has never and will never be there! Worryingly, there is little
-transparency on how researchers deal with missing data (Jankowska et
-al., 2015).
+(Barnett \& Onnela, 2016). Importantly, spatiotemporal measurements are
+often correlated in time and space. This means that common methods, such
+as mean imputation, are unsuitable. For example, imagine an individual
+who splits almost all her time between work and home. Suppose she spends
+a small amount of time commuting between the two along a circular path.
+Using mean imputation to estimate her missing coordinates, we impute her
+to be at the midpoint between home and work. She has never and will
+never be there! Worryingly, there is little transparency on how
+researchers deal with missing data (Jankowska et al., 2015).
 
@@
 computational and battery constraints. GPS is the most energy consuming
-sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
+sensor on most smartphones (Chen et al., 2006; LaMarca et al.,
 2005). In professional GPS trackers less than 80\% of measurements fall
 within 10 meters of the true location. GPS measures are most inaccurate
-in dense urban locations and indoors (S. Duncan et al., 2013;
+in dense urban locations and indoors (Duncan et al., 2013;
 Schipperijn et al., 2014). Unfortunately for researchers, this is where
@@
 Social scientists are most interested in aggregating spatiotemporal data
-to more socially relevant information, such as distance travelled. As we
-discussed earlier, aggregations without data processing can be highly
-biased. However, as an example we calculate time spent at home of the
-user for each day of the week in the month of February (Figure
-\ref{fig:aggrePlot}). Between individuals, time spent at home has been
-found to be a reliable predictor of extraversion (G. M. Harari et al.,
-2016).
+to more socially relevant information. For instance, Palmius et al.
+(2017) calculate several measures to predict bipolar depression, such as
+the total distance travelled, the amount of time spent at home,
+travelling and the regularity of diurnal movement. As an example we
+calculate two of these measures: time spent at home and distance
+travelled for the month of February. Figure \ref{fig:aggrePlot} shows
+how the aggregate measures are influenced by the noise in the
+measurements and the missingness. In the case of distance travelled,
+simply filtering out the most inaccurate measures leads to significantly
+lower estimates. This phenomenon is mapped out on Figure
+\ref{fig:accuracyPlot} . On the other hand, the estimated time spent at
+home on any given day can depend greatly on the imputation method due to
+the extent of missing data.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/timeUse} \caption{Proportion of time spent at home in February 2017.
-We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
+\includegraphics[width=1\linewidth]{img/aggPlot} \caption{Proportion of time spent at home  and distance travelled in February 2017.
+We estimate time spent at home by calculating the mean lattitude and longitude for every 5 minute time period in the month.
 The user is coded as at home if within 250 meters of the home coordiantes.
-Missing values are filled in by the previous observed value.
-Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on Saturday, due to travel.}\label{fig:aggrePlot}
+One can see that several days have long periods with missing data.
+More accurate estimates can be reached by imputing the missing measurements.
+As for distance travelled, it is evident that although the behavioural trends remain the same, the absolute estimate of distance travelled varies depending on the filter.}\label{fig:aggrePlot}
 \end{figure}
@@
 2014 ). The authors illustrate their example using rainfall data from
-gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
+gauging stations across Australia. Similarly, Zhang et al. (2017) use
 a variety of machine learning methods to impute missing values. The
@@
 Analyses were performed using R and a multitude of other statistical
-packages (Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J. Pebesma \&
-Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \& Francois,
-2016).
+packages (Arnold, 2017; Auguie, 2017; Aust, 2016; Bivand, Pebesma, \&
+Gomez-Rubio, 2013; Cheng, Karambelkar, \& Xie, 2017; Francois, 2017; E.
+J. Pebesma \& Bivand, 2005; R Core Team, 2017; RStudio Team, 2015;
+Thoen, 2017; Vaughan \& Dancho, 2017; Wickham, 2009, 2017; Wickham \&
+Henry, 2017; Wickham, Francois, Henry, \& Muller, 2017).
 
@@
 
-Three imputation methods were selected in order to cover a wide range of
+Four imputation methods were selected in order to cover a wide range of
 techniques applied in the literature:
@@
 Each imputation method used each of the three filtering methods as an
-input. Thus in the end we end up with eight methods to evaluate, three
+input. Thus in the end we have seven methods to evaluate: three
 for each filtering method as well as four for each imputation method.
@@
 \hypertarget{refs}{}
+\hypertarget{ref-ggthemes}{}
+Arnold, J. B. (2017). \emph{Ggthemes: Extra themes, scales and geoms for
+'ggplot2'}. Retrieved from
+\url{https://CRAN.R-project.org/package=ggthemes}
+
+\hypertarget{ref-gridExtra}{}
+Auguie, B. (2017). \emph{GridExtra: Miscellaneous functions for ``grid''
+graphics}. Retrieved from
+\url{https://CRAN.R-project.org/package=gridExtra}
+
+\hypertarget{ref-citr}{}
+Aust, F. (2016). \emph{Citr: 'RStudio' add-in to insert markdown
+citations}. Retrieved from \url{https://CRAN.R-project.org/package=citr}
+
 \hypertarget{ref-barnett_inferring_2016}{}
@@
 doi:\href{https://doi.org/10.4249/scholarpedia.30868}{10.4249/scholarpedia.30868}
+
+\hypertarget{ref-leaflet}{}
+Cheng, J., Karambelkar, B., \& Xie, Y. (2017). \emph{Leaflet: Create
+interactive web maps with the javascript 'leaflet' library}. Retrieved
+from \url{https://CRAN.R-project.org/package=leaflet}
 
@@
 doi:\href{https://doi.org/10.1016/j.jhydrol.2014.11.012}{10.1016/j.jhydrol.2014.11.012}
+
+\hypertarget{ref-bibtex}{}
+Francois, R. (2017). \emph{Bibtex: Bibtex parser}. Retrieved from
+\url{https://CRAN.R-project.org/package=bibtex}
 
@@
 
+\hypertarget{ref-rstudio}{}
+RStudio Team. (2015). \emph{RStudio: Integrated development environment
+for r}. Boston, MA: RStudio, Inc. Retrieved from
+\url{http://www.rstudio.com/}
+
 \hypertarget{ref-sadilek_far_2016}{}
@@
 
+\hypertarget{ref-padr}{}
+Thoen, E. (2017). \emph{Padr: Quickly get datetime data ready for
+analysis}. Retrieved from \url{https://CRAN.R-project.org/package=padr}
+
+\hypertarget{ref-tibbletime}{}
+Vaughan, D., \& Dancho, M. (2017). \emph{Tibbletime: Time aware
+tibbles}. Retrieved from
+\url{https://CRAN.R-project.org/package=tibbletime}
+
 \hypertarget{ref-wang_smartgpa:_2015}{}
@@
 
+\hypertarget{ref-scales}{}
+Wickham, H. (2017). \emph{Scales: Scale functions for visualization}.
+Retrieved from \url{https://CRAN.R-project.org/package=scales}
+
+\hypertarget{ref-tidyr}{}
+Wickham, H., \& Henry, L. (2017). \emph{Tidyr: Easily tidy data with
+'spread()' and 'gather()' functions}. Retrieved from
+\url{https://CRAN.R-project.org/package=tidyr}
+
 \hypertarget{ref-dplyr}{}
-Wickham, H., \& Francois, R. (2016). \emph{Dplyr: A grammar of data
-manipulation}. Retrieved from
+Wickham, H., Francois, R., Henry, L., \& Muller, K. (2017). \emph{Dplyr:
+A grammar of data manipulation}. Retrieved from
 \url{https://CRAN.R-project.org/package=dplyr}

### 2017-12-15 scripts/aggregation.R

@@ -46,6 +46,7 @@ x<- left_join(x, xDisthome,"time2")
 
 x <- left_join(x, xFiltered)
 
+
 exploreDates(x,"2017-02-15","2017-02-15")
 # time spent at home
 
@@ -53,44 +54,82 @@ exploreDates(x,"2017-02-15","2017-02-15")
 ggplot(x,aes(x = distanceHome, y = time2))+
   geom_point()
 
-athome<- x %>% mutate(atHome = case_when(distanceHome <0.25 ~ 1,
-                                distanceHome >0.25 ~ 0)) %>%
+athome<- x %>% mutate(atHome = case_when(distHome <0.25 ~ 1,
+                                distHome >0.25 ~ 0)) %>%
   thicken('day', colname = 'time') %>%
+  select(time,atHome) %>%
   group_by(time) %>%
-  summarise(atHome2 = sum(atHome)/288) %>%
-  mutate(day = strftime(time,"%A"))
+  summarise(atHomeProp = sum(atHome, na.rm = T)/288,
+            missingProp = sum(is.na(atHome))/288) %>%
+  mutate(day = strftime(time,"%a"),
+         propNotHome = 1-atHomeProp-missingProp)
 
-athome$day <- factor(athome$day, levels= c("Monday", 
-                                         "Tuesday", "Wednesday",
-                                         "Thursday", "Friday", "Saturday","Sunday"))
+athome$day <- factor(athome$day, levels= c("Mon", 
+                                         "Tue", "Wed",
+                                         "Thu", "Fri", "Sat","Sun"))
 
-timeUse<- ggplot(athome,aes(x = day, y = atHome2))+
-  geom_point()+theme_tufte()+xlab("Day of the week")+
-  ylab("Proportion of time\nspent at home")+ggtitle("Time spent at home - February 2017")
-
-ggsave(timeUse,filename = "../img/timeUse.png",device = "png",height = 10,width = 20, units = "cm")
+athm<- athome %>%
+  select(time,day,atHomeProp,propNotHome,missingProp) %>%
+  gather(type,proportion, atHomeProp:missingProp)%>%
+  as_tbl_time(index = time)
   
 
+timeUse<- ggplot(athm,aes(x = time, y = proportion, fill=type))+
+  geom_bar(stat="identity")+theme_tufte()+xlab("Day of the week")+
+  ylab("Proportion of time")+
+  ggtitle("Time spent at home",subtitle = "February 2017")+
[…]

### 2017-12-15 scripts/texhack.R

@@ -1,2 +1,2 @@
 # hacking tex
-tools::texi2pdf("thesisReportApa.tex")
+tools::texi2pdf("../thesisReportApa2.tex")

### 2018-02-09 scripts/auxFuns.R
Implemented Palmius et al filtering in R

@@ -1,11 +1,27 @@
-library(leaflet)
+# Often used auxilliary functions & packages
 
-Data_Start<- as.POSIXct("2015-09-07 01:55:00")
-Data_End  <- as.POSIXct("2015-09-10 01:59:00")
+# packages
 
+library(dplyr)
+library(tibbletime)
+library(sp)
+
+# functions
+
+shift.vec <- function(vec, shift){
+  if (length(vec) <= abs(shift)){
+    rep(NA ,length(vec))
+  } else {
+    if (shift >= 0) {
+      c(rep(NA, shift), vec[1:(length(vec) - shift)]) }
+    else {
+      c(vec[(abs(shift) + 1):length(vec)], rep(NA, abs(shift)))
+    }
+  }
+}
 
 exploreDates <- function(df = all, Data_Start= "2017-02-15", Data_End = "2017-02-17"){
-    
+  
   library(leaflet)
   library(dplyr)
   library(tibbletime)
@@ -40,15 +56,10 @@ exploreDates <- function(df = all, Data_Start= "2017-02-15", Data_End = "2017-02
   df %>%
     select(time,lat,lon,accuracy) %>%
     as_tbl_time(index = time) %>%
-    time_filter(time_formula = time_formula) %>%
+    filter_time(time_formula = time_formula) %>%
     leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
     addTiles() %>%
     addCircles(lng = ~lon, lat = ~lat,
                radius = ~accuracy, fillOpacity = 0.02,color = "#DF2935")%>%
     addProviderTiles(providers$CartoDB.Positron)
 }
-
-exploreDates(all,"2017-02-01","2017-02-01")
-
-+12:00:00
-+18:00:00

### 2018-02-09 implementingBipolar/implementationReport.Rmd
reorganised folders

---
title: "Reproducing Palmius’ filtering & imputation"
author: "Boaz Sobrado"
date: "2/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Implementation of filtering algorithm

I have implemented the filtering algorithm which is descriped in Palmius et al.'s paper. Nick was kind enough to provide pseudocode in addition to his matlab code. His pseudocode is below:
```
data: D(t) recorded data points at tmin ≤ t ≤ tmax. result: D′(t′) filtered data points at t′min ≤ t′ ≤ t′max.
begin
D′ ← D t′←t repeat
t∗ ← t′ where d′ D′ > 100kmh−1 dt
Dsrc ← D′(t∗ − 1)
Ddest ← D′(t∗)
Dˆsrc ← mode(Dsrc)
Dˆdest ← mode(Ddest)
if |Dˆsrc| ≥ |Dˆdest| and |Dˆsrc| > 10 then
Remove all occurrences of Dˆsrc from D′ else if |Dˆdest| ≥ 10 then
Remove all occurrences of Dˆdest from D′ end if
until |Dˆsrc| ≥ 10 and |Dˆdest| < 10 end
◃ Data points with high d′ D. dt
```
Based on that I implemented it and executed it in R. You can see the repository with all my code [here]().

## Preliminary evaluation of filtering algorithm using default parameters

As expected, the high speed between measurement is associated with high accuracy scores (that is, inaccurate measurements). Approximately x percent of measurements in all of 2017 were associated with such jumps.

What Nick does is exclude those points if they are often associated with big jumps (i.e. more than ten times).

For my own data in 2017 there was only one location which was excluded (a total of x measurements), which means it is not a very effective filtering algorithm using the original parameters. Moreover, the points which are filtered depend on the length of the data which is taken into account.

Further playing around with the parameters would be necessary to see what fits this data best.

## Down sampling method

In addition to filtering Nick also downsamples the data to 5 minute windows. The way he does this is that if the standard deviation within a fixed one hour window (e.g. from 2am to 3am) is less than 10 meters he simply takes a mean for the hour. If it is more than that he takes a 5 minute median window and applies it to the epoch in question.

## Results following

The red dots are the raw data, the blue dots are the filtered data for that day of mine on Feb 22.

```{}

```

## Missing data imputation

### 2018-02-09 report/thesisReportApa.tex
reorganised folders

--- 2017-12-15 thesisReportApa2.tex
@@
 Odoms-Young (2009) investigate the effects of the food environment on
-eating patterns. Harari et al. (2016) look at the movement
-correlates of personality, finding that extroverted individuals spend less
+eating patterns. G. M. Harari et al. (2016) look at the movement
+correlates of personality, finding that extrovers individuals spend less
 time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
@@
 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
-Other researchers report similar trends in different fields (e.g.
+Other researchers report similar trends in different fields (e.g. G. M.
 Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).
@@
 computational and battery constraints. GPS is the most energy consuming
-sensor on most smartphones (Chen et al., 2006; LaMarca et al.,
+sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
 2005). In professional GPS trackers less than 80\% of measurements fall
 within 10 meters of the true location. GPS measures are most inaccurate
-in dense urban locations and indoors (Duncan et al., 2013;
+in dense urban locations and indoors (S. Duncan et al., 2013;
 Schipperijn et al., 2014). Unfortunately for researchers, this is where
@@
 lower estimates. This phenomenon is mapped out on Figure
-\ref{fig:accuracyPlot} . On the other hand, the estimated time spent at
-home on any given day can depend greatly on the imputation method due to
-the extent of missing data.
+\ref{fig:accuracyPlot} . On the other hand, the time spent at home would
+be greatly influenced at times by the imputation method, as several days
+with long periods of missing data.
 
@@
 2014 ). The authors illustrate their example using rainfall data from
-gauging stations across Australia. Similarly, Zhang et al. (2017) use
+gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
 a variety of machine learning methods to impute missing values. The
@@
 Each imputation method used each of the three filtering methods as an
-input. Thus in the end we have seven methods to evaluate: three
+input. Thus in the end we end up with eight methods to evaluate, three
 for each filtering method as well as four for each imputation method.

### 2018-02-09 implementingBipolar/downSamplingAlgo.R

@@ -1,0 +1,105 @@
+# down sampling palmius algo
+
+# preceded by the filtering algo
+
+# divide in 1 hour epochs
+library(padr)
+
+# if lattitude and longitude sd both less than 0.01km
+# then replace with mean
+deg2rad <- function(deg) {(deg * pi) / (180)}
+
+#something fishy,figure out what it is
+
+d2<- d %>% thicken('hour') %>% 
+  group_by(time_hour) %>%
+  summarise(sdLat = sd(lat , na.rm = T),
+            sdLon = sd(lon , na.rm = T),
+            meanLat = mean(lat),
+            meanLon = mean(lon)) %>%
+  mutate(sdLatkm = sdLat*110.574,
+         sdLonkm = 111.320*cos(deg2rad(meanLat))
+         ) %>% 
+  pad() %>% 
+  fill_by_value( "NA") %>% 
+  filter(sdLatkm < 0.01 & sdLonkm < 0.01)
+
+summary(d2$sdLatkm)
+
+# Latitude: 1 deg = 110.574 km
+# Longitude: 1 deg = 111.320*cos(latitude) km
+
+
+#alternative version using pythagoras and distance via haversine forumla
+#in meters
+sqrt(10^2+10^2)
+
+d3<- d %>% thicken('hour') %>% 
+  group_by(time_hour) %>%
+  summarise(distancesMeterSd = sd(distancesMeter , na.rm = T),
+            meanLat = mean(lat),
+            meanLon = mean(lon)) %>% 
+  pad() %>% 
+  fill_by_value( "NA") %>%
+  filter(distancesMeterSd < 14.14) %>%
+  select(time_hour,meanLat,meanLon)
+
+dF<- d %>% thicken('hour') %>%
+  left_join(d3) %>%
+  mutate(
+    lonF1 = case_when(
+      is.na(meanLon) == F ~ meanLon
+    ),
+    latF1 = case_when(
+      is.na(meanLat) == F ~ meanLat
+    )
+  )
+
+dF2<- dF %>% thicken(by = "time","5 min",colname= "time_5min") %>% 
+  group_by(time_5min) %>% 
+  summarise(lonF2 = median(lon),
+            latF2 = median(lat)
+            )
+  
+d.filt <- dF %>% thicken(by = "time","5 min",colname= "time_5min") %>% 
+  left_join(dF2, by = "time_5min") %>%
[…]

### 2018-02-09 implementingBipolar/filteringAlgo.R

@@ -1,0 +1,89 @@
+# Palmius filtering algorithm implemented in R.
+
+# set wd
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+#read data
+data<- readRDS("../data/boaz/myLocationHistory09012018.rds")
+
+# load auxiliary functions
+source("../thesis/scripts/auxFuns.R")
+#filter down to a day
+
+dataExample <- select(data,time,lat,lon,accuracy) %>%
+  arrange(time) %>% 
+  as_tbl_time(time) %>% 
+  filter_time( ~ '2017') %>%
+  mutate( nextLon = shift.vec(lon,-1),
+          nextLat = shift.vec(lat,-1),
+          nextTimeSec = shift.vec(time,-1)-time,
+          index = 1:length(lat)
+)
+
+rm(data)
+
+d<- dataExample
+dPrimeSource<- 100
+dPrimeDest <- 1
+removed <- data.frame(lon = 1, lat = 1)
+
+while(dPrimeSource >= 10 & dPrimeDest < 10){
+  
+  #calculate distance between points
+  d$distancesMeter<- apply(d, 1, FUN = function(row) {
+    raster::pointDistance(c(as.numeric(as.character(row["lon"])),
+                            as.numeric(as.character(row["lat"]))),
+                          c(as.numeric(as.character(row["nextLon"])), as.numeric(as.character(row["nextLat"]))),
+                          lonlat = T) # Parameter 'lonlat' has to be TRUE!
+  })
+  
+  #add index and calculate speed between points
+  d <- mutate(d,
+              speedMeterSecond= distancesMeter/as.numeric(nextTimeSec),
+              index = 1:nrow(dataExample)
+  )
+  
+  # set speed parameter
+  maxSpeed <- 27.7 #equivalent to 100 kmh
+  
+  # extract source 
+  filteredSource <- d %>% 
+    filter(speedMeterSecond >= maxSpeed)
+  
+  # extract destination
+  filteredDestination <- d %>%
+    filter(index %in% as.numeric(filteredSource$index+1))
+  
+  # extract modes
+  
+  sourceMode<- filteredSource %>% 
+    count(lon, lat) %>%
+    slice(which.max(n))
+  
+  destMode<- filteredDestination %>% 
+    count(lon, lat) %>%
+    slice(which.max(n))
+  
+  if(sourceMode$n >= destMode$n & sourceMode$n >10){
[…]

### 2018-03-06 implementingBipolar/implementationReport.Rmd
all

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(dplyr)
+library(leaflet)
 ```
@@
 ```
-Based on that I implemented it and executed it in R. You can see the repository with all my code [here]().
+Based on that I implemented it and executed it in R.
+You can see the repository with all my code [here](https://github.com/sobradob/thesis/tree/master/implementingBipolar).
+Check out the filtering algo and the downsampling algo.
 
@@
 As expected, the high speed between measurement is associated with high accuracy scores (that is, inaccurate measurements).
-Approximately x percent of measurements in all of 2017 were associated with such jumps.
+Approximately 1.5% percent of measurements in all of 2017 were associated with such jumps.
 
@@
 
-For my own data in 2017 there was only one location which was excluded (a total of x measurements), which means it is not a very effective filtering algorithm using the original parameters.
+For my own data in 2017 there was only one location which was excluded (a total of 0.02% measurements), which means it is not a very effective filtering algorithm using the original parameters.
 Moreover, the points which are filtered depend on the length of the data which is taken into account.
@@
 Further playing around with the parameters would be necessary to see what fits this data best.
+
+The difference may be in the quality of the devices recording. I'll check to see the 2015 logs, perhaps it would be more efficient then.
 
@@
 
-## Results following
+## Results following filtering and downsampling
 
@@
 
-```{}
+```{r mapped, echo = F}
+
+raw<- readRDS("raw.rds")
+filtered<- readRDS("filtered.rds")
+
+leaflet() %>%
+  addTiles() %>%
+  addCircles(lng = ~lonF, lat = ~latF, data = raw, color = "#18206F", group = "Raw")%>%
+  addCircles(lng = ~lon, lat = ~lat, data=filtered,color = "#DF2935",group = "Filtered")%>%
+  addProviderTiles(providers$CartoDB.Positron) %>% 
+  addLayersControl(
+    overlayGroups = c("Raw", "Filtered"),
+    options = layersControlOptions(collapsed = FALSE)
+  )
 
@@
 ## Missing data imputation
+
+Palmius uses several imputation algorithms to impute missing data.
+
+```
+
+```

### 2018-03-06 implementingBipolar/algoA1clean.R

@@ -1,0 +1,54 @@
+# Palmius Algorithm A1
+imputePalmiusA1 <- function(distEndHome = distEndHome,
+                            distStartHome = distStartHome,
+                            delta_t = delta_t,
+                            hour24 = hour24,
+                            locMissStart = locMissStart,
+                            locMissEnd = locMissEnd,
+                            delta_dist = delta_dist){
+  # prior functions to massage the data before
+  imp <- NA
+  if (delta_dist < 1000 & (delta_t <= 6*60*60 | (hour24 >= 21 & delta_t <= 12 * 60 * 60))) {
+    cat("Midpoint interpolation \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    
+    
+  } else if (distStartHome < 750 & distEndHome < 750 & (hour24 > 21 & delta_t <= 18 * 60 * 60)) {
+    cat("Midpoint interpolation \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    
+  } else if (distStartHome < 250 & distEndHome < 250 & (hour24 >= 21 & delta_t <= 18 * 60 * 60)) {
+    cat("Midpoint interpolation \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    
+    
+  } else if (distStartHome > 750 & distEndHome < 750 & (delta_t < 6 * 60 * 60 | (hour24 >= 20 & delta_t <= 18 * 60 * 60))) {
+    cat("To Home interpolation \n")
+    imp <- imputeToHome(tMissStart = tMissStart, tMissEnd = tMissEnd, locMissStart = locMissStart, locMissEnd = locMissEnd,delta_t = delta_t,distStartHome = distStartHome)
+    
[…]

### 2018-03-06 implementingBipolar/algoA3.R

@@ -1,0 +1,36 @@
+# algorithm A3
+
+imputeToHome <- function(tMissStart = tMissStart,
+                         tMissEnd = tMissEnd,
+                         locMissStart = locMissStart,
+                         locMissEnd   = locMissEnd,
+                         delta_t = delta_t,
+                         distStartHome = distStartHome,...){
+  #time to midpoint from begin while travelling at 80kmh
+  t_prime<- distStartHome/22.23
+  
+  if(delta_t <= t_prime){
+    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
+  }else if(delta_t-t_prime<= 2*60*60){
+    
+    imp1 <- linearInterpolate(locMissStart,locMissStart,delta_t - t_prime)
+    imp2 <- linearInterpolate(locMissStart,locMissEnd,t_prime)
+    imp <- rbind(imp1,imp2)
+  }else{
+    imp3 <- linearInterpolate(locMissStart,locMissStart,2*60*60)
+    tMissStart <- tMissStart-2*60*60
+    if(tMissEnd- tMissStart-t_prime - 2*60*60 >30*60){
+      t_cross <- min(2*60*60,as.numeric(tMissEnd- tMissStart-t_prime - 2*60*60))
+      imp1 <- linearInterpolate(locMissStart,locMissStart,tMissEnd-t_cross)
+      tMissEnd <- t_cross
+    }else{
+      imp1 <- NULL
+    }
+    imp2 <- linearInterpolate(locMissStart,locMissEnd,t_prime)
+    imp <- rbind(imp1,imp2,imp3)
+    return(imp)
+  }
+  
+  return(imp)
+}
+

### 2018-03-06 implementingBipolar/downSamplingAlgo.R

@@ -75,11 +75,23 @@ d.filt <- dF %>% thicken(by = "time","5 min",colname= "time_5min") %>%
   )
 
 #summerise and filter and plot and shit
-d.filt %>% 
+raw<- d.filt %>% 
   select(time_5min, lonF,latF)%>%
   distinct() %>%
   as_tbl_time(time_5min) %>% 
-  filter_time(~ "2017-02-15") %>% 
+  filter_time(~ "2017-02-15")
+
+d.f1<- downSamplingPalmius(d) %>%
+  as_tbl_time(time_5min) %>% 
+  filter_time(~ "2017-02-15")
+
+d.f1 %>% 
+  leaflet() %>%
+  addTiles() %>%
+  addCircles(lng = ~lonF, lat = ~latF)%>%
+  addProviderTiles(providers$CartoDB.Positron)
+
+d.filt %>% 
   leaflet() %>%
   addTiles() %>%
   addCircles(lng = ~lonF, lat = ~latF)%>%
@@ -98,8 +110,28 @@ for(i in 1: nrow(test)){
   test[i,c("lat","lon")] <- latlon
 }
 
-test %>% 
+ 
 leaflet() %>%
   addTiles() %>%
   addCircles(lng = ~lon, lat = ~lat)%>%
   addProviderTiles(providers$CartoDB.Positron)
+
+filtered <- d.filt %>% as_tbl_time(time) %>% 
+  filter_time(~ "2017-02-15")
+
+raw <- dataExample %>% as_tbl_time(time) %>% 
+  filter_time(~ "2017-02-15")
+
+saveRDS(filtered,"filtered.rds")
+saveRDS(raw,"raw.rds")
+
+ 
+leaflet() %>%
+  addTiles() %>%
+  addCircles(lng = ~lonF, lat = ~latF, data = raw, color = "#18206F", group = "Raw")%>%
+  addCircles(lng = ~lon, lat = ~lat, data=filtered,color = "#DF2935",group = "Filtered")%>%
+  addProviderTiles(providers$CartoDB.Positron) %>% 
+  addLayersControl(
+    overlayGroups = c("Raw", "Filtered"),
+    options = layersControlOptions(collapsed = FALSE)
+  )

### 2018-03-06 implementingBipolar/downsamplePalmius.R

@@ -1,0 +1,66 @@
+# Function to downsampale
+
+downSamplingPalmius <- function(data, sd = 14.14){
+  
+  library(dplyr)
+  library(tibbletime)
+  library(padr)
+  
+  #sd value comes from sqrt(10^2+10^2)
+
+  # takes raw data, calculates the mean hourly latitude and longitude
+  # and the standard deviation in distance within the hour
+  # pads and filters out those which have a higher standard deviation
+  # d3 then has hourly data with mean latitude and longitude
+  
+  d3<- data %>% thicken('hour') %>% 
+    group_by(time_hour) %>%
+    summarise(distancesMeterSd = sd(distancesMeter , na.rm = T),
+              meanLat = mean(lat),
+              meanLon = mean(lon)) %>% 
+    pad() %>% 
+    fill_by_value( "NA") %>%
+    filter(distancesMeterSd < sd) %>%
+    select(time_hour,meanLat,meanLon)
+  
+  # takes the raw data, thickens it by hour then joins d3 containing mean hourly lattitude and longitude
+  # then it mutates it so that lonF1 is the mean longitude when the mean longitude is not missing
+  dF<- data %>% thicken('hour') %>%
+    left_join(d3) %>%
+    mutate(
+      lonF1 = case_when(
+        is.na(meanLon) == F ~ meanLon
+      ),
+      latF1 = case_when(
+        is.na(meanLat) == F ~ meanLat
+      )
+    )
+  
+  # thickens by five minutes and takes the median longitude and lattitude
+  dF2<- dF %>% thicken(by = "time","5 min",colname= "time_5min") %>% 
+    group_by(time_5min) %>% 
+    summarise(lonF2 = median(lon),
+              latF2 = median(lat)
+    )
+  # thickens by five minutes and adds the two filtered types of data together
+  d.filt <- dF %>% thicken(by = "time","5 min",colname= "time_5min") %>% 
+    left_join(dF2, by = "time_5min") %>%
+    select(time, time_5min, lat,lon, accuracy, latF1,lonF1,latF2,lonF2) %>% 
+    mutate(
+      lonF = case_when(
+        is.na(lonF1) ==F ~ lonF1,
+        is.na(lonF1) == T ~lonF2),
+      latF = case_when(
+        is.na(lonF1 ) == F ~ latF1,
+        is.na(lonF1 ) == T ~ latF2
+      )
+    )
+  
[…]

### 2018-03-06 implementingBipolar/draftScripts.R

@@ -1,0 +1,145 @@
+# Palmius Recreation Script workspace
+
+# set wd
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+#read data
+data<- readRDS("../../data/boaz/myLocationHistory09012018.rds")
+
+# load auxiliary functions
+source("../../thesis/scripts/auxFuns.R")
+
+library(tidyr)
+#filter down to a year
+
+dataExample <- select(data,time,lat,lon,accuracy) %>%
+  arrange(time) %>% 
+  as_tbl_time(time) %>% 
+  filter_time( ~ '2017') %>%
+  mutate( nextLon = shift.vec(lon,-1),
+          nextLat = shift.vec(lat,-1),
+          nextTimeSec = shift.vec(time,-1)-time,
+          index = 1:length(lat)
+  )
+
+filtered<- filteringPalmius(data = dataExample)
+downsampled <-  downSamplingPalmius(data = filtered) #change name of time column, lat and lon
+
+preImp <- downsampled %>% mutate( timestampMs = as.numeric(time_5min),
+                       tMissEnd = case_when(
+                         is.na(lonF) ==  T~ NA_real_,
+                         !is.na(lonF)== T ~ timestampMs),
+                       tMissStart = tMissEnd,
+                       lonMissEnd =case_when(
+                         is.na(lonF) ==  T~ NA_real_,
+                         !is.na(lonF)== T ~ lonF),
+                       lonMissStart = lonMissEnd,
+                       latMissEnd = case_when(
+                         is.na(lonF) ==  T~ NA_real_,
+                         !is.na(lonF)== T ~ latF),
+                       latMissStart = latMissEnd) %>% 
+  fill(tMissEnd,lonMissEnd,latMissEnd, .direction ="up") %>% 
+  fill(tMissStart,lonMissStart,latMissStart, .direction ="down")
+
+# filter down to periods when living in utrecht
+
+# create NA's for variables where it makes sene
+home<- c(5.113919,52.10421)
+
+preImp$distEndHome<- apply(preImp, 1, FUN = function(row) {
+  raster::pointDistance(home,
+                        c(as.numeric(as.character(row["lonMissEnd"])), as.numeric(as.character(row["latMissEnd"]))),
+                        lonlat = T) # Parameter 'lonlat' has to be TRUE!
+})
+
[…]

### 2018-03-06 implementingBipolar/filteringAlgo.R

@@ -4,11 +4,12 @@
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 #read data
-data<- readRDS("../data/boaz/myLocationHistory09012018.rds")
+data<- readRDS("../../data/boaz/myLocationHistory09012018.rds")
 
 # load auxiliary functions
-source("../thesis/scripts/auxFuns.R")
-#filter down to a day
+source("../../thesis/scripts/auxFuns.R")
+
+#filter down to a year
 
 dataExample <- select(data,time,lat,lon,accuracy) %>%
   arrange(time) %>% 
@@ -20,6 +21,7 @@ dataExample <- select(data,time,lat,lon,accuracy) %>%
           index = 1:length(lat)
 )
 
+readr::write_csv(dataExample,"balazs2017.csv")
 rm(data)
 
 d<- dataExample

### 2018-03-06 implementingBipolar/filteringPalmius.R

@@ -1,0 +1,62 @@
+# impute missing data
+
+# palmius to do: create functions to execute his downsampling and filtering methods
+# create functions to execute his imputation algos. 
+
+filteringPalmius <- function(data, n = 10, maxSpeed = 27.7 ){
+  
+  d<- data
+  dPrimeSource<- 100
+  dPrimeDest <- 1
+  removed <- data.frame(lon = 1, lat = 1)
+  
+  while(dPrimeSource >= n & dPrimeDest < n){
+    
+    #calculate distance between points
+    d$distancesMeter<- apply(d, 1, FUN = function(row) {
+      raster::pointDistance(c(as.numeric(as.character(row["lon"])),
+                              as.numeric(as.character(row["lat"]))),
+                            c(as.numeric(as.character(row["nextLon"])), as.numeric(as.character(row["nextLat"]))),
+                            lonlat = T) # Parameter 'lonlat' has to be TRUE!
+    })
+    
+    #add index and calculate speed between points
+    d <- mutate(d,
+                speedMeterSecond= distancesMeter/as.numeric(nextTimeSec),
+                index = 1:nrow(dataExample)
+    )
+    
+    
+    # extract source 
+    filteredSource <- d %>% 
+      filter(speedMeterSecond >= maxSpeed)
+    
+    # extract destination
+    filteredDestination <- d %>%
+      filter(index %in% as.numeric(filteredSource$index+1))
+    
+    # extract modes
+    
+    sourceMode<- filteredSource %>% 
+      count(lon, lat) %>%
+      slice(which.max(n))
+    
+    destMode<- filteredDestination %>% 
+      count(lon, lat) %>%
+      slice(which.max(n))
+    
+    if(sourceMode$n >= destMode$n & sourceMode$n >n){
+      d <- filter(d, lon != sourceMode$lon & lat != sourceMode$lat)
+      removed <- rbind(removed, sourceMode[,c("lon","lat")])
+    } else if(destMode$n >= 10){
+      d <- filter(d, lon != destMode$lon & lat != destMode$lat)
+      removed <- rbind(removed, sourceMode[,c("lon","lat")])
+    }
+    
+    dPrimeSource<- sourceMode$n
+    dPrimeDest <- destMode$n
+    
+  }
+  cat(paste0("Removed ", nrow(removed)-1, " unique locations"))
+  return(d)
+}

### 2018-03-06 implementingBipolar/imputeFromHome.R

@@ -1,0 +1,36 @@
+# Impute distance from Home
+
+imputeFromHome <- function(tMissStart = tMissStart,
+                         tMissEnd = tMissEnd,
+                         locMissStart = locMissStart,
+                         locMissEnd   = locMissEnd,
+                         delta_t = delta_t,
+                         distEndHome = distEndHome){
+  #time to midpoint from begin while travelling at 80kmh
+  t_prime<- distEndHome/22.23
+  
+  if(delta_t <= t_prime){
+    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
+  }else if(delta_t-t_prime<= 2*60*60){
+    
+    imp1 <- linearInterpolate(locMissStart,locMissEnd,delta_t - t_prime)
+    imp2 <- linearInterpolate(locMissEnd,locMissEnd,t_prime)
+    imp <- rbind(imp1,imp2)
+  }else{
+    imp3 <- linearInterpolate(locMissEnd,locMissEnd,2*60*60)
+    tMissEnd <- tMissEnd-2*60*60
+    if(tMissEnd- tMissStart-t_prime - 2*60*60 >30*60){
+      t_cross <- min(2*60*60,as.numeric(tMissEnd- tMissStart-t_prime - 2*60*60))
+      imp1 <- linearInterpolate(locMissStart,locMissStart,tMissEnd-t_cross)
+      tMissStart <- t_cross
+    }else{
+      imp1 <- NULL
+    }
+    imp2 <- linearInterpolate(locMissStart,locMissEnd,tMissEnd-tMissStart)
+    imp <- rbind(imp1,imp2,imp3)
+    return(imp[-1])
+  }
+  
+  #check order of imputation, is it fine?
+  return(imp)
+}

### 2018-03-06 implementingBipolar/imputeMidpoint.R

@@ -1,0 +1,26 @@
+imputeMidpointPalmius <- function(tMissStart = tMissStart,
+                                  tMissEnd = tMissEnd,
+                                  locMissStart = locMissStart,
+                                  locMissEnd   = locMissEnd,
+                                  delta_t = delta_t,
+                                  midpoint = midpoint,
+                                  ...){
+  
+  #time to midpoint from begin while travelling at 80kmh
+  t_prime<- raster::pointDistance(midpoint,
+                                  locMissStart,
+                                  lonlat = T)/22.23
+  if( delta_t <= 2* t_prime){
+    # evenly spaced points between missing two ends
+    cat("Simple interpolation \n")
+    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
+  }else{
+    # goes to middle
+    cat("Midpoint wait interpolation \n")
+    imp1 <- linearInterpolate(locMissStart,midpoint_lon,t_prime/2)
+    imp2 <- linearInterpolate(midpoint,midpoint,delta_t-t_prime)
+    imp3 <- linearInterpolate(midpoint,locMissEnd,t_prime/2)
+    imp <- rbind(imp1,imp2,imp3)
+  }
+  return(imp)
+}

### 2018-03-06 implementingBipolar/linearInterpolate.R

@@ -1,0 +1,19 @@
+# Linear interpolate
+# simple function with Euclidean distance
+# not suitable for larger distances
+
+linearInterpolate <- function(start,
+                              end,
+                              delta_t = delta_t,
+                              interval = 300){
+  
+  if(floor(delta_t/interval)==0){
+  return()  
+  }
+  
+  iLon<- as.numeric(start[1]) + as.numeric((end[1]-start[1])/delta_t)*(1:as.numeric(delta_t/interval))
+  iLat<- as.numeric(start[2]) + as.numeric((end[2]-start[2])/delta_t)*(1:as.numeric(delta_t/interval))
+  interpolated<- cbind(lon = iLon, lat =iLat)
+  colnames(interpolated) <- c("lon","lat")
+  return(interpolated)
+}

### 2018-03-06 implementingBipolar/palmiusImpute.R

@@ -1,0 +1,54 @@
+# unify missing fuction
+
+data<- d.f1
+
+which(is.na(data$lonF))
+library(tidyr)
+d1 <- data %>% mutate( timestampMs = as.numeric(time_5min),
+                 tMissEnd = case_when(
+                   is.na(lonF) ==  T~ NA_real_,
+                  !is.na(lonF)== T ~ timestampMs),
+                 tMissStart = tMissEnd,
+                 lonMissEnd =case_when(
+                   is.na(lonF) ==  T~ NA_real_,
+                   !is.na(lonF)== T ~ lonF),
+                 lonMissStart = lonMissEnd,
+                 latMissEnd = case_when(
+                   is.na(lonF) ==  T~ NA_real_,
+                   !is.na(lonF)== T ~ latF),
+                 latMissStart = latMissEnd) %>% 
+  fill(tMissEnd,lonMissEnd,latMissEnd, .direction ="up") %>% 
+  fill(tMissStart,lonMissStart,latMissStart, .direction ="down")
+
+# missing: distance to home
+
+imputePalmiusA1
+
+#testing script
+tMissStart <- as.numeric(d1[35,"tMissStart"])
+tMissEnd <- as.numeric(d1[35,"tMissEnd"])
+
+lonMissEnd <- as.numeric(d1[35,"lonMissEnd"])
+lonMissStart <- as.numeric(d1[35,"lonMissStart"])
+
+latMissEnd <- as.numeric(d1[35,"latMissEnd"])
+latMissStart <- as.numeric(d1[35,"latMissStart"])
+
+tMissStart <- as.numeric(d1[129,"tMissStart"])
+tMissEnd <- as.numeric(d1[129,"tMissEnd"])
+
+lonMissEnd <- as.numeric(d1[129,"lonMissEnd"])
+lonMissStart <- as.numeric(d1[129,"lonMissStart"])
+
+latMissEnd <- as.numeric(d1[129,"latMissEnd"])
+latMissStart <- as.numeric(d1[129,"latMissStart"])
+locMissStart = c(lonMissStart,latMissStart)
+locMissEnd   = c(lonMissEnd,latMissEnd)
+
+imputeMidpointPalmius(tMissStart = tMissStart,
+                      tMissEnd = tMissEnd,
+                      locMissStart = locMissStart,
+                      locMissEnd   = locMissEnd)
+
+# algorithm A1
+# needs distance from home

### 2018-03-06 scripts/downSample.R

@@ -1,0 +1,33 @@
+# Downsample Function
+# this function takes dataframe with google location data and downsamples it using desired methods and intervals
+# in addition adding X Y coordinates
+
+downSampleMean <- 
+  function(data,interval = '5 min'){
+    
+    library(dplyr)
+    library(tibbletime)
+    library(padr)
+    
+    # first summarise via interval
+    data1 <- data %>%
+      select(timestampMs,time,lat,lon,accuracy) %>% 
+      as_tbl_time(time) %>% 
+      arrange(time) %>% 
+      mutate(timestampMs = as.numeric(timestampMs)) %>% 
+      thicken(interval = interval, colname = 'time2') %>%
+      group_by(time2) %>%
+      summarise(lon = mean(lon),# summarising function
+                lat = mean(lat),
+                code = 1)
+    
+    # then add XY coords
+    data2<- cbind(data1,as.data.frame(LatLong2XY(data1$lat, data1$lon))) %>% 
+      pad() %>%
+      mutate(code = case_when(is.na(x_v) == T ~4,
+                              is.na(x_v) == F ~1),
+             timestampMs = as.numeric(time2)) %>% 
+      select(code,timestampMs,lon,lat,x_v,y_v)
+    return(data2)
+  }
+

### 2018-03-06 scripts/exploratoryRawDataViz.R

@@ -1,8 +1,3 @@
-library(leaflet)
-
-Data_Start<- as.POSIXct("2015-09-07 01:55:00")
-Data_End  <- as.POSIXct("2015-09-10 01:59:00")
-
 
 exploreDates <- function(df = all, Data_Start= "2017-02-15", Data_End = "2017-02-17"){
     
@@ -36,19 +31,18 @@ exploreDates <- function(df = all, Data_Start= "2017-02-15", Data_End = "2017-02
   }
   
   time_formula <- create_time_formula(Data_Start, Data_End)
-  # new scatterplot
+  
+  if(!"accuracy" %in% colnames(df)){
+   accuracy <- 30
+  }
+  
   df %>%
     select(time,lat,lon,accuracy) %>%
     as_tbl_time(index = time) %>%
-    time_filter(time_formula = time_formula) %>%
+    filter_time(time_formula = time_formula) %>%
     leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
     addTiles() %>%
     addCircles(lng = ~lon, lat = ~lat,
                radius = ~accuracy, fillOpacity = 0.02,color = "#DF2935")%>%
     addProviderTiles(providers$CartoDB.Positron)
 }
-
-exploreDates(all,"2017-02-01","2017-02-01")
-
-+12:00:00
-+18:00:00

### 2018-03-06 scripts/pauseFlight.R

@@ -1,0 +1,31 @@
+# this function converts matrices of a certain form to flight, pause, missing or in between missing
+
+pauseFlight <- 
+  function(data,r, w){
+    avgmat <- as.matrix(data)
+    #added line to stop error where end time is missing
+    indx <- which(avgmat[,1] == 4)
+    avgmat[indx,3] <- avgmat[indx+1,2] 
+
+    outmat=c()
+    curind=1
+    cat("Convert from X/Y to flights/pauses...\n") # here comes the amazing flight extracting algo
+    for(i in 1:nrow(avgmat)){
+      #ProgressBar(nrow(avgmat),i)
+      if(avgmat[i,1]==4){
+        outmat=rbind(outmat,ExtractFlights(avgmat[curind:(i-1),c(5,6,2)],r,w),
+                     c(avgmat[i,1],NA,NA,avgmat[i,2],NA,NA,avgmat[i,3]))
+        curind=i+1
+      }
+    }
+    if(curind<=nrow(avgmat)){  #extract last bit
+      outmat=rbind(outmat,ExtractFlights(avgmat[curind:nrow(avgmat),c(4,3,2)],r,w))# why not work?
+    }
+    rownames(outmat)=NULL
+    colnames(outmat)=c("Code","x0","y0","t0","x1","y1","t1") # code stands for missing (4), pause (2) or flight (1) 
+    
+    return(outmat)
+  }
+
+
+#still wierd, pauses don't overlap

### 2018-03-07 implementingBipolar/implementationReport.Rmd
finalised the report

@@
 author: "Boaz Sobrado"
-date: "2/7/2018"
 output: html_document
@@
 Palmius uses several imputation algorithms to impute missing data.
+It took me more time than expected, but I've managed to implement them.
+You can see the exact scripts in the github.
+The results are below.
+A few key takeaways:
+
+1. This type of simple imputation works, but is not infallable. It is only as powerful as the downsampling, which I see as a fragile step here. 
+2.
+There are still some bugs I have to work on.
+I was imputed as being in the Benin at one point (I've never been there), which is either because of time zone differences and mean based downsampling or because longitude and lattitude got mixed up somewhere.
+
+The map below shows the downsampled vs the imputed measurements for a period of time of my movement within the Netherlands. Interestingly, the averaging does seem to show the paths quite well.
+
+```{r finalmap, echo = F}
+u2<- readRDS("imputedPalmius.rds")
+
+leaflet() %>%
+  addTiles() %>%
+  addCircles(lng = ~lonF, lat = ~latF, data = u2, color = "#18206F", group = "Downsampled")%>%
+  addCircles(lng = ~impLon, lat = ~impLat, data=u2,color = "#DF2935",group = "Imputed")%>%
+  addProviderTiles(providers$CartoDB.Positron) %>% 
+  addLayersControl(
+    overlayGroups = c("Downsampled", "Imputed"),
+    options = layersControlOptions(collapsed = FALSE)
+  )
 
 ```
-
-```

### 2018-03-07 implementingBipolar/draftScripts.R

@@ -128,11 +128,13 @@ for(i in indxFirstMissing){
 }
 
 u2 <- utrecht %>% 
-filter(distEndHome < 500000)
+filter(distEndHome < 500000 & timestampMs < (1.489*10^9) & (impLat > 24 | is.na(impLat)))
 
 utrecht %>% 
   filter(!is.na(impLon)) %>%
   summary()
+
+saveRDS(u2,"imputedPalmius.rds")
 
 leaflet() %>%
   addTiles() %>%
@@ -143,3 +145,13 @@ leaflet() %>%
     overlayGroups = c("Downsampled", "Imputed"),
     options = layersControlOptions(collapsed = FALSE)
   )
+
+# to do:
+# add gitignore
+# push to git
+# debug benin 
+# finish document and send to Palmius & Lugtig
+# finish document for Barnett & Onella 
+# send to lugtig and Ian
+
+# look into neural netowrks. 

### 2018-03-13 implementingBipolar/implementationReport.Rmd
added points to path script and RMD

@@
 
-1. This type of simple imputation works, but is not infallable. It is only as powerful as the downsampling, which I see as a fragile step here. 
+1. This type of simple imputation works, but is not infallable. It is only as powerful as the downsampling/filtering, which I see as a fragile step here. 
 2.
@@
 I was imputed as being in the Benin at one point (I've never been there), which is either because of time zone differences and mean based downsampling or because longitude and lattitude got mixed up somewhere.
+3. In terms of imputed time periods it performed better than I expected it to. Of the 9764 missing windows it imputed all but 62.
 
@@
 
-```{r finalmap, echo = F}
+```{r finalmap, echo = F, warning=F,message=F}
 u2<- readRDS("imputedPalmius.rds")

### 2018-03-13 ownAlgo/points2Path.Rmd
added points to path script and RMD

---
title: "Paths to points"
author: "Boaz Sobrado"
date: "3/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(leaflet)
library(tidyr)
library(padr)

data<- readRDS("/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/data/boaz/myLocationHistory09012018.rds")

home<- c(5.113919,52.10421)
uithof <- c(5.1635562, 52.0853133)

test <- data %>% mutate( distanceHome = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                                        home,
                                                        longlat = T),
                           timestampMs = as.numeric(time))%>%
  filter(distanceHome < 300000) %>% 
    select(time,lon,lat,accuracy, timestampMs,distanceHome) %>%
  arrange(timestampMs) %>% mutate(
  distanceU = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                    uithof,
                                    longlat = T),
  tNextH = case_when(
    distanceHome < 200 ~ timestampMs,
    TRUE ~ NA_real_),
  tNextU = case_when(
    distanceU < 200 ~ timestampMs,
    TRUE ~ NA_real_),
  tPrevH = case_when(
    distanceHome < 200 ~ timestampMs,
    TRUE ~ NA_real_),
  tPrevU = case_when(
    distanceU < 200 ~ timestampMs,
    TRUE ~ NA_real_)
  ) %>% 
fill(tNextH,tNextU,.direction ="up") %>%
  fill(tPrevH,tPrevU,.direction ="down") %>%
  mutate(tNextHSec = as.numeric(tNextH - timestampMs),
         tNextUSec = as.numeric(tNextU - timestampMs),
         tPrevUSec = as.numeric(timestampMs - tPrevU),
         tPrevHSec = as.numeric(timestampMs - tPrevH)
         ) %>% thicken(by = "time","1 day")%>%
  mutate( trip = case_when(
    (tNextUSec < 60*60 & tPrevHSec < 60*60)~1,
    (tPrevUSec < 60*60 & tNextHSec < 60*60)~2,
    TRUE ~0
  )) %>% 
  filter(trip !=0) %>%
  filter(accuracy < 16) %>% 
  select(lon,lat,trip,time_day,time)

```

## Introduction

As discussed, the next steps for my own method are:

1) Extract clusters of highly dense points called endpoints (work, home, etc.)
2) Extract paths between endpoints.
3) Bin all points along paths and sort measurements into the points.

This is a pre-requirement for training the neural networks. The following section is regarding steps 2 and 3.

## Path extraction

Assuming two endpoints have been located, the data contains a set of sequences where the individual was at endpoint A and then went to endpoint B with no endpoints in between. Moreover, it will also contain a set of paths which go from endpoint B to endpoint A. These points will be defined as the path between A and B.

As an example, see the map below. In the control panel code 1 denotes from Home to Uithof, the return is coded 2. The day 2 - 2012-02-06 is should not be included as it is a Utrecht - Albert Heijn trip and stems from the heuristic I have used so far to extract the two locations.

```{r map}
test %>%
  mutate( trip2 = paste0(trip,' -',time_day)) %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat,
             radius = 30, fillOpacity = 0.02,color = "#18206F",label = ~time,group= ~as.factor(trip2)) %>% 
  addLayersControl(
    overlayGroups = ~as.factor(trip2),
    options = layersControlOptions(collapsed = FALSE)
  )
```

If I have a single path I can use [principle curves analysis](http://www.geocomputation.org/2007/1B-Algorithms_and_Architecture1/1B2.pdf) to generate the curve which fits the points. However, as you can see we have multiple paths. A potential alternative is to use [local principal curves](https://epub.ub.uni-muenchen.de/1701/1/paper_320.pdf) which can handle multiple paths. The problem is that the assumption made there is that all paths come from the same distribution. This is evidently wrong: some paths are less frequently travelled than others.

Hence, I need a way of extracting the multiple unique paths between two points. The difficulty is that they are irregularly sampled lines, which makes using clustering mechanisms like k-means inappropriate. A potential solution is to use dynamic time warping as it is a way to evaluate the path similarities of objects travelling at different speeds.

```{r dtw}
library(dtw)
sample<- split(test[,1:2], f=list(test$trip,test$time_day), drop = T)
#splitting it into a list of data frames based on day and type of trip
distMatrix <- dist(sample, method="DTW") #create a distance matrix
hc <- hclust(distMatrix, method="average") #hierarchical clustering
plot(hc)

```

As you can see, dynamic time warping has done a good job at separating trips from home and trips to home. Nonetheless, given that we've made the assumption that the trip is the same, we are more interested in visually inspecting wether the paths it clustered together are indeed the ones that should be clustered together. To remove the difference from the algorithm, we switch the order of the trips from home in order that they become trips from the uithof.

```{r dtw2}
sample<- split(test[,1:3], f=list(test$trip,test$time_day), drop = T)

for(i in 1:length(sample)){
  if(sample[[i]][1,3] == 1){
    sample[[i]] <- sample[[i]]%>% arrange(-row_number())  
  }
  sample[[i]] <- sample[[i]][,1:2]
}

distMatrix <- dist(sample, method="DTW") #create a distance matrix
hc <- hclust(distMatrix, method="single") #hierarchical clustering
plot(hc)
```

Here we get what we want, kind of. The algorithm has identified two bigger clusters, which are essentially lots of dots vs few dots. This worries me a bit, but for now I will ignore it. Then, it essentially identifies two larger clusters within the group of the many dots. These are the two paths we want (approximately). If we cut it at an arbitrary height *h*, and group them based on that we get a single path.

```{r path}
clust<- data.frame( code = cutree(hc, k = NULL, h = 0.2),
            trip = names(cutree(hc, k = NULL, h = 0.2)))

tripMat <- left_join( test %>%
             mutate( trip = paste0(trip,'.',time_day)),
           clust) %>% #merge the clusters with the data
  filter( code ==1) %>% #filter to cluster 1
  select(lon,lat) %>% 
  as.matrix()

library(LPCM)
  lpc(tripMat, h = 0.03, scaled = T,control = lpc.control(iter = 1000)) %>% 
  plot()

```

Now we have an object which is essentially a single line between the two points. Now it is easy to write a function which can take the line and just draw a point on the line every X meters away from an endpoint.

```{r filteredPath}

x<- lpc(tripMat, h = 0.03, scaled = T,control = lpc.control(iter = 1000))

data <- unscale(x)[["LPC"]]

spacePath <- function(data = x,start = home, end = uithof, space = 100){
  i <- 1
  points <- list()
  distanceFromEnd <- raster::pointDistance(data[1,], end, longlat = T)

  while(distanceFromEnd > space){
    distance <-  raster::pointDistance(data, start, longlat = T)
    indx <- which(distance/space>=1)[1]
    if(is.na(indx)){break}
    points[[i]] <- data[indx,]
    distanceFromEnd <- raster::pointDistance(data[indx,], end, longlat = T)
    start <- data[indx,]
    data <- data[indx:nrow(data),]
    i <-  i+1
  }
  points<- matrix(unlist(points),ncol = 2, byrow = T)
  return(points)
}

spaced<- spacePath(data = data, start = home, end = uithof, space = 100)
plot(spaced)

```

Creating binned points is not enough, every measured point has to be classified in one of the binned points. The way I can do this is as follows:

```{r binFin}
dists<- raster::pointDistance(tripMat,spaced,longlat = T)
trip<- data.frame(tripMat, index = apply(dists,MARGIN = 1, FUN = which.min))
spaced <- data.frame(spaced, index = 1:nrow(spaced))
trip<- left_join(trip,spaced)

library(ggplot2)
ggplot(trip, aes(x=lon, y=lat, color=factor(index))) +
  geom_point(shape=1,alpha = 0.2) +
  geom_point(data = spaced, aes( x = X1, y= X2)) +theme_bw()

```

## Concerns and things to take into account:

1. The parameters of the path clustering function.
2. The parameters of the line drawing function.
3. The parameter of the spacing function.
4. Computational difficulties in running this for all routes.

## What is left to do

1. Create binned points for all routes.
2. Merge binned points if they overlap.
3. Bin all measured points on the route into these binned points.
4. Train the RNN.

### 2018-04-27 manuscript/manuscript.Rmd
added a vast range of files and started editing the manuscript

---
title: "Manuscript"
author: "Boaz Sobrado"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
```

# Abstract
*Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underutilised in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurments and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.*

# Introduction

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that policymakers and social scientists in numerous fields are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. Economists study geographic labour mobility extensively [e.g. @tatsiramos] and the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. Researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017]. Indeed, within psychology when studying behavioural differences within individuals it has been often been neglected that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary accross environments.

Despite their importance, a large part of studies on human mobility are conducted with pen-and-paper travel diaries. Pen-and-paper diares have well-known methodological flaws. For instance, they are expensive to collect and present a high burden to respondents. This limits the duration of data collection and the span of the log accordingly. Another issue is that they rely on self-reported measures. It has been reported that short trips are frequently under-reported  [@wolf_impact_2003] and the self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility. Such data  can now be obtained using the Global Positioning System (GPS). GPS uses the distance between a device and several satellites to determine location. GPS measurements can be used to infer a vast range of socioeconomic and behavioural measures, including where the individual lives, how much time he or she spends at home and where (and how) they travel.

However, not all GPS measurements are made equal. It is useful to classify GPS logs into different categories depending on their origin because different types of logs pose different methodological problems. We argue that social science researchers should explore using automatically collected smartphone GPS logs.  In most studies participants receive a specialised GPS devices to track their movement. We call resulting logs *specialised logs*. @barnett_inferring_2016 point out several methodological issues with these studies. Collecting specialised logs is costly and places a high burden on participants. Besides, introducing a new device to the participant's life may bias their behaviour. Due to these drawbacks specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). The problem with this is that researchers have to develop a custom-made app, distribute it among research participants and enforce participation. Participants tend not to like these tracking apps because they drain the battery of the device. Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints ( these constraints will be detailed subsequently). These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with missing data and measurement noise.

| Feature        | Specialised Logs           | Custom logs  | Secondary logs |
| ------------- |:-------------:| -----:| -----:|
| Sampling Frequency | Regular | Regular |  Irregular |
| Sampling Frequency (Hz) | 0.1>      | Between 0.1 and 0.01| <0.01 |
| Typical Duration | 7 days    |    7-14 days | Years |
| Typical Accuracy Variance | Low    |    Medium | High |

In order to work with secondary logs, researchers need to be able to handle the data sparcity that leads to missing data. Missing data is a pervasive issue in secondary logs as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015]. In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparcity it is also evident.

```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of a secondary log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}

include_graphics("img/missingBoaz5minExample.png")
```

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are autocorrelated in both time and space. This means that common methods, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers. Android phones collect location information through a variety of methods. Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation. Different methods are used because of computational and battery constraints. GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy often (but not always) is  associated with movement (\@ref(fig:accuracyPlot2)). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. ",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. All journeys are marked with a red line. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. All other journeys are not shown in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

```

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. This is known as data drift (REFERENCE). If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure 1. The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

This paper wishes to achieve four objectives: First, we have argued that understanding human mobility is important. Secondly, we argued that social scientists should leverage data logs from smartphones to study human mobility, instead of relying on out-dated pen-and-paper questionnaires. Third, explained the practical difficulties in using smartphone location logs. In the following sections we would like to introduce Personal Map Matched Imputation (PPMI), a method for surmounting these difficulties. We will compare PPMI to existing methods in the literature.

# Related Work

This section aims to explore related work with regard to missing data in GPS logs. We will argue three points. First, there is a lack of established practices to follow with respect to missing data in human mobility logs. Second, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs. Third, we discuss in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**].

How have researchers dealt with missing data in GPS-based human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. Ostensibly this is because they are unaware of possible solutions. Often researchers simply downsample and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in GPS logs as they relate to human mobility is of @barnett_inferring_2016. Both of these methods will be detailed subsequently, after introducing exploring other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility patterns does not imply there is not a vast literature on modelling movement. The most widespread models are SSMs, therefore we shall detail a few examples and subsequently argue that they are nonetheless unsuited for long term human mobility logs. Ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex. @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013]. The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance, assumptions which are violated in the case of human movement patterns. For this reason we do not consider SSMs to be useful for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal autocorrelations. For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

>"If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".

Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm.

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by:

> "simulat[ing] flights and pauses over the period of missingness where the direction,
duration, and spatial length of each flight, the fraction of flights versus the fraction of
pauses, and the duration of pauses are sampled from observed data."

This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses. In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours (*circadian proximity*).

On the other hand, **Wu 2017** use what they call a Spatial Temporal Semantic Neural Network to predict future human movement. While the authors are concerned with prediction and not imputation, they devised an algorithm called the Spatial Temporal Semantic algorithm converts raw measurements to machine learning friendly discrete bins. This minimises the measurement error and paves the way for applying machine learning methods to human mobility problems.

# Methodology

## Notation

GPS measurements provide us with coordinates on the surface of the earth, which is ellipsoid shaped. Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion. For clarity, when we use the term distance we refer to the geodesic distances on an ellipsoid using the WGS84 ellipsoid parameters.

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidian plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

When several data sets are available from individuals living in overlapping areas we can construct a $t \times i$ matrix $M$ where the entry $M(t,i)$ contains $G(t)$ for the individual $i$.

### MYMETHOD algorithm

Our algorithm is designed to leverage the periodic nature of human movement along with the long span of the data to deal with measurement sparcity and inaccuracy.

#### Modelling assumptions

First, following @barnett_inferring_2016 we categorise all timepoints $t$ as either belonging to the set $P$ (pause) or set $F$ (flight). Conceptually pauses can be understood as periods of time where an individual spends singificant amount of continous time without moving. Flights are the times where the individual is moving. Let $t_a$ be a pause of length $n$.
$$t_a = t_i < ... < t_{i+n}$$
Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$:
$$t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j$$
Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
$$t_x =  t_{i+n} < ... < t_{j} |t_x \in F$$
We define pause locations $G(t_a), G(t_b) | t_a, t_b \in P$ as locations where an individual spends an extended amount of time in the same space (e.g. school, home, work, train station, barber shop, bar, gym). Importantly, our model assumes period and cyclic human movement such that there are many pauses $t_{a1},t_{a2},...,t_{an}$ such that $G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})$. Moreover,it is possible for $G(t_a) = G(t_b)$ such that $t_a \ne t_b$. For example, if the individual leaves home for a run and returns home without stopping anywhere else.

Let us define as $Flight^x_{ab}$ the set of all points belonging to a flight between $G(t_a)$ and $G(t_b)$ at timepoint $t_x$.
$$Flight^x_{ab}= G(t_x)|t_x \in F =  \{G(t_{i+n}),...,G(t_j)\}$$
Again, there are many flights $t_{x1},t_{x2},...,t_{xn}$ such that $Flight^{x1}_{ab} = Flight^{x2}_{ab}  = ... = Flight^{xn}_{ab}$. Then, we can define as $Path_{ab}$ the set of all flights between $G(t_a)$ and $G(t_b)$ at all timepoints. For simplicity, we assume that $Path_{ab}=Path_{ba}$.

In addition, we consider all measurements $D(t)$ to be imperfect measurements of $G(t)$:

$$ G(t) = D(t) + \text{Measurement Error}  $$

### MYMETHOD algorithm

Our algorithm performs the following steps:

1. Extract from measurements $D$ all pause locations and path clusters. Following **Wu 2017 (NOTE: ADD REFERENCE)** we use a modified spatial-temporal-semantic (STS) feature extraction algorithm to transform pause and path locations into machine learning friendly discrete location sequences.
2. Assign each measurement $D$ to a unique discrete location. 
3. Use neural networks to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.

In this paper we have chosen relatively simple functions as component functions of MYMETHOD. It is clear each of them can be improved. We aim to illustrate not the efficacy of the components themselves but rather how they can work together to impute missing data.

#### Extract pause locations

There are multiple ways of extracting pause locations in the literature (**examples & citations needed: see Wu 2017**). As far as we know there are only two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these processes pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. In particular, @barnett_inferring_2016 use a version of @rhee_human_2007's rectangular method to classify pauses and determine flight sub-sections. MYMETHOD uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

First the measurements $D$ are filtered such that only measurements with an accuracy value lower than $a_{\text{P lim}}$ remain within the sample. Then, a measurement $D_t$ belongs to a pause if and only if:

1. The next measurement $D_{t+1}$ is within $t_{\text{Pause lim}}$ amount of seconds (so it is not missing)
2. The next measurement $D_{t+1}$ is within $d_{\text{Pause lim}}$ meters.
3. The duration of the pause is more than $\delta_{\text{Pause lim}}$ seconds.
4. Let the points of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$. These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a cluster.

#### Extract paths

To extract path clusters we apply a modified version of the spatial-temporal-semantic (STS) feature extraction presented in **Wu 2017 (NOTE: ADD REFERENCE)**.

The key modification is that whilst **Wu 2017 (NOTE: ADD REFERENCE)** use map matching as a ground truth to assign noisy measurements into discrete bins along paths, we use the total mobility history of the individual to do so. This leverages the periodic nature of human movement as well as the long span of secondary logs. Therefore we consider each measurement at $D_x$ as a sample of $Path_{ab}$. In practice, for each observed path, MYMETHOD uses the following algorithm to estimate $Path_{ab}$:

1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
2. Create a distance matrix for each point and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.

At this point all empirically observed path clusters and pause clusters are extracted. However, there may be some overlap between pause clusters and path clusters. Thus, the clusters are clustered again, such that the pause clusters retain priority. This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed. The reasoning for this is twofold: first, that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. Secondly, due to data drift it can often appear that there is movement when there is none.The end result is a discretised map which contains pause and flight clusters based on the history of the individual.

Subsequently, all measurements $D$ were are then assigned to a discrete cluster. We used a simple assigning function, whereby the measurements where assigned to the cluster nearest to the measurement. We shall discuss the merits of choosing a more sophisticated assigning function subsequently.

#### Neural network

An artificial neural network was trained to predict the missing values. The input variables to the neural net are:

1. The previous and subsequent observed cluster.
2. The distance in time to the next & previous cluster.
3. The time of the day encoded as a cyclical two-dimensional feature **NOTE: add reference**
4. The day of the month encoded as a cyclical two-dimensional feature **NOTE: add reference**
5. The month of the year as a binary class matrix.

For more information on the precise architecture of the artificial neural network please consult the appendix.

For a missing timepoint at $D_t$, the output of the model is a set of probability estimates associated with every location cluster. That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per cluster) associated with where the individual is.

### Datasets & Analyses

The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (table 1).

```{r datadetailsTable, echo = F, eval = F}
datadetails <- readRDS("tempdata/datadescriptives.rds")
datadetails <- datadetails[,-7] #remove last column for now
kable(datadetails,digits = 2,col.names = c("Log duration",
                                           "Logged days",
                                           "Observations",
                                           "Missing days",
                                           "Missing data",
                                           "Mean Accuracy"),
      caption = "Table with descriptives about the data sets used to build the imputation methods. Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.")
```

Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2]. **add packages**

### Results & Evaluation Metrics

The results will consist of multiple steps:

1. A descriptive documentation of the characteristics of missingness in secondary logs.
2. Evaluating the performance of the modified STS algorithm in extracting paths as well as two rival methods.
3. Comparing the performance of MYMETHOD using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.

#### Missingness

With respect to the characteristics of missingness we will report how often data is missing and whether missingness displays temporal or spatial patterns.

#### Binning evaluation

In order to evalute the modified STS algorithm we will first visually evaluate the paths and pause locations. A visual evaluation of paths is an established way to heuristically check their accuracy (*add reference*). Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $d$. With respect to the deviation distance, we expect:

1. A positive relationship between the deviation distance and the accuracy of each measurement.
2. Roughly 67% of the deviation distances $d$ are within accuracy $a$ of each measurement.

#### Imputation algorithm performance

We treat missing data imputation as a classification problem, taking 25% of all measurements at random and eliminating them from the sample, recalculating features and computing classification accuracy. Moreover, we evaluate the performance of the imputation algorithm for the entire duration of the log with in comparison to a naive baseline models where for each missing period we predict the prevous non-missing cluster.

To compare the performance of the different missing data imputation algorithms we will remove 25% of measured time intervals at random within a four week period. We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.

To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location). For MYMETHOD's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally. This is necessary because Barnett's and Palmius' model estimate far fewer values than MYMETHOD. While Barnett's and Palmius' models estimate 12 measurements for each missing hour, MYMETHOD estimates as many measurements as there are in the sample. When individuals go to unfrequently travelled locations they tend to look use their phone's location services more, which leads to more measurements, hence more imputed values. Hence, if the weights are not used MYMETHOD is succeptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.

In addition, other measures we will report for MYMETHOD are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance. The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster. For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.

Finally, we will take objective real world data and compare it to predicted values. We will use information from the Dutch public transportation card, which provides users with timestamped data of when and where they board, change lines or leave public transportation.

## Results

### A descriptive documentation of the characteristics of missingness

This is an example of a smartphone secondary log. About 54% of the data is missing for the entire duration of the log. This may be misleading as there are several long periods with no measurements whatsoever (see Figure \@ref(fig:longMeasurementsPerDay)). For days which were not entirely missing, approximately 22% of all five minute segments were missing. The structure of missingness of a day with measurements is shown in Figure \@ref(fig:measurementsPerDay). As you can see, there are several long periods over the course of the log for which there are no measurements. Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case. The median sampling frequency per day for non-missing days is around 0.006 Hz.

```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data on February 15th 2017. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingdayBoaz5min.png")

```

### Modified STS performance in comparison to other binning methods

Before we can evaluate the model, it is essential to evaluate how the noisy data has been restructured. Without this evaluation model accuracy measures could be misleading, as a good performance may simply be modelling noise. Therefore we will assess the performance of the modified STS binning method and compare it to alternatives.

We used the following parameters to extract pauses: an accuracy limit $a_{\text{Pause lim}}$ of 250 meters, a time limit $t_{\text{Pause lim}}$ of 300 seconds, a distance limit of $d_{\text{Pause lim}}$ 50 meters and a minimum pause duration limit $\delta_{\text{Pause lim}}$ of 100 seconds. Moreover, to extract path clusters we used the parameters: $a_{\text{Path Lim}} = 150  \text{meters}$ and $d_{\text{Path Lim}} =300  \text{meters}$.

Importantly, we did not use the entire length of the log, but rather only the time periods when the individual was in the Netherlands. This was done to reduce computation time.

When selecting parameters there is a trade off is between resolution and accuracy. This is because an increase in resolution comes at the expense of accuracy. For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster. This is illustrated in Figures \@ref(fig:pauseLocUithof1) and \@ref(fig:pauseLocUithof2).

```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
#fix width
include_graphics("img/mapPauseRes1.png")
```

```{r, pauseLocUithof2,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
include_graphics("img/mapPauseRes2.png")
```

The end result of step 3 of MYMETHOD results in a personalised map with pause and path clusters. An excerpt can be seen in Figure \@ref(fig:mapClusterMap). As expected, MYMETHODs path extraction yields greater accuracy for frequently ocurring paths. It is important to remind the reader that despite the close overlap with features on the map,  such as roads and railway lines, MYMETHOD is map agnostic and uses no information from the map to recreate these features.

```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map if an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
#fix width
include_graphics("img/mapClusterMap.png")
```

For the entire period when the individual is in the Netherlands, there is a mean deviance of 40 meters and a median deviance of 15 meters. Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected.

For the narrower period of March 2017, figure \@ref(fig:devianceComp) illustrates the relationship between the deviance and accuracy. Approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected. The mean and median deviance are 38 and 14 meters respectively.

```{r, devianceComp,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
# add figure with deviance comparisons three columns each method once
```

In comparison, Palmius' method has a median deviance of 3 meters, with a mean of 115 due to high deviance outliers. Over 90% of deviance values are within accuracy, which suggests that the model is binning noise. In Figure \@ref(fig:palmiusVme) the key difference is visible. The Palmius method is much more sensitive to inaccurate values, such as the one at the top right of the figure.

On the other hand, Barnett's method has mean deviance of 343 meters and a median deviance of 8 meters. Barnett's deviance is neccessarily higher than Palmius', as they downsample and subsequently aggregate into pauses and linear flights.

```{r, palmiusVsMe1,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
#fix width
include_graphics("img/palmiusVme.png")
```

```{r, palmiusVsMe2,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
include_graphics("img/binComp.png")
```

### Comparison to alternative models

```{r,resulttable, echo = F}
library(knitr)
library(kableExtra)

resultsDf <- data.frame( mean5min = c(82,43,269,426),
                         median5min = c(4,0,0,0),
                         meanexpDist5min = c(NA,NA,414,NA),
                         medianexpDist5min = c(NA,NA,0,NA),
                         mean1hr = c(345,497,908,1502),
                         median1hr = c(6,4,0,0),
                         meanexpDist1hour = c(NA,NA,916, NA),
                         medianexpDis1hour = c(NA,NA,0, NA),
                         mean1day = c(9273,NA,5757,14266),
                         median1day = c(12,NA,0,1288),
                         meanexpDist1Day = c(NA,NA,5757, NA),
                         medianexpDis1Day = c(NA,NA,0, NA))

rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive")
colnames(resultsDf) <- c("Mean","Median","Expected Mean","Expected Median",
                         "Mean","Median","Expected Mean","Expected Median",
                         "Mean","Median","Expected Mean","Expected Median")

# median 12 meters mean 9273
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '-')
kable(resultsDf, "html") %>%
  kable_styling("striped") %>%
  add_header_above(c(" " = 1, "Five Minutes" = 4, "One Hour" = 4,"One Day" = 4))

```

In terms of coverage, it is important to note that the Palmius model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests. Palmius' model failed to impute a single value for the day tests. Barnett's failed to impute 11% of the missing values. MYMETHOD made an imputation for all missing values.

As for MYMETHOD's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively. As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively.

### Comparison with objective data

The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters. On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a median distance of 5637. The expected distance has a mean of 6432 and a median of 1037.

Palmius's model failed to impute 38 out of 97 periods. For those it did impute, it showed a mean distance of 1517 meters and a median distance of 1617. Barnett's model failed to impute 18 out of 97 periods. For those it did impute, it had a mean of 5506 and a median of 1342.

### Effect on aggregate measures

Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled. As we discussed earlier, aggregations without data processing can be highly biased. However, as an example we calculate time spent at home of the user for each day of the week in the month of February (\@ref(fig:aggrePlot)). Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].

```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017. We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month. Missing values are filled in by the previous observed value. Behavioural trends are evident, the user spends more time at home on weekends than on weekdays. Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
include_graphics("img/aggPlot.png")
```

## Discussion

Overall the MYMETHOD performed well, although not substantially better than the naive model.  In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs. Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation.

In addition to higher accuracy under the conditions of secondary logs, the advantages of MYMETHOD are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smoothen movement noise.

MYMETHOD's increased coverage and flexibility comes from its ability to make complex non-linear predictions. For instance, in a given missing period it might make sense to predict that the individual is either at home, or at the office, or at a shop with equal probability. While MYMETHOD can make such an imputation, none of the alternative methods can do this. Moreover, the ability to take the prediction probability values from the neural network also helps in dealing with uncertainty. A known-drawback of single imputation is that it takes an imputed value and treats it as observed (*RUBIN*). Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods. With MYMETHOD it is possible to construct confidence intervals using the predicted probabilities of each estimate.

With respect to irregular sampling, alternative methods use temporally based downsampling in order to reduce noise. This leads to deterioration in resolution not only over space, but also over time. A combination of irregular sampling with the fluctuating accuracy values can lead to nonsensical results. For instance, consider a case where there are two inaccurate measurements in movement at 12:00:01 and 12:04:59. Downsampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not. MYMETHOD instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations.

While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for paths. Hence all paths are calculated based on sensor input only. This may be feasible with high frequency regularly sampled short duration logs but creates noise with secondary logs. Moreover, with secondary logs it is feasible to spatially "average out" multiple samples of the same path in order to recreate it in its entirety. For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (*report value here*) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.

There are multiple methodological limitations in this paper. Most importantly, the evaluation methods are imperfect. The golden standard would be to use at least one highly accurate professional grade GPS device with high sampling frequency to compare our data to. Until that is available, the use of public transport data and crossvalidation is just a substitute.

Furthermore, MYMETHOD also has several limitations. The pause extracting function, the path clustering function, the classification model and the assignment function remain simplistic and could be improved. Indeed, our intention is not to develop an optimal classification model or any of the other components of MYMETHOD, but rather to show a new way in which secondary logs can be made useful by social scientists.

With regards to pause extraction, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location. These factors are not taken into account in the pause extraction function. Improved methods would do well to do so.

As for the path clustering function, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density. It does not take into account the length of the path as well as the average sampling frequency of the path. This is an issue because it can lead to clusters to which data points are seldomly assigned. For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements. This leads to difficulties in the machine learning part of the model, as unfrequently observed clusters are hard for the model to predict.

Similarly, a limitation of the current assignment function is that it only takes into account the distance between the observed measurement and all the clusters when assigning each measurement to a cluster. It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on. For instance, assume that it is known that an individual is travelling from point A to point B along path AB, and there is an inaccurate measurement closest to a cluster which belongs to path AC. By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC. An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and the ANN model. This could take the form of a Kalman filter. In this case the state side of the model would represent a probabilistic representation of where the individual could be based on what the ANN has extracted from the individuals movement history. The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.

As for the simplicity of the neural network which was used to generate predictions. A more sophisticated recurrant neural network (RNN), or a long short-term memory recurrant neural network (LSTM) would likely perform significantly better.

## Conclusion

Advantages of MYMETHOD:

1. Full imputation coverage.
2. Probability of cluster as measure of uncertainty.
3. Reduction of data drift through discretisation.
4. Robust to irregular sampling.
5. Few assumptions.

Disadvantages:

1. Weak in unfrequently sampled locations. Perhaps use alternative methods there (e.g. Kalman filter)
2. Some sacrifice of accuracy for consistency. See high vs low resolution pause maps.

Further research:

1. More sophisticated pause/path extracted algorithms.
2. More sophisiticated ML models.
3. Introducing explicit assumptions about paths.
4. Run a study where individuals have a high frequency sampling high accuracy GPS receiver on them (test out of sample performance).
5. Path extraction for train stuff.

*To do: add info on paragraphs and sections/subsections.*

### 2018-04-27 implementingBipolar/algoA1clean.R

@@ -1,54 +1,0 @@
-# Palmius Algorithm A1
-imputePalmiusA1 <- function(distEndHome = distEndHome,
-                            distStartHome = distStartHome,
-                            delta_t = delta_t,
-                            hour24 = hour24,
-                            locMissStart = locMissStart,
-                            locMissEnd = locMissEnd,
-                            delta_dist = delta_dist){
-  # prior functions to massage the data before
-  imp <- NA
-  if (delta_dist < 1000 & (delta_t <= 6*60*60 | (hour24 >= 21 & delta_t <= 12 * 60 * 60))) {
-    cat("Midpoint interpolation \n")
-    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
-                                 tMissEnd = tMissEnd,
-                                 locMissStart = locMissStart,
-                                 locMissEnd = locMissEnd,
-                                 midpoint = midpoint,
-                                 delta_t = delta_t)
-    
-    
-  } else if (distStartHome < 750 & distEndHome < 750 & (hour24 > 21 & delta_t <= 18 * 60 * 60)) {
-    cat("Midpoint interpolation \n")
-    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
-                                 tMissEnd = tMissEnd,
-                                 locMissStart = locMissStart,
-                                 locMissEnd = locMissEnd,
-                                 midpoint = midpoint,
-                                 delta_t = delta_t)
-    
-  } else if (distStartHome < 250 & distEndHome < 250 & (hour24 >= 21 & delta_t <= 18 * 60 * 60)) {
-    cat("Midpoint interpolation \n")
-    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
-                                 tMissEnd = tMissEnd,
-                                 locMissStart = locMissStart,
-                                 locMissEnd = locMissEnd,
-                                 midpoint = midpoint,
-                                 delta_t = delta_t)
-    
-    
-  } else if (distStartHome > 750 & distEndHome < 750 & (delta_t < 6 * 60 * 60 | (hour24 >= 20 & delta_t <= 18 * 60 * 60))) {
-    cat("To Home interpolation \n")
-    imp <- imputeToHome(tMissStart = tMissStart, tMissEnd = tMissEnd, locMissStart = locMissStart, locMissEnd = locMissEnd,delta_t = delta_t,distStartHome = distStartHome)
-    
[…]

### 2018-04-27 implementingBipolar/draftScripts.R

@@ -46,6 +46,7 @@ preImp <- downsampled %>% mutate( timestampMs = as.numeric(time_5min),
 # create NA's for variables where it makes sene
 home<- c(5.113919,52.10421)
 
+#add additional features needed for the imputation
 preImp$distEndHome<- apply(preImp, 1, FUN = function(row) {
   raster::pointDistance(home,
                         c(as.numeric(as.character(row["lonMissEnd"])), as.numeric(as.character(row["latMissEnd"]))),
@@ -150,8 +151,3 @@ leaflet() %>%
 # add gitignore
 # push to git
 # debug benin 
-# finish document and send to Palmius & Lugtig
-# finish document for Barnett & Onella 
-# send to lugtig and Ian
-
-# look into neural netowrks. 

### 2018-04-27 implementingBipolar/fullPalmiusResults.R

@@ -1,0 +1,147 @@
+# Full Palmius Script Results
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+data<- readRDS("../ownalgo/finalBinnedDataNL.rds")
+
+# load auxiliary functions
+source("../../thesis/scripts/auxFuns.R")
+
+file.sources <- list.files(c("functions"), 
+                           pattern="*.R$", full.names=TRUE, 
+                           ignore.case=TRUE)
+sapply(file.sources,source,.GlobalEnv)
+
+#select the month of march
+
+dataExample <- select(data,time,lat,lon,accuracy,nextMeas) %>%
+  arrange(time) %>% 
+  mutate( nextLon = lead(lon),
+          nextLat = lead(lat),
+          index = 1:length(lat),
+          nextTimeSec = nextMeas
+  ) %>% 
+  as_tbl_time(time) %>% 
+  filter_time( ~ '2017-03')
+
+filtered<- filteringPalmius(data = dataExample) #no unique locations removed
+
+downsampled <-  downSamplingPalmius(data = filtered) #change name of time column, lat and lon
+
+## calculate deviance
+
+test <- left_join(dataExample %>% thicken("5 min"),
+          downsampled,
+          by =c("time_5_min" = "time_5min")) %>% 
+  select(lon,lat,lonF,latF,accuracy, time_5_min)
+
+test$dist <- raster::pointDistance(as.matrix(cbind(test$lon,test$lat),ncol = 2),
+                      as.matrix(cbind(test$lonF,test$latF),ncol = 2),
+                      lonlat = TRUE)
+
+summary(test$dist)
+
+test %>%  summarise(mean(accuracy >= dist)) #each measurement
+# by bin
+test %>% group_by(time_5_min) %>% summarise(distMean = mean(dist),
+                                            distSD   = sd(dist)) %>% summarise(mean(distMean))
+library(ggplot2)
+library(ggthemes)
+ggplot(test, aes(x=accuracy, y=dist)) +
+  geom_point()+theme_tufte()+
+  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0,
+              na.rm = FALSE, show.legend = NA)
+
+
+#there is something wierd happening with these outliers
+test %>% 
+  filter(dist>100000) %>% 
+  leaflet() %>% 
+  addTiles() %>% 
[…]

### 2018-04-27 implementingBipolar/functions/algoA1clean.R

@@ -1,0 +1,67 @@
+# Palmius Algorithm A1
+imputePalmiusA1 <- function(distEndHome = distEndHome,
+                            distStartHome = distStartHome,
+                            delta_t = delta_t,
+                            hour24 = hour24,
+                            locMissStart = locMissStart,
+                            locMissEnd = locMissEnd,
+                            delta_dist = delta_dist,
+                            midpoint = midpoint,
+                            tMissEnd = tMissEnd,
+                            tMissStart = tMissStart){
+  # prior functions to massage the data before
+  imp <- NA
+  type <- "None"
+  if (delta_dist < 1000 & (delta_t <= 6*60*60 | (hour24 >= 21 & delta_t <= 12 * 60 * 60))) {
+    cat("Midpoint interpolation 1 \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    type <- "Midpoint Imputatation 1"
+    
+    
+  } else if (distStartHome < 750 & distEndHome < 750 & (hour24 > 21 & delta_t <= 18 * 60 * 60)) {
+    cat("Midpoint interpolation 2 \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    type <- "Midpoint Imputatation 2"
+  } else if (distStartHome < 250 & distEndHome < 250 & (hour24 >= 21 & delta_t <= 18 * 60 * 60)) {
+    cat("Midpoint interpolation 3 \n")
+    imp <- imputeMidpointPalmius(tMissStart = tMissStart,
+                                 tMissEnd = tMissEnd,
+                                 locMissStart = locMissStart,
+                                 locMissEnd = locMissEnd,
+                                 midpoint = midpoint,
+                                 delta_t = delta_t)
+    type <- "Midpoint Imputatation 3"
+    
+  } else if (distStartHome > 750 & distEndHome < 750 & (delta_t < 6 * 60 * 60 | (hour24 >= 20 & delta_t <= 18 * 60 * 60))) {
[…]

### 2018-04-27 implementingBipolar/functions/featureExtractPalmius.R

@@ -1,0 +1,54 @@
+# feature extract palmius
+
+#change the data format as to make the imputation possible
+featureExtractPalmius <- function(data){
+  preImp <- data %>% mutate( timestampMs = as.numeric(time_5min),
+                             tMissEnd = case_when(
+                               is.na(lonF) ==  T~ NA_real_,
+                               !is.na(lonF)== T ~ timestampMs),
+                             tMissStart = tMissEnd,
+                             lonMissEnd =case_when(
+                               is.na(lonF) ==  T~ NA_real_,
+                               !is.na(lonF)== T ~ lonF),
+                             lonMissStart = lonMissEnd,
+                             latMissEnd = case_when(
+                               is.na(lonF) ==  T~ NA_real_,
+                               !is.na(lonF)== T ~ latF),
+                             latMissStart = latMissEnd) %>% 
+    fill(tMissEnd,lonMissEnd,latMissEnd, .direction ="up") %>% 
+    fill(tMissStart,lonMissStart,latMissStart, .direction ="down") %>% 
+    mutate(code = case_when(
+      is.na(lonF) ==  T~ 0,
+      !is.na(lonF)== T ~ 1),
+      priorCode = lag(code),
+      hour24 = as.numeric(strftime(time_5min, "%H")),
+      firstMissing = case_when(
+        code == 0 & priorCode == 1 ~ 1,
+        TRUE ~ 0))
+  
+  #how many missing? 
+  preImp %>% summarise(mean(is.na(lonF))) #about 12 percent
+  preImp %>% summarise(mean(code)) #non missing
+  
+  #add additional features needed for the imputation
+  home<- c(5.113919,52.10421)
+  preImp$distEndHome<- apply(preImp, 1, FUN = function(row) {
+    raster::pointDistance(home,
+                          c(as.numeric(as.character(row["lonMissEnd"])), as.numeric(as.character(row["latMissEnd"]))),
+                          lonlat = T) # Parameter 'lonlat' has to be TRUE!
+  })
+  
+  preImp$distStartHome<- apply(preImp, 1, FUN = function(row) {
+    raster::pointDistance(home,
+                          c(as.numeric(as.character(row["lonMissStart"])), as.numeric(as.character(row["latMissStart"]))),
+                          lonlat = T) # Parameter 'lonlat' has to be TRUE!
+  })
+  
+  preImp$delta_dist<- apply(preImp, 1, FUN = function(row) {
[…]

### 2018-04-27 implementingBipolar/functions/imputeFromHome.R

@@ -1,0 +1,38 @@
+# Impute distance from Home
+
+imputeFromHome <- function(tMissStart = tMissStart,
+                         tMissEnd = tMissEnd,
+                         locMissStart = locMissStart,
+                         locMissEnd   = locMissEnd,
+                         delta_t = delta_t,
+                         distEndHome = distEndHome){
+  #time to midpoint from begin while travelling at 80kmh
+  t_prime<- distEndHome/22.23
+  
+  if(delta_t <= t_prime){
+    c("type 1 \n")
+    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
+  }else if(delta_t-t_prime<= 2*60*60){
+    c("type 2 \n")
+    imp1 <- linearInterpolate(locMissStart,locMissEnd,delta_t - t_prime)
+    imp2 <- linearInterpolate(locMissEnd,locMissEnd,t_prime)
+    imp <- rbind(imp1,imp2)
+  }else{
+    imp3 <- linearInterpolate(locMissEnd,locMissEnd,2*60*60)
+    tMissEnd <- tMissEnd-2*60*60
+    if(tMissEnd- tMissStart-t_prime - 2*60*60 >30*60){
+      t_cross <- min(2*60*60,as.numeric(tMissEnd- tMissStart-t_prime - 2*60*60))
+      imp1 <- linearInterpolate(locMissStart,locMissStart,tMissEnd-t_cross)
+      tMissStart <- t_cross
+    }else{
+      imp1 <- NULL
+    }
+    imp2 <- linearInterpolate(locMissStart,locMissEnd,tMissEnd-tMissStart)
+    c("type 3 \n")
+    imp <- rbind(imp1,imp2,imp3)
+    return(imp[-1,])
+  }
+  
+  #check order of imputation, is it fine?
+  return(imp)
+}

### 2018-04-27 implementingBipolar/functions/imputeMidpoint.R

@@ -1,0 +1,26 @@
+imputeMidpointPalmius <- function(tMissStart = tMissStart,
+                                  tMissEnd = tMissEnd,
+                                  locMissStart = locMissStart,
+                                  locMissEnd   = locMissEnd,
+                                  delta_t = delta_t,
+                                  midpoint = midpoint,
+                                  ...){
+  
+  #time to midpoint from begin while travelling at 80kmh
+  t_prime<- raster::pointDistance(midpoint,
+                                  locMissStart,
+                                  lonlat = T)/22.23
+  if( delta_t <= 2* t_prime){
+    # evenly spaced points between missing two ends
+    cat("Simple interpolation \n")
+    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
+  }else{
+    # goes to middle
+    cat("Midpoint wait interpolation \n")
+    imp1 <- linearInterpolate(locMissStart,midpoint,t_prime/2)
+    imp2 <- linearInterpolate(midpoint,midpoint,delta_t-t_prime)
+    imp3 <- linearInterpolate(midpoint,locMissEnd,t_prime/2)
+    imp <- rbind(imp1,imp2,imp3)
+  }
+  return(imp)
+}

### 2018-04-27 implementingBipolar/functions/palmiusImputeLoop.R

@@ -1,0 +1,71 @@
+palmiusImputeLoop <- function(inputDF){
+  
+  timeNow <- Sys.time()
+  indxFirstMissing<- which(inputDF$firstMissing == 1)
+  
+  inputDF$impLon <- NA
+  inputDF$impLat <- NA
+  inputDF$impType <- NA
+  mismatch <- 0
+  mismatchList <- list()
+  for(i in indxFirstMissing){
+    cat(i)
+    data<- inputDF[i,]
+    # extract variables
+    tMissEnd<- data["tMissEnd"]
+    tMissStart<- data["tMissStart"]
+    
+    #distance to end etc
+    distEndHome <- data["distEndHome"]
+    distStartHome <- data["distStartHome"]
+    
+    #location missing end and beginning
+    locMissEnd<- data[c("lonMissEnd", "latMissEnd")]
+    locMissStart<- data[c("lonMissStart", "latMissStart")]
+    
+    
+    #Hour in PM needed
+    hour24 <- data["hour24"]
+    delta_dist <- data["delta_dist"]
+    
+    # Calculate time elapsed
+    delta_t <- tMissEnd - tMissStart
+    # length of missing intervals
+    lengthMiss <- (delta_t/interval)-1 
+    
+    # Calculate midpoints
+    midpoint <- colMeans(rbind(locMissStart,setNames(locMissEnd, names(locMissStart))))
+    
+    #impute 
+    imputed <- imputePalmiusA1(distEndHome = distEndHome,
+                               distStartHome = distStartHome,
+                               delta_t = delta_t,
+                               hour24 = hour24,
+                               locMissStart = locMissStart,
+                               locMissEnd = locMissEnd,
+                               delta_dist = delta_dist,
+                               midpoint = midpoint,
+                               tMissEnd = tMissEnd,
+                               tMissStart = tMissStart)
+    inputDF[(i:as.numeric(i+lengthMiss-1)),"impType"] <- imputed[[2]]
+    imputed <- imputed[[1]]
+    #end result for loop
+    #NA breaks it
+    if(is.na(imputed)){next}
+    
+    if(dim(imputed)!= dim(inputDF[i:as.numeric(i+lengthMiss-1),c("impLon","impLat")])){
+      cat(paste0("dimension mismatch \n"))
+      mismatch <- mismatch + 1
[…]

### 2018-04-27 implementingBipolar/imputeMidpoint.R

@@ -1,26 +1,0 @@
-imputeMidpointPalmius <- function(tMissStart = tMissStart,
-                                  tMissEnd = tMissEnd,
-                                  locMissStart = locMissStart,
-                                  locMissEnd   = locMissEnd,
-                                  delta_t = delta_t,
-                                  midpoint = midpoint,
-                                  ...){
-  
-  #time to midpoint from begin while travelling at 80kmh
-  t_prime<- raster::pointDistance(midpoint,
-                                  locMissStart,
-                                  lonlat = T)/22.23
-  if( delta_t <= 2* t_prime){
-    # evenly spaced points between missing two ends
-    cat("Simple interpolation \n")
-    imp <- linearInterpolate(locMissStart,locMissEnd,delta_t)
-  }else{
-    # goes to middle
-    cat("Midpoint wait interpolation \n")
-    imp1 <- linearInterpolate(locMissStart,midpoint_lon,t_prime/2)
-    imp2 <- linearInterpolate(midpoint,midpoint,delta_t-t_prime)
-    imp3 <- linearInterpolate(midpoint,locMissEnd,t_prime/2)
-    imp <- rbind(imp1,imp2,imp3)
-  }
-  return(imp)
-}

### 2018-04-27 manuscript/tableCreation.R

@@ -1,0 +1,28 @@
+# results df
+
+resultsDf <- data.frame( mean5min = c(82,43,269,426),
+                         median5min = c(4,0,0,0),
+                         meanexpDist5min = c(NA,NA,414,NA),
+                         medianexpDist5min = c(NA,NA,0,NA),
+                         mean1hr = c(345,497,908,1502),
+                         median1hr = c(6,4,0,0),
+                         meanexpDist1hour = c(NA,NA,916, NA),
+                         medianexpDis1hour = c(NA,NA,0, NA),
+                         mean1day = c(9273,NA,5757,14266),
+                         median1day = c(12,NA,0,1288),
+                         meanexpDist1Day = c(NA,NA,5757, NA),
+                         medianexpDis1Day = c(NA,NA,0, NA))
+
+rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive")
+colnames(resultsDf) <- c("Mean","Median","Expected Mean","Expected Median",
+                         "Mean","Median","Expected Mean","Expected Median",
+                         "Mean","Median","Expected Mean","Expected Median")
+
+# median 12 meters mean 9273
+library(knitr)
+library(kableExtra)
+options(knitr.kable.NA = '-')
+kable(resultsDf, "html") %>%
+  kable_styling("striped") %>%
+  add_header_above(c(" " = 1, "Five Minutes" = 4, "One Hour" = 4,"One Day" = 4))
+

### 2018-04-27 ownAlgo/Clusters.R

@@ -21,6 +21,7 @@ rm(data)
 
 # analyse time spent in one area
 
+# to do : use accuracy better e.g. take its value to modify distlim
 test2 <- clusterFunc(data = test,
                      timeLim = 200,
                      distLim = 100,
@@ -35,6 +36,11 @@ toClust <- test2 %>% filter( isPause == 1 & duration >= 300) %>%
 # merge all clusters within 400 meters of each other  
 clusterMap <- mergePoints(toClust, d = 400)
 
+clusterMap %>%
+  leaflet() %>%
+  addTiles() %>% 
+  addCircles(lng = ~lon, lat = ~lat)
+
 # all measurements binned into a pause cluster
 test3<- binMeasurements(test,clusterMap)
 
@@ -48,7 +54,12 @@ routeMap <- test3 %>% mutate(timeDay = as.Date(time)) %>%
   filter( (nextPauseClust != prevPausClust)) %>% 
   arrange(desc(n))
 
-checkPath(test3, start = 33, end = 32,clusterMap)
+checkPath(test3, start = 44, end = 32,clusterMap)
+
+mergePointsRoute(data= test3, start = 44, end = 32, d = 150, maxAcc = 100) %>% 
+  leaflet() %>% 
+  addTiles() %>% 
+  addCircles(lng = ~lon, lat = ~lat)
 
 #clusterPaths
 routePoints<- extractAllPaths(test3,routeMap)
@@ -87,8 +98,7 @@ pal <- colorFactor(c("navy", "red"), domain = c(0, 1))
 allPoints %>% 
   leaflet() %>% 
   addTiles() %>% 
-  addCircleMarkers(lng =~lon, lat = ~lat, color = ~pal(type),radius = ~ifelse(type == 0, 6, 10))
-  addCircleMarkers()
+  addCircleMarkers(lng =~lon, lat = ~lat, color = ~pal(type),radius = ~ifelse(type == 0, 6, 10), label = ~as.character(clust))
 
   library(leaflet.extras)
   routePoints%>% 
@@ -96,3 +106,34 @@ allPoints %>%
     addTiles() %>%
     addHeatmap(lng = ~lon, lat = ~lat, intensity = ~visits)
   
+  # SINGLE PATH
+  start <- 32
+  end <- 44
+  tripRaw <- fin %>% 
+    filter((nextPauseClust == start &  prevPausClust == end)|
+             (nextPauseClust == end &  prevPausClust == start))
+  
[…]

### 2018-04-27 ownAlgo/Results.R

@@ -1,0 +1,115 @@
+# Results bit:
+
+#Sampling frequency in herz for march
+
+fin %>%as_tbl_time(time) %>%
+  filter_time(~"2017-03") %>% 
+  thicken("day") %>% 
+  group_by(time_day) %>% 
+  summarise(total = n()/(24*60*60)) %>%
+  pull(total) %>% hist()
+
+# baseline models
+
+# MODEL: always previous
+
+## accuracy of the entire log:
+fin %>% mutate(lagClust = lag(clust)) %>%
+  summarise( prev = mean(clust == lagClust, na.rm= T))
+
+## mean deviation distance
+fin %>% mutate(lagClust = lag(clust)) %>% select(clust, lagClust) %>% 
+  left_join(allPoints,by="clust") %>% 
+  left_join(allPoints,by=c("lagClust" = "clust"),suffix = c(".clust",".predC")) %>% 
+  na.omit() %>% 
+  filter(clust != lagClust) %>% 
+  mutate( dist = raster::pointDistance(matrix(c(lon.clust,lat.clust),ncol = 2),
+                                                matrix(c(lon.predC,lat.predC),ncol = 2),
+                                                longlat = T)) %>% 
+  summarise(distMedian = median(dist),
+            distMean = mean(dist))
+
+# model: dailyMode
+
+Mode <- function(x) {
+  ux <- unique(x)
+  ux[which.max(tabulate(match(x, ux)))]
+}
+
+dailyMode<- fin %>%
+  thicken('day') %>%
+  group_by(time_day) %>% 
+  summarise(modeClust = Mode(clust))
+
+left_join(fin %>% thicken("day"),dailyMode, by="time_day") %>% 
+  select(clust,modeClust) %>% 
+  summarise( prev = mean(clust == modeClust, na.rm= T))
+
+left_join(fin %>% thicken("day"),dailyMode, by="time_day") %>% 
+  select(clust,modeClust) %>% 
+  left_join(allPoints,by="clust") %>% 
+  left_join(allPoints,by=c("modeClust" = "clust"),suffix = c(".clust",".predC")) %>% 
+  na.omit() %>% 
+  filter(clust != modeClust) %>% 
+  mutate( dist = raster::pointDistance(matrix(c(lon.clust,lat.clust),ncol = 2),
+                                       matrix(c(lon.predC,lat.predC),ncol = 2),
+                                       longlat = T)) %>% 
+  summarise(distMedian = median(dist),
+            distMean = mean(dist))
+
+
+# Check results
[…]

### 2018-04-27 ownAlgo/extractMissing.R

@@ -1,4 +1,6 @@
 # extract missing values\
+# unfinished, using the nnFeatureFunc to do this
+
 
 library(padr)
 library(tidyr)
@@ -18,3 +20,13 @@ data2 %>% top_n(250) %>%
 
 
 # turn the above into a function
+
+
+# take the time and the timestamp, with NA's
+
+train <- data  %>%
+  slice(remove_ind) %>%
+  select(time, timestampMs)
+  
+  
+  

### 2018-04-27 ownAlgo/functions/clusterFunc.R

@@ -1,0 +1,61 @@
+# cluster extraction funcion
+
+clusterFunc <- function(data, timeLim, accuracyLim,distLim,minPause){
+  
+  c<- data %>% filter(accuracy <accuracyLim) %>% 
+    mutate( prevLon = lag(lon),
+            prevLat = lag(lat),
+            nextLon = lead(lon),
+            nextLat = lead(lat),
+            nextDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                             matrix(c(nextLon,nextLat),ncol = 2),
+                                             longlat = T),
+            prevDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                             matrix(c(prevLon,prevLat),ncol = 2),
+                                             longlat = T),
+            nextMeas = as.numeric(timestampMs-lead(timestampMs)),
+            prevMeas = as.numeric(lag(timestampMs)-timestampMs),
+            isPause = case_when(prevMeas <= timeLim &nextMeas <= timeLim & nextDist <= distLim~1,
+                                TRUE ~0)
+    ) %>%
+    group_by(run = {run = rle(isPause); rep(seq_along(run$lengths), run$lengths)}) %>%
+    summarize(isPause = mean(isPause),
+              meanLon = mean(lon),
+              meanLat = mean(lat),
+              meanAcc = mean(accuracy),
+              t0  = min(timestampMs),
+              t1  = max(timestampMs),
+              t0TS = min(time),
+              t1TS = max(time),
+              duration = t1-t0,
+              minLon = min(lon),
+              minLonsLat = lat[which.min(lon)],
+              maxLon = max(lon),
+              maxLonsLat = lat[which.max(lon)],
+              minLat = min(lat),
+              minLatsLon = lon[which.min(lat)],
+              maxLat = max(lat),
+              maxLatsLon = lon[which.max(lat)],
+              distMeanNorth = raster::pointDistance(matrix(c(meanLon,meanLat),ncol = 2),
+                                                    matrix(c(maxLon,maxLonsLat),ncol = 2),
+                                                    longlat = T),
+              distMeanSouth = raster::pointDistance(matrix(c(meanLon,meanLat),ncol = 2),
+                                                    matrix(c(minLon,minLonsLat),ncol = 2),
+                                                    longlat = T),
+              distMeanWest = raster::pointDistance(matrix(c(meanLon,meanLat),ncol = 2),
+                                                    matrix(c(minLatsLon,minLat),ncol = 2),
[…]

### 2018-04-27 ownAlgo/functions/extractAllPathsFunc.R

@@ -1,0 +1,58 @@
+# Extract all paths & bin
+
+# bin measurements
+
+# to do: code all measurements into bins.
+
+# be clever about pause bins & data drift
+
+# write up stuff & send in one week
+extractAllPaths<- function(test3,routeMap, d = 150, maxAcc = 100, minVisit = 2){
+  allRoute <- list()
+  allRouteClusts <- list()
+  for(i in 1:nrow(routeMap)){
+    
+    start <- routeMap[[i,1]]
+    end <- routeMap[[i,2]]
+    route <- paste0(
+      min(start,end),"-",
+      max(start,end))
+    cat(paste0("Running ",route," which is ",i," out of ", nrow(routeMap),"\n"))
+    if((route %in% allRoute) |routeMap[[i,3]]==1){
+      cat("skipped...\n")
+      next}
+    rPoints<- mergePointsRoute(data= test3, start = start, end = end, d = d, maxAcc = maxAcc, minVisit = minVisit) %>%
+      mutate(route = route)
+    allRouteClusts[[i]] <- rPoints
+    allRoute[[i]] <- route
+  }
+  allRouteClusts <- bind_rows(allRouteClusts)
+  return(allRouteClusts)
+}
+
+
+#not doen
+extractAllPaths2<- function(toClust, d = 150, maxAcc = 100){
+  
+  #BIN BY CLUSTER and then work on clusters
+  allRoute <- list()
+  allRouteClusts <- list()
+  for(i in 1:nrow(routeMap)){
+    
+    start <- routeMap[[i,1]]
+    end <- routeMap[[i,2]]
+    route <- paste0(
+      min(start,end),"-",
+      max(start,end))
+    cat(paste0("Running ",route," which is ",i," out of ", nrow(routeMap),"\n"))
+    if((route %in% allRoute) |routeMap[[i,3]]==1){
+      cat("skipped...\n")
+      next}
+    rPoints<- mergePointsRoute(data= test3, start = start, end = end, d = d, maxAcc = maxAcc, minVisit = minVisit) %>%
+      mutate(route = route)
+    allRouteClusts[[i]] <- rPoints
+    allRoute[[i]] <- route
+  }
+  allRouteClusts <- bind_rows(allRouteClusts)
+  return(allRouteClusts)
+}

### 2018-04-27 ownAlgo/functions/featureFunc.R

@@ -1,0 +1,76 @@
+#Calculates features for the ML model
+
+featureFunc <- function(data, timeLim = 200, distLim = 100,minPause = 120, accuracyLim = 50){
+  
+  #calculate wether it is within accuracy
+  # add other metrics
+  data2<- data %>%
+    arrange(timestampMs) %>% mutate(
+      withinAcc = case_when(accuracy >= distClustMin ~1,
+                            TRUE ~ 0),
+      prevLon = lag(lon),
+      prevLat = lag(lat),
+      nextLon = lead(lon),
+      nextLat = lead(lat),
+      nextDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                       matrix(c(nextLon,nextLat),ncol = 2),
+                                       longlat = T),
+      prevDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                       matrix(c(prevLon,prevLat),ncol = 2),
+                                       longlat = T),
+      nextMeas = as.numeric(lead(timestampMs)-timestampMs),
+      prevMeas = as.numeric(timestampMs-lag(timestampMs)),
+      isPause = case_when(prevMeas <= timeLim &nextMeas <= timeLim & nextDist <= distLim~1,
+                          TRUE ~0),
+      pauseClust = case_when(isPause == 1 & withinAcc == 1~ as.numeric(clust),
+                             TRUE ~NA_real_),
+      nextPauseClust = pauseClust,
+      prevPausClust = pauseClust
+    ) %>% 
+    fill(nextPauseClust,.direction ="up") %>%
+    fill(prevPausClust,.direction ="down")
+  
+  return(data2)
+}
+
+featureFunc2 <- function(data,pauseExtracted,distCoef=2){
+pauseStart<- pauseExtracted %>% filter(isPause == 1) %>%
+select(t0TS,t1TS,meanAcc,distMeanEast,distMeanWest,distMeanSouth,distMeanNorth) %>%
+filter(!(distMeanEast > distCoef*meanAcc |
+distMeanWest >distCoef*meanAcc |
+distMeanNorth >distCoef*meanAcc |
+distMeanSouth >distCoef*meanAcc)) %>%
+select(t0TS,t1TS) %>% pull(t0TS)
[…]

### 2018-04-27 ownAlgo/functions/getRemoveIndexFunc.R

@@ -1,0 +1,42 @@
+
+getRemoveIndex <- function(fin,thickenParam="5 min",seed=1234){
+  set.seed(seed)
+  removed<- fin%>%
+    as_tbl_time(time) %>% 
+    filter_time(~"2017-03") %>% 
+    thicken(thickenParam,colname = "removed") %>%
+    distinct(removed) %>% 
+    sample_frac(.25)  %>% 
+    pull(removed) %>%
+    as.character()
+  
+  index <- fin%>%    as_tbl_time(time) %>% 
+    thicken(thickenParam,colname = "removed")%>% pull(removed) %>% as.character()
+
+  remove_ind <- which(index %in% removed)
+  cat(head(as.character(removed)))
+  
+return(list(remove_ind,removed))
+  
+}
+
+getRemoveIndexDay <- function(fin,thickenParam="day",seed=1234){
+  set.seed(seed)
+  removed<- fin%>%
+    as_tbl_time(time) %>% 
+    filter_time(~"2017-03") %>% 
+    thicken(thickenParam,colname = "removed") %>%
+    distinct(removed) %>% 
+    sample_frac(.25)  %>% 
+    pull(removed) %>%
+    as.Date() %>% as.character()
+  
+  index <- fin%>%    as_tbl_time(time) %>% 
+    thicken(thickenParam,colname = "removed")%>% pull(removed) %>% as.Date()%>% as.character()
+  
+  remove_ind <- which(index %in% removed)
+  cat(head(as.character(removed)))
+  
+  return(list(remove_ind,removed))
+  
+}

### 2018-04-27 ownAlgo/functions/nnFeatureFunc.R

@@ -1,0 +1,85 @@
+# extract features for the neural nets
+getTestTrain <- function(data, remove_ind,nClust = 1222){
+  seconds_in_day <-  24*60*60
+  
+  train <- data[-remove_ind,]
+  test <- data[remove_ind,]
+  
+  # issue: train is right, test isn't!
+  # changed it, check to see if it works!
+  train<- train %>% 
+    select(time,timestampMs,clust) %>% 
+    mutate( day  = strftime(time,"%u"),
+            month = strftime(time,"%m"),
+            secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
+            sinTime = sin(2*pi*secMidnight/seconds_in_day),
+            cosTime = cos(2*pi*secMidnight/seconds_in_day),
+            nextMeas = as.numeric(lead(timestampMs)-timestampMs),
+            prevMeas = as.numeric(timestampMs-lag(timestampMs)),
+            lagClust = lag(clust),
+            leadClust = lead(clust),
+            timestampLast = as.numeric(time),
+            timestampNext = as.numeric(time),
+            clustNext = clust,
+            clustPrev = clust)
+  
+  test<- test %>% 
+    select(time,timestampMs,clust) %>% 
+    mutate( day  = strftime(time,"%u"),
+            month = strftime(time,"%m"),
+            secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
+            sinTime = sin(2*pi*secMidnight/seconds_in_day),
+            cosTime = cos(2*pi*secMidnight/seconds_in_day),
+            nextMeas = NA,
+            prevMeas = NA,
+            lagClust = NA,
+            leadClust = NA,
+            timestampLast = NA,
+            timestampNext = NA,
+            clustNext = NA,
+            clustPrev = NA)
+
+  all <- rbind(train,test) %>%
+    arrange(timestampMs) %>% 
+    fill(timestampLast,clustPrev, .direction = "down") %>% 
+    fill(timestampNext,clustNext, .direction = "up") %>%
+    mutate( prevMeas = timestampMs - timestampLast,
+            nextMeas = timestampNext - timestampMs,
+            lagClust = clustPrev,
+            leadClust = clustNext) %>% 
+    mutate(nextMeas = scales::rescale(nextMeas),
+           prevMeas = scales::rescale(prevMeas))
+  
+  test <- all %>% slice(remove_ind)
+  train <- all %>% slice(-remove_ind)
+  
[…]

### 2018-04-27 ownAlgo/functions/nnResults.R

@@ -1,0 +1,303 @@
+#neural nets march test results
+# to be run after neuralNets.R 
+
+remove_ind_5min<- getRemoveIndex(fin,"5 min")
+
+# missing: the features have to include the fact that they are missing, so they have to be recalcuated
+# the extracted ones are wrong!!
+
+testTrain5min<- getTestTrain(fin,remove_ind_5min[[1]])
+
+# get baseline distance measure
+testTrain5min[[1]] %>%
+  select(clust,lagClust) %>% 
+  left_join(allPoints,by="clust") %>% 
+  left_join(allPoints,by=c("lagClust" = "clust"),suffix = c(".clust",".predC")) %>% 
+  na.omit() %>% 
+  mutate( dist = raster::pointDistance(matrix(c(lon.clust,lat.clust),ncol = 2),
+                                       matrix(c(lon.predC,lat.predC),ncol = 2),
+                                       longlat = T)) %>% 
+  summary()
+  
+  
+dataList5min<- getModelInput(test = testTrain5min[[1]], train = testTrain5min[[2]])
+
+x_train <- dataList5min[[1]]
+x_test  <- dataList5min[[2]]
+y_train <- dataList5min[[3]]
+y_test  <- dataList5min[[4]]
+
+remove_ind_5min <- remove_ind_5min[[1]]
+
+model <- keras_model_sequential() 
+model %>% 
+  layer_dense(units = 250, activation = 'relu', input_shape = ncol(x_test)) %>%
+  layer_dropout(rate = 0.4) %>%
+  layer_dense(units = 120, activation = 'relu') %>%
+  layer_dropout(rate = 0.3) %>%
+  layer_dense(units = nClust, activation = 'softmax')
+
+summary(model)
+
+model %>% compile(
+  loss = 'categorical_crossentropy',
+  optimizer = optimizer_rmsprop(),
+  metrics = c('accuracy')
+)
+
+history <- model %>% fit(
+  as.matrix(x_train), as.matrix(y_train), 
+  epochs = 10, batch_size = 200, 
+  validation_split = 0.2
+)
+#plot(history)
+
+model %>% evaluate(as.matrix(x_test),as.matrix(y_test)) #88 percent accuracy
+
+# distance matrix of all clusters
+distMatrix<- geosphere::distm(allPoints[,c("lon","lat")])
+
+t<- fin %>% slice(remove_ind_5min)
+
[…]

### 2018-04-27 ownAlgo/modifiedSTS.R

@@ -1,0 +1,83 @@
+# Master data Prep
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+
+#load required packages
+library(dplyr)
+library(leaflet)
+
+#read the data in
+Sys.time()
+data<- readRDS("/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/data/boaz/myLocationHistory09012018.rds")
+
+home<- c(5.113919,52.10421)
+# filter to measurements within 300 km from home in Utrecht
+
+test <- data %>% mutate( distanceHome = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                                              home,
+                                                              longlat = T),
+                         timestampMs = as.numeric(time))%>%
+  filter(distanceHome < 300000) %>% 
+  select(time,lon,lat,accuracy, timestampMs,distanceHome) %>%
+  arrange(timestampMs)
+
+rm(data)
+
+# load up custom functions
+file.sources <- list.files(c("functions"), 
+                           pattern="*.R$", full.names=TRUE, 
+                           ignore.case=TRUE)
+sapply(file.sources,source,.GlobalEnv)
+
+# this extracts pauses 
+test2 <- clusterFunc(data = test,
+                     timeLim = 300,
+                     distLim = 50,
+                     minPause = 100,
+                     accuracyLim = 250)
+
+# Extract the pause locations
+toClust <- test2 %>% filter( isPause == 1  & !(distMeanEast > 2*meanAcc |
+                                                 distMeanWest >2*meanAcc |
+                                                 distMeanNorth >2*meanAcc |
+                                                 distMeanSouth >2*meanAcc)) %>%
+  select(meanLon,meanLat) %>% as.matrix()
+
+colnames(toClust) <- c("lon","lat")
+
+# merge all clusters within 300 meters of each other  
+clusterMap <- mergePoints(toClust, d = 300)
+
+# all measurements binned into a pause cluster
+test3<- binMeasurements(test,clusterMap)
+
+# extract additional features
+test3 <- featureFunc2(test3,test2)
+# all points method of using rasters for computational help
+
+pathClust<- gridFunc(test3)
+
[…]

### 2018-04-27 ownAlgo/neuralNets.R

@@ -5,7 +5,7 @@ library(keras)
 library(tibbletime)
 library(dplyr)
 data2 <- fin
-data2<- readRDS('nnData.rds')
+#data2<- readRDS('nnData.rds')
 # reshape
 seconds_in_day <-  24*60*60
 
@@ -16,16 +16,17 @@ d <- data2 %>%
           secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
           sinTime = sin(2*pi*secMidnight/seconds_in_day),
           cosTime = cos(2*pi*secMidnight/seconds_in_day)) %>% 
-  select(time,clust,nextDist,prevDist,nextMeas,prevMeas,day,sinTime,cosTime,month) %>% 
-  mutate(prevDist = scales::rescale(prevDist),
-         nextDist = scales::rescale(nextDist),
-         nextMeas = scales::rescale(nextMeas),
+  select(time,clust,nextMeas,prevMeas,day,sinTime,cosTime,month) %>% 
+  mutate(nextMeas = scales::rescale(nextMeas),
          prevMeas = scales::rescale(prevMeas),
          lagClust = lag(clust),
          leadClust = lead(clust))
 
 d <- d %>% na.omit()# remove first and previous lag difference
 
+# add feature: distance of previous measurement to all other clusters
+dClustM<- distm(allPoints[,c("lon","lat")])
+dClustMr<- apply(X = dClustM,MARGIN = 1,FUN = scales::rescale)
 # If random sample
 ## 75% of the sample size
 smp_size <- floor(0.75 * nrow(d))
@@ -33,40 +34,59 @@ train_ind <- sample(seq_len(nrow(d)), size = smp_size)
 
 train <- d[train_ind,]
 test <- d[-train_ind,]
-#atlernative
 
+#atlernative
 train <- d %>% as_tbl_time(time)%>%
-  filter_time(("start"~ "2017-03-15"))
+  filter_time(("start"~ "2017-03-12"))
 
 train <- rbind(train, d %>% as_tbl_time(time)%>%
-                 filter_time("2017-03-17" ~ "end"))
+                 filter_time("2017-03-14" ~ "end"))
 
 test <- d %>% as_tbl_time(time)%>%
-  filter_time((~ "2017-03-16"))
-
+  filter_time((~ "2017-03-13"))
 # alternative end
-nClust<- (length(unique(fin$clust))+1)
 
[…]

### 2018-04-27 ownAlgo/nnLSTM.R

@@ -1,0 +1,121 @@
+# my own neural net lstm
+
+# need to look into this
+# got the data massaged into the right list (ish)
+# but the lstm doesn't work
+
+# constants
+data_dim <- 2
+timesteps <- 2
+num_classes <- 10
+
+test<- data.frame(clust = rep(1:10,10), index = 1:100, feature= runif(100))
+
+nClust <- 10+1
+x_train <- array(unlist(test), dim = c(100, timesteps, data_dim))
+
+smp_size <- floor(0.90 * nrow(test))
+train_ind <- sample(seq_len(nrow(test)), size = smp_size)
+train_ind_prev <- train_ind-1
+train_ind_next <- train_ind+1
+
+# create list and then to array
+selectedRandom <- 4
+
+listMat <- list()
+i <- 1
+for( selectedRandom in train_ind){
+  listMat[[i]]<- test[c(selectedRandom-1,selectedRandom+1),] %>% as.matrix()
+  i <- i+1
+}
+# doesn't work! but input can be a list too. 
+a<- array(as.numeric(unlist(listMat)), dim=c(2, 3, 90))
+
+
+# define and compile model
+# expected input data shape: (batch_size, timesteps, data_dim)
+model <- keras_model_sequential() 
+model %>% 
+  layer_lstm(units = 250, return_sequences = TRUE, input_shape = c(timesteps, data_dim)) %>% 
+  layer_lstm(units = 32) %>% # return a single vector dimension 32
+  layer_dense(units = nClust, activation = 'softmax') %>% 
+  compile(
+    loss = 'categorical_crossentropy',
+    optimizer = 'rmsprop',
+    metrics = c('accuracy')
+  )
+
+# generate training data
+
+# prepare NL data for ML
+d <- fin %>%
+  mutate( day  = strftime(time,"%u"),
+          month = strftime(time,"%m"),
+          secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
+          sinTime = sin(2*pi*secMidnight/seconds_in_day),
+          cosTime = cos(2*pi*secMidnight/seconds_in_day)) %>% 
+  select(time,clust,nextMeas,prevMeas,day,sinTime,cosTime,month) %>% 
+  mutate(nextMeas = scales::rescale(nextMeas),
+         prevMeas = scales::rescale(prevMeas)
+         )
+
[…]

### 2018-04-27 ownAlgo/objectiveResults.R

@@ -1,0 +1,218 @@
+# objective data comparison
+
+# read in data
+Sys.time()# 14:46:00
+
+dir()
+library(readr)
+objLog<- read_csv2("objectiveLog.csv")
+objKey<- read_csv2("objectiveKey.csv") %>% select(Location,Cluster)
+
+objKey<- left_join(objKey,allPoints, by = c("Cluster" = "clust")) %>% select(Location,Cluster,lon,lat)
+
+strftime(paste0(objLog$Date," ",objLog$Time),"%d-%m-%Y %H:%M:%S")
+
+
+
+objLog <- objLog %>%
+  mutate( time = as.POSIXct(paste(Date, Time), format="%d-%m-%Y %H:%M:%S")) %>% 
+  as_tbl_time(time)
+
+objLog <- left_join(objLog,objKey)
+
+# compare MYMETHOD to the objective data
+Sys.time()# 15:02:03
+
+test<- objLog %>% mutate(timestampMs = as.numeric(time), clust = Cluster)
+seconds_in_day <-  24*60*60
+
+test<- test %>% 
+  select(time,timestampMs,clust) %>% 
+  mutate( day  = strftime(time,"%u"),
+          month = strftime(time,"%m"),
+          secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
+          sinTime = sin(2*pi*secMidnight/seconds_in_day),
+          cosTime = cos(2*pi*secMidnight/seconds_in_day),
+          nextMeas = NA,
+          prevMeas = NA,
+          lagClust = NA,
+          leadClust = NA,
+          timestampLast = NA,
+          timestampNext = NA,
+          clustNext = NA,
+          clustPrev = NA,
+          isMissing = 1)
+
+
+train<- fin %>% 
+  select(time,timestampMs,clust) %>% 
+  mutate( day  = strftime(time,"%u"),
+          month = strftime(time,"%m"),
+          secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
+          sinTime = sin(2*pi*secMidnight/seconds_in_day),
+          cosTime = cos(2*pi*secMidnight/seconds_in_day),
+          nextMeas = as.numeric(lead(timestampMs)-timestampMs),
+          prevMeas = as.numeric(timestampMs-lag(timestampMs)),
+          lagClust = lag(clust),
+          leadClust = lead(clust),
[…]

### 2018-04-27 ownAlgo/rastering.R

@@ -1,0 +1,31 @@
+# rasterize
+
+gridFunc <- function(test3, minAccuracy = 150, cs = 0.5, d = 300){
+  library(raster)
+  toClust <- test3 %>% filter(is.na(isPause) & accuracy <=minAccuracy) %>%
+    select(lon,lat) %>% 
+    as.matrix()
+  
+  xy <- SpatialPointsDataFrame(
+    toClust, data.frame(ID=seq(1:nrow(toClust))),
+    proj4string=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))
+  cat("In Spatial Data frame Object \n")
+  
+  bb <- bbox(xy)
+  myPointsGrid <- GridTopology(cellcentre.offset =  bb[, 1] + (cs/2), cellsize = c(1, 1), cells.dim = ceiling(diff(t(bb))/cs))
+  myPointsGrid <- SpatialGrid(myPointsGrid, proj4string = "+proj=longlat +ellps=WGS84 +datum=WGS84")
+  res<- cbind(toClust,grid = as.vector(over(xy,myPointsGrid)))
+  cat(paste0("Put into ",length(unique(res[,"grid"]))," grids"))
+  
+  allRoute <- list()
+  gridBins<- unique(res[,3])
+  i <- 1
+  for(grid in gridBins){
+    cat(paste0(Sys.time()," doing grid ", i, "\n total rows ", nrow(res[which(res[,3] == grid),1:2])),"\n")
+    allRoute[[i]]<- mergePoints(res[which(res[,3] == grid),1:2], d = d)
+    i <- i+1
+  }
+  tBins<- bind_rows(allRoute)
+  return(tBins)  
+}
+

### 2018-04-27 ownAlgo/resultsBinning.R

@@ -1,0 +1,77 @@
+# results for evalutating binning
+# read in allpoints and end results of binning if not already loaded
+
+library(ggplot2)
+library(ggthemes)
+
+# calculate percentage within accuracy
+fin %>%
+  summarise(mean(accuracy >= distClustMin))
+
+# calculate minimum deviance
+fin %>%
+  select(distClustMin) %>% summary(d)
+
+#generate plots
+fin %>% 
+  ggplot(aes(x = accuracy, y = distClustMin))+
+  geom_point()+
+  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0,
+              na.rm = FALSE, show.legend = NA, colour = "grey")+
+  theme_tufte()+
+  ggtitle("Deviance vs Accuracy",subtitle = "Within 300km Utrecht")+
+  xlab("Accuracy")+ylab("Deviance")
+
+# calculate percentage March
+fin %>%
+  as_tbl_time(time) %>% 
+  filter_time(~"2017-03") %>% 
+  summarise(mean(accuracy >= distClustMin))
+
+# calculate deviance March
+fin %>%
+  as_tbl_time(time) %>% 
+  filter_time(~"2017-03") %>% 
+  select(distClustMin) %>% summary(d)
+
+fin %>%
+  as_tbl_time(time) %>% 
+  filter_time(~"2017-03") %>% 
+  ggplot(aes(x = accuracy, y = distClustMin))+
+  geom_point()+
+  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0,
+              na.rm = FALSE, show.legend = NA, colour = "grey")+
+  theme_tufte()+
+  ggtitle("Deviance vs Accuracy",subtitle = "March 2017")+
+  xlab("Accuracy")+ylab("Deviance")
+
+# Create a plot compared to Palmius
+plotBin<- left_join(fin %>% as_tbl_time(time) %>% filter_time(~"2017-03-01"),allPoints,by="clust", suffix=c(".measured",".cluster"))
+
+plotBin <- left_join(plotBin %>% thicken("5 min"),
+          downsampled,
+          by =c("time_5_min" = "time_5min"))
+
+plotBin %>% 
+leaflet() %>% 
+  addCircles(lng = ~jitter(lon.measured), lat = ~jitter(lat.measured), color = "#18206F") %>% 
[…]

### 2018-04-27 ownAlgo/resultsIan.R

@@ -1,0 +1,142 @@
+# full Ian results 
+directory <- "/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/harvard/Example"
+setwd(directory)
+
+
+#from finite
+data <- fin %>% as_tbl_time(time) %>% filter_time(~"2017-03")%>% select(timestampMs,time,lat,lon,accuracy)
+
+# Load needed functions
+source("/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/thesis/scripts/downSample.R")
+source("/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/thesis/scripts/pauseFlight.R")
+file.sources = list.files(pattern="*.R",path = "gpsmobility/R/")
+setwd("gpsmobility/R/")
+sapply(file.sources,source,.GlobalEnv)
+
+downsampledIan <- downSampleMean(data,interval = '300 sec') #downsamples and adds XY coordinates
+
+
+# remove data for results
+
+# 5 minute bits
+
+d2<- downsampledIan %>%
+  mutate( time = as.POSIXct(timestampMs, origin="1970-01-01")) %>% 
+  pull(time)
+
+remove_ind <- which(as.character(d2) %in% remove_ind_5min[[2]])
+
+downsampled5minRemoved<- downsampledIan
+downsampled5minRemoved[remove_ind,c("lon","lat","x_v","y_v")] <- NA
+downsampled5minRemoved[remove_ind,c("code")] <- 4
+
+#params
+minpausedur=120
+minpausedist=50
+
+mobmatmiss<- pauseFlight(downsampled5minRemoved, r = sqrt(300), w = mean(data$accuracy)) # extracts pauses and flights using algo
+mobmatmiss <- mobmatmiss[-nrow(mobmatmiss),] #clean a bit
+
+# Guessing Pauses
+#with example data first day
+mobmat = GuessPause(mobmatmiss,mindur=minpausedur,r=minpausedist)
+
+obj=InitializeParams(mobmat)
+spread_pars=c(10,1)
+wtype <- "TL"
+out3 <- SimulateMobilityGaps(mobmat,obj,wtype,spread_pars) # it is returning NA's only for
+
+evalDf<- evalIan(out3,downsampledIan = downsampledIan)
+
+# 5 minute thing
+
[…]

### 2018-04-27 ownAlgo/testBinning.R

@@ -1,0 +1,192 @@
+# test path & pause clustering
+# this is a draft script created while coding.
+
+library(dplyr)
+library(leaflet)
+library(ggplot2)
+
+data<- readRDS("/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/data/boaz/myLocationHistory09012018.rds")
+
+home<- c(5.113919,52.10421)
+#extract measurements within 300 km from home in Utrecht
+test <- data %>% mutate( distanceHome = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
+                                                              home,
+                                                              longlat = T),
+                         timestampMs = as.numeric(time))%>%
+  filter(distanceHome < 300000) %>% 
+  select(time,lon,lat,accuracy, timestampMs,distanceHome) %>%
+  arrange(timestampMs)
+
+rm(data)
+
+# analyse time spent in one area
+
+# load up custom functions
+file.sources <- list.files(c("functions"), 
+                          pattern="*.R$", full.names=TRUE, 
+                          ignore.case=TRUE)
+sapply(file.sources,source,.GlobalEnv)
+
+# this extracts pauses 
+# to do : use accuracy better e.g. take its value to modify distlim
+test2 <- clusterFunc(data = test,
+                     timeLim = 300,
+                     distLim = 50,
+                     minPause = 100,
+                     accuracyLim = 250)
+
+# Extract the pause locations
+toClust <- test2 %>% filter( isPause == 1  & !(distMeanEast > 2*meanAcc |
+                                                                 distMeanWest >2*meanAcc |
+                                                                 distMeanNorth >2*meanAcc |
+                                                                 distMeanSouth >2*meanAcc)) %>%
+  select(meanLon,meanLat) %>% as.matrix()
+
+colnames(toClust) <- c("lon","lat")
+# merge all clusters within 400 meters of each other  
+clusterMap <- mergePoints(toClust, d = 300)
+
+# all measurements binned into a pause cluster
[…]

### 2018-04-27 scripts/downSample.R

@@ -9,9 +9,9 @@ downSampleMean <-
     library(tibbletime)
     library(padr)
     
-    # first summarise via interval
-    data1 <- data %>%
-      select(timestampMs,time,lat,lon,accuracy) %>% 
+    #  summarise via interval and add coordinates
+    data1 <- cbind(data,as.data.frame(LatLong2XY(data$lat, data$lon)))%>%
+      select(timestampMs,time,lat,lon,accuracy,x_v,y_v) %>% 
       as_tbl_time(time) %>% 
       arrange(time) %>% 
       mutate(timestampMs = as.numeric(timestampMs)) %>% 
@@ -19,15 +19,16 @@ downSampleMean <-
       group_by(time2) %>%
       summarise(lon = mean(lon),# summarising function
                 lat = mean(lat),
-                code = 1)
-    
-    # then add XY coords
-    data2<- cbind(data1,as.data.frame(LatLong2XY(data1$lat, data1$lon))) %>% 
+                x_v = mean(x_v),
+                y_v = mean(y_v),
+                code = 1) %>% 
       pad() %>%
       mutate(code = case_when(is.na(x_v) == T ~4,
                               is.na(x_v) == F ~1),
              timestampMs = as.numeric(time2)) %>% 
       select(code,timestampMs,lon,lat,x_v,y_v)
-    return(data2)
+    
+    
+    return(data1)
   }
 

### 2018-04-27 manuscript/manuscript.Rmd
added text

@@
 How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010].
-Therefore it is unsurprising that policymakers and social scientists in numerous fields are interested in human mobility measures.
+Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures.
 The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip].
-Economists study geographic labour mobility extensively [e.g.
-@tatsiramos] and the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission].
-Researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017].
-Indeed, within psychology when studying behavioural differences within individuals it has been often been neglected that individuals vary not only over time but also over space.
+In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g.
+@villanueva].
+Moreover, economists study geographic labour mobility extensively [e.g.
+@tatsiramos].
+In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission].
+The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes.
+Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017].
+Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough.
+When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space.
 To fully understand behaviour we must understand how behaviour can vary accross environments.
 
-Despite their importance, a large part of studies on human mobility are conducted with pen-and-paper travel diaries.
-Pen-and-paper diares have well-known methodological flaws.
-For instance, they are expensive to collect and present a high burden to respondents.
-This limits the duration of data collection and the span of the log accordingly.
-Another issue is that they rely on self-reported measures.
-It has been reported that short trips are frequently under-reported  [@wolf_impact_2003] and the self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].
+Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries.
+Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting.
+The burdensomeness stems from the fact that participans must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness.
+This makes the data expensive to collect and limits the extent of data collection in practice.
+In addition, the frequent reporting duties of the participant may bias the participants behaviour.
+Participant forgetfulness also limits the accuracy of self reported measures.
+There is clear evidence that participants are systematically biased when self-reporting mobility measures.
+For instance, participants underreport the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017].
 These obstacles can be overcome by using objective data on human mobility.
-Such data  can now be obtained using the Global Positioning System (GPS).
-GPS uses the distance between a device and several satellites to determine location.
-GPS measurements can be used to infer a vast range of socioeconomic and behavioural measures, including where the individual lives, how much time he or she spends at home and where (and how) they travel.
+
+Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones.
+Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times.
+@barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour.
+Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016.
+Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers.
+Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times.
+These smartphones can collect and store hundreds of location measurements a day.
+For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017].
+Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017].
+Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millians of individuals spanning several years at little cost and without a significant burden of the participants.
+
+This paper wishes to achieve four objectives: First, we have argued that understanding human mobility is important.
+Secondly, we argued that social scientists should leverage data logs from smartphones to study human mobility, instead of relying on out-dated pen-and-paper questionnaires.
+Now we will explore the practical difficulties in using smartphone location logs.
+Finally we will introduce Personal Map Matched Imputation (PPMI), a method for surmounting these difficulties.
+We will compare PPMI to existing methods in the literature.
+
+# Background
+
+Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements.
+A GPS sensor uses the distance between a device and several satellites to determine the location of the device.
+Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
+In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone triangulation.
 
@@
 We call resulting logs *specialised logs*.
-@barnett_inferring_2016 point out several methodological issues with these studies.
-Collecting specialised logs is costly and places a high burden on participants.
-Besides, introducing a new device to the participant's life may bias their behaviour.
-Due to these drawbacks specialised logs usually span a short amount of time.
+
 @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
@@
 Participants tend not to like these tracking apps because they drain the battery of the device.
+
 Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) .
@@
 Android phones collect location information through a variety of methods.
-Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation.
-Different methods are used because of computational and battery constraints.
-GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
+
+*removed somethign here*
+
  In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location.
@@
 @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.
-
-This paper wishes to achieve four objectives: First, we have argued that understanding human mobility is important.
-Secondly, we argued that social scientists should leverage data logs from smartphones to study human mobility, instead of relying on out-dated pen-and-paper questionnaires.
-Third, explained the practical difficulties in using smartphone location logs.
-In the following sections we would like to introduce Personal Map Matched Imputation (PPMI), a method for surmounting these difficulties.
-We will compare PPMI to existing methods in the literature.
 

### 2018-04-27 manuscript/manuscript.Rmd
updated methods

@@
 Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
-In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone triangulation.
+In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation.
+Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values.
+The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017].
+In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.
 
-However, not all GPS measurements are made equal.
-It is useful to classify GPS logs into different categories depending on their origin because different types of logs pose different methodological problems.
-We argue that social science researchers should explore using automatically collected smartphone GPS logs.
-In most studies participants receive a specialised GPS devices to track their movement.
-We call resulting logs *specialised logs*.
+Researchers often develop custom-made tracking applications which participants are instructed to download on their phone.
+Alternatively participants are given a phone to use for a given period of time with the custom-made tracking app pre-installed.
+We call location logs resulting from these custom-made apps *custom logs*.
+The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor with which they are made.
+The disadvantage with this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation.
+Participants may dislike tracking apps because they view them as more intrusive and these apps regularly drain the battery of the device [@harari_using_2016].
+Moreover, researchers have distribute this application among research participants and convince them not to turn it off.
 
-@barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*).
-The problem with this is that researchers have to develop a custom-made app, distribute it among research participants and enforce participation.
-Participants tend not to like these tracking apps because they drain the battery of the device.
-
-Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) .
-For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017].
-Often, secondary logs span several years.
-By law, secondary logs are accessible to users for free [@commission_protecting_2017].
-Yet, secondary logs also present methodological challenges.
-They were created for non-academic purposes under engineering constraints ( these constraints will be detailed subsequently).
-These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate.
+We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*) .
+The advantages are clear: repositories such as Google's Location History contains information on millions of users spanning years [@location_history_timeline_2017], participants can share the data by the click of a button and there can be no behavioural changes due to participation in the study as the participant share past data.
+The disadvantage is that researchers have no control over the tracking parameters, often resulting in logs with sparse and inaccurate measurements.
 Hence, two important challenges are dealing with missing data and measurement noise.
-
-| Feature        | Specialised Logs           | Custom logs  | Secondary logs |
-| ------------- |:-------------:| -----:| -----:|
-| Sampling Frequency | Regular | Regular |  Irregular |
-| Sampling Frequency (Hz) | 0.1>      | Between 0.1 and 0.01| <0.01 |
-| Typical Duration | 7 days    |    7-14 days | Years |
-| Typical Accuracy Variance | Low    |    Medium | High |
 
@@
 include_graphics("img/missingBoaz5minExample.png")
+#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.
+
 ```
@@
 Importantly, spatiotemporal data measurements are autocorrelated in both time and space.
-This means that common methods, such as mean imputation, are unsuitable.
+This means that best practices with other types of data, such as mean imputation, are unsuitable.
 For example, imagine an individual who splits almost all her time between work and home.
 Suppose she spends a small amount of time commuting between the two along a circular path.
-Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.
-She has never and will never be there!
+Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there.
 Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].
@@
 Another methodological problem is related to the noise in the measurements that are collected.
-The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers.
-Android phones collect location information through a variety of methods.
-
-*removed somethign here*
-
+The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors.
 In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location.
@@
 Unfortunately for researchers, this is where people in the developed world spend most of their time.
-
-Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle.
-@palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
-\@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location.
-Inaccurate measures are often followed by more accurate measures.
-Most notably, low accuracy often (but not always) is  associated with movement (\@ref(fig:accuracyPlot2)).
-Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.
-There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.
+Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location.
+Most notably, low accuracy is often (but not always) associated with movement (\@ref(fig:accuracyPlot2)).
 
@@
 The red line shows the path with all measurements included.
-",out.width = '100%'}
+@palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+Inaccurate measures are often followed by more accurate measures.
+There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
 include_graphics("img/journeyTillMiddayBoaz.png")
@@
 ```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day.
-All journeys are marked with a red line.
 The first peak corresponds to the first journey from the user's home to a gym around 8am.
@@
 Both journeys can be seen in Figure 1.
-All other journeys are not shown in Figure 1.
 The large jump between journey 5 and 6 is measurement error.
 The lower chart shows the accuracy over the course of the day.
-The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
+The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.
+Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.",out.width = '100%'}
 include_graphics("img/accuracyLocShift.png")
@@
 But if there is noise, the coordinates will vary even though the individual is not moving.
-This is known as data drift (REFERENCE).
 If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
-This issue is also visualised in Figure 1.
+This issue is also visualised in Figure X.
 The problem is further complicated because missing data and noisy measurements are related.
@@
 
+In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs.
+We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simualted as missing and their imputed  counterparts.
+Finally we will calculate time spent at home as a function of the imputation method.
+
 # Related Work
 
-This section aims to explore related work with regard to missing data in GPS logs.
-We will argue three points.
-First, there is a lack of established practices to follow with respect to missing data in human mobility logs.
-Second, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs.
-Third, we discuss in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**].
-
-How have researchers dealt with missing data in GPS-based human mobility logs thus far?
+How have researchers dealt with missing data in human mobility logs thus far?
 Unfortunately there is no golden standard in how to deal with this type of missing data.
 Researchers are generally vague about what practices they follow [@jankowska_framework_2015].
-Ostensibly this is because they are unaware of possible solutions.
-Often researchers simply downsample and remove missing observations or use some sort of rule-based common sense imputations (e.g.
+This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons].
+The vaugeness is possibly also due to the fact that most researchers are unfamiliar with possible solutions.
+Most researchers simply downsample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g.
 @palmius_detecting_2017).
-The only principled approach that we know of that aims to solve the issue of missing data in GPS logs as they relate to human mobility is of @barnett_inferring_2016.
-Both of these methods will be detailed subsequently, after introducing exploring other spatiotemporal methods.
+The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016.
+We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.
 
@@
 These solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
-Moreover, these models often assume time and space invariance, assumptions which are violated in the case of human movement patterns.
-For this reason we do not consider SSMs to be useful for imputing missing data in this case.
+Moreover, these models often assume time and space invariance (location is not a direct function of time or space).
+These mathematical assumptions are violated in the case of human movement patterns.
+For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.
 
@@
 These often come from climate or geological research and rely on spatiotemporal autocorrelations.
-For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ].
+For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ].
 The authors illustrate their example using rainfall data from gauging stations across Australia.
@@
 
-On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**].
+On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu].
 Importantly, none of them worked with secondary logs.
@@
 Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm.
+In cases where the previous conditions are not met no values are imputed.
 
@@
 @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes.
-In the words of the authors, @barnett_inferring_2016 handle missing data by:
+In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses.
+Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move.
+Subsequently, the authors impute missing data by:
 
@@
 
-On the other hand, **Wu 2017** use what they call a Spatial Temporal Semantic Neural Network to predict future human movement.
-While the authors are concerned with prediction and not imputation, they devised an algorithm called the Spatial Temporal Semantic algorithm converts raw measurements to machine learning friendly discrete bins.
-This minimises the measurement error and paves the way for applying machine learning methods to human mobility problems.
+On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement.
+While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins.
+Working with high-frequency measurements, @wu's method downsamples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways.
+This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.
+
+In this section we have argued that there is a lack of established practices to follow with respect to missing data in human mobility logs.
+Moreover, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs.
+Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu].
 
@@
 Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
-Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude.
+For simplicity, we can assume that the x-axis is the longitude and the y-axis is the lattitude.
+Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the lattitude and longitude.
 In addition, let $a$ denote the estimated accuracy of the recorded data.
@@
 
-### MYMETHOD algorithm
+### Personalised Map Matching Imputation
 
-Our algorithm is designed to leverage the periodic nature of human movement along with the long span of the data to deal with measurement sparcity and inaccuracy.
+Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparcity and inaccuracy.
 
@@
 
-### MYMETHOD algorithm
+### Personalised Map Matching Imputation algorithm
 
@@
 
-1.
-Extract from measurements $D$ all pause locations and path clusters.
-Following **Wu 2017 (NOTE: ADD REFERENCE)** we use a modified spatial-temporal-semantic (STS) feature extraction algorithm to transform pause and path locations into machine learning friendly discrete location sequences.
-2. Assign each measurement $D$ to a unique discrete location. 
-3. Use neural networks to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.
+1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
+2. Binning: Assign each measurement $D$ to a unique discrete location. 
+3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.
 
-In this paper we have chosen relatively simple functions as component functions of MYMETHOD.
-It is clear each of them can be improved.
-We aim to illustrate not the efficacy of the components themselves but rather how they can work together to impute missing data.
+#### Map building
 
-#### Extract pause locations
-
-There are multiple ways of extracting pause locations in the literature (**examples & citations needed: see Wu 2017**).
-As far as we know there are only two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
-Both of these processes pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity.
-In particular, @barnett_inferring_2016 use a version of @rhee_human_2007's rectangular method to classify pauses and determine flight sub-sections.
-MYMETHOD uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
+Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences.
+There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;].
+We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
+Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity.
+For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
+However, there is no reason why more sophisticated methods cannot be used.
 
@@
 4.
-Let the points of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$.
+Let the measurements of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$.
 These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.
@@
 
-#### Extract paths
-
-To extract path clusters we apply a modified version of the spatial-temporal-semantic (STS) feature extraction presented in **Wu 2017 (NOTE: ADD REFERENCE)**.
-
-The key modification is that whilst **Wu 2017 (NOTE: ADD REFERENCE)** use map matching as a ground truth to assign noisy measurements into discrete bins along paths, we use the total mobility history of the individual to do so.
-This leverages the periodic nature of human movement as well as the long span of secondary logs.
-Therefore we consider each measurement at $D_x$ as a sample of $Path_{ab}$.
-In practice, for each observed path, MYMETHOD uses the following algorithm to estimate $Path_{ab}$:
+For all remaining points we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:
 
 1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
-2. Create a distance matrix for each point and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.
+2.
+Create a distance matrix for all remaining measurements $D_t \in F$ and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.
 
@@
 This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed.
-The reasoning for this is twofold: first, that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster.
-Secondly, due to data drift it can often appear that there is movement when there is none.The end result is a discretised map which contains pause and flight clusters based on the history of the individual.
+The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster.
+The end result is a discretised map which contains pause and flight clusters based on the history of the individual.
 
-Subsequently, all measurements $D$ were are then assigned to a discrete cluster.
-We used a simple assigning function, whereby the measurements where assigned to the cluster nearest to the measurement.
-We shall discuss the merits of choosing a more sophisticated assigning function subsequently.
+### Binning
 
-#### Neural network
+@wu's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads.
+In other words, in addition to the measured data they also use a geographic database that contains information about the area in which the individual is (e.g.
+where precisely the roads are), and sort measurements into bins based on both the measurement and the geographic data base.
+For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu's method will assume that the individual is on the road A.
 
-An artificial neural network was trained to predict the missing values. The input variables to the neural net are:
+The key modification in PMMI is that whilst @wu uses data from outside the persons location logs, we use the total location history of the individual to create a personalised map which will be then used to bin measurements.
+This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long.
+To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_ab$.
 
-1. The previous and subsequent observed cluster.
-2. The distance in time to the next & previous cluster.
-3. The time of the day encoded as a cyclical two-dimensional feature **NOTE: add reference**
-4. The day of the month encoded as a cyclical two-dimensional feature **NOTE: add reference**
+Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map.
+This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$.
+Again, for simplicity we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
+
+#### Classification
+
+At this point the objective of PMMI is to take all the information available about the mobility history of an individual and impute the missing value.
+In this implementation we trained an artificial neural network (ANN) to do so.
+The input variables to the ANN are:
+
+1. The previous and subsequent observed bin.
+2. The distance in time to the next & previous bin.
+3. The time of the day encoded as a cyclical two-dimensional feature.
+4. The day of the month as a binary class matrix.
 5. The month of the year as a binary class matrix.
+
+For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london].
+This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.
 
@@
 
-For a missing timepoint at $D_t$, the output of the model is a set of probability estimates associated with every location cluster.
+For a missing timepoint at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster.
 That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per cluster) associated with where the individual is.

### 2018-04-27 manuscript/manuscript.Rmd
finished methods

@@
 In this implementation we trained an artificial neural network (ANN) to do so.
+For more information on the precise architecture of the artificial neural network please consult the appendix.
 The input variables to the ANN are:
@@
 
-For more information on the precise architecture of the artificial neural network please consult the appendix.
-
 For a missing timepoint at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster.
-That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per cluster) associated with where the individual is.
+That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per bin) associated with where the individual is.
 
@@
 
-The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (table 1).
-
-```{r datadetailsTable, echo = F, eval = F}
-datadetails <- readRDS("tempdata/datadescriptives.rds")
-datadetails <- datadetails[,-7] #remove last column for now
-kable(datadetails,digits = 2,col.names = c("Log duration",
-                                           "Logged days",
-                                           "Observations",
-                                           "Missing days",
-                                           "Missing data",
-                                           "Mean Accuracy"),
-      caption = "Table with descriptives about the data sets used to build the imputation methods.
-Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.")
-```
-
-Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2]. **add packages**
-
-### Results & Evaluation Metrics
-
-The results will consist of multiple steps:
-
-1. A descriptive documentation of the characteristics of missingness in secondary logs.
-2. Evaluating the performance of the modified STS algorithm in extracting paths as well as two rival methods.
-3.
-Comparing the performance of MYMETHOD using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.
-
-#### Missingness
-
-With respect to the characteristics of missingness we will report how often data is missing and whether missingness displays temporal or spatial patterns.
-
-#### Binning evaluation
-
-In order to evalute the modified STS algorithm we will first visually evaluate the paths and pause locations.
-A visual evaluation of paths is an established way to heuristically check their accuracy (*add reference*).
-Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $d$.
-With respect to the deviation distance, we expect:
-
-1. A positive relationship between the deviation distance and the accuracy of each measurement.
-2. Roughly 67% of the deviation distances $d$ are within accuracy $a$ of each measurement.
-
-#### Imputation algorithm performance
-
-We treat missing data imputation as a classification problem, taking 25% of all measurements at random and eliminating them from the sample, recalculating features and computing classification accuracy.
-Moreover, we evaluate the performance of the imputation algorithm for the entire duration of the log with in comparison to a naive baseline models where for each missing period we predict the prevous non-missing cluster.
-
-To compare the performance of the different missing data imputation algorithms we will remove 25% of measured time intervals at random within a four week period.
-We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day.
-In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.
-
-To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location).
-For MYMETHOD's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally.
-This is necessary because Barnett's and Palmius' model estimate far fewer values than MYMETHOD.
-While Barnett's and Palmius' models estimate 12 measurements for each missing hour, MYMETHOD estimates as many measurements as there are in the sample.
-When individuals go to unfrequently travelled locations they tend to look use their phone's location services more, which leads to more measurements, hence more imputed values.
-Hence, if the weights are not used MYMETHOD is succeptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.
-
-In addition, other measures we will report for MYMETHOD are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance.
-The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster.
-For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.
-
-Finally, we will take objective real world data and compare it to predicted values.
-We will use information from the Dutch public transportation card, which provides users with timestamped data of when and where they board, change lines or leave public transportation.
-
-## Results
-
-### A descriptive documentation of the characteristics of missingness
-
-This is an example of a smartphone secondary log.
+The secondary location log used to train the imputation methods was collected between 2013 and 2017 on different Android devices from a single individual.
 About 54% of the data is missing for the entire duration of the log.
@@
 include_graphics("img/missingdayBoaz5min.png")
-
 ```
 
-### Modified STS performance in comparison to other binning methods
+For simplicity, we only used a time period when the indivdual was living in the Netherlands in the model.
+Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].
+**add packages** All the code is available on a public repository.
 
-Before we can evaluate the model, it is essential to evaluate how the noisy data has been restructured.
-Without this evaluation model accuracy measures could be misleading, as a good performance may simply be modelling noise.
-Therefore we will assess the performance of the modified STS binning method and compare it to alternatives.
+### Results & Evaluation Metrics
+
+The results will consist of multiple steps:
+
+1. Evaluating the performance of the map building and assigning functions.
+2.
+Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.
+
+#### Map building & binning evaluation
+
+Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been cleaned.
+Otherwise we run the risk of overfitting the model in the sense that we will measure the extent to which an imputation method can correctly impute measurement noise within $D_t$ instead of true location $G(t)$.
+
+In order to evalute the map building and binning we will first visually evaluate the paths and pause locations.
+A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (*add reference*).
+Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$.
+With respect to the deviation distance $\delta_{\text{dev}}$, we expect:
+
+1. A positive relationship between the deviation distance and the accuracy of each measurement.
+2. Roughly 67% of the deviation distances $\delta_{\text{dev}}$ are within accuracy $a$ of each measurement.
+
+#### Imputation algorithm performance
+
+We will compare the peformance of PMMI, @palmius_detecting_2017 and  @barnett_inferring_2016.
+In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value.
+To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period.
+We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day.
+In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.
+
+To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location).
+For PMMI's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally.
+This is necessary because the other two model estimate far fewer values than PMMI.
+While @barnett_inferring_2016's and @palmius_detecting_2017's models estimate 12 measurements for each missing hour, PMMI estimates as many measurements as there are in the log.
+When individuals go to unfrequently travelled locations they tend to use their phone's location services more than 100 more times, which leads to more measurements, and hence more imputed values.
+Hence, if the weights are not used PMMI is succeptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.
+
+In addition, other measures we will report for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance.
+The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster.
+For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.
+
+Finally, we will take objective real world data and compare it to predicted values.
+We will use information from the Dutch public transportation card as objective real world data.
+The Dutch public transportation card provides users with timestamped data of when and where they board, change lines or leave public transportation.
+
+## Results
+
+### Map building & binning evaluation
 
@@
 
-Importantly, we did not use the entire length of the log, but rather only the time periods when the individual was in the Netherlands. This was done to reduce computation time.
+The selection of these parameters was more-or-less heuristic.
 

### 2018-04-29 manuscript/apaManu.Rmd
APA manuscript

--- 2018-04-27 manuscript/manuscript.Rmd
@@
 ---
-title: "Manuscript"
-author: "Boaz Sobrado"
-output:
-  pdf_document: default
-  html_document: default
----
+title             : "Handling missing data in smartphone location logs"
+shorttitle        : "Missing Data in Smartphone Location Logs"
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
-library(ggthemes)
-library(ggplot2)
-library(knitr)
-```
+author: 
+  - name          : "Boaz Sobrado"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : ""
+    email         : "boaz@boazsobrado.com"
 
-# Abstract
-*Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health.
+affiliation:
+  - id            : "1"
+    institution   : "Utrecht University"
+
+author_note: |
+  Department of Methodology & Statistics
+
+  Submitted as a research report conforming to APA manuscript guidelines (6th edition).
+
+abstract: |
+  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health.
 The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data.
@@
 By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals.
-When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.*
+When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.
+
+keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
+wordcount         : "2297"
+
+bibliography      : 
+  - thesisRef.bib
+  - r-references.bib
+
+figsintext        : no
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : yes
+mask              : no
+
+class             : "man"
+output            : papaja::apa6_pdf
+
+---
+
+```{r load_packages, include = FALSE}
+library("papaja")
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
+library(ggthemes)
+library(ggplot2)
+library(knitr)
+library(kableExtra)
+
+```
 
@@
 In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g.
-@villanueva].
+@villanueva_association_2013].
 Moreover, economists study geographic labour mobility extensively [e.g.
@@
 The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes.
-Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017].
+Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017].
 Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough.
 When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space.
-To fully understand behaviour we must understand how behaviour can vary accross environments.
+To fully understand behaviour we must understand how behaviour can vary across environments.
 
@@
 Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting.
-The burdensomeness stems from the fact that participans must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness.
+The burdensomeness stems from the fact that participants must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness.
 This makes the data expensive to collect and limits the extent of data collection in practice.
@@
 There is clear evidence that participants are systematically biased when self-reporting mobility measures.
-For instance, participants underreport the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017].
+For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017].
 These obstacles can be overcome by using objective data on human mobility.
@@
 Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017].
-Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millians of individuals spanning several years at little cost and without a significant burden of the participants.
+Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.
 
@@
 A GPS sensor uses the distance between a device and several satellites to determine the location of the device.
-Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
+Although using a GPS sensor is the most accurate way to establish location on a smartphone, the GPS sensor is also the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
 In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation.
-Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values.
+Smartphone location logs contain measurements from all of these sources, usually in the form of time-stamped latitude, longitude and accuracy values.
 The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017].
@@
 
-In order to work with secondary logs, researchers need to be able to handle the data sparcity that leads to missing data.
+In order to work with secondary logs, researchers need to be able to handle the data sparsity that leads to missing data.
 Missing data is a pervasive issue in secondary logs as it can arise due to several reasons.
@@
 @harari_using_2016;@jankowska_framework_2015].
-In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparcity it is also evident.
+In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparsity it is also evident.
 
@@
 For this day there were several periods with no information.
-These points are filled with red and lie on the x-axis.",out.width = '100%'}
+These points lie on the x-axis.",out.width = '100%'}
 
-include_graphics("img/missingBoaz5minExample.png")
-#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.
-
+include_graphics("img/missin5minExample2018.png")
 ```
@@
 There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016].
-Importantly, spatiotemporal data measurements are autocorrelated in both time and space.
+Importantly, spatiotemporal data measurements are auto-correlated in both time and space.
 This means that best practices with other types of data, such as mean imputation, are unsuitable.
@@
 Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location.
-Most notably, low accuracy is often (but not always) associated with movement (\@ref(fig:accuracyPlot2)).
+Most notably, low accuracy is often (but not always) associated with movement (see Figure \@ref(fig:accuracyPlot2)).
 
@@
 The red line shows the path with all measurements included.
-@palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+In smartphone logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
 Inaccurate measures are often followed by more accurate measures.
-There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
+There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cellphone tower triangulation.",out.width = '100%'}
 include_graphics("img/journeyTillMiddayBoaz.png")
@@
 If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
-This issue is also visualised in Figure X.
+This issue is also visualised in Figure \@ref(fig:palmiusVme1).
 The problem is further complicated because missing data and noisy measurements are related.
@@
 In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs.
-We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simualted as missing and their imputed  counterparts.
+We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simulated as missing and their imputed  counterparts.
 Finally we will calculate time spent at home as a function of the imputation method.
@@
 Researchers are generally vague about what practices they follow [@jankowska_framework_2015].
-This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons].
-The vaugeness is possibly also due to the fact that most researchers are unfamiliar with possible solutions.
-Most researchers simply downsample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g.
+This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons_false-positive_2011].
+The vagueness is possibly also due to the fact that most researchers are unfamiliar with possible solutions.
+Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g.
 @palmius_detecting_2017).
@@
 Because SSMs are based on the Markov property, they cannot incorporate this information.
-In other words, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
+In other words, the estimated location $G(t)$ at time-point $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
 Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models.
@@
 In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods.
-These often come from climate or geological research and rely on spatiotemporal autocorrelations.
+These often come from climate or geological research and rely on spatiotemporal auto-correlations.
 For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ].
@@
 Nonetheless we will detail what they did as informative examples.
-@palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
+@palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
@@
 
-@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data.
+@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only principled approach to dealing with missing data in human mobility data.
 @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes.
@@
 While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins.
-Working with high-frequency measurements, @wu's method downsamples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways.
+Working with high-frequency measurements, @wu's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways.
 This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.
@@
 
-GPS measurements provide us with coordinates on the surface of the earth, which is ellipsoid shaped.
+Location measurements, such as those produced by GPS sensors, provide us with coordinates (latitude and longitude) on the surface of the earth, which is ellipsoid shaped.
 Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion.
@@
 
-Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidian plane.
+Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidean plane.
 Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
-For simplicity, we can assume that the x-axis is the longitude and the y-axis is the lattitude.
-Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the lattitude and longitude.
+For simplicity, we can assume that the x-axis is the longitude and the y-axis is the latitude.
+Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the latitude and longitude.
 In addition, let $a$ denote the estimated accuracy of the recorded data.
-Furthermore, $G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ...
+Furthermore, $G(t)$, $D$ and $a$ are indexed by time labeled by the countable set $t = t_1 < ...
 < t_{n+1}$.
@@
 
-### Personalised Map Matching Imputation
+## Personalised Map Matching Imputation
 
-Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparcity and inaccuracy.
+Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparsity and inaccuracy.
 
@@
 
-First, following @barnett_inferring_2016 we categorise all timepoints $t$ as either belonging to the set $P$ (pause) or set $F$ (flight).
-Conceptually pauses can be understood as periods of time where an individual spends singificant amount of continous time without moving.
+First, following @barnett_inferring_2016 we categorise all time-points $t$ as either belonging to the set $P$ (pause) or set $F$ (flight).
+Conceptually pauses can be understood as periods of time where an individual spends significant amount of continuous time without moving.
 Flights are the times where the individual is moving.
@@
 
-Let us define as $Flight^x_{ab}$ the set of all points belonging to a flight between $G(t_a)$ and $G(t_b)$ at timepoint $t_x$.
+Let us define as $Flight^x_{ab}$ the set of all points belonging to a flight between $G(t_a)$ and $G(t_b)$ at time-point $t_x$.
 $$Flight^x_{ab}= G(t_x)|t_x \in F =  \{G(t_{i+n}),...,G(t_j)\}$$
@@
 = Flight^{xn}_{ab}$.
-Then, we can define as $Path_{ab}$ the set of all flights between $G(t_a)$ and $G(t_b)$ at all timepoints.
+Then, we can define as $Path_{ab}$ the set of all flights between $G(t_a)$ and $G(t_b)$ at all time-points.
 For simplicity, we assume that $Path_{ab}=Path_{ba}$.
@@
 
-### Personalised Map Matching Imputation algorithm
+## Personalised Map Matching Imputation algorithm
 
@@
 
-1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
-2. Binning: Assign each measurement $D$ to a unique discrete location. 
-3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.
+1. *Map building*: Extract from measurements $D$ all pause location bins and path location bins to create a personalised map.
+2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
+3. *Imputing*: Use a classification method to predict missing measurements based on all the available information.
 
-#### Map building
+### Map building
 
 Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences.
-There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;].
+There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007;].
 We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
 Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity.
-For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
-However, there is no reason why more sophisticated methods cannot be used.
+This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
 
@@
 
-This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location.
-Each pause location is a cluster.
+This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.
 
-For all remaining points we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:
+For all remaining measurements we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:
 
@@
 2.
-Create a distance matrix for all remaining measurements $D_t \in F$ and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.
+Create a distance matrix for all remaining measurements $D_t \in F$ and hierarchically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.
 
-At this point all empirically observed path clusters and pause clusters are extracted.
-However, there may be some overlap between pause clusters and path clusters.
-Thus, the clusters are clustered again, such that the pause clusters retain priority.
-This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed.
+At this point all empirically observed path bins and pause bins are extracted.
+However, there may be some overlap between pause bins and path bins.
+Thus, the bins are clustered again, such that the pause bins retain priority.
+This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed.
 The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster.
-The end result is a discretised map which contains pause and flight clusters based on the history of the individual.
+The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.
 
@@
 
-The key modification in PMMI is that whilst @wu uses data from outside the persons location logs, we use the total location history of the individual to create a personalised map which will be then used to bin measurements.
+PMMI uses a similar logic, but without using any external geographic database.
+The key modification in PMMI is that whilst @wu uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map.
+This map is subsequently used to bin measurements.
 This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long.
-To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_ab$.
+To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.
 
 Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map.
-This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$.
-Again, for simplicity we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
+This includes previous measurements which were discarded from the map building exercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$.
+In this implementation we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
 
-#### Classification
+### Classification
 
@@
 
-1. The previous and subsequent observed bin.
+1. The previous and subsequent observed bin as a binary class matrix.
 2. The distance in time to the next & previous bin.
@@
 
-For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london].
+For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london_encoding_2016].
 This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.
+Moreover we scaled the non-binary values to occupy a range between 0 and 1 in order to ensure convergence.
 
-For a missing timepoint at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster.
-That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per bin) associated with where the individual is.
+For a missing time-point at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster.
+That is, for each missing time-point the model returns a vector of probability estimates (with one estimate per bin) associated with where the individual is.
 
-### Datasets & Analyses
+## Datasets & Analyses
 
@@
 About 54% of the data is missing for the entire duration of the log.
-This may be misleading as there are several long periods with no measurements whatsoever (see Figure \@ref(fig:longMeasurementsPerDay)).
+This may be misleading as there are several long periods with no measurements whatsoever.
 For days which were not entirely missing, approximately 22% of all five minute segments were missing.
@@
 As you can see, there are several long periods over the course of the log for which there are no measurements.
-Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.
 The median sampling frequency per day for non-missing days is around 0.006 Hz.
 
-```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data on February 15th 2017.
+```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data for the entire duration of the log.
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
@@
 include_graphics("img/missingdayBoaz5min.png")
+my_citations <- cite_r(file = "r-references.bib")
 ```
 
-For simplicity, we only used a time period when the indivdual was living in the Netherlands in the model.
-Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].
-**add packages** All the code is available on a public repository.
+For simplicity, we subsequently used a time period when the individual was living in the Netherlands.
+This leaves us with 156,000 measurements over a period of less than six months.
+Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages `r my_citations`.
+All the code is available on a public repository [@mygithub].
 
-### Results & Evaluation Metrics
+## Results & Evaluation Metrics
 
@@
 2.
-Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.
+Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data (public transportation time-stamps).
 
-#### Map building & binning evaluation
+### Map building & binning evaluation
 
 Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been cleaned.
-Otherwise we run the risk of overfitting the model in the sense that we will measure the extent to which an imputation method can correctly impute measurement noise within $D_t$ instead of true location $G(t)$.
+Otherwise we run the risk of over-fitting the model in the sense that we will measure the extent to which an imputation method can correctly impute measurement noise within $D_t$ instead of true location $G(t)$.
 
-In order to evalute the map building and binning we will first visually evaluate the paths and pause locations.
-A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (*add reference*).
+In order to evaluate the map building and binning we will first visually evaluate the paths and pause locations.
+A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (e.g.
+@lcm).
 Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$.
@@
 
-#### Imputation algorithm performance
+### Imputation algorithm performance
 
-We will compare the peformance of PMMI, @palmius_detecting_2017 and  @barnett_inferring_2016.
+We will compare the performance of PMMI, @palmius_detecting_2017 and @barnett_inferring_2016.
 In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value.
+The naive model will serve as a baseline model.
 To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period.
@@
 In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.
+
+For the @barnett_inferring_2016 and PMMI models we will use all the available data to train the models with the exception of the time periods being investigated.
+@palmius_detecting_2017's model does not require training.
 
@@
 While @barnett_inferring_2016's and @palmius_detecting_2017's models estimate 12 measurements for each missing hour, PMMI estimates as many measurements as there are in the log.
-When individuals go to unfrequently travelled locations they tend to use their phone's location services more than 100 more times, which leads to more measurements, and hence more imputed values.
-Hence, if the weights are not used PMMI is succeptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.
+When individuals go to infrequently travelled locations they tend to use their phone's location services more than 100 more times, which leads to more measurements, and hence more imputed values.
+Hence, if the weights are not used PMMI is susceptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.
 
-In addition, other measures we will report for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance.
-The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster.
-For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.
+In addition, other measures of interest for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*.
+The confidence is the probability with which the model predicts the most likely cluster.
+For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say it has high confidence in the prediction.
+Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster.
+For example, if the true location of an individual is bin A (bin A is 10 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 1 meter.
 
 Finally, we will take objective real world data and compare it to predicted values.
-We will use information from the Dutch public transportation card as objective real world data.
-The Dutch public transportation card provides users with timestamped data of when and where they board, change lines or leave public transportation.
+We will use information from the Dutch public transportation card.
+The Dutch public transportation service provides users with time-stamped data of when and where they board, change lines or leave public transportation.
+To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a time-stamped measurement is available.
+Then we will use each model to impute the location of the individual within that period.
 
-## Results
+# Results
 
-### Map building & binning evaluation
+## Map building & binning evaluation
 
@@
 
-The selection of these parameters was more-or-less heuristic.
+The selection of these parameters was more-or-less heuristically driven based on their on their ability to extract a meaningful personalised map.
+When selecting parameters there is a trade off is between bias and precision.
+This is because an increase in precision in the form of a higher for high density locations resolution comes at the expense of precision as assigning measurements to bins becomes more difficult.
+For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster.
+This is illustrated in Figure \@ref(fig:pauseLocUithof1).
 
-When selecting parameters there is a trade off is between resolution and accuracy.
-This is because an increase in resolution comes at the expense of accuracy.
-For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster.
-This is illustrated in Figures \@ref(fig:pauseLocUithof1) and \@ref(fig:pauseLocUithof2).
-
-```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
+```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof university campus using 150 meters (blue) and 400 meters (red) as clustering parameters.",out.width = '100%'}
 #fix width
-include_graphics("img/mapPauseRes1.png")
+include_graphics("img/distanceParam.png")
 ```
 
-```{r, pauseLocUithof2,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
-include_graphics("img/mapPauseRes2.png")
-```
+Map building results in a personalised map with pause and path clusters.
+An excerpt can be seen in Figure \@ref(fig:mapClusterMap).
+It is important to remind the reader that PMMI is map agnostic and uses no information from the map.
+Therefore, the close overlap with features on the map, such as pause bins at relevant buildings and transportation clusters, as well as the flight bins following roads and railway lines indicate a high degree of precision in personal map building.
+As expected, PMMI's path extraction yields greater accuracy for frequently occurring paths.
+For example, the frequently travelled Amsterdam-Utrecht railway line has been extracted almost perfectly, while the less frequently travelled Utrecht-Enschede line is far sparser.
 
-The end result of step 3 of MYMETHOD results in a personalised map with pause and path clusters.
-An excerpt can be seen in Figure \@ref(fig:mapClusterMap).
-As expected, MYMETHODs path extraction yields greater accuracy for frequently ocurring paths.
-It is important to remind the reader that despite the close overlap with features on the map,  such as roads and railway lines, MYMETHOD is map agnostic and uses no information from the map to recreate these features.
-
-```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map if an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
+```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
 #fix width
@@
 
-For the entire period when the individual is in the Netherlands, there is a mean deviance of 40 meters and a median deviance of 15 meters.
+For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters.
 Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected.
+Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}}$.
 
-For the narrower period of March 2017, figure \@ref(fig:devianceComp) illustrates the relationship between the deviance and accuracy.
-Approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected.
-The mean and median deviance are 38 and 14 meters respectively.
+For the narrower period of March 2017 approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected.
+The raw unweighted mean and median deviance are 38 and 14 meters respectively.
+If we weight them such that each 5 minute interval is weighted equally to ease comparison with the other two methods, we find a mean deviance of 35 meters with a median of 12.8.
 
-```{r, devianceComp,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
-#fix width
-# add figure with deviance comparisons three columns each method once
-```
+In comparison, Palmius' method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers.
+On the other hand, Barnett's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
+Barnett's deviance is necessarily higher than Palmius', as they down-sample temporally (like Palmius) and subsequently aggregate into pauses and linear flights.
 
-In comparison, Palmius' method has a median deviance of 3 meters, with a mean of 115 due to high deviance outliers.
-Over 90% of deviance values are within accuracy, which suggests that the model is binning noise.
-In Figure \@ref(fig:palmiusVme) the key difference is visible.
-The Palmius method is much more sensitive to inaccurate values, such as the one at the top right of the figure.
-
-On the other hand, Barnett's method has mean deviance of 343 meters and a median deviance of 8 meters.
-Barnett's deviance is neccessarily higher than Palmius', as they downsample and subsequently aggregate into pauses and linear flights.
+The key difference between temporal and spatial down-sampling is shown in Figure \@ref(fig:palmiusVme1).
+Temporal down-sampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods.
+Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the down-sampled values.
+Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each down-sampled interval.
+The fact that over 90% of deviance values are within accuracy (much higher than the expected theoretical 67%) confirms that temporal down-sampling is not sufficiently filtering out the noise.
 
@@
 #fix width
-include_graphics("img/palmiusVme.png")
+include_graphics("img/tempVsspace.png")
 ```
 
-```{r, palmiusVsMe2,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
-include_graphics("img/binComp.png")
-```
+### Imputation evaluation
 
-### Comparison to alternative models
+The @palmius_detecting_2017 model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests.
+@palmius_detecting_2017' model failed to impute a single value for the day tests.
+Similarly, @barnett_inferring_2016's method failed to impute over 11% of the missing values.
+PPMI made an imputation for all missing values.
+Table 1 shows the results of the distance metrics for each method.
 
-```{r,resulttable, echo = F}
-library(knitr)
-library(kableExtra)
+```{r,resulttable, echo = F, results='asis'}
+# results df
 
@@
                          median5min = c(4,0,0,0),
-                         meanexpDist5min = c(NA,NA,414,NA),
-                         medianexpDist5min = c(NA,NA,0,NA),
                          mean1hr = c(345,497,908,1502),
                          median1hr = c(6,4,0,0),
-                         meanexpDist1hour = c(NA,NA,916, NA),
-                         medianexpDis1hour = c(NA,NA,0, NA),
                          mean1day = c(9273,NA,5757,14266),
-                         median1day = c(12,NA,0,1288),
-                         meanexpDist1Day = c(NA,NA,5757, NA),
-                         medianexpDis1Day = c(NA,NA,0, NA))
+                         median1day = c(12,NA,0,1288)
+                         )
 
-rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive")
-colnames(resultsDf) <- c("Mean","Median","Expected Mean","Expected Median",
-                         "Mean","Median","Expected Mean","Expected Median",
-                         "Mean","Median","Expected Mean","Expected Median")
+rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "PPMI","Naive Baseline")
+colnames(resultsDf) <- c("Mean","Median",
+                         "Mean","Median",
+                         "Mean","Median")
 
-# median 12 meters mean 9273
-library(knitr)
-library(kableExtra)
+# median 7 meters mean 60 not imputed 282 NA's Ian all data
 options(knitr.kable.NA = '-')
-kable(resultsDf, "html") %>%
-  kable_styling("striped") %>%
-  add_header_above(c(" " = 1, "Five Minutes" = 4, "One Hour" = 4,"One Day" = 4))
-
+apa_table(resultsDf, col_spanners = list(`Five minutes` = c(2, 3), `One Hour` = c(4, 5),`One Day` = c(6,7)), placement = "htb", digits = 0, caption = "Distance in meters between the removed time period and the imputed value.")
 ```
 
-In terms of coverage, it is important to note that the Palmius model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests.
-Palmius' model failed to impute a single value for the day tests.
-Barnett's failed to impute 11% of the missing values.
-MYMETHOD made an imputation for all missing values.
-
-As for MYMETHOD's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
+In terms of PPMI's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
 As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively.
+The distance expectation values were very similar to the distance scores, albeit approximately 5% higher.
+Confidence scores ranged from 0.006 to 1.
 
@@
 
+Once we removed the measurements within the 5 minute period of each time-stamp, we employed the models to predict the location of the individual at the time-stamped period.
 The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters.
-On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a median distance of 5637.
-The expected distance has a mean of 6432 and a median of 1037.
+On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a mean distance of 5637.
+Again the expectation of the distance was quite similar to the distance with a mean of 6432 and a median of 1037 meters.
 
@@
 
-### Effect on aggregate measures
+### Example: effect on aggregate measures
 
-Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled.
-As we discussed earlier, aggregations without data processing can be highly biased.
-However, as an example we calculate time spent at home of the user for each day of the week in the month of February (\@ref(fig:aggrePlot)).
-Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as the amount of time spent at home.
+As an example we calculated the time spent at home of the user in the month of March (Figure \@ref(fig:aggrePlot)) without any model and with all three of the investigated methods.
 
-```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017.
-We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
-Missing values are filled in by the previous observed value.
-Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
-include_graphics("img/aggPlot.png")
+```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in March 2017.
+The raw values are estimated by downsampling temporally the lattitude and longitude for every 5 minute time period in the month.
+We used each method's own binning method and classified as at home if the downsampled measurement was within 250 meters from home.",out.width = '100%'}
+
+include_graphics("img/timeUse.png")
 ```
 
-## Discussion
+Interestingly, all three models suggest that the user spent approximately 60% of their time at home.
+However, without using any form of missing data imputation approximately 12% of the time is unaccounted for.
+Given that the amount of time spent at home has been found to be a reliable predictor of extroversion [@harari_using_2016] and the onset of depressive episodes in bipolar patients [@palmius_detecting_2017] obtaining an accurate value for mobility metrics like this is highly important to social scientists.
 
-Overall the MYMETHOD performed well, although not substantially better than the naive model.
+## Discussion & Conclusion
+
+Overall the PMMI performed better than the alternative models, particularly during longer missing periods and with objective data.
+However, it did not perform substantially better than the naive baseline model.
 In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs.
-Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation.
+Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone GPS logs.
 
-In addition to higher accuracy under the conditions of secondary logs, the advantages of MYMETHOD are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smoothen movement noise.
+In addition to higher accuracy under the conditions typical of secondary logs, the advantages of PPMI are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smooth movement noise.
 
-MYMETHOD's increased coverage and flexibility comes from its ability to make complex non-linear predictions.
+PPMI's increased coverage and flexibility comes from its ability to make complex non-linear predictions.
 For instance, in a given missing period it might make sense to predict that the individual is either at home, or at the office, or at a shop with equal probability.
-While MYMETHOD can make such an imputation, none of the alternative methods can do this.
+While PPMI can make such an imputation, none of the alternative methods can do this.
 Moreover, the ability to take the prediction probability values from the neural network also helps in dealing with uncertainty.
-A known-drawback of single imputation is that it takes an imputed value and treats it as observed (*RUBIN*).
+A known-drawback of single imputation is that it takes an imputed value and treats it as observed.
 Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods.
-With MYMETHOD it is possible to construct confidence intervals using the predicted probabilities of each estimate.
+With PPMI it is possible to model uncertainty using the predicted probabilities of each estimate.
+For instance, in the previous example, we could choose to only take estimates with a high degree of confidence, thus creating confidence intervals by adding and subtracting the amount of cases where the location of the individual is ambiguous.
 
-With respect to irregular sampling, alternative methods use temporally based downsampling in order to reduce noise.
+With respect to irregular sampling, alternative methods use temporally based down-sampling in order to reduce noise.
 This leads to deterioration in resolution not only over space, but also over time.
@@
 For instance, consider a case where there are two inaccurate measurements in movement at 12:00:01 and 12:04:59.
-Downsampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not.
-MYMETHOD instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations.
+Down-sampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not.
+PPMI instead down-samples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations, not just the few that happen within a single period.
 
-While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for paths.
-Hence all paths are calculated based on sensor input only.
-This may be feasible with high frequency regularly sampled short duration logs but creates noise with secondary logs.
+While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for non-pause locations.
+This may be feasible with high frequency, regularly sampled short duration logs but creates noise with secondary logs.
 Moreover, with secondary logs it is feasible to spatially "average out" multiple samples of the same path in order to recreate it in its entirety.
-For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (*report value here*) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
+For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (about 0.01 Hz) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
 
@@
 The golden standard would be to use at least one highly accurate professional grade GPS device with high sampling frequency to compare our data to.
-Until that is available, the use of public transport data and crossvalidation is just a substitute.
+Until that is available, the use of public transport data and cross-validation is just a substitute.
 
-Furthermore, MYMETHOD also has several limitations.
-The pause extracting function, the path clustering function, the classification model and the assignment function remain simplistic and could be improved.
-Indeed, our intention is not to develop an optimal classification model or any of the other components of MYMETHOD, but rather to show a new way in which secondary logs can be made useful by social scientists.
+Furthermore, PPMI can be further developed. The map building function, the assignment function and the classification model remain simplistic and could be improved.
 
-With regards to pause extraction, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location.
+In map building, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location.
 These factors are not taken into account in the pause extraction function.
 Improved methods would do well to do so.
+As for paths, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
+It does not take into account the length of the path as well as the average sampling frequency of the path.
+This is an issue because it can lead to bins to which data points are seldom assigned.
+For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements.
+This leads to difficulties in the classification part of the model, as infrequently observed clusters are hard for the model to predict.
 
-As for the path clustering function, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
-It does not take into account the length of the path as well as the average sampling frequency of the path.
-This is an issue because it can lead to clusters to which data points are seldomly assigned.
-For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements.
-This leads to difficulties in the machine learning part of the model, as unfrequently observed clusters are hard for the model to predict.
-
-Similarly, a limitation of the current assignment function is that it only takes into account the distance between the observed measurement and all the clusters when assigning each measurement to a cluster.
+The current assignment function simply assigns each measurement to the nearest bin.
 It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on.
@@
 By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC.
-An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and the ANN model.
-This could take the form of a Kalman filter.
-In this case the state side of the model would represent a probabilistic representation of where the individual could be based on what the ANN has extracted from the individuals movement history.
+An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and a model of the individuals movement history.
+In terms of state-space models like the Kalman filter the state equation would represent a probabilistic representation of where the individual could be based on the individuals entire movement history.
 The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.
 
-As for the simplicity of the neural network which was used to generate predictions.
-A more sophisticated recurrant neural network (RNN), or a long short-term memory recurrant neural network (LSTM) would likely perform significantly better.
+As for the simplicity of the classification method, the neural network which was used to generate predictions used no information on sequence patterns longer than the previous and next bin.
+A more sophisticated recurrent neural network (RNN), or a long short-term memory recurrent neural network (LSTM) would likely perform significantly better.
 
-## Conclusion
+That there is room for improvement is unsurprising given that it is still early days for using smartphone location measurements in social science.
+Nonetheless, the methodological advantages are clear: millions of individuals have a location log containing objective measurements spanning years which can be accessed for free.
+This is vastly superior to alternatives such questionnaires which rely on accurate self-reporting.
+Social science researchers must take advantage of regulatory changes in data portability regulations and put the vast wealth of data collected by commercial entities to public use.
 
-Advantages of MYMETHOD:
+\newpage
 
-1. Full imputation coverage.
-2. Probability of cluster as measure of uncertainty.
-3. Reduction of data drift through discretisation.
-4. Robust to irregular sampling.
-5. Few assumptions.
+# References
+```{r create_r-references}
+r_refs(file = "r-references.bib")
+```
 
-Disadvantages:
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
 
-1. Weak in unfrequently sampled locations. Perhaps use alternative methods there (e.g. Kalman filter)
-2. Some sacrifice of accuracy for consistency. See high vs low resolution pause maps.
-
-Further research:
-
-1. More sophisticated pause/path extracted algorithms.
-2. More sophisiticated ML models.
-3. Introducing explicit assumptions about paths.
-4. Run a study where individuals have a high frequency sampling high accuracy GPS receiver on them (test out of sample performance).
-5. Path extraction for train stuff.
-
-*To do: add info on paragraphs and sections/subsections.*
+<div id = "refs"></div>
+\endgroup

### 2018-04-29 manuscript/apaManu.tex
APA manuscript

\documentclass[man]{apa6}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

% Table formatting
\usepackage{longtable, booktabs}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}

\newenvironment{lltable}
  {\begin{landscape}\begin{center}\begin{ThreePartTable}}
  {\end{ThreePartTable}\end{center}\end{landscape}}

  \usepackage{ifthen} % Only add declarations when endfloat package is loaded
  \ifthenelse{\equal{\string man}{\string man}}{%
   \DeclareDelayedFloatFlavor{ThreePartTable}{table} % Make endfloat play with longtable
   % \DeclareDelayedFloatFlavor{ltable}{table} % Make endfloat play with lscape
   \DeclareDelayedFloatFlavor{lltable}{table} % Make endfloat play with lscape & longtable
  }{}%

% The following enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand\getlongtablewidth{%
 \begingroup
  \ifcsname LT@\roman{LT@tables}\endcsname
  \global\longtablewidth=0pt
  \renewcommand\LT@entry[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}%
  \@nameuse{LT@\roman{LT@tables}}%
  \fi
\endgroup}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            pdfauthor={},
            pdftitle={Handling missing data in smartphone location logs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setlength{\parindent}{0pt}
%\setlength{\parskip}{0pt plus 0pt minus 0pt}

\setlength{\emergencystretch}{3em}  % prevent overfull lines

% Manuscript styling
\captionsetup{font=singlespacing,justification=justified}
\usepackage{csquotes}
\usepackage{upgreek}

 % Line numbering
  \usepackage{lineno}
  \linenumbers

\usepackage{tikz} % Variable definition to generate author note

% fix for \tightlist problem in pandoc 1.14
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Essential manuscript parts
  \title{Handling missing data in smartphone location logs}

  \shorttitle{Missing Data in Smartphone Location Logs}

  \author{Boaz Sobrado\textsuperscript{1}}

  % \def\affdep{{""}}%
  % \def\affcity{{""}}%

  \affiliation{
    \vspace{0.5cm}
          \textsuperscript{1} Utrecht University  }

  \authornote{
    Department of Methodology \& Statistics

    Submitted as a research report conforming to APA manuscript guidelines
    (6th edition).

    Correspondence concerning this article should be addressed to Boaz
    Sobrado, . E-mail:
    \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
  }

  \abstract{Personal mobility, or how people move in their environment, is
associated with a vast range of behavioural traits and outcomes, such as
socioeconomic status, personality and mental health. The widespread
adoption of location-sensor equipped smartphones has generated a wealth
of objective personal mobility data. Nonetheless, smartphone collected
personal mobility data has remained underutilised in behavioural
research, partly due to the practical difficulties associated with
obtaining the data and partly because of the methodological complexity
associated with analysing it. Recent changes in European regulation have
made it easier for researchers to obtain this data, but the
methodological difficulties remain. The difficulty lies in that
smartphone location data is irregularly sampled, sparse and often
inaccurate. This results in a high proportion of missing data and
significant noise. In this paper we present a method called Personal Map
Matched Imputation (PPMI) to deal with missing data and noise in
smartphone location logs. The main innovation of PPMI is that it creates
a personalised spatial map for each individual based on all the
available data. In doing so PPMI leverages the regularity of human
mobility in order to smoothen noisy measurments and impute missing data
values. By simulating missing periods in real data we find that a simple
implementation of PPMI performs as well as existing methods for short (5
minute) missing intervals and substantially better for longer (1 day)
missing intervals. When imputing a subset of real missing data where
travel logs are available as a reference points, we find that PPMI
performs substantially better than existing models.}
  \keywords{Missing Data, Measurement Bias, GPS, Human Mobility \\

    \indent Word count: 2297
  }

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

\maketitle

\setcounter{secnumdepth}{0}

\section{Introduction}\label{introduction}

Why is human mobility important? How people move about in their
environment affects a wide range of outcomes, such as health, income and
social capital (Goodchild \& Janelle, 2010). Therefore it is
unsurprising that social scientists in numerous fields and even
policymakers are interested in human mobility measures. The most widely
administered personality questionnaires ask individuals to what extent
they agree with statements such as \enquote{I love large parties},
\enquote{I prefer going to the movies to watching videos at home} and
\enquote{I love to travel to places that I have never been before}
({\textbf{???}}). In economic research, the postal code of an
individual's home address is often used as a proxy for socioeconomic
status (e.g. Villanueva \& Aggarwal, 2013). Moreover, economists study
geographic labour mobility extensively (e.g. {\textbf{???}}). In fact,
the European Commission considers labour mobility important enough to
warrant its own directorate ({\textbf{???}}). The extensive use of
measures like these across different domains strongly suggests that
mobility metrics are linked to real world outcomes. Perhaps it is
unsurprising that behavioural researchers have found that mobility
measures can be used to predict academic performance (Wang, Harari, Hao,
Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz, \&
Odoms-Young, 2009) or even the onset of a depressive episode in bipolar
depression patients (Palmius et al., 2017). Indeed, there is an argument
to be made that perhaps psychologists have not been studying mobility
enough. When studying behavioural differences within individuals
behavioural scientists have often neglected the fact that individuals
vary not only over time but also over space. To fully understand
behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these
measures are obtained through the use of questionnaires (such as the
IPIP) or specifically through pen-and-paper travel diaries.
Questionnaires and travel diaries are have well-known methodological
flaws: they are burdensome to collect and rely on accurate
self-reporting. The burdensomeness stems from the fact that participants
must be explicitly asked to report on their movement patterns at
frequent intervals due to forgetfulness. This makes the data expensive
to collect and limits the extent of data collection in practice. In
addition, the frequent reporting duties of the participant may bias the
participants behaviour. Participant forgetfulness also limits the
accuracy of self reported measures. There is clear evidence that
participants are systematically biased when self-reporting mobility
measures. For instance, participants under-report the frequency of short
trips (Wolf, Oliveira, \& Thompson, 2003) and underestimate the duration
of regular commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
These obstacles can be overcome by using objective data on human
mobility.

Social scientists now have the unprecedented opportunity to easily
obtain objective data on human mobility from smartphones. Before
smartphones the only way to collect objective data on human mobility
involved giving participants an expensive professional-grade location
sensor and convincing them to take it with themselves at all times.
Barnett and Onnela (2016) points out that introducing a new device to
the participant's life may bias their behaviour. Moreover, collecting
data in such a way is costly, places a high burden on participants and
therefore the logs do not span a long time {[}Barnett and Onnela (2016).
Today millions of individuals carry smartphones with themselves every
day and do not need to be encouraged to do so by researchers.
Smartphones are equipped with a range of sensors that can be used to
track the location of the device at all times. These smartphones can
collect and store hundreds of location measurements a day. For instance,
Google Location History contains movement information on millions of
users, often spanning years (Location History, 2017). Moreover, recent
changes in EU regulations with regard to consumer data-portability
rights ensure that a willing participant should be able to easily share
this information with researchers at no cost to the either the
participant or the researchers (Commission, 2017). Taken together, this
means that researchers now have at their disposal the ability to easily
access the objective human mobility data of millions of individuals
spanning several years at little cost and without a significant burden
of the participants.

This paper wishes to achieve four objectives: First, we have argued that
understanding human mobility is important. Secondly, we argued that
social scientists should leverage data logs from smartphones to study
human mobility, instead of relying on out-dated pen-and-paper
questionnaires. Now we will explore the practical difficulties in using
smartphone location logs. Finally we will introduce Personal Map Matched
Imputation (PPMI), a method for surmounting these difficulties. We will
compare PPMI to existing methods in the literature.

\section{Background}\label{background}

Smartphone location measures are obtained primarily (but not
exclusively) by Global Positioning System (GPS) measurements. A GPS
sensor uses the distance between a device and several satellites to
determine the location of the device. Although using a GPS sensor is the
most accurate way to establish location on a smartphone, the GPS sensor
is also the most energy consuming sensor on most smartphones (M. Y. Chen
et al., 2006; LaMarca et al., 2005). In order to avoid battery depletion
and to overcome computational constraints smartphones also use
less-accurate heuristics such as WiFi access points and cellphone tower
triangulation. Smartphone location logs contain measurements from all of
these sources, usually in the form of time-stamped latitude, longitude
and accuracy values. The accuracy \(a\) of any given measurement is
given in meters such that it represents the radius of a 67\% confidence
circle (Location History, 2017). In other words, the true location of a
device should be within the radius \(a\) of the measurement 67\% of the
time.

Researchers often develop custom-made tracking applications which
participants are instructed to download on their phone. Alternatively
participants are given a phone to use for a given period of time with
the custom-made tracking app pre-installed. We call location logs
resulting from these custom-made apps \emph{custom logs}. The advantage
of custom logs is that the researchers can adjust tracking parameters,
such as the the frequency of measurements and the sensor with which they
are made. The disadvantage with this approach is that researchers have
to develop or adapt a custom-made tracking application (which is not
easy given hundreds of different types of smartphone models), distribute
it among research participants and enforce participation. Participants
may dislike tracking apps because they view them as more intrusive and
these apps regularly drain the battery of the device (G. M. Harari et
al., 2016). Moreover, researchers have distribute this application among
research participants and convince them not to turn it off.

We focus on another solution, which is to take advantage of existing
smartphone location logs (\emph{secondary logs}) . The advantages are
clear: repositories such as Google's Location History contains
information on millions of users spanning years (Location History,
2017), participants can share the data by the click of a button and
there can be no behavioural changes due to participation in the study as
the participant share past data. The disadvantage is that researchers
have no control over the tracking parameters, often resulting in logs
with sparse and inaccurate measurements. Hence, two important challenges
are dealing with missing data and measurement noise.

In order to work with secondary logs, researchers need to be able to
handle the data sparsity that leads to missing data. Missing data is a
pervasive issue in secondary logs as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. G. M.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015). In Figure
\ref{fig:longMeasurementsPerDay} shows that despite the long duration of
the log the sparsity it is also evident.

\begin{figure}
\includegraphics[width=1\linewidth]{img/missin5minExample2018} \caption{Example of missing data over the entire duration of a secondary log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points lie on the x-axis.}\label{fig:longMeasurementsPerDay}
\end{figure}

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
are auto-correlated in both time and space. This means that best
practices with other types of data, such as mean imputation, are
unsuitable. For example, imagine an individual who splits almost all her
time between work and home. Suppose she spends a small amount of time
commuting between the two along a circular path. Using mean imputation
to estimate her missing coordinates, we impute her to be at the midpoint
between home and work, even though she has never been there. Worryingly,
there is little transparency on how researchers deal with missing data
(Jankowska et al., 2015).

Another methodological problem is related to the noise in the
measurements that are collected. The accuracy of smartphone location
measurements is substantially lower than that of professional GPS
location trackers because smartphones often use less accurate sensors.
In professional GPS trackers less than 80\% of measurements fall within
10 meters of the true location. GPS measures are most inaccurate in
dense urban locations and indoors (S. Duncan et al., 2013; Schipperijn
et al., 2014). Unfortunately for researchers, this is where people in
the developed world spend most of their time. Figure
\ref{fig:accuracyPlot} shows how accuracy can vary as a function of user
behaviour, time and location. Most notably, low accuracy is often (but
not always) associated with movement (see Figure
\ref{fig:accuracyPlot2}).

\begin{figure}
\includegraphics[width=1\linewidth]{img/journeyTillMiddayBoaz} \caption{Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. In smartphone logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cellphone tower triangulation.}\label{fig:accuracyPlot}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.}\label{fig:accuracyPlot2}
\end{figure}

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distance between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! This issue is also visualised in Figure
\ref{fig:palmiusVme1}. The problem is further complicated because
missing data and noisy measurements are related. Methods used by
researchers to reduce noise, such as throwing out inaccurate
measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing
data and measurement error in secondary location logs. We will compare
PPMI to similar solutions in the literature by evaluating the distance
between points which were simulated as missing and their imputed
counterparts. Finally we will calculate time spent at home as a function
of the imputation method.

\section{Related Work}\label{related-work}

How have researchers dealt with missing data in human mobility logs thus
far? Unfortunately there is no golden standard in how to deal with this
type of missing data. Researchers are generally vague about what
practices they follow (Jankowska et al., 2015). This vagueness is
worrisome as it invites solutions which contain significant researcher
degrees of freedom (Simmons, Nelson, \& Simonsohn, 2011). The vagueness
is possibly also due to the fact that most researchers are unfamiliar
with possible solutions. Most researchers simply down-sample temporally
and remove missing observations or use some sort of rule-based common
sense imputations (e.g. Palmius et al. (2017)). The only principled
approach that we know of that aims to solve the issue of missing data in
location logs as they relate to human mobility is that of Barnett and
Onnela (2016). We will explore the methods of Barnett and Onnela (2016)
and Palmius et al. (2017) in detail subsequently, after introducing
exploring other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility
patterns does not imply there is not a vast literature on modelling
movement. The most widespread models are SSMs, therefore we shall detail
a few examples and subsequently argue that they are nonetheless unsuited
for long term human mobility logs. Ecologists have used SSMs to explain
how animals interact with their environment (Patterson, Thomas, Wilcox,
Ovaskainen, \& Matthiopoulos, 2008). These models can be quite complex.
Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. The extended Kalman filter is the de facto standard for GPS
navigation (Z. Chen \& Brown, 2013). The advantage of state space models
is that they are flexible, deal with measurement inaccuracy, include
information from different sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that
they ignore movement routines. For instance, humans tend to go to work
on weekdays and sleep at night. Because SSMs are based on the Markov
property, they cannot incorporate this information. In other words, the
estimated location \(G(t)\) at time-point \(t\) is often based only upon
measurements \(D_t\), \(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\).
Hierarchical structuring and conditioning on a larger context have been
suggested as ways to add periodicity to Markovian models. These
solutions are often computationally intractable or unfeasible (Sadilek
\& Krumm, 2016). Moreover, these models often assume time and space
invariance (location is not a direct function of time or space). These
mathematical assumptions are violated in the case of human movement
patterns. For this reason we do not consider existing SSMs to be useful
for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous
missing data imputation methods. These often come from climate or
geological research and rely on spatiotemporal auto-correlations. For
instance, the CUTOFF method estimates missing values by incorporating
similar observed temporal information from the value's nearest spatial
neighbors (Feng, Nowak, O'Neill, \& Welsh, 2014 ). The authors
illustrate their example using rainfall data from gauging stations
across Australia. Similarly, Z. Zhang et al. (2017) use a variety of
machine learning methods to impute missing values. The example provided
relates to underground water data. Generally these models assume fixed
measurement stations (such as rainfall gauging stations). For this
reason they cannot be easily applied to missing mobility tracks without
significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns ({\textbf{???}}; Barnett \&
Onnela, 2016; Palmius et al., 2017 ). Importantly, none of them worked
with secondary logs. Nonetheless we will detail what they did as
informative examples. Palmius et al. (2017) deal with the measurement
inaccuracy of \(D\) in custom logs by removing from the data set all
unique low-accuracy \(a\) data points that had
\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
sample the data to a sample rate of 12 per hour using a median filter.
Moreover, Palmius et al. (2017) explain:

\begin{quote}
\enquote{If the standard deviation of {[}\(D\){]} in both latitude and
longitude within a 1 h epoch was less than 0.01 km, then all samples
within the hour were set to the mean value of the recorded data,
otherwise a 5 min median filter window was applied to the recorded
latitude and longitude in the epoch}.
\end{quote}

Missing data was imputed using the mean of measurements close in time if
the participant was recorded within 500m of either end of a missing
section and the missing section had a length of \(\leq 2h\) or
\(\leq 12h\) after 9pm. In cases where the previous conditions are not
met no values are imputed.

Barnett and Onnela (2016) follow a different approach which is, to the
best of our knowledge, the only principled approach to dealing with
missing data in human mobility data. Barnett and Onnela (2016) work with
custom logs where location is measured for 2 minutes and subsequently
not measured for 10 minutes. In the words of the authors, Barnett and
Onnela (2016) handle missing data by first converting data to mobility
traces, which are defined as a sequence of flights and pauses. Flights
are segments of linear movements and pauses corresponding to periods of
time where a person does not move. Subsequently, the authors impute
missing data by:

\begin{quote}
\enquote{simulat{[}ing{]} flights and pauses over the period of
missingness where the direction, duration, and spatial length of each
flight, the fraction of flights versus the fraction of pauses, and the
duration of pauses are sampled from observed data.}
\end{quote}

This method can be extended to imputing the data based on temporally,
spatially or periodically close flights and pauses. In other words, for
a given missing period, the individual's mobility can be estimated based
on measured movements in that area, at that point in time or movements
in the last 24 hours (\emph{circadian proximity}).

On the other hand, ({\textbf{???}}) use what they call a Spatial
Temporal Semantic Neural Network (STS-NN) to predict future human
movement. While the authors are concerned with prediction and not
imputation, they devised a method called the Spatial Temporal Semantic
(STS) algorithm which converts raw measurements to machine learning
friendly discrete bins. Working with high-frequency measurements,
({\textbf{???}})'s method down-samples the raw data temporally and
map-matches the resulting bins to discrete points along pre-established
geographical features such as roads and highways. This minimises
measurement error and paves the way for applying machine learning
methods to human mobility problems.

In this section we have argued that there is a lack of established
practices to follow with respect to missing data in human mobility logs.
Moreover, extensively used spatiotemporal methods, such as state space
models (SSMs), are not well suited to deal with human mobility patterns
in secondary logs. Finally we discussed in detail three approaches which
deal explicitly with mobility patterns from custom or secondary logs
({\textbf{???}}; Barnett \& Onnela, 2016; Palmius et al., 2017 ).

\section{Methodology}\label{methodology}

\subsection{Notation}\label{notation}

Location measurements, such as those produced by GPS sensors, provide us
with coordinates (latitude and longitude) on the surface of the earth,
which is ellipsoid shaped. Projecting three dimensional measurements in
\(\mathbb{R}^3\) onto a two dimensional plane in \(\mathbb{R}^2\)
results in distortion. For clarity, when we use the term distance we
refer to the geodesic distances on an ellipsoid using the WGS84
ellipsoid parameters.

Subsequently let us simplify by assuming that a persons location is on
two-dimensional Euclidean plane. Let a person's true location on this
two-dimensional plane be \(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and
\(G_y(t)\) denote the location of the individual at time \(t\) on the
x-axis and y-axis respectively. For simplicity, we can assume that the
x-axis is the longitude and the y-axis is the latitude. Moreover, let
\(D \in \mathbb{R}^2\) be the recorded data containing the latitude and
longitude. In addition, let \(a\) denote the estimated accuracy of the
recorded data. Furthermore, \(G(t)\), \(D\) and \(a\) are indexed by
time labeled by the countable set \(t = t_1 < ... < t_{n+1}\). The
measure of accuracy \(a_t\) is given in meters such that it represents
the radius of a 67\% confidence circle. If \(D_t = \emptyset\) it is
considered \emph{missing} and it is not missing otherwise.

When several data sets are available from individuals living in
overlapping areas we can construct a \(t \times i\) matrix \(M\) where
the entry \(M(t,i)\) contains \(G(t)\) for the individual \(i\).

\subsection{Personalised Map Matching
Imputation}\label{personalised-map-matching-imputation}

Our algorithm is designed to leverage the periodic nature of human
movement along with the long span of secondary to deal with measurement
sparsity and inaccuracy.

\paragraph{Modelling assumptions}\label{modelling-assumptions}

First, following Barnett and Onnela (2016) we categorise all time-points
\(t\) as either belonging to the set \(P\) (pause) or set \(F\)
(flight). Conceptually pauses can be understood as periods of time where
an individual spends significant amount of continuous time without
moving. Flights are the times where the individual is moving. Let
\(t_a\) be a pause of length \(n\). \[t_a = t_i < ... < t_{i+n}\] Let
\(t_b\) be a pause of length \(m\) such that there is no temporal
overlap between \(t_a\) and \(t_b\):
\[t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j\] Then it follows that between
the two pauses there must be a flight indexed by \(t_x\) of length
\(j-i+n\). \[t_x =  t_{i+n} < ... < t_{j} |t_x \in F\] We define pause
locations \(G(t_a), G(t_b) | t_a, t_b \in P\) as locations where an
individual spends an extended amount of time in the same space
(e.g.~school, home, work, train station, barber shop, bar, gym).
Importantly, our model assumes period and cyclic human movement such
that there are many pauses \(t_{a1},t_{a2},...,t_{an}\) such that
\(G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})\). Moreover,it is possible for
\(G(t_a) = G(t_b)\) such that \(t_a \ne t_b\). For example, if the
individual leaves home for a run and returns home without stopping
anywhere else.

Let us define as \(Flight^x_{ab}\) the set of all points belonging to a
flight between \(G(t_a)\) and \(G(t_b)\) at time-point \(t_x\).
\[Flight^x_{ab}= G(t_x)|t_x \in F =  \{G(t_{i+n}),...,G(t_j)\}\] Again,
there are many flights \(t_{x1},t_{x2},...,t_{xn}\) such that
\(Flight^{x1}_{ab} = Flight^{x2}_{ab} = ... = Flight^{xn}_{ab}\). Then,
we can define as \(Path_{ab}\) the set of all flights between \(G(t_a)\)
and \(G(t_b)\) at all time-points. For simplicity, we assume that
\(Path_{ab}=Path_{ba}\).

In addition, we consider all measurements \(D(t)\) to be imperfect
measurements of \(G(t)\):

\[ G(t) = D(t) + \text{Measurement Error}  \]

\subsection{Personalised Map Matching Imputation
algorithm}\label{personalised-map-matching-imputation-algorithm}

Our algorithm performs the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{Map building}: Extract from measurements \(D\) all pause
  location bins and path location bins to create a personalised map.
\item
  \emph{Binning}: Assign each measurement \(D\) to a unique discrete
  location bin.
\item
  \emph{Imputing}: Use a classification method to predict missing
  measurements based on all the available information.
\end{enumerate}

\subsubsection{Map building}\label{map-building}

Following ({\textbf{???}})'s spatial-temporal-semantic (STS) feature
extraction algorithm our aim is to transform pause and path locations
into machine learning friendly discrete location sequences. There are
multiple ways of extracting such measurement clusters in the literature,
such as Spatio-Temporal Density-Based Spatial Clustering of Applications
with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
{[}(``ST-DBSCAN,'' 2007);{]}. We will focus on two methods which
explicitly with mobility patterns from unevenly sampled smartphone logs
(Barnett \& Onnela, 2016; Palmius et al., 2017 ). Both of these methods
pre-process the data and subsequently use two steps to extract pause
locations: first they extract pauses and their corresponding locations,
then they cluster pause locations based on spatial proximity. This
implementation of PMMI uses a stricter version of Barnett and Onnela
(2016)'s approach to extract pauses.

First the measurements \(D\) are filtered such that only measurements
with an accuracy value lower than \(a_{\text{P lim}}\) remain within the
sample. Then, a measurement \(D_t\) belongs to a pause if and only if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The next measurement \(D_{t+1}\) is within \(t_{\text{Pause lim}}\)
  amount of seconds (so it is not missing)
\item
  The next measurement \(D_{t+1}\) is within \(d_{\text{Pause lim}}\)
  meters.
\item
  The duration of the pause is more than \(\delta_{\text{Pause lim}}\)
  seconds.
\item
  Let the measurements of a possible pause which fit the aforementioned
  criteria be \(D_{t,t+1,...,t+n}\). These points are only a pause if
  the distance between the mean coordinates of \(D_{t,t+1,...,t+n}\) and
  the furthest away points of \(D_{t,t+1,...,t+n}\) is within 2 times
  the mean accuracy \(a\) of \(D_{t,t+1,...,t+n}\).
\end{enumerate}

This set of points were then hierarchically clustered using a distance
matrix, such that all points within \(d\) meters of each other were
clustered into a pause location. Each pause location is a bin.

For all remaining measurements we assume that they belong to paths. In
this implementation of PMMI we use the following algorithm to estimate
paths:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Take all measurements which are not pauses, filter them based on an
  accuracy threshold \(a_{\text{Path Lim}}\).
\item
  Create a distance matrix for all remaining measurements \(D_t \in F\)
  and hierarchically cluster it accordingly, such that all points within
  \(d_{\text{Path Lim}}\) meters of each other are clustered into a
  single pause point.
\end{enumerate}

At this point all empirically observed path bins and pause bins are
extracted. However, there may be some overlap between pause bins and
path bins. Thus, the bins are clustered again, such that the pause bins
retain priority. This means that if a pause bin and a path bin are
within less than \(d\) meters of each other, the path bin is removed.
The reasoning for this is that the threshold for not being in a pause
cluster should be higher, as individuals spend the majority of time at a
pause cluster. The end result is a discretised map which contains pause
and flight bins based on the entire log history of the individual.

\subsubsection{Binning}\label{binning}

({\textbf{???}})'s spatial-temporal-semantic (STS) feature extraction
algorithm uses map matching as a ground truth to assign noisy
measurements into discrete bins along roads. In other words, in addition
to the measured data they also use a geographic database that contains
information about the area in which the individual is (e.g.~where
precisely the roads are), and sort measurements into bins based on both
the measurement and the geographic data base. For example, if an
individual is measured as moving closely in parallel to a road A in an
area where there is no other parallel road, ({\textbf{???}})'s method
will assume that the individual is on the road A.

PMMI uses a similar logic, but without using any external geographic
database. The key modification in PMMI is that whilst ({\textbf{???}})
uses a map from outside the persons location logs, we use the total
location history of the individual to create a personalised map. This
map is subsequently used to bin measurements. This is feasible for two
reasons: humans tend to have repetitive movement habits and secondary
logs tend to be long. To put it in simpler terms, we consider each
measurement at \(D_x\) as a sample of \(Path_{ab}\), and by aggregating
many measurements we can use them to map out \(Path_{ab}\).

Thus, all measurements \(D\) were are then assigned to a discrete bin on
the personal map. This includes previous measurements which were
discarded from the map building exercise due to an accuracy \(a\) value
which exceeded \(a_{\text{P lim}}\) or \(a_{\text{F lim}}\). In this
implementation we used a simple assigning function, whereby the
measurements where assigned to the bin nearest to the measurement.

\subsubsection{Classification}\label{classification}

At this point the objective of PMMI is to take all the information
available about the mobility history of an individual and impute the
missing value. In this implementation we trained an artificial neural
network (ANN) to do so. For more information on the precise architecture
of the artificial neural network please consult the appendix. The input
variables to the ANN are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The previous and subsequent observed bin as a binary class matrix.
\item
  The distance in time to the next \& previous bin.
\item
  The time of the day encoded as a cyclical two-dimensional feature.
\item
  The day of the month as a binary class matrix.
\item
  The month of the year as a binary class matrix.
\end{enumerate}

For the encoding of the time of day we took the cosine and the sine
transforms of the amount of seconds that have elapsed after midnight
(London, 2016). This is necessary so that the model can understand that
one second past midnight and one second before midnight are in fact two
seconds away from each other. Moreover we scaled the non-binary values
to occupy a range between 0 and 1 in order to ensure convergence.

For a missing time-point at \(D_t \in \emptyset\), the output of the
model is a set of probability estimates associated with every location
cluster. That is, for each missing time-point the model returns a vector
of probability estimates (with one estimate per bin) associated with
where the individual is.

\subsection{Datasets \& Analyses}\label{datasets-analyses}

The secondary location log used to train the imputation methods was
collected between 2013 and 2017 on different Android devices from a
single individual. About 54\% of the data is missing for the entire
duration of the log. This may be misleading as there are several long
periods with no measurements whatsoever. For days which were not
entirely missing, approximately 22\% of all five minute segments were
missing. The structure of missingness of a day with measurements is
shown in Figure \ref{fig:measurementsPerDay}. As you can see, there are
several long periods over the course of the log for which there are no
measurements. The median sampling frequency per day for non-missing days
is around 0.006 Hz.

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingdayBoaz5min} \caption{Missing data for the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

For simplicity, we subsequently used a time period when the individual
was living in the Netherlands. This leaves us with 156,000 measurements
over a period of less than six months. Analyses as well as
implementations and adaptions Palmius et al. (2017)'s and Barnett and
Onnela (2016)'s model were performed using R and a multitude of other
statistical packages R (Version 3.4.3; R Core Team, 2017) and the
R-packages \emph{ggplot2} (Version 2.2.1; Wickham, 2009),
\emph{ggthemes} (Version 3.4.0; Arnold, 2017), \emph{knitr} (Version
1.20; Xie, 2015), and \emph{papaja} (Version 0.1.0.9709; Aust \& Barth,
2018). All the code is available on a public repository
({\textbf{???}}).

\subsection{Results \& Evaluation
Metrics}\label{results-evaluation-metrics}

The results will consist of multiple steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluating the performance of the map building and assigning
  functions.
\item
  Comparing the performance of PMMI using a) baseline models b)
  performance with randomly removed data in comparison to the
  aforementioned methods (Barnett \& Onnela, 2016; Palmius et al., 2017)
  and c) objective ground-truth data (public transportation
  time-stamps).
\end{enumerate}

\subsubsection{Map building \& binning
evaluation}\label{map-building-binning-evaluation}

Before we can evaluate the accuracy of the imputations, it is essential
to evaluate how well noise in the data has been cleaned. Otherwise we
run the risk of over-fitting the model in the sense that we will measure
the extent to which an imputation method can correctly impute
measurement noise within \(D_t\) instead of true location \(G(t)\).

In order to evaluate the map building and binning we will first visually
evaluate the paths and pause locations. A visual evaluation of paths
superimposed on is an established way to heuristically check their
accuracy (e.g. ({\textbf{???}})). Then, let the average distance between
the actual measured point and the binned point be the \emph{deviation
distance} \(\delta_{\text{dev}}\). With respect to the deviation
distance \(\delta_{\text{dev}}\), we expect:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A positive relationship between the deviation distance and the
  accuracy of each measurement.
\item
  Roughly 67\% of the deviation distances \(\delta_{\text{dev}}\) are
  within accuracy \(a\) of each measurement.
\end{enumerate}

\subsubsection{Imputation algorithm
performance}\label{imputation-algorithm-performance}

We will compare the performance of PMMI, Palmius et al. (2017) and
Barnett and Onnela (2016). In addition, we will also compute a naive
model, which simply imputes as the missing value the previous observed
value. The naive model will serve as a baseline model. To compare the
performance of these methods we will remove 25\% of measured time
intervals at random within a four week period. We will make our
comparisons for intervals of 5 minutes, 1 hour and 1 day. In other
words, we will remove 25\% of time intervals at random, while varying
the size of the time intervals removed.

For the Barnett and Onnela (2016) and PMMI models we will use all the
available data to train the models with the exception of the time
periods being investigated. Palmius et al. (2017)'s model does not
require training.

To compare all methods with each other we will compute a distance
measure (how far was the removed location from the predicted location).
For PMMI's imputations we will use a weighted mean for the distance
measures whereby each 5 minute period is weighted equally. This is
necessary because the other two model estimate far fewer values than
PMMI. While Barnett and Onnela (2016)'s and Palmius et al. (2017)'s
models estimate 12 measurements for each missing hour, PMMI estimates as
many measurements as there are in the log. When individuals go to
infrequently travelled locations they tend to use their phone's location
services more than 100 more times, which leads to more measurements, and
hence more imputed values. Hence, if the weights are not used PMMI is
susceptible to error overestimates which are related to measurement
frequency in a way that the other two methods are not.

In addition, other measures of interest for PMMI are accuracy (in what
percentage of the cases was the appropriate cluster predicted), the
\emph{confidence} and the \emph{distance expectation}. The confidence is
the probability with which the model predicts the most likely cluster.
For instance, if the model predicts the missing bin to be bin A with a
probability of 0.9 then we can say it has high confidence in the
prediction. Similarly, the distance expectation is the cross-product of
the estimated probabilities that the individual is at any of all given
clusters and the distances between the clusters to the true cluster. For
example, if the true location of an individual is bin A (bin A is 10
meters away from bin B) and the model assigns a probability of 0.9 at
bin A and 0.1 at bin B, then the distance expectation would be 1 meter.

Finally, we will take objective real world data and compare it to
predicted values. We will use information from the Dutch public
transportation card. The Dutch public transportation service provides
users with time-stamped data of when and where they board, change lines
or leave public transportation. To be able to make a comparison between
models we will remove all measurements from within the 5 minute period
that a time-stamped measurement is available. Then we will use each
model to impute the location of the individual within that period.

\section{Results}\label{results}

\subsection{Map building \& binning
evaluation}\label{map-building-binning-evaluation-1}

We used the following parameters to extract pauses: an accuracy limit
\(a_{\text{Pause lim}}\) of 250 meters, a time limit
\(t_{\text{Pause lim}}\) of 300 seconds, a distance limit of
\(d_{\text{Pause lim}}\) 50 meters and a minimum pause duration limit
\(\delta_{\text{Pause lim}}\) of 100 seconds. Moreover, to extract path
clusters we used the parameters:
\(a_{\text{Path Lim}} = 150 \text{meters}\) and
\(d_{\text{Path Lim}} =300 \text{meters}\).

The selection of these parameters was more-or-less heuristically driven
based on their on their ability to extract a meaningful personalised
map. When selecting parameters there is a trade off is between bias and
precision. This is because an increase in precision in the form of a
higher for high density locations resolution comes at the expense of
precision as assigning measurements to bins becomes more difficult. For
instance, by increasing the clustering distance parameter
\(d_{\text{Pause lim}}\) we can extract more valid pause locations at
the expense of falsely categorising certain measurements to the wrong
cluster. This is illustrated in Figure \ref{fig:pauseLocUithof1}.

\begin{figure}
\includegraphics[width=1\linewidth]{img/distanceParam} \caption{Example of pause locations in De Uithof university campus using 150 meters (blue) and 400 meters (red) as clustering parameters.}\label{fig:pauseLocUithof1}
\end{figure}

Map building results in a personalised map with pause and path clusters.
An excerpt can be seen in Figure \ref{fig:mapClusterMap}. It is
important to remind the reader that PMMI is map agnostic and uses no
information from the map. Therefore, the close overlap with features on
the map, such as pause bins at relevant buildings and transportation
clusters, as well as the flight bins following roads and railway lines
indicate a high degree of precision in personal map building. As
expected, PMMI's path extraction yields greater accuracy for frequently
occurring paths. For example, the frequently travelled Amsterdam-Utrecht
railway line has been extracted almost perfectly, while the less
frequently travelled Utrecht-Enschede line is far sparser.

\begin{figure}
\includegraphics[width=1\linewidth]{img/mapClusterMap} \caption{Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.}\label{fig:mapClusterMap}
\end{figure}

For the entire period examined period the we find a deviance of 40
meters and a median deviance of 15 meters. Around 69\% of the deviance
values are within their corresponding accuracy value, which is close to
the theoretical \(\frac{2}{3}\) value that is expected. Approximately
9\% of values were not taken into account when creating the bins,given
that their the accuracy \(a\) exceeded \(a_{\text{Path Lim}}\).

For the narrower period of March 2017 approximately 74\% of the deviance
values are within their corresponding accuracy values, which is close to
the theoretical \(\frac{2}{3}\) value that is expected. The raw
unweighted mean and median deviance are 38 and 14 meters respectively.
If we weight them such that each 5 minute interval is weighted equally
to ease comparison with the other two methods, we find a mean deviance
of 35 meters with a median of 12.8.

In comparison, Palmius' method has a median deviance
\(\delta_{\text{dev}}\) of 3 meters, with a mean of 115 due to high
deviance outliers. On the other hand, Barnett's method has mean deviance
\(\delta_{\text{dev}}\) of 343 meters and a median deviance of 8 meters.
Barnett's deviance is necessarily higher than Palmius', as they
down-sample temporally (like Palmius) and subsequently aggregate into
pauses and linear flights.

The key difference between temporal and spatial down-sampling is shown
in Figure \ref{fig:palmiusVme1}. Temporal down-sampling is much more
sensitive to noise in sparsely measured periods because it averages out
values within five minute periods. Often there are only a few noisy
measurements in those periods (see Figure
\ref{fig:longMeasurementsPerDay} ), which leads to a noise in the
down-sampled values. Unsurprisingly, there is a positive relationship
between deviance and the amount of measurements in each down-sampled
interval. The fact that over 90\% of deviance values are within accuracy
(much higher than the expected theoretical 67\%) confirms that temporal
down-sampling is not sufficiently filtering out the noise.

\begin{figure}
\includegraphics[width=1\linewidth]{img/tempVsspace} \caption{Excerpt. MISSING CAPTION HERE}\label{fig:palmiusVsMe1}
\end{figure}

\subsubsection{Imputation evaluation}\label{imputation-evaluation}

The Palmius et al. (2017) model failed to impute 3\% of all removed
values for both the 5 minute and over 10\% for the 1 hour tests. Palmius
et al. (2017)' model failed to impute a single value for the day tests.
Similarly, Barnett and Onnela (2016)'s method failed to impute over 11\%
of the missing values. PPMI made an imputation for all missing values.
Table 1 shows the results of the distance metrics for each method.

\begin{table}[htb]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:resulttable}Distance in meters between the removed time period and the imputed value.}
\begin{tabular}{lllllll}
\toprule
 & \multicolumn{2}{c}{Five minutes} & \multicolumn{2}{c}{One Hour} & \multicolumn{2}{c}{One Day} \\
\cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
 & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median}\\
\midrule
Barnett \& Onella & 82 & 4 & 345 & 6 & 9,273 & 12\\
Palmius & 43 & 0 & 497 & 4 & NA & NA\\
PPMI & 269 & 0 & 908 & 0 & 5,757 & 0\\
Naive Baseline & 426 & 0 & 1,502 & 0 & 14,266 & 1,288\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

In terms of PPMI's accuracy in this period, the prediction accuracy was
88\%, 73\% and 47\% for the 5 minute, 1 hour and 1 day periods
respectively. As a comparison, the naive model's accuracy ratings were
87\%, 68\% and 24\% respectively. The distance expectation values were
very similar to the distance scores, albeit approximately 5\% higher.
Confidence scores ranged from 0.006 to 1.

\subsubsection{Comparison with objective
data}\label{comparison-with-objective-data}

Once we removed the measurements within the 5 minute period of each
time-stamp, we employed the models to predict the location of the
individual at the time-stamped period. The naive model reaches an
accuracy of 16\%, with a median distance of 555 meters and mean distance
of 3303 meters. On the other hand MYMETHOD has an accuracy of 24\%, with
a median distance of 1037 meters and a mean distance of 5637. Again the
expectation of the distance was quite similar to the distance with a
mean of 6432 and a median of 1037 meters.

Palmius's model failed to impute 38 out of 97 periods. For those it did
impute, it showed a mean distance of 1517 meters and a median distance
of 1617. Barnett's model failed to impute 18 out of 97 periods. For
those it did impute, it had a mean of 5506 and a median of 1342.

\subsubsection{Example: effect on aggregate
measures}\label{example-effect-on-aggregate-measures}

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information, such as the amount of time spent
at home. As an example we calculated the time spent at home of the user
in the month of March (Figure \ref{fig:aggrePlot}) without any model and
with all three of the investigated methods.

\begin{figure}
\includegraphics[width=1\linewidth]{img/timeUse} \caption{Proportion of time spent at home in March 2017. The raw values are estimated by downsampling temporally the lattitude and longitude for every 5 minute time period in the month. We used each method's own binning method and classified as at home if the downsampled measurement was within 250 meters from home.}\label{fig:aggrePlot}
\end{figure}

Interestingly, all three models suggest that the user spent
approximately 60\% of their time at home. However, without using any
form of missing data imputation approximately 12\% of the time is
unaccounted for. Given that the amount of time spent at home has been
found to be a reliable predictor of extroversion (G. M. Harari et al.,
2016) and the onset of depressive episodes in bipolar patients (Palmius
et al., 2017) obtaining an accurate value for mobility metrics like this
is highly important to social scientists.

\subsection{Discussion \& Conclusion}\label{discussion-conclusion}

Overall the PMMI performed better than the alternative models,
particularly during longer missing periods and with objective data.
However, it did not perform substantially better than the naive baseline
model. In addition, the comparison to the performance of the
Barnett\&Onella and Palmius models is somewhat unfair, as they were
created for custom logs, not secondary logs. Nonetheless, the comparison
remains valid as they are the closest we found to a missing data
imputation methods in smartphone GPS logs.

In addition to higher accuracy under the conditions typical of secondary
logs, the advantages of PPMI are increased coverage and flexibility for
missing data imputation, robustness to irregular sampling, the ability
to model complex non-linear interactions in its imputations, and the
ability to use historical records to smooth movement noise.

PPMI's increased coverage and flexibility comes from its ability to make
complex non-linear predictions. For instance, in a given missing period
it might make sense to predict that the individual is either at home, or
at the office, or at a shop with equal probability. While PPMI can make
such an imputation, none of the alternative methods can do this.
Moreover, the ability to take the prediction probability values from the
neural network also helps in dealing with uncertainty. A known-drawback
of single imputation is that it takes an imputed value and treats it as
observed. Simple rule-based methods such as Palmius' are essentially
algorithmic single imputation methods. With PPMI it is possible to model
uncertainty using the predicted probabilities of each estimate. For
instance, in the previous example, we could choose to only take
estimates with a high degree of confidence, thus creating confidence
intervals by adding and subtracting the amount of cases where the
location of the individual is ambiguous.

With respect to irregular sampling, alternative methods use temporally
based down-sampling in order to reduce noise. This leads to
deterioration in resolution not only over space, but also over time. A
combination of irregular sampling with the fluctuating accuracy values
can lead to nonsensical results. For instance, consider a case where
there are two inaccurate measurements in movement at 12:00:01 and
12:04:59. Down-sampling over 5 minute periods will lead to a value that
will be the mean of the two inaccurate samples, which is likely to be a
location the individual is certainly not. PPMI instead down-samples
spatially, which ensures that the binned location is one which is
composed of the mean of hundreds of observations, not just the few that
happen within a single period.

While both Barnett and Onnela (2016) and Palmius et al. (2017) use
historical data to smoothen pause locations by clustering pause
locations with with a close degree of spatial proximity, neither of them
do the same for non-pause locations. This may be feasible with high
frequency, regularly sampled short duration logs but creates noise with
secondary logs. Moreover, with secondary logs it is feasible to
spatially \enquote{average out} multiple samples of the same path in
order to recreate it in its entirety. For instance, although the mean
sampling frequency during train travels on the Amsterdam-Utrecht line is
low (about 0.01 Hz) the personalised map manages to recreate the train
line almost perfectly, despite being completely map agnostic.

There are multiple methodological limitations in this paper. Most
importantly, the evaluation methods are imperfect. The golden standard
would be to use at least one highly accurate professional grade GPS
device with high sampling frequency to compare our data to. Until that
is available, the use of public transport data and cross-validation is
just a substitute.

Furthermore, PPMI can be further developed. The map building function,
the assignment function and the classification model remain simplistic
and could be improved.

In map building, the probability of a pause at a given location is
certainly related to other factors, such as the time of the day as well
as the prior history of pauses at that location. These factors are not
taken into account in the pause extraction function. Improved methods
would do well to do so. As for paths, a drawback of the current method
is that the density of the clusters is a function of the clustering
parameter \(d\), the distance between the observed points and their
sampling density. It does not take into account the length of the path
as well as the average sampling frequency of the path. This is an issue
because it can lead to bins to which data points are seldom assigned.
For example, while the Amsterdam-Utrecht line has been mapped out almost
perfectly, many of the clusters along the route have only been assigned
few measurements. This leads to difficulties in the classification part
of the model, as infrequently observed clusters are hard for the model
to predict.

The current assignment function simply assigns each measurement to the
nearest bin. It does not take into account any contextual information
that can be gleaned from the entire movement history of the information,
such as what path they are on. For instance, assume that it is known
that an individual is travelling from point A to point B along path AB,
and there is an inaccurate measurement closest to a cluster which
belongs to path AC. By only taking distance into account, the
measurement can get assigned to the wrong cluster on path AC. An
improvement would be use a Bayesian method, whereby assignment is a
function of both the measurement and a model of the individuals movement
history. In terms of state-space models like the Kalman filter the state
equation would represent a probabilistic representation of where the
individual could be based on the individuals entire movement history.
The space side of the model would be a measurement equation representing
the measurement and the uncertainty surrounding it in the form of \(a\).

As for the simplicity of the classification method, the neural network
which was used to generate predictions used no information on sequence
patterns longer than the previous and next bin. A more sophisticated
recurrent neural network (RNN), or a long short-term memory recurrent
neural network (LSTM) would likely perform significantly better.

That there is room for improvement is unsurprising given that it is
still early days for using smartphone location measurements in social
science. Nonetheless, the methodological advantages are clear: millions
of individuals have a location log containing objective measurements
spanning years which can be accessed for free. This is vastly superior
to alternatives such questionnaires which rely on accurate
self-reporting. Social science researchers must take advantage of
regulatory changes in data portability regulations and put the vast
wealth of data collected by commercial entities to public use.

\newpage

\section{References}\label{references}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{r_refs}\NormalTok{(}\DataTypeTok{file =} \StringTok{"r-references.bib"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begingroup
\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\hypertarget{ref-R-ggthemes}{}
Arnold, J. B. (2017). \emph{Ggthemes: Extra themes, scales and geoms for
'ggplot2'}. Retrieved from
\url{https://CRAN.R-project.org/package=ggthemes}

\hypertarget{ref-R-papaja}{}
Aust, F., \& Barth, M. (2018). \emph{papaja: Create APA manuscripts with
R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}

\hypertarget{ref-barnett_inferring_2016}{}
Barnett, I., \& Onnela, J.-P. (2016). Inferring mobility measures from
GPS traces with missing data. \emph{arXiv:1606.06328 {[}Stat{]}}.
Retrieved from \url{http://arxiv.org/abs/1606.06328}

\hypertarget{ref-chen_practical_2006}{}
Chen, M. Y., Sohn, T., Chmelev, D., Haehnel, D., Hightower, J., Hughes,
J., \ldots{} Varshavsky, A. (2006). Practical metropolitan-scale
positioning for GSM phones. In \emph{UbiComp 2006: Ubiquitous computing}
(pp. 225--242). Springer, Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11853565_14}{10.1007/11853565\_14}

\hypertarget{ref-chen_state_2013}{}
Chen, Z., \& Brown, E. N. (2013). State space model.
\emph{Scholarpedia}, \emph{8}(3), 30868.
doi:\href{https://doi.org/10.4249/scholarpedia.30868}{10.4249/scholarpedia.30868}

\hypertarget{ref-commission_protecting_2017}{}
Commission, E. (2017). \emph{Protecting your data: Your rights -
european commission}. Retrieved from
\url{http://ec.europa.eu/justice/data-protection/individuals/rights/index_en.htm}

\hypertarget{ref-delclos-alio_keeping_2017}{}
Delclòs-Alió, X., Marquet, O., \& Miralles-Guasch, C. (2017). Keeping
track of time: A smartphone-based analysis of travel time perception in
a suburban environment. \emph{Travel Behaviour and Society},
\emph{9}(Supplement C), 1--9.
doi:\href{https://doi.org/10.1016/j.tbs.2017.07.001}{10.1016/j.tbs.2017.07.001}

\hypertarget{ref-duncan_portable_2013}{}
Duncan, S., Stewart, T. I., Oliver, M., Mavoa, S., MacRae, D., Badland,
H. M., \& Duncan, M. J. (2013). Portable global positioning system
receivers: Static validity and environmental conditions. \emph{American
Journal of Preventive Medicine}, \emph{44}(2), e19--29.
doi:\href{https://doi.org/10.1016/j.amepre.2012.10.013}{10.1016/j.amepre.2012.10.013}

\hypertarget{ref-feng_cutoff:_2014}{}
Feng, L., Nowak, G., O'Neill, T., \& Welsh, A. (2014). CUTOFF: A
spatio-temporal imputation method. \emph{Journal of Hydrology},
\emph{519}, 3591--3605.
doi:\href{https://doi.org/10.1016/j.jhydrol.2014.11.012}{10.1016/j.jhydrol.2014.11.012}

\hypertarget{ref-goodchild_toward_2010}{}
Goodchild, M. F., \& Janelle, D. G. (2010). Toward critical spatial
thinking in the social sciences and humanities. \emph{GeoJournal},
\emph{75}(1), 3--13.
doi:\href{https://doi.org/10.1007/s10708-010-9340-3}{10.1007/s10708-010-9340-3}

\hypertarget{ref-grunerbl_smartphone-based_2015}{}
Grünerbl, A., Muaremi, A., Osmani, V., Bahle, G., Ohler, S., Tröster,
G., \ldots{} Lukowicz, P. (2015). Smartphone-based recognition of states
and state changes in bipolar disorder patients. \emph{IEEE Journal of
Biomedical and Health Informatics}, \emph{19}(1), 140--148.
doi:\href{https://doi.org/10.1109/JBHI.2014.2343154}{10.1109/JBHI.2014.2343154}

\hypertarget{ref-harari_using_2016}{}
Harari, G. M., Lane, N. D., Wang, R., Crosier, B. S., Campbell, A. T.,
\& Gosling, S. D. (2016). Using smartphones to collect behavioral data
in psychological science: Opportunities, practical considerations, and
challenges. \emph{Perspectives on Psychological Science}, \emph{11}(6),
838--854.
doi:\href{https://doi.org/10.1177/1745691616650285}{10.1177/1745691616650285}

\hypertarget{ref-jankowska_framework_2015}{}
Jankowska, M. M., Schipperijn, J., \& Kerr, J. (2015). A framework for
using GPS data in physical activity and sedentary behavior studies.
\emph{Exercise and Sport Sciences Reviews}, \emph{43}(1), 48--56.
doi:\href{https://doi.org/10.1249/JES.0000000000000035}{10.1249/JES.0000000000000035}

\hypertarget{ref-lamarca_place_2005}{}
LaMarca, A., Chawathe, Y., Consolvo, S., Hightower, J., Smith, I.,
Scott, J., \ldots{} Schilit, B. (2005). Place lab: Device positioning
using radio beacons in the wild. In \emph{Pervasive computing} (pp.
116--133). Springer, Berlin, Heidelberg.
doi:\href{https://doi.org/10.1007/11428572_8}{10.1007/11428572\_8}

\hypertarget{ref-location_history_timeline_2017}{}
Location History, G. (2017). \emph{Timeline}. Retrieved from
\url{https://www.google.com/maps/timeline?pb}

\hypertarget{ref-london_encoding_2016}{}
London, I. (2016). \emph{Encoding cyclical continuous features - 24-hour
time. Ian london's blog}. Retrieved April 27, 2018, from
\url{//ianlondon.github.io/blog/encoding-cyclical-features-24hour-time/}

\hypertarget{ref-palmius_detecting_2017}{}
Palmius, N., Tsanas, A., Saunders, K. E. A., Bilderbeck, A. C., Geddes,
J. R., Goodwin, G. M., \& Vos, M. D. (2017). Detecting bipolar
depression from geographic location data. \emph{IEEE Transactions on
Biomedical Engineering}, \emph{64}(8), 1761--1771.
doi:\href{https://doi.org/10.1109/TBME.2016.2611862}{10.1109/TBME.2016.2611862}

\hypertarget{ref-patterson_statespace_2008}{}
Patterson, T. A., Thomas, L., Wilcox, C., Ovaskainen, O., \&
Matthiopoulos, J. (2008). State--space models of individual animal
movement. \emph{Trends in Ecology \& Evolution}, \emph{23}(2), 87--94.
doi:\href{https://doi.org/10.1016/j.tree.2007.10.009}{10.1016/j.tree.2007.10.009}

\hypertarget{ref-preisler_modeling_2004}{}
Preisler, H. K., Ager, A. A., Johnson, B. K., \& Kie, J. G. (2004).
Modeling animal movements using stochastic differential equations.
\emph{Environmetrics 15: P. 643-657}. Retrieved from
\url{https://www.fs.usda.gov/treesearch/pubs/33038}

\hypertarget{ref-R-base}{}
R Core Team. (2017). \emph{R: A language and environment for statistical
computing}. Vienna, Austria: R Foundation for Statistical Computing.
Retrieved from \url{https://www.R-project.org/}

\hypertarget{ref-sadilek_far_2016}{}
Sadilek, A., \& Krumm, J. (2016). Far out: Predicting long-term human
mobility. \emph{Microsoft Research}. Retrieved from
\url{https://www.microsoft.com/en-us/research/publication/far-predicting-long-term-human-mobility/}

\hypertarget{ref-saeb_mobile_2015}{}
Saeb, S., Zhang, M., Karr, C. J., Schueller, S. M., Corden, M. E.,
Kording, K. P., \& Mohr, D. C. (2015). Mobile phone sensor correlates of
depressive symptom severity in daily-life behavior: An exploratory
study. \emph{Journal of Medical Internet Research}, \emph{17}(7), e175.
doi:\href{https://doi.org/10.2196/jmir.4273}{10.2196/jmir.4273}

\hypertarget{ref-schipperijn_dynamic_2014}{}
Schipperijn, J., Kerr, J., Duncan, S., Madsen, T., Klinker, C. D., \&
Troelsen, J. (2014). Dynamic accuracy of GPS receivers for use in health
research: A novel method to assess GPS accuracy in real-world settings.
\emph{Frontiers in Public Health}, \emph{2}, 21.
doi:\href{https://doi.org/10.3389/fpubh.2014.00021}{10.3389/fpubh.2014.00021}

\hypertarget{ref-simmons_false-positive_2011}{}
Simmons, J. P., Nelson, L. D., \& Simonsohn, U. (2011). False-positive
psychology: Undisclosed flexibility in data collection and analysis
allows presenting anything as significant. \emph{Psychological Science},
\emph{22}(11), 1359--1366.
doi:\href{https://doi.org/10.1177/0956797611417632}{10.1177/0956797611417632}

\hypertarget{ref-_st-dbscan:_2007}{}
ST-DBSCAN: An algorithm for clustering spatial--temporal data. (2007).
\emph{Data \& Knowledge Engineering}, \emph{60}(1), 208--221.
doi:\href{https://doi.org/10.1016/j.datak.2006.01.013}{10.1016/j.datak.2006.01.013}

\hypertarget{ref-villanueva_association_2013}{}
Villanueva, C., \& Aggarwal, B. (2013). The association between
neighborhood socioeconomic status and clinical outcomes among patients 1
year after hospitalization for cardiovascular disease. \emph{Journal of
Community Health}, \emph{38}(4), 690--697.
doi:\href{https://doi.org/10.1007/s10900-013-9666-0}{10.1007/s10900-013-9666-0}

\hypertarget{ref-wang_smartgpa:_2015}{}
Wang, R., Harari, G., Hao, P., Zhou, X., \& Campbell, A. T. (2015).
SmartGPA: How smartphones can assess and predict academic performance of
college students. In \emph{Proceedings of the 2015 ACM international
joint conference on pervasive and ubiquitous computing} (pp. 295--306).
New York, NY, USA: ACM.
doi:\href{https://doi.org/10.1145/2750858.2804251}{10.1145/2750858.2804251}

\hypertarget{ref-R-ggplot2}{}
Wickham, H. (2009). \emph{Ggplot2: Elegant graphics for data analysis}.
Springer-Verlag New York. Retrieved from \url{http://ggplot2.org}

\hypertarget{ref-wolf_impact_2003}{}
Wolf, J., Oliveira, M., \& Thompson, M. (2003). Impact of underreporting
on mileage and travel time estimates: Results from global positioning
system-enhanced household travel survey. \emph{Transportation Research
Record: Journal of the Transportation Research Board}, \emph{1854},
189--198. doi:\href{https://doi.org/10.3141/1854-21}{10.3141/1854-21}

\hypertarget{ref-R-knitr}{}
Xie, Y. (2015). \emph{Dynamic documents with R and knitr} (2nd ed.).
Boca Raton, Florida: Chapman; Hall/CRC. Retrieved from
\url{https://yihui.name/knitr/}

\hypertarget{ref-zenk_how_2009}{}
Zenk, S. N., Schulz, A. J., \& Odoms-Young, A. (2009). How neighborhood
environments contribute to obesity. \emph{The American Journal of
Nursing}, \emph{109}(7), 61--64.
doi:\href{https://doi.org/10.1097/01.NAJ.0000357175.86507.c8}{10.1097/01.NAJ.0000357175.86507.c8}

\hypertarget{ref-zhang_application_2017}{}
Zhang, Z., Yang, X., Li, H., Li, W., Yan, H., \& Shi, F. (2017).
Application of a novel hybrid method for spatiotemporal data imputation:
A case study of the minqin county groundwater level. \emph{Journal of
Hydrology}, \emph{553}(Supplement C), 384--397.
doi:\href{https://doi.org/10.1016/j.jhydrol.2017.07.053}{10.1016/j.jhydrol.2017.07.053}

\endgroup

\end{document}

### 2018-04-29 manuscript/manuscript.Rmd
APA manuscript

@@
 ---
-title: "Manuscript"
+title: "Personalised Map Matched Imputation: an approach to impute missing data in smartphone location logs"
 author: "Boaz Sobrado"
@@
 A GPS sensor uses the distance between a device and several satellites to determine the location of the device.
-Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
+Although using a GPS sensor is the most accurate way to establish location on a smartphone, the GPS sensor is also the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006].
 In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation.
@@
 
-include_graphics("img/missingBoaz5minExample.png")
-#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.
-
+include_graphics("img/missin5minExample2018.png")
 ```
@@
 The red line shows the path with all measurements included.
-@palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
+P note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour.
 Inaccurate measures are often followed by more accurate measures.
-There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
+There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cellphone tower triangulation.",out.width = '100%'}
 include_graphics("img/journeyTillMiddayBoaz.png")
@@
 
-GPS measurements provide us with coordinates on the surface of the earth, which is ellipsoid shaped.
+Location measurements, such as those produced by GPS sensors, provide us with coordinates (lattitude and longitude) on the surface of the earth, which is ellipsoid shaped.
 Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion.
@@
 
-1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
-2. Binning: Assign each measurement $D$ to a unique discrete location. 
-3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.
+1. *Map building*: Extract from measurements $D$ all pause location bins and path locatin bins to create a personalised map.
+2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
+3. *Imputing*: Use a classification method to predict missing measurements based on all the availble information.
 
@@
 Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity.
-For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
-However, there is no reason why more sophisticated methods cannot be used.
+This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.
 
@@
 
-At this point all empirically observed path clusters and pause clusters are extracted.
-However, there may be some overlap between pause clusters and path clusters.
-Thus, the clusters are clustered again, such that the pause clusters retain priority.
-This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed.
+At this point all empirically observed path bins and pause bins are extracted.
+However, there may be some overlap between pause bins and path bins.
+Thus, the bins are clustered again, such that the pause bins retain priority.
+This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed.
 The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster.
-The end result is a discretised map which contains pause and flight clusters based on the history of the individual.
+The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.
 
-### Binning
+#### Binning
 
@@
 
-The key modification in PMMI is that whilst @wu uses data from outside the persons location logs, we use the total location history of the individual to create a personalised map which will be then used to bin measurements.
+PMMI uses a similar logic, but without using any external geographic database.
+The key modification in PMMI is that whilst @wu uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map.
+This map is subsequently used to bin measurements.
 This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long.
-To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_ab$.
+To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_{ab}$.
 
@@
 This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$.
-Again, for simplicity we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
+In this implementation we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
 
@@
 This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.
+Moreover we scaled the non-binary values to occupy a range between 0 and 1 in order to ensure convergence.
 
@@
 For simplicity, we only used a time period when the indivdual was living in the Netherlands in the model.
+This leaves us with 156000 measurements over a period of less than six months.
 Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].
-**add packages** All the code is available on a public repository.
+All the code is available on a public repository [@mygithub].
 
@@
 2.
-Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data.
+Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data (public transportation timestamps).
 
@@
 In order to evalute the map building and binning we will first visually evaluate the paths and pause locations.
-A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (*add reference*).
+A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (e.g.
+@lcm).
 Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$.
@@
 In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value.
+The naive model will serve as a baseline model.
 To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period.
@@
 
-In addition, other measures we will report for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted) and the expected distance.
-The expected distance is the crossproduct of the estimated probabilities that the individual is at a given cluster and the distances between the clusters to the true cluster.
-For example, if the true location of an individual is cluster A (cluster A is 10 meters away from cluster B) and the model assigns a probability of 0.9 at cluster A and 0.1 at cluster B, then the expected distance would be 1 meter.
+In addition, other measures of interest for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*.
+The confidence is the probability with which the model predicts the most likely cluster.
+For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say it has high confidence in the prediction.
+Similarly, the distance expectation is the crossproduct of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster.
+For example, if the true location of an individual is bin A (bin A is 10 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 1 meter.
 
 Finally, we will take objective real world data and compare it to predicted values.
-We will use information from the Dutch public transportation card as objective real world data.
-The Dutch public transportation card provides users with timestamped data of when and where they board, change lines or leave public transportation.
+We will use information from the Dutch public transportation card.
+The Dutch public transportation service provides users with timestamped data of when and where they board, change lines or leave public transportation.
+To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a timestamped measurement is available.
+Then we will use each model to impute the location of the individual within that period.
 
@@
 
-The selection of these parameters was more-or-less heuristic.
+The selection of these parameters was more-or-less heuristically driven based on their on their ability to extract a meaningful personalised map.
+When selecting parameters there is a trade off is between bias and precision.
+This is because an increase in precision in the form of a higher for high density locations resolution comes at the expense of precision as assigning measurements to bins becomes more difficult.
+For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster.
+This is illustrated in Figure \@ref(fig:pauseLocUithof1).
 
-When selecting parameters there is a trade off is between resolution and accuracy.
-This is because an increase in resolution comes at the expense of accuracy.
-For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster.
-This is illustrated in Figures \@ref(fig:pauseLocUithof1) and \@ref(fig:pauseLocUithof2).
-
-```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
+```{r, pauseLocUithof1,echo = FALSE,fig.cap="Example of pause locations in De Uithof university campus using 150 meters (blue) and 400 meters (red) as clustering parameters.",out.width = '100%'}
 #fix width
-include_graphics("img/mapPauseRes1.png")
+include_graphics("img/distanceParam.png")
 ```
 
-```{r, pauseLocUithof2,echo = FALSE,fig.cap="Example of pause locations in De Uithof using 150 meters and 400 meters as clustering parameters respectively.",out.width = '100%'}
-include_graphics("img/mapPauseRes2.png")
-```
+Map building results in a personalised map with pause and path clusters.
+An excerpt can be seen in Figure \@ref(fig:mapClusterMap).
+It is important to remind the reader that PMMI is map agnostic and uses no information from the map.
+Therefore, the close overlap with features on the map, such as pause bins at relevant buildings and transportation clusters, as well as the flight bins following roads and railway lines indicate a high degree of precision in personal map building.
+As expected, PMMI's path extraction yields greater accuracy for frequently ocurring paths.
+For example, the frequently travelled Amsterdam-Utrecht railway line has been extracted almost perfectly, while the less frequently travelled Utrecht-Enschede line is far sparser.
 
-The end result of step 3 of MYMETHOD results in a personalised map with pause and path clusters.
-An excerpt can be seen in Figure \@ref(fig:mapClusterMap).
-As expected, MYMETHODs path extraction yields greater accuracy for frequently ocurring paths.
-It is important to remind the reader that despite the close overlap with features on the map,  such as roads and railway lines, MYMETHOD is map agnostic and uses no information from the map to recreate these features.
-
-```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map if an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
+```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
 #fix width
@@
 
-For the entire period when the individual is in the Netherlands, there is a mean deviance of 40 meters and a median deviance of 15 meters.
+For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters.
 Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected.
+Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}}$.
 
-For the narrower period of March 2017, figure \@ref(fig:devianceComp) illustrates the relationship between the deviance and accuracy.
-Approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected.
-The mean and median deviance are 38 and 14 meters respectively.
+For the narrower period of March 2017 approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected.
+The raw unweighted mean and median deviance are 38 and 14 meters respectively.
+If we weight them such that each 5 minute interval is weighted equally to ease comparison with the other two methods, we find a mean deviance of 35 meters with a median of 12.8.
 
-```{r, devianceComp,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
-#fix width
-# add figure with deviance comparisons three columns each method once
-```
+In comparison, Palmius' method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers.
+On the other hand, Barnett's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
+Barnett's deviance is neccessarily higher than Palmius', as they downsample temporally (like Palmius) and subsequently aggregate into pauses and linear flights.
 
-In comparison, Palmius' method has a median deviance of 3 meters, with a mean of 115 due to high deviance outliers.
-Over 90% of deviance values are within accuracy, which suggests that the model is binning noise.
-In Figure \@ref(fig:palmiusVme) the key difference is visible.
-The Palmius method is much more sensitive to inaccurate values, such as the one at the top right of the figure.
-
-On the other hand, Barnett's method has mean deviance of 343 meters and a median deviance of 8 meters.
-Barnett's deviance is neccessarily higher than Palmius', as they downsample and subsequently aggregate into pauses and linear flights.
+The key difference between temporal and spatial downsampling is shown in Figure \@ref(fig:palmiusVme1).
+Temporal downsampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods.
+Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the downsampled values.
+Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each downsampled interval.
+The fact that over 90% of deviance values are within accuracy (much higher than the expected theoretical 67%) confirms that temporal downsampling is not sufficiently filtering out the noise.
 
@@
 #fix width
-include_graphics("img/palmiusVme.png")
+include_graphics("img/tempVsspace.png")
 ```
 
-```{r, palmiusVsMe2,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
-include_graphics("img/binComp.png")
-```
+### Imputation evaluation
 
-### Comparison to alternative models
+The @palmius_detecting_2017 model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests.
+@palmius_detecting_2017' model failed to impute a single value for the day tests.
+Similarly, @barnett_inferring_2016's method failed to impute over 11% of the missing values. PPMI made an imputation for all missing values.
 
 ```{r,resulttable, echo = F}
-library(knitr)
-library(kableExtra)
+# results df
 
@@
                          median5min = c(4,0,0,0),
-                         meanexpDist5min = c(NA,NA,414,NA),
-                         medianexpDist5min = c(NA,NA,0,NA),
                          mean1hr = c(345,497,908,1502),
                          median1hr = c(6,4,0,0),
-                         meanexpDist1hour = c(NA,NA,916, NA),
-                         medianexpDis1hour = c(NA,NA,0, NA),
                          mean1day = c(9273,NA,5757,14266),
-                         median1day = c(12,NA,0,1288),
-                         meanexpDist1Day = c(NA,NA,5757, NA),
-                         medianexpDis1Day = c(NA,NA,0, NA))
+                         median1day = c(12,NA,0,1288)
+                         )
 
-rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive")
-colnames(resultsDf) <- c("Mean","Median","Expected Mean","Expected Median",
-                         "Mean","Median","Expected Mean","Expected Median",
-                         "Mean","Median","Expected Mean","Expected Median")
+rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "Sobrado","Naive Baseline")
+colnames(resultsDf) <- c("Mean","Median",
+                         "Mean","Median",
+                         "Mean","Median")
 
-# median 12 meters mean 9273
+# median 7 meters mean 60 not imputed 282 NA's Ian all data
+
 library(knitr)
@@
 options(knitr.kable.NA = '-')
-kable(resultsDf, "html") %>%
-  kable_styling("striped") %>%
-  add_header_above(c(" " = 1, "Five Minutes" = 4, "One Hour" = 4,"One Day" = 4))
+kable(resultsDf, "latex") %>%
+  kable_styling(latex_options = "striped") %>%
+  add_header_above(c(" " = 1, "Five Minutes" = 2, "One Hour" = 2,"One Day" = 2))
 
@@
 
-In terms of coverage, it is important to note that the Palmius model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests.
-Palmius' model failed to impute a single value for the day tests.
-Barnett's failed to impute 11% of the missing values.
-MYMETHOD made an imputation for all missing values.
-
-As for MYMETHOD's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
+As for PPMI's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
 As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively.
+The distance expectation values were very similar to the distance scores, albeit approximatley 5% higher.
+Confidence scores ranged from 0.006 to 1.
 
@@
 
+Once we removed the measurements within the 5 minute period of each timestamp, we employed the models to predict the location of the individual at the timestamped period.
 The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters.
-On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a median distance of 5637.
-The expected distance has a mean of 6432 and a median of 1037.
+On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a mean distance of 5637.
+Again the expectation of the distance was quite similar to the distance with a mean of 6432 and a median of 1037 meters.
 
@@
 
-### Effect on aggregate measures
+### Example: effect on aggregate measures
 
-Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled.
-As we discussed earlier, aggregations without data processing can be highly biased.
-However, as an example we calculate time spent at home of the user for each day of the week in the month of February (\@ref(fig:aggrePlot)).
-Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as the amount of time spent at home.
+As an example we calculated the time spent at home of the user in the month of March (\@ref(fig:aggrePlot)) without any model and with all three of the investigated methods.
 
-```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017.
-We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month.
-Missing values are filled in by the previous observed value.
-Behavioural trends are evident, the user spends more time at home on weekends than on weekdays.
-Moreover, the variance is greater on weekends, due to travel.",out.width = '100%'}
-include_graphics("img/aggPlot.png")
+```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in March 2017.
+The raw values are estimated by downsampling temporally the lattitude and longitude for every 5 minute time period in the month.
+We used each method's own binning method and classified as at home if the downsampled measurement was within 250 meters from home.",out.width = '100%'}
+
+include_graphics("img/timeUse.png")
 ```
 
-## Discussion
+Interestingly, all three models suggest that the user spent approximately 60% of their time at home.
+However, without using any form of missing data imputation approximately 12% of the time is unaccounted for.
+Given that the amount of time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016] and the onset of depressive episodes in bipolar patients [@palmius_detecting_2017] obtaining an accurate value for mobility metrics like this is highly important to social scientists.
 
-Overall the MYMETHOD performed well, although not substantially better than the naive model.
+## Discussion & Conclusion
+
+Overall the PMMI performed better than the alternative models, particularly during longer missing periods and with objective data.
+However, it did not perform substantially better than the naive baseline model.
 In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs.
-Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation.
+Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone GPS logs.
 
-In addition to higher accuracy under the conditions of secondary logs, the advantages of MYMETHOD are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smoothen movement noise.
+In addition to higher accuracy under the conditions typical of secondary logs, the advantages of PPMI are increased coverage and flexibility for missing data imputation, robustness to irregular sampling, the ability to model complex non-linear interactions in its imputations, and the ability to use historical records to smooth movement noise.
 
-MYMETHOD's increased coverage and flexibility comes from its ability to make complex non-linear predictions.
+PPMI's increased coverage and flexibility comes from its ability to make complex non-linear predictions.
 For instance, in a given missing period it might make sense to predict that the individual is either at home, or at the office, or at a shop with equal probability.
-While MYMETHOD can make such an imputation, none of the alternative methods can do this.
+While PPMI can make such an imputation, none of the alternative methods can do this.
 Moreover, the ability to take the prediction probability values from the neural network also helps in dealing with uncertainty.
-A known-drawback of single imputation is that it takes an imputed value and treats it as observed (*RUBIN*).
+A known-drawback of single imputation is that it takes an imputed value and treats it as observed.
 Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods.
-With MYMETHOD it is possible to construct confidence intervals using the predicted probabilities of each estimate.
+With PPMI it is possible to model uncertainty using the predicted probabilities of each estimate.
+For instance, in the previous example, we could choose to only take estimates with a high degree of confidence, thus creating confidence intervals by adding and substracting the amount of cases where the location of the individual is ambiguous.
 
@@
 Downsampling over 5 minute periods will lead to a value that will be the mean of the two inaccurate samples, which is likely to be a location the individual is certainly not.
-MYMETHOD instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations.
+PPMI instead downsamples spatially, which ensures that the binned location is one which is composed of the mean of hundreds of observations, not just the few that happen within a single period.
 
-While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for paths.
-Hence all paths are calculated based on sensor input only.
-This may be feasible with high frequency regularly sampled short duration logs but creates noise with secondary logs.
+While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for non-pause locations.
+This may be feasible with high frequency, regularly sampled short duration logs but creates noise with secondary logs.
 Moreover, with secondary logs it is feasible to spatially "average out" multiple samples of the same path in order to recreate it in its entirety.
-For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (*report value here*) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
+For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (about 0.01 Hz) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
 
@@
 
-Furthermore, MYMETHOD also has several limitations.
-The pause extracting function, the path clustering function, the classification model and the assignment function remain simplistic and could be improved.
-Indeed, our intention is not to develop an optimal classification model or any of the other components of MYMETHOD, but rather to show a new way in which secondary logs can be made useful by social scientists.
+Furthermore, PPMI can be further developed. The map building function, the assignment function and the classification model remain simplistic and could be improved.
 
-With regards to pause extraction, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location.
+In map building, the probability of a pause at a given location is certainly related to other factors, such as the time of the day as well as the prior history of pauses at that location.
 These factors are not taken into account in the pause extraction function.
 Improved methods would do well to do so.
+As for paths, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
+It does not take into account the length of the path as well as the average sampling frequency of the path.
+This is an issue because it can lead to bins to which data points are seldomly assigned.
+For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements.
+This leads to difficulties in the classification part of the model, as unfrequently observed clusters are hard for the model to predict.
 
-As for the path clustering function, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
-It does not take into account the length of the path as well as the average sampling frequency of the path.
-This is an issue because it can lead to clusters to which data points are seldomly assigned.
-For example, while the Amsterdam-Utrecht line has been mapped out almost perfectly, many of the clusters along the route have only been assigned few measurements.
-This leads to difficulties in the machine learning part of the model, as unfrequently observed clusters are hard for the model to predict.
-
-Similarly, a limitation of the current assignment function is that it only takes into account the distance between the observed measurement and all the clusters when assigning each measurement to a cluster.
+The current assignment function simply assigns each masurement to the nearest bin.
 It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on.
@@
 By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC.
-An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and the ANN model.
-This could take the form of a Kalman filter.
-In this case the state side of the model would represent a probabilistic representation of where the individual could be based on what the ANN has extracted from the individuals movement history.
+An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and a model of the individuals movement history.
+In terms of state-space models like the Kalman filter the state equation would represent a probabilistic representation of where the individual could be based on the individuals entire movement history.
 The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.
 
-As for the simplicity of the neural network which was used to generate predictions.
+As for the simplicity of the classification method, the neural network which was used to generate predictions used no information on sequence patterns longer than the previous and next bin.
 A more sophisticated recurrant neural network (RNN), or a long short-term memory recurrant neural network (LSTM) would likely perform significantly better.
 
-## Conclusion
-
-Advantages of MYMETHOD:
-
-1. Full imputation coverage.
-2. Probability of cluster as measure of uncertainty.
-3. Reduction of data drift through discretisation.
-4. Robust to irregular sampling.
-5. Few assumptions.
-
-Disadvantages:
-
-1. Weak in unfrequently sampled locations. Perhaps use alternative methods there (e.g. Kalman filter)
-2. Some sacrifice of accuracy for consistency. See high vs low resolution pause maps.
-
-Further research:
-
-1. More sophisticated pause/path extracted algorithms.
-2. More sophisiticated ML models.
-3. Introducing explicit assumptions about paths.
-4. Run a study where individuals have a high frequency sampling high accuracy GPS receiver on them (test out of sample performance).
-5. Path extraction for train stuff.
-
-*To do: add info on paragraphs and sections/subsections.*
+That there is room for improvement is unsurprising given that it is still early days for using smartphone location measurements in social science.
+Nonetheless, the methodological advantages are clear: millions of individuals have a location log containing objective measurements spanning years which can be accessed for free.
+This is vastly superior to alternatives such as bias heavy questionnaires or expensive research.
+Social science researchers must take advantage of regulatory changes in data portability regulations and put the vast wealth of data collected by commercial entities to public use.

### 2018-04-29 report/thesisReportApa.tex
APA manuscript

@@
 
-  \def\affdep{{""}}%
-  \def\affcity{{""}}%
+  % \def\affdep{{""}}%
+  % \def\affcity{{""}}%
 
@@
   \authornote{
-    \newcounter{author}
     Department of Methodology \& Statistics
@@
 
-                      Correspondence concerning this article should be addressed to Boaz Sobrado. E-mail: \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
+    Correspondence concerning this article should be addressed to Boaz
+    Sobrado, . E-mail:
+    \href{mailto:boaz@boazsobrado.com}{\nolinkurl{boaz@boazsobrado.com}}
   }
@@
 \usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}{Lemma}[section]
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{corollary}{Corollary}
-\newtheorem{proposition}{Proposition}
+\newtheorem{definition}{Definition}[section]
+\newtheorem{corollary}{Corollary}[section]
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
-\newtheorem{example}{Example}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
-\newtheorem{exercise}{Exercise}
+\newtheorem{exercise}{Exercise}[section]
 \theoremstyle{remark}

### 2018-04-29 implementingBipolar/aggregateResults.R

@@ -1,0 +1,138 @@
+# results: aggregation time spent at home
+
+leaflet(allPoints) %>% 
+  addTiles() %>% 
+  addCircles(lng = ~lon, lat = ~lat, label = ~as.character(clust))
+
+home <- c(5.114051,52.10415)
+
+test <- fin %>% 
+  as_tbl_time(index = time) %>%
+  filter_time(~ "2017-03") %>% 
+  select(time, clust, lon, lat) %>% 
+  thicken("5 min", "thick") %>% 
+  left_join(., allPoints, by = "clust", suffix = c(".measured",".clust")) %>% 
+  mutate( distHClust = raster::pointDistance(p1 = home, p2 = as.matrix(cbind(lon.clust,lat.clust)), lonlat = T))
+
+library(ggplot2)
+plot(test %>%as_tbl_time(time) %>% 
+       filter_time(~"2017-03-07") %>% pull(distH))
+
+test %>%as_tbl_time(time) %>% 
+  filter_time(~"2017-03-07") %>%
+  filter(distH<400) %>% 
+leaflet() %>% 
+  addTiles() %>% 
+  addCircles(lng = ~lon.clust, lat = ~lat.clust, label = ~as.character(clust))
+
+#clusts 49, 348 and 336
+
+test2 <- test %>% group_by(thick) %>% 
+  summarise(lon  = mean(lon.measured),
+            lat  = mean(lat.measured),
+            clust = Mode(clust),
+            home1 = clust %in% c(49,348,336),
+            home2 = raster::pointDistance(p1 = home, p2 = as.matrix(cbind(lon,lat)), lonlat = T))
+
+test2 %>%  group_by(home1) %>% summarise(m = min(home2), mean = mean(home2), max = max(home2))  
+
+test2 %>%  filter ( home1 == T) %>% 
+  leaflet() %>% 
+  addTiles() %>% 
+  addCircles(lng = ~lon, lat = ~lat)
+
+
+test2 %>% pad() %>% summarise(atHomeProp = sum(home1, na.rm = T)/nrow(.),
+                              missingProp = sum(is.na(home1))/nrow(.))
+
+# predicted values
+t <- test2 %>%  pad()
+t[which(t$thick %in% rems),"clust"] <- as.integer(model %>% predict_classes(pred))
+
+# model prediction
+t %>% mutate( home1 = clust %in% c(49,348,336)) %>% summarise(atHomeProp = sum(home1)/nrow(.))
+
[…]

### 2018-04-29 implementingBipolar/fullPalmiusResults.R

@@ -12,7 +12,7 @@ file.sources <- list.files(c("functions"),
 sapply(file.sources,source,.GlobalEnv)
 
 #select the month of march
-
+data <-  fin
 dataExample <- select(data,time,lat,lon,accuracy,nextMeas) %>%
   arrange(time) %>% 
   mutate( nextLon = lead(lon),

### 2018-04-29 ownAlgo/getTrainFreq.R

@@ -1,0 +1,45 @@
+# train line sampling frequency
+
+library(readr)
+objLog<- read_csv2("objectiveLog.csv")
+
+objLog <- objLog %>%
+  mutate( time = as.POSIXct(paste(Date, Time), format="%d-%m-%Y %H:%M:%S")) %>% 
+  as_tbl_time(time)
+
+objLog %>% filter(Location %in% c("Utrecht Centraal", "Amsterdam Centraal"))
+
+ts<- objLog %>% arrange(time) %>% 
+  mutate( nextStop = lead(Location),
+          nextType = lead(Type),
+          nextTime = lead(time)) %>% 
+  filter( (Location == "Utrecht Centraal" & nextStop == "Amsterdam Centraal")|
+          (Location == "Amsterdam Centraal" & nextStop == "Utrecht Centraal")  
+           ) %>% 
+  mutate(duration = nextTime - time) %>% 
+  filter(Type == "Check-in") %>% 
+  select(time,nextTime)
+
+timeInput <- ts[1,1][[1]]
+nextTime <- ts[1,2][[1]]
+
+getSampFreq <- function(timeInput,nextTime){
+  duration <- as.numeric(nextTime)-as.numeric(timeInput)
+  fin %>% 
+    filter(timestampMs >= as.numeric(timeInput) & timestampMs <= as.numeric(nextTime)) %>% 
+    summarise(count = n(),
+              hz = count/duration) %>% pull(hz)->sampFreq
+  return(sampFreq)
+}
+
+hzs<- vector()
+for( i in 1:nrow(ts)){
+  hzs[i]<- getSampFreq(ts[i,1][[1]],ts[i,2][[1]])
+}
+
+mean(hzs)
+
+r_refs(file = "../manuscript/r-references.bib")
+my_citations <- cite_r(file = "r-references.bib")
+
+,"r-references.bib"

### 2018-04-29 ownAlgo/imgCreation.R

@@ -1,0 +1,3 @@
+# image creation
+install.packages("magick")
+library(rmagick)

### 2018-04-29 ownAlgo/modifiedSTS.R

@@ -18,7 +18,7 @@ test <- data %>% mutate( distanceHome = raster::pointDistance(matrix(c(lon,lat),
                                                               home,
                                                               longlat = T),
                          timestampMs = as.numeric(time))%>%
-  filter(distanceHome < 300000) %>% 
+  filter(distanceHome < 200000) %>% 
   select(time,lon,lat,accuracy, timestampMs,distanceHome) %>%
   arrange(timestampMs)
 
@@ -47,7 +47,9 @@ toClust <- test2 %>% filter( isPause == 1  & !(distMeanEast > 2*meanAcc |
 colnames(toClust) <- c("lon","lat")
 
 # merge all clusters within 300 meters of each other  
-clusterMap <- mergePoints(toClust, d = 300)
+#clusterMap <- mergePoints(toClust, d = 300)
+clusterMap1 <- mergePoints(toClust, d = 150)
+clusterMap2 <- mergePoints(toClust, d = 450)
 
 # all measurements binned into a pause cluster
 test3<- binMeasurements(test,clusterMap)
@@ -56,7 +58,7 @@ test3<- binMeasurements(test,clusterMap)
 test3 <- featureFunc2(test3,test2)
 # all points method of using rasters for computational help
 
-pathClust<- gridFunc(test3)
+pathClust<- gridFunc(test3,d = 500)
 
 routePoints <- pathClust %>% mutate(clust = 1:nrow(pathClust))
 allPoints <- routePoints %>%
@@ -80,4 +82,4 @@ rm(pause,notpause,routePoints,pathClust,test3,clusterMap,toClust,test2)
 Sys.time()
 
 saveRDS(fin,file = "finalBinnedDataNL.rds")
-saveRDS(allPoints, file = "finalAllClusts.rds")
+saveRDS(allPoints, file = "finalAllClusts2.rds")

### 2018-04-29 ownAlgo/objectiveResults.R

@@ -1,4 +1,5 @@
 # objective data comparison
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # read in data
 Sys.time()# 14:46:00
@@ -9,8 +10,6 @@ objLog<- read_csv2("objectiveLog.csv")
 objKey<- read_csv2("objectiveKey.csv") %>% select(Location,Cluster)
 
 objKey<- left_join(objKey,allPoints, by = c("Cluster" = "clust")) %>% select(Location,Cluster,lon,lat)
-
-strftime(paste0(objLog$Date," ",objLog$Time),"%d-%m-%Y %H:%M:%S")
 
 
 
@@ -216,3 +215,4 @@ imputedP %>% slice(remove_ind) %>%
   left_join(., objLog %>%thicken("5 min", "thick5") %>% mutate(timestampMs = as.numeric(thick5)), by = "timestampMs") %>% 
   mutate( dist = raster::pointDistance(p1 = as.matrix(cbind(impLon,impLat)), p2 = as.matrix(cbind(lon,lat)),lonlat = T )) %>% 
   summary()
+

### 2018-04-29 ownAlgo/resultsIan.R

@@ -13,6 +13,7 @@ file.sources = list.files(pattern="*.R",path = "gpsmobility/R/")
 setwd("gpsmobility/R/")
 sapply(file.sources,source,.GlobalEnv)
 
+data <- fin %>% select(timestampMs,time,lat,lon,accuracy)
 downsampledIan <- downSampleMean(data,interval = '300 sec') #downsamples and adds XY coordinates
 
 
@@ -20,30 +21,34 @@ downsampledIan <- downSampleMean(data,interval = '300 sec') #downsamples and add
 
 # 5 minute bits
 
+Sys.time()
 d2<- downsampledIan %>%
   mutate( time = as.POSIXct(timestampMs, origin="1970-01-01")) %>% 
   pull(time)
 
 remove_ind <- which(as.character(d2) %in% remove_ind_5min[[2]])
 
-downsampled5minRemoved<- downsampledIan
+downsampleSys.time()
+d5minRemoved<- downsampledIan
 downsampled5minRemoved[remove_ind,c("lon","lat","x_v","y_v")] <- NA
 downsampled5minRemoved[remove_ind,c("code")] <- 4
 
 #params
 minpausedur=120
 minpausedist=50
-
+Sys.time()
 mobmatmiss<- pauseFlight(downsampled5minRemoved, r = sqrt(300), w = mean(data$accuracy)) # extracts pauses and flights using algo
+Sys.time()
 mobmatmiss <- mobmatmiss[-nrow(mobmatmiss),] #clean a bit
 
 # Guessing Pauses
-#with example data first day
+Sys.time()
 mobmat = GuessPause(mobmatmiss,mindur=minpausedur,r=minpausedist)
-
+Sys.time()
 obj=InitializeParams(mobmat)
 spread_pars=c(10,1)
 wtype <- "TL"
+Sys.time()
 out3 <- SimulateMobilityGaps(mobmat,obj,wtype,spread_pars) # it is returning NA's only for
 
 evalDf<- evalIan(out3,downsampledIan = downsampledIan)
@@ -132,11 +137,7 @@ wtype <- "TL"
 out3 <- SimulateMobilityGaps(mobmat,obj,wtype,spread_pars)
 evalDf<- evalIan(out3,downsampledIan = downsampledIan)
 
-evalDf[remove_ind,] %>% summary() # median 12 meters mean 9273
[…]

### 2018-04-29 scripts/dailyMissPlot.R

@@ -1,10 +1,10 @@
 # daily missing plot 
 
 # by 5 minute version 2
-m5minv2<- all %>%
+m5minv2<- fin %>%
   select(time,accuracy) %>%
   as_tbl_time(index = time) %>%
-  time_filter(2000-01 ~ 2017-06) %>% 
+  filter_time("2000-01" ~ "2017-06") %>% 
   thicken('5 min') %>%
   group_by(time_5_min) %>%
   summarise(measurements = n(),
@@ -23,19 +23,19 @@ m5minv2<- all %>%
 
 exampleMiss<- m5minv2 %>%
   as_tbl_time(index = time2)%>%
-  time_filter(2017-02-15 ~ 2017-02-15) %>%
-  ggplot( aes(x = hour, y=measurements, colour = factor(missing)))+
+  filter_time("2017-02-15" ~ "2017-02-15") %>%
+  ggplot( aes(x = hour, y=measurements))+
   geom_point()+theme_tufte()+
   xlab("")+
   ylab("Measurements")+
   ggtitle("Measurements per 5 minute window",subtitle = "February 15 2017")+
   scale_x_datetime(breaks = date_breaks("4 hour"),
                    minor_breaks=date_breaks("2 hour"),
-                   labels=date_format("%H:%M:%S", tz = "Asia/Singapore"))+
-  scale_colour_manual(values=c("black",muted("red")),labels=c("Not Missing", "Missing"))+
-  labs(x = NULL, colour = "")+xlab("Time")
+                   labels=date_format("%H:%M", tz = "Europe/Budapest"))+
+  labs(x = NULL, colour = "")+xlab("Time")+
+  scale_y_continuous(breaks = seq(0, 9, by = 2), name = "Measurements")
 
-ggsave(exampleMiss,filename = "../img/missingBoaz5minExample.png",device = "png",height = 10, width = 20, units = "cm")
+ggsave(exampleMiss,filename = "../manuscript/img/missin5minExample2018.png",device = "png",height = 10, width = 20, units = "cm")
 
 # randomly select two days
 sample(x = c(13,14,15,16,17),1)#YEAR

### 2018-05-03 manuscript/apaManu.Rmd
last commit before preparation of research archive

@@
 ---
-title             : "Handling missing data in smartphone location logs"
+title             : "Personalised Map Matched Imputation: imputing missing data from smartphone location logs"
 shorttitle        : "Missing Data in Smartphone Location Logs"
@@
 keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
-wordcount         : "2297"
+wordcount         : "8319"
 
@@
 library(kableExtra)
+library(sp)
+library(leaflet)
+library(raster)
+library(geosphere)
 
@@
 Why is human mobility important?
-How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010].
-Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures.
-The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip].
+Human mobility measures are quantified metrics of how people move about in their environment.
+Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010].
+The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@goldberg_international_2006].
+At their root these questions are mobility measures.
 In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g.
 @villanueva_association_2013].
-Moreover, economists study geographic labour mobility extensively [e.g.
-@tatsiramos].
-In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission].
-The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes.
+Economists are interested not only in where people live, but also where they go to work: geographic labour mobility is related to the income of individuals, the well-being of a country's economy and even informs the policy of bodies such as the European Comission [@tatsiramos_geographic_2009].
+The extensive use of measures like these across different domains in social science strongly suggests that mobility metrics are linked to important real world outcomes.
 Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017].
@@
 
-Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries.
-Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting.
-The burdensomeness stems from the fact that participants must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness.
-This makes the data expensive to collect and limits the extent of data collection in practice.
+Despite the importance of mobility measures, the majority of these metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries.
+As a thought experiment we encourage the reader to try to remember where exactly he or she was five weeks ago on Saturday at midday.
+Most people find this sort of task very difficult (if not impossible), which illustrates that questionnaires and travel diaries are have well-known methodological flaws.
+This method of collecting data is difficult and relies on accurate self-reporting.
+Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals due to human forgetfulness.
+This makes the data collection expensive and limits the duration of data collection in practice.
 In addition, the frequent reporting duties of the participant may bias the participants behaviour.
-Participant forgetfulness also limits the accuracy of self reported measures.
-There is clear evidence that participants are systematically biased when self-reporting mobility measures.
+The limits of human cognition also limit the accuracy of self-reported measures.
+There is evidence that participants are systematically biased when self-reporting mobility measures.
 For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017].
-These obstacles can be overcome by using objective data on human mobility.
+These obstacles can only be overcome by using objective data on human mobility.
 
@@
 @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour.
-Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016.
+Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016].
 Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers.
@@
 The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre.
-Both journeys can be seen in Figure 1.
+Both journeys can be seen in Figure 2.
 The large jump between journey 5 and 6 is measurement error.
@@
 If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
-This issue is also visualised in Figure \@ref(fig:palmiusVme1).
+This issue is also visualised in Figure \@ref(fig:palmiusVsMe1).
 The problem is further complicated because missing data and noisy measurements are related.
@@
 These often come from climate or geological research and rely on spatiotemporal auto-correlations.
-For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ].
+For instance, @feng_cutoff:_2014 estimate missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors.
 The authors illustrate their example using rainfall data from gauging stations across Australia.
@@
 
-On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu].
+On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @@wu_spatial-temporal-semantic_2017].
 Importantly, none of them worked with secondary logs.
@@
 This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses.
-In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours (*circadian proximity*).
+In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours.
 
-On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement.
+It is also worth mentioning that there is a body of work that incorporates road features and other background information to generate predictions or imputations.
+For instance, @liao_learning_2007 use hierarchical Markov models in combination with knowledge about the transportation system to bridge the gap between raw GPS sensor measurements and high level information such as a user's destination and mode of transportation.
+This can be then used to impute or predict missing steps.
+Along similar lines @wu_spatial-temporal-semantic_2017 use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement.
 While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins.
-Working with high-frequency measurements, @wu's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways.
+Working with high-frequency measurements, @wu_spatial-temporal-semantic_2017's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways.
 This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.
+Subsequently we will focus on methods which do not incorporate external information as they are more generally applicable.
 
@@
 Moreover, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs.
-Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu].
+Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [@palmius_detecting_2017 ;@barnett_inferring_2016; @wu_spatial-temporal-semantic_2017].
 
@@
 
-Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences.
-There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007;].
+Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences.
+There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007].
 We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
@@
 
-@wu's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads.
+@wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads.
 In other words, in addition to the measured data they also use a geographic database that contains information about the area in which the individual is (e.g.
 where precisely the roads are), and sort measurements into bins based on both the measurement and the geographic data base.
-For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu's method will assume that the individual is on the road A.
+For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu_spatial-temporal-semantic_2017's method will assume that the individual is on the road A.
 
 PMMI uses a similar logic, but without using any external geographic database.
-The key modification in PMMI is that whilst @wu uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map.
+The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map.
 This map is subsequently used to bin measurements.
@@
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
-For this day there were several periods with no information.
+The entire log contains several long periods with no information.
 These points are filled with red and lie on the x-axis.",out.width = '100%'}
@@
 include_graphics("img/missingdayBoaz5min.png")
-my_citations <- cite_r(file = "r-references.bib")
+my_citations <- cite_r(file = "r-references.bib", footnote = TRUE)
 ```
 
-For simplicity, we subsequently used a time period when the individual was living in the Netherlands.
-This leaves us with 156,000 measurements over a period of less than six months.
-Analyses as well as implementations and adaptions @palmius_detecting_2017's and @barnett_inferring_2016's model were performed using R and a multitude of other statistical packages `r my_citations`.
-All the code is available on a public repository [@mygithub].
+For simplicity, we subsequently used a time period when the individual was living in the Netherlands. This subset contains 156,000 measurements over a period of less than six months.
+
+@palmius_detecting_2017's algorithm was implemented in R based on the original MATLAB code and pseudocode kindly provided by the author.
+@barnett_inferring_2016's method was slightly adapted in R to fit the the data structure from the original R code provided by the author.
+When executing their models we used the same parameters as the authors did.
+To represent @barnett_inferring_2016's model we used the variant where movements were sampled based on spatial proximity.
+
+All analyses were performed using `r my_citations$r`.
+
+`r my_citations$pkgs`
+
+All the code is available on a public repository [@sobrado_my_2018].
 
@@
 1. Evaluating the performance of the map building and assigning functions.
-2.
-Comparing the performance of PMMI using a) baseline models b) performance with randomly removed data in comparison to the aforementioned methods [@palmius_detecting_2017; @barnett_inferring_2016] and c) objective ground-truth data (public transportation time-stamps).
+2. Comparing the performance of PMMI compared to alternative methods [[@palmius_detecting_2017; @barnett_inferring_2016]] and a baseline model using cross-validation.
+3. Comparing the performance of PPMI compared to the aformentioned methods using objective ground-truth data (public transportation time-stamps locations).
 
@@
 
-Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been cleaned.
-Otherwise we run the risk of over-fitting the model in the sense that we will measure the extent to which an imputation method can correctly impute measurement noise within $D_t$ instead of true location $G(t)$.
+Before we can evaluate the accuracy of the imputations, it is essential to evaluate how well noise in the data has been smoothed.
+All methods smoothen noise differently in an attempt to extract true location $G(t)$ from measurements $D_t$.
+In the subsequent cross-validation step we will remove random smoothened blocks and compare the imputed value to the smoothened value.
+For this reason we want to ensure that the smoothened method actually does reflect the true location $G(t)$.
+Otherwise we run the risk of over-estimating the accuracy of a model in the sense that we will compute the extent to which an imputation method can correctly impute a smoothened value which has little to do with the true location $G(t)$.
 
 In order to evaluate the map building and binning we will first visually evaluate the paths and pause locations.
-A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy (e.g.
-@lcm).
+A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy [e.g.
+@brunsdon_path_2007].
 Then, let the average distance between the actual measured point and the binned point be the *deviation distance* $\delta_{\text{dev}}$.
@@
 
-### Imputation algorithm performance
+### Imputation algorithm performance in cross-validation.
 
 We will compare the performance of PMMI, @palmius_detecting_2017 and @barnett_inferring_2016.
-In addition, we will also compute a naive model, which simply imputes as the missing value the previous observed value.
-The naive model will serve as a baseline model.
-To compare the performance of these methods we will remove 25% of measured time intervals at random within a four week period.
+In addition, we will also compute two baseline models.
+The first one is a naive model, which simply imputes the previous observed value.
+The second one we will call the "home" model, which imputes that the individual is at home at all missing timepoints.
+Using these two models as baselines makes sense given the high degree of autocorrelation in the data and that individuals tend to spend most of their time at home.
+To compare the performance of the aforementioned methods we will remove 25% of measured time intervals at random within a four week period.
 We will make our comparisons for intervals of 5 minutes, 1 hour and 1 day.
-In other words, we will remove 25% of time intervals at random, while varying the size of the time intervals removed.
+In other words, we will remove 25% of time intervals at random, while varying the duration of the time intervals removed.
 
@@
 
-To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location).
+To compare all methods with each other we will compute a distance measure (how far was the removed location from the predicted location) and the coverage.
+The coverage refers to the percentage of missing cases which were imputed.
 For PMMI's imputations we will use a weighted mean for the distance measures whereby each 5 minute period is weighted equally.
-This is necessary because the other two model estimate far fewer values than PMMI.
-While @barnett_inferring_2016's and @palmius_detecting_2017's models estimate 12 measurements for each missing hour, PMMI estimates as many measurements as there are in the log.
-When individuals go to infrequently travelled locations they tend to use their phone's location services more than 100 more times, which leads to more measurements, and hence more imputed values.
-Hence, if the weights are not used PMMI is susceptible to error overestimates which are related to measurement frequency in a way that the other two methods are not.
+This is necessary because the other two models downsample temporally to 5 minutes and because the frequency of measurements in a 5 minute period is correlated with imprecise measurements and visiting infrequently travelled locations.
+Hence weights must be used to avoid inflating error rates for PPMI in comparison to the other models.
 
-In addition, other measures of interest for PMMI are accuracy (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*.
+In addition, other measures of interest for PMMI are *accuracy* (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*.
 The confidence is the probability with which the model predicts the most likely cluster.
-For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say it has high confidence in the prediction.
+For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say this is a high confidence prediction.
 Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster.
-For example, if the true location of an individual is bin A (bin A is 10 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 1 meter.
+For example, if the true location of an individual is bin A (bin A is 1000 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 100 meters.
 
-Finally, we will take objective real world data and compare it to predicted values.
-We will use information from the Dutch public transportation card.
-The Dutch public transportation service provides users with time-stamped data of when and where they board, change lines or leave public transportation.
+### Imputation algorithm performance using external benchmarks.
+
+We will compare the performance of PMMI, @palmius_detecting_2017, @barnett_inferring_2016 and the naive baseline model with regards to imputing real missing measurments.
+To do will use information from the Dutch public transportation card.
+The Dutch public transportation service provides users with time-stamped location data of when and where they board, change lines or leave public transportation.
+In this paper we used 97 such events in a single month.
+
 To be able to make a comparison between models we will remove all measurements from within the 5 minute period that a time-stamped measurement is available.
-Then we will use each model to impute the location of the individual within that period.
+Then we will use each model to impute the location of the individual within that period and compare it to the true location.
 
@@
 
-The selection of these parameters was more-or-less heuristically driven based on their on their ability to extract a meaningful personalised map.
+The model was tuned based on a visual inspection of the resulting personal map meeting the following criteria:
+
+1. Plausible extraction of pause locations. For example, a pause in a field with no path leading to it is less plausible than a pause at a train station, airport, or office building.
+2. A suitable extraction of paths. For instance, are ensuring that individual paths are complete and not merged unneccessarily in the case of closely parallel paths.
+3.
+Less than a fifth of extracted clusters are assigned only one measurement.
+Unique clusters are unavoidable as people sometimes visit a location briefly and only once, nonetheless they pose problems for the classification algorithms.
+
 When selecting parameters there is a trade off is between bias and precision.
-This is because an increase in precision in the form of a higher for high density locations resolution comes at the expense of precision as assigning measurements to bins becomes more difficult.
-For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the expense of falsely categorising certain measurements to the wrong cluster.
+This is because an increase in precision (in the form of a higher resolution of locations location) comes at the expense of precision (assigning measurements to bins becomes more difficult).
+For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the risk of falsely categorising certain measurements to the wrong bin.
 This is illustrated in Figure \@ref(fig:pauseLocUithof1).
@@
 For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters.
-Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical $\frac{2}{3}$ value that is expected.
-Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}}$.
+Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical 67% value that is expected.
+Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}} = 300$.
 
-For the narrower period of March 2017 approximately 74% of the deviance values are within their corresponding accuracy values, which is close to the theoretical $\frac{2}{3}$ value that is expected.
-The raw unweighted mean and median deviance are 38 and 14 meters respectively.
-If we weight them such that each 5 minute interval is weighted equally to ease comparison with the other two methods, we find a mean deviance of 35 meters with a median of 12.8.
-
-In comparison, Palmius' method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers.
-On the other hand, Barnett's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
-Barnett's deviance is necessarily higher than Palmius', as they down-sample temporally (like Palmius) and subsequently aggregate into pauses and linear flights.
+In comparison, @palmius_detecting_2017's method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers.
+On the other hand, @barnett_inferring_2016's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
+@barnett_inferring_2016's deviance is necessarily higher than @palmius_detecting_2017', as they down-sample temporally (like @palmius_detecting_2017) and subsequently aggregate into pauses and linear flights.
 
@@
 Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each down-sampled interval.
-The fact that over 90% of deviance values are within accuracy (much higher than the expected theoretical 67%) confirms that temporal down-sampling is not sufficiently filtering out the noise.
+The fact that over 90% of deviance values are within accuracy (substantially higher than the expected theoretical 67%) confirms that temporal down-sampling is not sufficiently filtering out the noise.
 
-```{r, palmiusVsMe1,echo = FALSE,fig.cap="Excerpt. MISSING CAPTION HERE",out.width = '100%'}
+```{r, palmiusVsMe1,echo = FALSE,fig.cap="The difference between temporal and spatial downsampling.
+The blue circles are raw measurments, the yellow circles are temporally downsampled locations.
+Spatially downsampled locations are in red.
+Due to measurement sparcity and inaccuracy many of the temporally downsampled locations are in unfeasable locations.",out.width = '100%'}
 #fix width
@@
 
-### Imputation evaluation
+### Imputation algorithm performance evaluation in cross-validation.
 
-The @palmius_detecting_2017 model failed to impute 3% of all removed values for both the 5 minute and over 10% for the 1 hour tests.
-@palmius_detecting_2017' model failed to impute a single value for the day tests.
-Similarly, @barnett_inferring_2016's method failed to impute over 11% of the missing values.
-PPMI made an imputation for all missing values.
-Table 1 shows the results of the distance metrics for each method.
+The first factor we must take into account is the coverage of each imputation model.
+The results show us that while PPMI and the naive baseline model impute all missing values, the other two models have difficulties doing so (Table 1).
+In particular, the models of @palmius_detecting_2017 and @barnett_inferring_2016's fail to impute an increasing amount of missing values with an increase in the duration of the missing period.
+@palmius_detecting_2017's method in particular failed to impute a single missing time period.
+The longer a log, the higher the probability that long time periods will be missing.
+Secondary logs are typically long.
+Hence missing data imputation methods designed to work with secondary logs must be capable of imputing longer time periods.
 
@@
 
-resultsDf <- data.frame( mean5min = c(82,43,269,426),
-                         median5min = c(4,0,0,0),
-                         mean1hr = c(345,497,908,1502),
-                         median1hr = c(6,4,0,0),
-                         mean1day = c(9273,NA,5757,14266),
-                         median1day = c(12,NA,0,1288)
+resultsDf <- data.frame( mean5min = c(240,43,269,426,5599),
+                         median5min = c(5,9,0,0,0),
+                         imputed5min = c("97%","97%","100%","100%","100%"),
+                         mean1hr = c(33,497,908,1502,5667),
+                         median1hr = c(6,4,0,0,0),
+                         imputed1hr = c("88%","89%","100%","100%","100%"),
+                         mean1day = c(62,NA,5757,14266,5757),
+                         median1day = c(6,NA,0,1288,0),
+                         imputed1day = c("85%","0%","100%","100%","100%")
                          )
 
-rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "PPMI","Naive Baseline")
-colnames(resultsDf) <- c("Mean","Median",
-                         "Mean","Median",
-                         "Mean","Median")
+rownames(resultsDf) <- c("Barnett & Onella", "Palmius", "PPMI","Naive Baseline","Home Baseline")
+colnames(resultsDf) <- c("Mean","Median","Coverage",
+                         "Mean","Median","Coverage",
+                         "Mean","Median","Coverage")
 
-# median 7 meters mean 60 not imputed 282 NA's Ian all data
 options(knitr.kable.NA = '-')
-apa_table(resultsDf, col_spanners = list(`Five minutes` = c(2, 3), `One Hour` = c(4, 5),`One Day` = c(6,7)), placement = "htb", digits = 0, caption = "Distance in meters between the removed time period and the imputed value.")
+apa_table(resultsDf, col_spanners = list(`Five minutes` = c(2, 4), `One Hour` = c(5, 7),`One Day` = c(8,10)), placement = "htb", digits = 0, caption = "Distance in meters between the removed time period and the imputed value.",landscape = T)
 ```
 
-In terms of PPMI's accuracy in this period, the prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
-As a comparison, the naive model's accuracy ratings were 87%, 68% and 24% respectively.
-The distance expectation values were very similar to the distance scores, albeit approximately 5% higher.
-Confidence scores ranged from 0.006 to 1.
+With respect to the distance metrics, Table 1 shows the results of the distance metrics for each method.
+We can see that PPMI predicted the location perfectly for over half of the removed values in all three time categories.
+To be precise PPMI's prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
+As a comparison, the baseline naive model's prediction accuracy was 87%, 68% and 22%, and the home baseline model's were 46%, 52% and 40%.
+Although the home baseline model and PPMI achieved similar distance metrics in the 1 day condition, this does not mean that PPMI defaulted to simply predicting the individual is home at all times (only 40% of predictions were home).
+PPMI outperformed the alternative models when taking into account the median scores, but not the mean.
+The explanation is that the alternative models failed to impute more difficult time periods.
 
-### Comparison with objective data
+On the other hand, the confidence score for each time period were heavily skewed left, withh a median of 1.
+Accordingly, the distance expectation (the crossproduct of the model probabilities of an estimate with the distances of the true value to the estimates) values were quite similar to the distance scores, albeit approximately 5% higher.
+Based on this we can say that this implementation of PPMI was overconfident in its predictions.
 
-Once we removed the measurements within the 5 minute period of each time-stamp, we employed the models to predict the location of the individual at the time-stamped period.
-The naive model reaches an accuracy of 16%, with a median distance of 555 meters and mean distance of 3303 meters.
-On the other hand MYMETHOD has an accuracy of 24%, with a median distance of 1037 meters and a mean distance of 5637.
-Again the expectation of the distance was quite similar to the distance with a mean of 6432 and a median of 1037 meters.
+### Imputation algorithm performance evaluation using external benchmarks.
 
-Palmius's model failed to impute 38 out of 97 periods.
-For those it did impute, it showed a mean distance of 1517 meters and a median distance of 1617.
-Barnett's model failed to impute 18 out of 97 periods.
-For those it did impute, it had a mean of 5506 and a median of 1342.
+Once we removed the measurements within the 5 minute period of each time-stamp in the public transportation log, we employed each of the aforementioned methods to impute the location of the individual at the time-stamped period.
+Then we calculated the distance metrics between the true location (based on the public transport log) and the imputed location.
+We did not compute the home baseline model as it would have always been wrong for this subset of data (travelling and being at home are mutually exclusive).
+
+First, the median distance between the imputed value and the true value for the baseline naive model was 555 meters (with a mean of 3303).
+For this method, the prediction accuracy was 16%.
+This means that the binning method places the individual at the location reported by the public transportation travel log in the five minutes prior to the measurement 16% of the time.
+Needless to say the coverage was 100% as all missing values were imputed.
+
+For the PPMI the median distance metric was 1037 meters (mean 5637).
+The prediction accuracy was higher than that of the baseline naive model at 24%.
+Although PPMI imputed more points correctly than the naive baseline model, those it imputed incorrectly were significantly more incorrect.
+All missing values were imputed, hence resulting in a coverage of 100%.
+
+The same could not be said about @palmius_detecting_2017's method, as it failed to impute almost 40% of the time periods.
+This results in a coverage of 60%.
+For the time periods the method did impute, there was a median distance of 1617 meters between the true location and the imputed value (with a mean of 1517 meters).
+
+@barnett_inferring_2016's method performed better in terms of coverage at 81%. The median distance metric was 1342 meters and the mean value was 5506 meters.
 
@@
 
-Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as the amount of time spent at home.
-As an example we calculated the time spent at home of the user in the month of March (Figure \@ref(fig:aggrePlot)) without any model and with all three of the investigated methods.
+Social scientists are most interested in aggregating spatiotemporal data to more socially relevant metrics, such as the amount of time spent at home, frequency of travels to new locations, the amount of trips made or distance covered.
+As an example we calculated the time spent at home of the user without imputing any missing data and with all three of the investigated methods.
+The results can be seen in Figure \@ref(fig:aggrePlot).
+Without any form of missing imputation the individual's time spent at home is unknown for 12% of the time, which adds up to almost four days over the course of a 30 day month.
 
@@
 
-Interestingly, all three models suggest that the user spent approximately 60% of their time at home.
-However, without using any form of missing data imputation approximately 12% of the time is unaccounted for.
-Given that the amount of time spent at home has been found to be a reliable predictor of extroversion [@harari_using_2016] and the onset of depressive episodes in bipolar patients [@palmius_detecting_2017] obtaining an accurate value for mobility metrics like this is highly important to social scientists.
+The amount of time spent at home has been found to be a reliable predictor of extroversion [@harari_using_2016] and the onset of depressive episodes in bipolar patients [@palmius_detecting_2017].
+The 12% of time that remains unaccounted for could mean the difference between an introvert and an extrovert, or a healthy individual and an individual suffering of depression.
+Filling in these gaps is the exactly the sort of problem that can be solved using missing data imputation.
+Interestingly, despite their radically different imputation methods, all three examined models suggest that the user spent approximately 60% of their time at home.
 
@@
 
-Overall the PMMI performed better than the alternative models, particularly during longer missing periods and with objective data.
-However, it did not perform substantially better than the naive baseline model.
-In addition, the comparison to the performance of the Barnett&Onella and Palmius models is somewhat unfair, as they were created for custom logs, not secondary logs.
+Overall the PMMI performed better than the alternative models, particularly during longer missing periods and with objective data gathered from the Dutch public transport service.
+However, PPMI did not perform substantially better than the baseline models for all time periods and with the objective Dutch public transport data, suggesting that the classification method can be improved.
+In addition, the comparison to the performance of the @barnett_inferring_2016 and @palmius_detecting_2017 models is somewhat unfair, as they were created for custom logs, not secondary logs.
 Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone GPS logs.
@@
 A known-drawback of single imputation is that it takes an imputed value and treats it as observed.
-Simple rule-based methods such as Palmius' are essentially algorithmic single imputation methods.
+Simple rule-based methods such as @palmius_detecting_2017' are essentially algorithmic single imputation methods.
 With PPMI it is possible to model uncertainty using the predicted probabilities of each estimate.
@@
 For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (about 0.01 Hz) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
+Map agnisticism makes the model more flexible as it can be used for areas where no good geographical databases are available.
 
@@
 An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and a model of the individuals movement history.
-In terms of state-space models like the Kalman filter the state equation would represent a probabilistic representation of where the individual could be based on the individuals entire movement history.
+In terms of state-space models the state equation would represent a probabilistic representation of where the individual could be at that given time based on the individuals entire movement history.
 The space side of the model would be a measurement equation representing the measurement and the uncertainty surrounding it in the form of $a$.
+This would be a similar implementation to that used in @liao_learning_2007, but rather taking the personalised map instead of information about road networks.
 
 As for the simplicity of the classification method, the neural network which was used to generate predictions used no information on sequence patterns longer than the previous and next bin.
-A more sophisticated recurrent neural network (RNN), or a long short-term memory recurrent neural network (LSTM) would likely perform significantly better.
+It also has no contextual information about how far away the bins are from each other.
+A model with more input variables and a more sophisticated design, such as a recurrent neural network (RNN), or a long short-term memory recurrent neural network (LSTM) would likely perform significantly better.
+Moreover, more inputs can be added from alternative sensors in smartphones such as the accelerometer, barometer, measures of phone activity and so on.
+These can be used to supplement or improve the raw location measurements, or indeed inform the predictive model.
 
-That there is room for improvement is unsurprising given that it is still early days for using smartphone location measurements in social science.
-Nonetheless, the methodological advantages are clear: millions of individuals have a location log containing objective measurements spanning years which can be accessed for free.
-This is vastly superior to alternatives such questionnaires which rely on accurate self-reporting.
-Social science researchers must take advantage of regulatory changes in data portability regulations and put the vast wealth of data collected by commercial entities to public use.
+In conclusion, with this paper we aim to start filling the noticeable gap in the social science methodology literature in using smartphone location logs to study human movement.
+Clearly there is room for improvement, but this is to be expected with new methods.
+We are in the early days of using smartphone location measurements in social science.
+Nonetheless, the methodological advantages are clear: millions of individuals have years long location logs containing objective measurements.
+In addition, these measurements can be easily obtained.
+This is an unprecedented opportunity.
+However, with great opportunity comes a great responsibility, as this data raises difficult questions about privacy that researchers must be prepared to answer.
+Privacy concerns should be addressed, but should not dissuade researchers from developing new methods.
+Objective location logs are vastly superior to alternatives, such as questionnaires and travel diares, which rely on accurate self-reporting.
+Social science researchers must take advantage of regulatory changes in with regard to data portability and put the vast wealth of data collected by commercial entities to scientific use.
 
@@
 # References
-```{r create_r-references}
-r_refs(file = "r-references.bib")
+```{r create_r-references, echo=FALSE}
 ```

### 2018-05-03 ownAlgo/functions/expDist.R

@@ -1,5 +1,6 @@
 # calculate expected distance
 
+dClustM <- distMatrix
 expDist <- function(model,x_test,dClustM){
   
   probs <- model %>% predict_proba(as.matrix(x_test))

### 2018-05-03 ownAlgo/functions/nnFeatureFunc.R

@@ -1,197 +1,0 @@
-# extract features for the neural nets
-getTestTrain <- function(data, remove_ind,nClust = 1222){
-  seconds_in_day <-  24*60*60
-  
-  train <- data[-remove_ind,]
-  test <- data[remove_ind,]
-  
-  # issue: train is right, test isn't!
-  # changed it, check to see if it works!
-  train<- train %>% 
-    select(time,timestampMs,clust) %>% 
-    mutate( day  = strftime(time,"%u"),
-            month = strftime(time,"%m"),
-            secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
-            sinTime = sin(2*pi*secMidnight/seconds_in_day),
-            cosTime = cos(2*pi*secMidnight/seconds_in_day),
-            nextMeas = as.numeric(lead(timestampMs)-timestampMs),
-            prevMeas = as.numeric(timestampMs-lag(timestampMs)),
-            lagClust = lag(clust),
-            leadClust = lead(clust),
-            timestampLast = as.numeric(time),
-            timestampNext = as.numeric(time),
-            clustNext = clust,
-            clustPrev = clust)
-  
-  test<- test %>% 
-    select(time,timestampMs,clust) %>% 
-    mutate( day  = strftime(time,"%u"),
-            month = strftime(time,"%m"),
-            secMidnight = lubridate::period_to_seconds(lubridate::hms(strftime(time,"%T"))),
-            sinTime = sin(2*pi*secMidnight/seconds_in_day),
-            cosTime = cos(2*pi*secMidnight/seconds_in_day),
-            nextMeas = NA,
-            prevMeas = NA,
-            lagClust = NA,
-            leadClust = NA,
-            timestampLast = NA,
-            timestampNext = NA,
-            clustNext = NA,
-            clustPrev = NA)
-
-  all <- rbind(train,test) %>%
-    arrange(timestampMs) %>% 
-    fill(timestampLast,clustPrev, .direction = "down") %>% 
-    fill(timestampNext,clustNext, .direction = "up") %>%
-    mutate( prevMeas = timestampMs - timestampLast,
-            nextMeas = timestampNext - timestampMs,
-            lagClust = clustPrev,
-            leadClust = clustNext) %>% 
-    mutate(nextMeas = scales::rescale(nextMeas, from = c(0,3600000)),
-           prevMeas = scales::rescale(prevMeas, from = c(0,3600000)))
-  
-  test <- all %>% slice(remove_ind)
[…]

### 2018-05-03 ownAlgo/modifiedSTS.R

@@ -83,3 +83,8 @@ Sys.time()
 
 saveRDS(fin,file = "finalBinnedDataNL.rds")
 saveRDS(allPoints, file = "finalAllClusts2.rds")
+
+# load 
+
+fin <- readRDS("finalBinnedDataNL.rds")
+allPoints <- readRDS("finalAllClusts2.rds")

### 2018-05-03 ownAlgo/nnResults.R

@@ -1,0 +1,444 @@
+#neural nets march test results
+# to be run after neuralNets.R 
+
+remove_ind_5min<- getRemoveIndex(fin,"5 min")
+
+# missing: the features have to include the fact that they are missing, so they have to be recalcuated
+# the extracted ones are wrong!!
+
+testTrain5min<- getTestTrain(fin,remove_ind_5min[[1]])
+
+# get baseline distance measure
+testTrain5min[[1]] %>%
+  select(clust,lagClust) %>% 
+  left_join(allPoints,by="clust") %>% 
+  left_join(allPoints,by=c("lagClust" = "clust"),suffix = c(".clust",".predC")) %>% 
+  na.omit() %>% 
+  mutate( dist = raster::pointDistance(matrix(c(lon.clust,lat.clust),ncol = 2),
+                                       matrix(c(lon.predC,lat.predC),ncol = 2),
+                                       longlat = T)) %>% 
+  summary()
+  
+  
+dataList5min<- getModelInput(test = testTrain5min[[1]], train = testTrain5min[[2]])
+
+x_train <- dataList5min[[1]]
+x_test  <- dataList5min[[2]]
+y_train <- dataList5min[[3]]
+y_test  <- dataList5min[[4]]
+
+nClust <- 1222
+
+model <- keras_model_sequential() 
+model %>% 
+  layer_dense(units = 250, activation = 'relu', input_shape = ncol(x_test)) %>%
+  layer_dropout(rate = 0.4) %>%
+  layer_dense(units = 120, activation = 'relu') %>%
+  layer_dropout(rate = 0.3) %>%
+  layer_dense(units = nClust, activation = 'softmax')
+
+summary(model)
+
+model %>% compile(
+  loss = 'categorical_crossentropy',
+  optimizer = optimizer_rmsprop(),
+  metrics = c('accuracy')
+)
+
+model %>% fit(
+  as.matrix(x_train), as.matrix(y_train), 
+  epochs = 10, batch_size = 200, 
+  validation_split = 0.2
+)
+#plot(history)
+
+model %>% evaluate(as.matrix(x_test),as.matrix(y_test)) #88 percent accuracy
+
+# distance matrix of all clusters
+distMatrix<- geosphere::distm(allPoints[,c("lon","lat")])
+
+t<- fin %>% slice(remove_ind_5min[[1]])
+
+#t <- t[-c(1,nrow(t)),] #remove first and due to the feature extraction loss using lead and lag
+
[…]

### 2018-05-03 ownAlgo/resultsBinning.R

@@ -198,3 +198,13 @@ plotBin %>%
   addCircles(lng = ~lon.cluster,lat = ~lat.cluster, color = "#DF2935") %>% 
   addCircles(lng = ~lonF,lat = ~latF, color = "yellow",label = ~as.character(time_5_min))
 
+# CHECK how many measurements are clustered
+lowClust <- (fin %>% group_by(clust) %>% summarise( n = n()) %>% filter(n  == max(n)) %>% pull(clust))
+
+fin %>% 
+  filter( clust %in% lowClust) %>%
+  left_join(.,allPoints,by="clust",suffix=c(".measured",".cluster")) %>% 
+  leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
+  addProviderTiles(providers$CartoDB.Positron) %>%
+  addCircles(lng = ~lon.cluster,lat = ~lat.cluster, color = "#DF2935")
+  

### 2018-05-03 ownAlgo/resultsIan.R

@@ -1,4 +1,5 @@
-# full Ian results 
+# full Ian results take 2
+
 directory <- "/Users/boazsobrado/Desktop/Academic/Utrecht/year2/thesisFiles/harvard/Example"
 setwd(directory)
 
@@ -14,7 +15,7 @@ setwd("gpsmobility/R/")
 sapply(file.sources,source,.GlobalEnv)
 
 data <- fin %>% select(timestampMs,time,lat,lon,accuracy)
-downsampledIan <- downSampleMean(data,interval = '300 sec') #downsamples and adds XY coordinates
+downsampledIan <- downSampleMean(data,interval = '5 min') #downsamples and adds XY coordinates
 
 
 # remove data for results
@@ -28,8 +29,8 @@ d2<- downsampledIan %>%
 
 remove_ind <- which(as.character(d2) %in% remove_ind_5min[[2]])
 
-downsampleSys.time()
-d5minRemoved<- downsampledIan
+Sys.time()
+downsampled5minRemoved<- downsampledIan
 downsampled5minRemoved[remove_ind,c("lon","lat","x_v","y_v")] <- NA
 downsampled5minRemoved[remove_ind,c("code")] <- 4
 
@@ -57,6 +58,11 @@ evalDf<- evalIan(out3,downsampledIan = downsampledIan)
 
 evalDf[remove_ind,] %>% summary() # median 5 meters mean 141
 
+evalDf %>% filter( timestampMs %in% (downsampledIan[remove_ind,"timestampMs"] %>% pull())) %>% 
+  unique() %>%
+  summary()
+
+1-(65/length(remove_ind))
 
 #distA is the deviance score
 
@@ -87,6 +93,10 @@ summary(devianceIan) # mean 343 median 8.21
 
 
 # 1 hour thing
+
+# TO DO: SEPARATE INTO BEFORE MARCH, MARCH AND MARCH.
+
+# march
 removedHours<- remove_ind_hour[[2]]
 
 
@@ -94,24 +104,34 @@ d2<- downsampledIan %>%
   mutate( time = as.POSIXct(timestampMs, origin="1970-01-01")) %>% 
   thicken("hour","hour")
 
-remove_ind <- which(as.character(d2$hour) %in% removedHours)
+remove_ind1hr <- which(as.character(d2$hour) %in% removedHours)
 
 downsampled1hrRemoved<- downsampledIan
[…]

### 2018-05-03 scripts/downSample.R

@@ -10,12 +10,13 @@ downSampleMean <-
     library(padr)
     
     #  summarise via interval and add coordinates
-    data1 <- cbind(data,as.data.frame(LatLong2XY(data$lat, data$lon)))%>%
-      select(timestampMs,time,lat,lon,accuracy,x_v,y_v) %>% 
-      as_tbl_time(time) %>% 
-      arrange(time) %>% 
+    data1 <- data %>%
+      thicken(interval = interval, colname = 'time2')
+  
+      data1 <- cbind(data1,as.data.frame(LatLong2XY(data1$lat, data1$lon)))%>%
+      select(timestampMs,time,time2,lat,lon,accuracy,x_v,y_v) %>% 
+      arrange(timestampMs) %>% 
       mutate(timestampMs = as.numeric(timestampMs)) %>% 
-      thicken(interval = interval, colname = 'time2') %>%
       group_by(time2) %>%
       summarise(lon = mean(lon),# summarising function
                 lat = mean(lat),

### 2018-05-03 Thesis Manuscript/apaManu.tex
restructure and organise

--- 2018-04-29 manuscript/apaManu.tex
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
 
@@
             pdfauthor={},
-            pdftitle={Handling missing data in smartphone location logs},
+            pdftitle={Personalised Map Matched Imputation: imputing missing data from smartphone location logs},
             colorlinks=true,
@@
 % Essential manuscript parts
-  \title{Handling missing data in smartphone location logs}
+  \title{Personalised Map Matched Imputation: imputing missing data from
+smartphone location logs}
 
@@
 
-    \indent Word count: 2297
+    \indent Word count: 8319
   }
@@
 
-Why is human mobility important? How people move about in their
-environment affects a wide range of outcomes, such as health, income and
-social capital (Goodchild \& Janelle, 2010). Therefore it is
-unsurprising that social scientists in numerous fields and even
-policymakers are interested in human mobility measures. The most widely
-administered personality questionnaires ask individuals to what extent
-they agree with statements such as \enquote{I love large parties},
-\enquote{I prefer going to the movies to watching videos at home} and
-\enquote{I love to travel to places that I have never been before}
-({\textbf{???}}). In economic research, the postal code of an
-individual's home address is often used as a proxy for socioeconomic
-status (e.g. Villanueva \& Aggarwal, 2013). Moreover, economists study
-geographic labour mobility extensively (e.g. {\textbf{???}}). In fact,
-the European Commission considers labour mobility important enough to
-warrant its own directorate ({\textbf{???}}). The extensive use of
-measures like these across different domains strongly suggests that
-mobility metrics are linked to real world outcomes. Perhaps it is
-unsurprising that behavioural researchers have found that mobility
-measures can be used to predict academic performance (Wang, Harari, Hao,
-Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz, \&
-Odoms-Young, 2009) or even the onset of a depressive episode in bipolar
-depression patients (Palmius et al., 2017). Indeed, there is an argument
-to be made that perhaps psychologists have not been studying mobility
-enough. When studying behavioural differences within individuals
-behavioural scientists have often neglected the fact that individuals
-vary not only over time but also over space. To fully understand
-behaviour we must understand how behaviour can vary across environments.
+Why is human mobility important? Human mobility measures are quantified
+metrics of how people move about in their environment. Human mobility
+affects a wide range of outcomes, such as health, income and social
+capital (Goodchild \& Janelle, 2010). The most widely administered
+personality questionnaires ask individuals to what extent they agree
+with statements such as \enquote{I love large parties}, \enquote{I
+prefer going to the movies to watching videos at home} and \enquote{I
+love to travel to places that I have never been before} (Goldberg et
+al., 2006). At their root these questions are mobility measures. In
+economic research, the postal code of an individual's home address is
+often used as a proxy for socioeconomic status (e.g. Villanueva \&
+Aggarwal, 2013). Economists are interested not only in where people
+live, but also where they go to work: geographic labour mobility is
+related to the income of individuals, the well-being of a country's
+economy and even informs the policy of bodies such as the European
+Comission (Tatsiramos, 2009). The extensive use of measures like these
+across different domains in social science strongly suggests that
+mobility metrics are linked to important real world outcomes. Perhaps it
+is unsurprising that behavioural researchers have found that mobility
+measures can be used to predict academic performance (R. Wang, Harari,
+Hao, Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz,
+\& Odoms-Young, 2009) or even the onset of a depressive episode in
+bipolar depression patients (Palmius et al., 2017). Indeed, there is an
+argument to be made that perhaps psychologists have not been studying
+mobility enough. When studying behavioural differences within
+individuals behavioural scientists have often neglected the fact that
+individuals vary not only over time but also over space. To fully
+understand behaviour we must understand how behaviour can vary across
+environments.
 
 Despite the importance of mobility measures, the majority of these
-measures are obtained through the use of questionnaires (such as the
-IPIP) or specifically through pen-and-paper travel diaries.
-Questionnaires and travel diaries are have well-known methodological
-flaws: they are burdensome to collect and rely on accurate
-self-reporting. The burdensomeness stems from the fact that participants
-must be explicitly asked to report on their movement patterns at
-frequent intervals due to forgetfulness. This makes the data expensive
-to collect and limits the extent of data collection in practice. In
+metrics are obtained through the use of questionnaires (such as the
+aforementioned personality questionnaire) or specifically through
+pen-and-paper travel diaries. As a thought experiment we encourage the
+reader to try to remember where exactly he or she was five weeks ago on
+Saturday at midday. Most people find this sort of task very difficult
+(if not impossible), which illustrates that questionnaires and travel
+diaries are have well-known methodological flaws. This method of
+collecting data is difficult and relies on accurate self-reporting.
+Travel diaries are burdensome to collect because participants must be
+explicitly asked to write down on their movement patterns at frequent
+intervals due to human forgetfulness. This makes the data collection
+expensive and limits the duration of data collection in practice. In
 addition, the frequent reporting duties of the participant may bias the
-participants behaviour. Participant forgetfulness also limits the
-accuracy of self reported measures. There is clear evidence that
-participants are systematically biased when self-reporting mobility
-measures. For instance, participants under-report the frequency of short
-trips (Wolf, Oliveira, \& Thompson, 2003) and underestimate the duration
-of regular commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
-These obstacles can be overcome by using objective data on human
+participants behaviour. The limits of human cognition also limit the
+accuracy of self-reported measures. There is evidence that participants
+are systematically biased when self-reporting mobility measures. For
+instance, participants under-report the frequency of short trips (Wolf,
+Oliveira, \& Thompson, 2003) and underestimate the duration of regular
+commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017). These
+obstacles can only be overcome by using objective data on human
 mobility.
@@
 data in such a way is costly, places a high burden on participants and
-therefore the logs do not span a long time {[}Barnett and Onnela (2016).
+therefore the logs do not span a long time (Barnett \& Onnela, 2016).
 Today millions of individuals carry smartphones with themselves every
@@
 did not move at all! This issue is also visualised in Figure
-\ref{fig:palmiusVme1}. The problem is further complicated because
+\ref{fig:palmiusVsMe1}. The problem is further complicated because
 missing data and noisy measurements are related. Methods used by
@@
 for long term human mobility logs. Ecologists have used SSMs to explain
-how animals interact with their environment (Patterson, Thomas, Wilcox,
-Ovaskainen, \& Matthiopoulos, 2008). These models can be quite complex.
-Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
+how animals interact with their environment (T. A. Patterson, Thomas,
+Wilcox, Ovaskainen, \& Matthiopoulos, 2008). These models can be quite
+complex. Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
 processes to characterise the effect of roads, food patches and streams
@@
 geological research and rely on spatiotemporal auto-correlations. For
-instance, the CUTOFF method estimates missing values by incorporating
-similar observed temporal information from the value's nearest spatial
-neighbors (Feng, Nowak, O'Neill, \& Welsh, 2014 ). The authors
-illustrate their example using rainfall data from gauging stations
-across Australia. Similarly, Z. Zhang et al. (2017) use a variety of
-machine learning methods to impute missing values. The example provided
-relates to underground water data. Generally these models assume fixed
-measurement stations (such as rainfall gauging stations). For this
-reason they cannot be easily applied to missing mobility tracks without
-significant pre-processing.
+instance, Feng, Nowak, O'Neill, and Welsh (2014) estimate missing values
+by incorporating similar observed temporal information from the value's
+nearest spatial neighbors. The authors illustrate their example using
+rainfall data from gauging stations across Australia. Similarly, Z.
+Zhang et al. (2017) use a variety of machine learning methods to impute
+missing values. The example provided relates to underground water data.
+Generally these models assume fixed measurement stations (such as
+rainfall gauging stations). For this reason they cannot be easily
+applied to missing mobility tracks without significant pre-processing.
 
 On the other hand, a few researchers have explicitly attempted to impute
-missing data from human mobility patterns ({\textbf{???}}; Barnett \&
-Onnela, 2016; Palmius et al., 2017 ). Importantly, none of them worked
-with secondary logs. Nonetheless we will detail what they did as
-informative examples. Palmius et al. (2017) deal with the measurement
-inaccuracy of \(D\) in custom logs by removing from the data set all
-unique low-accuracy \(a\) data points that had
+missing data from human mobility patterns (Barnett \& Onnela, 2016;
+Palmius et al., 2017 ; @ Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
+none of them worked with secondary logs. Nonetheless we will detail what
+they did as informative examples. Palmius et al. (2017) deal with the
+measurement inaccuracy of \(D\) in custom logs by removing from the data
+set all unique low-accuracy \(a\) data points that had
 \(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
@@
 on measured movements in that area, at that point in time or movements
-in the last 24 hours (\emph{circadian proximity}).
-
-On the other hand, ({\textbf{???}}) use what they call a Spatial
-Temporal Semantic Neural Network (STS-NN) to predict future human
-movement. While the authors are concerned with prediction and not
-imputation, they devised a method called the Spatial Temporal Semantic
-(STS) algorithm which converts raw measurements to machine learning
-friendly discrete bins. Working with high-frequency measurements,
-({\textbf{???}})'s method down-samples the raw data temporally and
-map-matches the resulting bins to discrete points along pre-established
-geographical features such as roads and highways. This minimises
-measurement error and paves the way for applying machine learning
-methods to human mobility problems.
+in the last 24 hours.
+
+It is also worth mentioning that there is a body of work that
+incorporates road features and other background information to generate
+predictions or imputations. For instance, Liao, Patterson, Fox, and
+Kautz (2007) use hierarchical Markov models in combination with
+knowledge about the transportation system to bridge the gap between raw
+GPS sensor measurements and high level information such as a user's
+destination and mode of transportation. This can be then used to impute
+or predict missing steps. Along similar lines Wu et al. (2017) use what
+they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict
+future human movement. While the authors are concerned with prediction
+and not imputation, they devised a method called the Spatial Temporal
+Semantic (STS) algorithm which converts raw measurements to machine
+learning friendly discrete bins. Working with high-frequency
+measurements, Wu et al. (2017)'s method down-samples the raw data
+temporally and map-matches the resulting bins to discrete points along
+pre-established geographical features such as roads and highways. This
+minimises measurement error and paves the way for applying machine
+learning methods to human mobility problems. Subsequently we will focus
+on methods which do not incorporate external information as they are
+more generally applicable.
 
@@
 deal explicitly with mobility patterns from custom or secondary logs
-({\textbf{???}}; Barnett \& Onnela, 2016; Palmius et al., 2017 ).
+(Barnett \& Onnela, 2016; Palmius et al., 2017 ; Wu et al., 2017).
 
@@
 
-Following ({\textbf{???}})'s spatial-temporal-semantic (STS) feature
+Following Wu et al. (2017)'s spatial-temporal-semantic (STS) feature
 extraction algorithm our aim is to transform pause and path locations
@@
 with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
-{[}(``ST-DBSCAN,'' 2007);{]}. We will focus on two methods which
-explicitly with mobility patterns from unevenly sampled smartphone logs
-(Barnett \& Onnela, 2016; Palmius et al., 2017 ). Both of these methods
-pre-process the data and subsequently use two steps to extract pause
-locations: first they extract pauses and their corresponding locations,
-then they cluster pause locations based on spatial proximity. This
-implementation of PMMI uses a stricter version of Barnett and Onnela
-(2016)'s approach to extract pauses.
+(``ST-DBSCAN,'' 2007). We will focus on two methods which explicitly
+with mobility patterns from unevenly sampled smartphone logs (Barnett \&
+Onnela, 2016; Palmius et al., 2017 ). Both of these methods pre-process
+the data and subsequently use two steps to extract pause locations:
+first they extract pauses and their corresponding locations, then they
+cluster pause locations based on spatial proximity. This implementation
+of PMMI uses a stricter version of Barnett and Onnela (2016)'s approach
+to extract pauses.
 
@@
 
-({\textbf{???}})'s spatial-temporal-semantic (STS) feature extraction
+Wu et al. (2017)'s spatial-temporal-semantic (STS) feature extraction
 algorithm uses map matching as a ground truth to assign noisy
@@
 individual is measured as moving closely in parallel to a road A in an
-area where there is no other parallel road, ({\textbf{???}})'s method
+area where there is no other parallel road, Wu et al. (2017)'s method
 will assume that the individual is on the road A.
@@
 PMMI uses a similar logic, but without using any external geographic
-database. The key modification in PMMI is that whilst ({\textbf{???}})
+database. The key modification in PMMI is that whilst Wu et al. (2017)
 uses a map from outside the persons location logs, we use the total
@@
 The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window.
-For this day there were several periods with no information.
+The entire log contains several long periods with no information.
 These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
@@
 For simplicity, we subsequently used a time period when the individual
-was living in the Netherlands. This leaves us with 156,000 measurements
-over a period of less than six months. Analyses as well as
-implementations and adaptions Palmius et al. (2017)'s and Barnett and
-Onnela (2016)'s model were performed using R and a multitude of other
-statistical packages R (Version 3.4.3; R Core Team, 2017) and the
-R-packages \emph{ggplot2} (Version 2.2.1; Wickham, 2009),
-\emph{ggthemes} (Version 3.4.0; Arnold, 2017), \emph{knitr} (Version
-1.20; Xie, 2015), and \emph{papaja} (Version 0.1.0.9709; Aust \& Barth,
-2018). All the code is available on a public repository
-({\textbf{???}}).
+was living in the Netherlands. This subset contains 156,000 measurements
+over a period of less than six months.
+
+Palmius et al. (2017)'s algorithm was implemented in R based on the
+original MATLAB code and pseudocode kindly provided by the author.
+Barnett and Onnela (2016)'s method was slightly adapted in R to fit the
+the data structure from the original R code provided by the author. When
+executing their models we used the same parameters as the authors did.
+To represent Barnett and Onnela (2016)'s model we used the variant where
+movements were sampled based on spatial proximity.
+
+All analyses were performed using R (Version 3.4.3; R Core Team,
+2017)\footnote{We, furthermore, used the R-packages \emph{bindrcpp}
+  (Version 0.2; Müller, 2017), \emph{dbplyr} (Version 1.2.0; Wickham \&
+  Ruiz, 2018), \emph{dplyr} (Version 0.7.4; Wickham, Francois, Henry, \&
+  Müller, 2017), \emph{geosphere} (Version 1.5.7; Hijmans, 2017a),
+  \emph{ggplot2} (Version 2.2.1; Wickham, 2009), \emph{ggthemes}
+  (Version 3.4.0; Arnold, 2017), \emph{kableExtra} (Version 0.7.0; Zhu,
+  2018), \emph{keras} (Version 2.1.4; Allaire \& Chollet, 2018),
+  \emph{knitr} (Version 1.20; Xie, 2015), \emph{leaflet} (Version 2.0.0;
+  Cheng, Karambelkar, \& Xie, n.d.), \emph{padr} (Version 0.4.0; Thoen,
+  2017), \emph{papaja} (Version 0.1.0.9709; Aust \& Barth, 2018),
+  \emph{raster} (Version 2.6.7; Hijmans, 2017b), \emph{RColorBrewer}
+  (Version 1.1.2; Neuwirth, 2014), \emph{readr} (Version 1.1.1; Wickham,
+  Hester, \& Francois, 2017), \emph{rgdal} (Version 1.2.16; R. Bivand,
+  Keitt, \& Rowlingson, 2017), \emph{scales} (Version 0.5.0; Wickham,
+  2017), \emph{sp} (Version 1.2.7; Hijmans, 2017a; Pebesma \& Bivand,
+  2005), \emph{tibbletime} (Version 0.1.0; Vaughan \& Dancho, 2018), and
+  \emph{tidyr} (Version 0.8.0; Wickham \& Henry, 2018).}.
+
+All the code is available on a public repository (Sobrado, 2018).
 
@@
 \item
-  Comparing the performance of PMMI using a) baseline models b)
-  performance with randomly removed data in comparison to the
-  aforementioned methods (Barnett \& Onnela, 2016; Palmius et al., 2017)
-  and c) objective ground-truth data (public transportation
-  time-stamps).
+  Comparing the performance of PMMI compared to alternative methods
+  {[}(Barnett \& Onnela, 2016; Palmius et al., 2017){]} and a baseline
+  model using cross-validation.
+\item
+  Comparing the performance of PPMI compared to the aformentioned
+  methods using objective ground-truth data (public transportation
+  time-stamps locations).
 \end{enumerate}
@@
 Before we can evaluate the accuracy of the imputations, it is essential
-to evaluate how well noise in the data has been cleaned. Otherwise we
-run the risk of over-fitting the model in the sense that we will measure
-the extent to which an imputation method can correctly impute
-measurement noise within \(D_t\) instead of true location \(G(t)\).
+to evaluate how well noise in the data has been smoothed. All methods
+smoothen noise differently in an attempt to extract true location
+\(G(t)\) from measurements \(D_t\). In the subsequent cross-validation
+step we will remove random smoothened blocks and compare the imputed
+value to the smoothened value. For this reason we want to ensure that
+the smoothened method actually does reflect the true location \(G(t)\).
+Otherwise we run the risk of over-estimating the accuracy of a model in
+the sense that we will compute the extent to which an imputation method
+can correctly impute a smoothened value which has little to do with the
+true location \(G(t)\).
 
@@
 superimposed on is an established way to heuristically check their
-accuracy (e.g. ({\textbf{???}})). Then, let the average distance between
+accuracy (e.g. Brunsdon, 2007). Then, let the average distance between
 the actual measured point and the binned point be the \emph{deviation
@@
 
-\subsubsection{Imputation algorithm
-performance}\label{imputation-algorithm-performance}
+\subsubsection{Imputation algorithm performance in
+cross-validation.}\label{imputation-algorithm-performance-in-cross-validation.}
 
 We will compare the performance of PMMI, Palmius et al. (2017) and
-Barnett and Onnela (2016). In addition, we will also compute a naive
-model, which simply imputes as the missing value the previous observed
-value. The naive model will serve as a baseline model. To compare the
-performance of these methods we will remove 25\% of measured time
-intervals at random within a four week period. We will make our
-comparisons for intervals of 5 minutes, 1 hour and 1 day. In other
-words, we will remove 25\% of time intervals at random, while varying
-the size of the time intervals removed.
+Barnett and Onnela (2016). In addition, we will also compute two
+baseline models. The first one is a naive model, which simply imputes
+the previous observed value. The second one we will call the
+\enquote{home} model, which imputes that the individual is at home at
+all missing timepoints. Using these two models as baselines makes sense
+given the high degree of autocorrelation in the data and that
+individuals tend to spend most of their time at home. To compare the
+performance of the aforementioned methods we will remove 25\% of
+measured time intervals at random within a four week period. We will
+make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In
+other words, we will remove 25\% of time intervals at random, while
+varying the duration of the time intervals removed.
 
@@
 To compare all methods with each other we will compute a distance
-measure (how far was the removed location from the predicted location).
-For PMMI's imputations we will use a weighted mean for the distance
-measures whereby each 5 minute period is weighted equally. This is
-necessary because the other two model estimate far fewer values than
-PMMI. While Barnett and Onnela (2016)'s and Palmius et al. (2017)'s
-models estimate 12 measurements for each missing hour, PMMI estimates as
-many measurements as there are in the log. When individuals go to
-infrequently travelled locations they tend to use their phone's location
-services more than 100 more times, which leads to more measurements, and
-hence more imputed values. Hence, if the weights are not used PMMI is
-susceptible to error overestimates which are related to measurement
-frequency in a way that the other two methods are not.
-
-In addition, other measures of interest for PMMI are accuracy (in what
-percentage of the cases was the appropriate cluster predicted), the
+measure (how far was the removed location from the predicted location)
+and the coverage. The coverage refers to the percentage of missing cases
+which were imputed. For PMMI's imputations we will use a weighted mean
+for the distance measures whereby each 5 minute period is weighted
+equally. This is necessary because the other two models downsample
+temporally to 5 minutes and because the frequency of measurements in a 5
+minute period is correlated with imprecise measurements and visiting
+infrequently travelled locations. Hence weights must be used to avoid
+inflating error rates for PPMI in comparison to the other models.
+
+In addition, other measures of interest for PMMI are \emph{accuracy} (in
+what percentage of the cases was the appropriate cluster predicted), the
 \emph{confidence} and the \emph{distance expectation}. The confidence is
@@
 For instance, if the model predicts the missing bin to be bin A with a
-probability of 0.9 then we can say it has high confidence in the
-prediction. Similarly, the distance expectation is the cross-product of
-the estimated probabilities that the individual is at any of all given
+probability of 0.9 then we can say this is a high confidence prediction.
+Similarly, the distance expectation is the cross-product of the
+estimated probabilities that the individual is at any of all given
 clusters and the distances between the clusters to the true cluster. For
-example, if the true location of an individual is bin A (bin A is 10
+example, if the true location of an individual is bin A (bin A is 1000
 meters away from bin B) and the model assigns a probability of 0.9 at
-bin A and 0.1 at bin B, then the distance expectation would be 1 meter.
-
-Finally, we will take objective real world data and compare it to
-predicted values. We will use information from the Dutch public
-transportation card. The Dutch public transportation service provides
-users with time-stamped data of when and where they board, change lines
-or leave public transportation. To be able to make a comparison between
-models we will remove all measurements from within the 5 minute period
-that a time-stamped measurement is available. Then we will use each
-model to impute the location of the individual within that period.
+bin A and 0.1 at bin B, then the distance expectation would be 100
+meters.
+
+\subsubsection{Imputation algorithm performance using external
+benchmarks.}\label{imputation-algorithm-performance-using-external-benchmarks.}
+
+We will compare the performance of PMMI, Palmius et al. (2017), Barnett
+and Onnela (2016) and the naive baseline model with regards to imputing
+real missing measurments. To do will use information from the Dutch
+public transportation card. The Dutch public transportation service
+provides users with time-stamped location data of when and where they
+board, change lines or leave public transportation. In this paper we
+used 97 such events in a single month.
+
+To be able to make a comparison between models we will remove all
+measurements from within the 5 minute period that a time-stamped
+measurement is available. Then we will use each model to impute the
+location of the individual within that period and compare it to the true
+location.
 
@@
 
-The selection of these parameters was more-or-less heuristically driven
-based on their on their ability to extract a meaningful personalised
-map. When selecting parameters there is a trade off is between bias and
-precision. This is because an increase in precision in the form of a
-higher for high density locations resolution comes at the expense of
-precision as assigning measurements to bins becomes more difficult. For
+The model was tuned based on a visual inspection of the resulting
+personal map meeting the following criteria:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Plausible extraction of pause locations. For example, a pause in a
+  field with no path leading to it is less plausible than a pause at a
+  train station, airport, or office building.
+\item
+  A suitable extraction of paths. For instance, are ensuring that
+  individual paths are complete and not merged unneccessarily in the
+  case of closely parallel paths.
+\item
+  Less than a fifth of extracted clusters are assigned only one
+  measurement. Unique clusters are unavoidable as people sometimes visit
+  a location briefly and only once, nonetheless they pose problems for
+  the classification algorithms.
+\end{enumerate}
+
+When selecting parameters there is a trade off is between bias and
+precision. This is because an increase in precision (in the form of a
+higher resolution of locations location) comes at the expense of
+precision (assigning measurements to bins becomes more difficult). For
 instance, by increasing the clustering distance parameter
 \(d_{\text{Pause lim}}\) we can extract more valid pause locations at
-the expense of falsely categorising certain measurements to the wrong
-cluster. This is illustrated in Figure \ref{fig:pauseLocUithof1}.
+the risk of falsely categorising certain measurements to the wrong bin.
+This is illustrated in Figure \ref{fig:pauseLocUithof1}.
 
@@
 values are within their corresponding accuracy value, which is close to
-the theoretical \(\frac{2}{3}\) value that is expected. Approximately
-9\% of values were not taken into account when creating the bins,given
-that their the accuracy \(a\) exceeded \(a_{\text{Path Lim}}\).
-
-For the narrower period of March 2017 approximately 74\% of the deviance
-values are within their corresponding accuracy values, which is close to
-the theoretical \(\frac{2}{3}\) value that is expected. The raw
-unweighted mean and median deviance are 38 and 14 meters respectively.
-If we weight them such that each 5 minute interval is weighted equally
-to ease comparison with the other two methods, we find a mean deviance
-of 35 meters with a median of 12.8.
-
-In comparison, Palmius' method has a median deviance
-\(\delta_{\text{dev}}\) of 3 meters, with a mean of 115 due to high
-deviance outliers. On the other hand, Barnett's method has mean deviance
-\(\delta_{\text{dev}}\) of 343 meters and a median deviance of 8 meters.
-Barnett's deviance is necessarily higher than Palmius', as they
-down-sample temporally (like Palmius) and subsequently aggregate into
+the theoretical 67\% value that is expected. Approximately 9\% of values
+were not taken into account when creating the bins,given that their the
+accuracy \(a\) exceeded \(a_{\text{Path Lim}} = 300\).
+
+In comparison, Palmius et al. (2017)\enquote{s method has a median
+deviance \(\delta_{\text{dev}}\) of 3 meters, with a mean of 115 due to
+high deviance outliers. On the other hand, Barnett and Onnela (2016)'s
+method has mean deviance \(\delta_{\text{dev}}\) of 343 meters and a
+median deviance of 8 meters. Barnett and Onnela (2016)'s deviance is
+necessarily higher than Palmius et al. (2017)}, as they down-sample
+temporally (like Palmius et al. (2017)) and subsequently aggregate into
 pauses and linear flights.
@@
 interval. The fact that over 90\% of deviance values are within accuracy
-(much higher than the expected theoretical 67\%) confirms that temporal
-down-sampling is not sufficiently filtering out the noise.
+(substantially higher than the expected theoretical 67\%) confirms that
+temporal down-sampling is not sufficiently filtering out the noise.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/tempVsspace} \caption{Excerpt. MISSING CAPTION HERE}\label{fig:palmiusVsMe1}
+\includegraphics[width=1\linewidth]{img/tempVsspace} \caption{The difference between temporal and spatial downsampling.
+The blue circles are raw measurments, the yellow circles are temporally downsampled locations.
+Spatially downsampled locations are in red.
+Due to measurement sparcity and inaccuracy many of the temporally downsampled locations are in unfeasable locations.}\label{fig:palmiusVsMe1}
 \end{figure}
 
-\subsubsection{Imputation evaluation}\label{imputation-evaluation}
-
-The Palmius et al. (2017) model failed to impute 3\% of all removed
-values for both the 5 minute and over 10\% for the 1 hour tests. Palmius
-et al. (2017)' model failed to impute a single value for the day tests.
-Similarly, Barnett and Onnela (2016)'s method failed to impute over 11\%
-of the missing values. PPMI made an imputation for all missing values.
-Table 1 shows the results of the distance metrics for each method.
-
-\begin{table}[htb]
-\begin{center}
-\begin{threeparttable}
-\caption{\label{tab:resulttable}Distance in meters between the removed time period and the imputed value.}
-\begin{tabular}{lllllll}
+\subsubsection{Imputation algorithm performance evaluation in
+cross-validation.}\label{imputation-algorithm-performance-evaluation-in-cross-validation.}
+
+The first factor we must take into account is the coverage of each
+imputation model. The results show us that while PPMI and the naive
+baseline model impute all missing values, the other two models have
+difficulties doing so (Table 1). In particular, the models of Palmius et
+al. (2017) and Barnett and Onnela (2016)'s fail to impute an increasing
+amount of missing values with an increase in the duration of the missing
+period. Palmius et al. (2017)'s method in particular failed to impute a
+single missing time period. The longer a log, the higher the probability
+that long time periods will be missing. Secondary logs are typically
+long. Hence missing data imputation methods designed to work with
+secondary logs must be capable of imputing longer time periods.
+
+\begin{lltable}
+\begin{longtable}{llllllllll}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:resulttable}Distance in meters between the removed time period and the imputed value.}\\
 \toprule
- & \multicolumn{2}{c}{Five minutes} & \multicolumn{2}{c}{One Hour} & \multicolumn{2}{c}{One Day} \\
-\cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
- & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median}\\
+ & \multicolumn{3}{c}{Five minutes} & \multicolumn{3}{c}{One Hour} & \multicolumn{3}{c}{One Day} \\
+\cmidrule(r){2-4} \cmidrule(r){5-7} \cmidrule(r){8-10}
+ & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Coverage} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Coverage} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{Median} & \multicolumn{1}{c}{Coverage}\\
 \midrule
-Barnett \& Onella & 82 & 4 & 345 & 6 & 9,273 & 12\\
-Palmius & 43 & 0 & 497 & 4 & NA & NA\\
-PPMI & 269 & 0 & 908 & 0 & 5,757 & 0\\
-Naive Baseline & 426 & 0 & 1,502 & 0 & 14,266 & 1,288\\
+Barnett \& Onella & 240 & 5 & 97\% & 33 & 6 & 88\% & 62 & 6 & 85\%\\
+Palmius & 43 & 9 & 97\% & 497 & 4 & 89\% & NA & NA & 0\%\\
+PPMI & 269 & 0 & 100\% & 908 & 0 & 100\% & 5,757 & 0 & 100\%\\
+Naive Baseline & 426 & 0 & 100\% & 1,502 & 0 & 100\% & 14,266 & 1,288 & 100\%\\
+Home Baseline & 5,599 & 0 & 100\% & 5,667 & 0 & 100\% & 5,757 & 0 & 100\%\\
 \bottomrule
-\end{tabular}
-\end{threeparttable}
-\end{center}
-\end{table}
-
-In terms of PPMI's accuracy in this period, the prediction accuracy was
-88\%, 73\% and 47\% for the 5 minute, 1 hour and 1 day periods
-respectively. As a comparison, the naive model's accuracy ratings were
-87\%, 68\% and 24\% respectively. The distance expectation values were
-very similar to the distance scores, albeit approximately 5\% higher.
-Confidence scores ranged from 0.006 to 1.
-
-\subsubsection{Comparison with objective
-data}\label{comparison-with-objective-data}
+\end{longtable}
+\end{lltable}
+
+With respect to the distance metrics, Table 1 shows the results of the
+distance metrics for each method. We can see that PPMI predicted the
+location perfectly for over half of the removed values in all three time
+categories. To be precise PPMI's prediction accuracy was 88\%, 73\% and
+47\% for the 5 minute, 1 hour and 1 day periods respectively. As a
+comparison, the baseline naive model's prediction accuracy was 87\%,
+68\% and 22\%, and the home baseline model's were 46\%, 52\% and 40\%.
+Although the home baseline model and PPMI achieved similar distance
+metrics in the 1 day condition, this does not mean that PPMI defaulted
+to simply predicting the individual is home at all times (only 40\% of
+predictions were home). PPMI outperformed the alternative models when
+taking into account the median scores, but not the mean. The explanation
+is that the alternative models failed to impute more difficult time
+periods.
+
+On the other hand, the confidence score for each time period were
+heavily skewed left, withh a median of 1. Accordingly, the distance
+expectation (the crossproduct of the model probabilities of an estimate
+with the distances of the true value to the estimates) values were quite
+similar to the distance scores, albeit approximately 5\% higher. Based
+on this we can say that this implementation of PPMI was overconfident in
+its predictions.
+
+\subsubsection{Imputation algorithm performance evaluation using
+external
+benchmarks.}\label{imputation-algorithm-performance-evaluation-using-external-benchmarks.}
 
 Once we removed the measurements within the 5 minute period of each
-time-stamp, we employed the models to predict the location of the
-individual at the time-stamped period. The naive model reaches an
-accuracy of 16\%, with a median distance of 555 meters and mean distance
-of 3303 meters. On the other hand MYMETHOD has an accuracy of 24\%, with
-a median distance of 1037 meters and a mean distance of 5637. Again the
-expectation of the distance was quite similar to the distance with a
-mean of 6432 and a median of 1037 meters.
-
-Palmius's model failed to impute 38 out of 97 periods. For those it did
-impute, it showed a mean distance of 1517 meters and a median distance
-of 1617. Barnett's model failed to impute 18 out of 97 periods. For
-those it did impute, it had a mean of 5506 and a median of 1342.
+time-stamp in the public transportation log, we employed each of the
+aforementioned methods to impute the location of the individual at the
+time-stamped period. Then we calculated the distance metrics between the
+true location (based on the public transport log) and the imputed
+location. We did not compute the home baseline model as it would have
+always been wrong for this subset of data (travelling and being at home
+are mutually exclusive).
+
+First, the median distance between the imputed value and the true value
+for the baseline naive model was 555 meters (with a mean of 3303). For
+this method, the prediction accuracy was 16\%. This means that the
+binning method places the individual at the location reported by the
+public transportation travel log in the five minutes prior to the
+measurement 16\% of the time. Needless to say the coverage was 100\% as
+all missing values were imputed.
+
+For the PPMI the median distance metric was 1037 meters (mean 5637). The
+prediction accuracy was higher than that of the baseline naive model at
+24\%. Although PPMI imputed more points correctly than the naive
+baseline model, those it imputed incorrectly were significantly more
+incorrect. All missing values were imputed, hence resulting in a
+coverage of 100\%.
+
+The same could not be said about Palmius et al. (2017)'s method, as it
+failed to impute almost 40\% of the time periods. This results in a
+coverage of 60\%. For the time periods the method did impute, there was
+a median distance of 1617 meters between the true location and the
+imputed value (with a mean of 1517 meters).
+
+Barnett and Onnela (2016)'s method performed better in terms of coverage
+at 81\%. The median distance metric was 1342 meters and the mean value
+was 5506 meters.
 
@@
 Social scientists are most interested in aggregating spatiotemporal data
-to more socially relevant information, such as the amount of time spent
-at home. As an example we calculated the time spent at home of the user
-in the month of March (Figure \ref{fig:aggrePlot}) without any model and
-with all three of the investigated methods.
+to more socially relevant metrics, such as the amount of time spent at
+home, frequency of travels to new locations, the amount of trips made or
+distance covered. As an example we calculated the time spent at home of
+the user without imputing any missing data and with all three of the
+investigated methods. The results can be seen in Figure
+\ref{fig:aggrePlot}. Without any form of missing imputation the
+individual's time spent at home is unknown for 12\% of the time, which
+adds up to almost four days over the course of a 30 day month.
 
@@
 
-Interestingly, all three models suggest that the user spent
-approximately 60\% of their time at home. However, without using any
-form of missing data imputation approximately 12\% of the time is
-unaccounted for. Given that the amount of time spent at home has been
-found to be a reliable predictor of extroversion (G. M. Harari et al.,
-2016) and the onset of depressive episodes in bipolar patients (Palmius
-et al., 2017) obtaining an accurate value for mobility metrics like this
-is highly important to social scientists.
+The amount of time spent at home has been found to be a reliable
+predictor of extroversion (G. M. Harari et al., 2016) and the onset of
+depressive episodes in bipolar patients (Palmius et al., 2017). The 12\%
+of time that remains unaccounted for could mean the difference between
+an introvert and an extrovert, or a healthy individual and an individual
+suffering of depression. Filling in these gaps is the exactly the sort
+of problem that can be solved using missing data imputation.
+Interestingly, despite their radically different imputation methods, all
+three examined models suggest that the user spent approximately 60\% of
+their time at home.
 
@@
 Overall the PMMI performed better than the alternative models,
-particularly during longer missing periods and with objective data.
-However, it did not perform substantially better than the naive baseline
-model. In addition, the comparison to the performance of the
-Barnett\&Onella and Palmius models is somewhat unfair, as they were
-created for custom logs, not secondary logs. Nonetheless, the comparison
-remains valid as they are the closest we found to a missing data
-imputation methods in smartphone GPS logs.
+particularly during longer missing periods and with objective data
+gathered from the Dutch public transport service. However, PPMI did not
+perform substantially better than the baseline models for all time
+periods and with the objective Dutch public transport data, suggesting
+that the classification method can be improved. In addition, the
+comparison to the performance of the Barnett and Onnela (2016) and
+Palmius et al. (2017) models is somewhat unfair, as they were created
+for custom logs, not secondary logs. Nonetheless, the comparison remains
+valid as they are the closest we found to a missing data imputation
+methods in smartphone GPS logs.
 
@@
 of single imputation is that it takes an imputed value and treats it as
-observed. Simple rule-based methods such as Palmius' are essentially
-algorithmic single imputation methods. With PPMI it is possible to model
-uncertainty using the predicted probabilities of each estimate. For
-instance, in the previous example, we could choose to only take
-estimates with a high degree of confidence, thus creating confidence
-intervals by adding and subtracting the amount of cases where the
-location of the individual is ambiguous.
+observed. Simple rule-based methods such as Palmius et al. (2017)' are
+essentially algorithmic single imputation methods. With PPMI it is
+possible to model uncertainty using the predicted probabilities of each
+estimate. For instance, in the previous example, we could choose to only
+take estimates with a high degree of confidence, thus creating
+confidence intervals by adding and subtracting the amount of cases where
+the location of the individual is ambiguous.
 
@@
 low (about 0.01 Hz) the personalised map manages to recreate the train
-line almost perfectly, despite being completely map agnostic.
+line almost perfectly, despite being completely map agnostic. Map
+agnisticism makes the model more flexible as it can be used for areas
+where no good geographical databases are available.
 
@@
 function of both the measurement and a model of the individuals movement
-history. In terms of state-space models like the Kalman filter the state
-equation would represent a probabilistic representation of where the
-individual could be based on the individuals entire movement history.
+history. In terms of state-space models the state equation would
+represent a probabilistic representation of where the individual could
+be at that given time based on the individuals entire movement history.
 The space side of the model would be a measurement equation representing
 the measurement and the uncertainty surrounding it in the form of \(a\).
+This would be a similar implementation to that used in Liao et al.
+(2007), but rather taking the personalised map instead of information
+about road networks.
 
@@
 which was used to generate predictions used no information on sequence
-patterns longer than the previous and next bin. A more sophisticated
-recurrent neural network (RNN), or a long short-term memory recurrent
-neural network (LSTM) would likely perform significantly better.
-
-That there is room for improvement is unsurprising given that it is
-still early days for using smartphone location measurements in social
-science. Nonetheless, the methodological advantages are clear: millions
-of individuals have a location log containing objective measurements
-spanning years which can be accessed for free. This is vastly superior
-to alternatives such questionnaires which rely on accurate
-self-reporting. Social science researchers must take advantage of
-regulatory changes in data portability regulations and put the vast
-wealth of data collected by commercial entities to public use.
+patterns longer than the previous and next bin. It also has no
+contextual information about how far away the bins are from each other.
+A model with more input variables and a more sophisticated design, such
+as a recurrent neural network (RNN), or a long short-term memory
+recurrent neural network (LSTM) would likely perform significantly
+better. Moreover, more inputs can be added from alternative sensors in
+smartphones such as the accelerometer, barometer, measures of phone
+activity and so on. These can be used to supplement or improve the raw
+location measurements, or indeed inform the predictive model.
+
+In conclusion, with this paper we aim to start filling the noticeable
+gap in the social science methodology literature in using smartphone
+location logs to study human movement. Clearly there is room for
+improvement, but this is to be expected with new methods. We are in the
+early days of using smartphone location measurements in social science.
+Nonetheless, the methodological advantages are clear: millions of
+individuals have years long location logs containing objective
+measurements. In addition, these measurements can be easily obtained.
+This is an unprecedented opportunity. However, with great opportunity
+comes a great responsibility, as this data raises difficult questions
+about privacy that researchers must be prepared to answer. Privacy
+concerns should be addressed, but should not dissuade researchers from
+developing new methods. Objective location logs are vastly superior to
+alternatives, such as questionnaires and travel diares, which rely on
+accurate self-reporting. Social science researchers must take advantage
+of regulatory changes in with regard to data portability and put the
+vast wealth of data collected by commercial entities to scientific use.
 
@@
 \section{References}\label{references}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{r_refs}\NormalTok{(}\DataTypeTok{file =} \StringTok{"r-references.bib"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-R-keras}{}
+Allaire, J., \& Chollet, F. (2018). \emph{Keras: R interface to
+'keras'}. Retrieved from \url{https://CRAN.R-project.org/package=keras}
+
 \hypertarget{ref-R-ggthemes}{}
@@
 Retrieved from \url{http://arxiv.org/abs/1606.06328}
+
+\hypertarget{ref-R-rgdal}{}
+Bivand, R., Keitt, T., \& Rowlingson, B. (2017). \emph{Rgdal: Bindings
+for the 'geospatial' data abstraction library}. Retrieved from
+\url{https://CRAN.R-project.org/package=rgdal}
+
+\hypertarget{ref-brunsdon_path_2007}{}
+Brunsdon, C. (2007). Path estimation from GPS tracks. \emph{Proceedings
+of the 9th International Conference on GeoComputation}. Retrieved from
+\url{http://eprints.maynoothuniversity.ie/6148/}
 
@@
 
+\hypertarget{ref-R-leaflet}{}
+Cheng, J., Karambelkar, B., \& Xie, Y. (n.d.). \emph{Leaflet: Create
+interactive web maps with the javascript 'leaflet' library}. Retrieved
+from \url{http://rstudio.github.io/leaflet/}
+
 \hypertarget{ref-commission_protecting_2017}{}
@@
 doi:\href{https://doi.org/10.1016/j.jhydrol.2014.11.012}{10.1016/j.jhydrol.2014.11.012}
+
+\hypertarget{ref-goldberg_international_2006}{}
+Goldberg, L. R., Johnson, J. A., Eber, H. W., Hogan, R., Ashton, M. C.,
+Cloninger, C. R., \& Gough, H. G. (2006). The international personality
+item pool and the future of public-domain personality measures.
+\emph{Journal of Research in Personality}, \emph{40}(1), 84--96.
 
@@
 
+\hypertarget{ref-R-geosphere}{}
+Hijmans, R. J. (2017a). \emph{Geosphere: Spherical trigonometry}.
+Retrieved from \url{https://CRAN.R-project.org/package=geosphere}
+
+\hypertarget{ref-R-raster}{}
+Hijmans, R. J. (2017b). \emph{Raster: Geographic data analysis and
+modeling}. Retrieved from
+\url{https://CRAN.R-project.org/package=raster}
+
 \hypertarget{ref-jankowska_framework_2015}{}
@@
 
+\hypertarget{ref-liao_learning_2007}{}
+Liao, L., Patterson, D. J., Fox, D., \& Kautz, H. (2007). Learning and
+inferring transportation routines. \emph{Artificial Intelligence},
+\emph{171}(5), 311--331.
+doi:\href{https://doi.org/10.1016/j.artint.2007.01.006}{10.1016/j.artint.2007.01.006}
+
 \hypertarget{ref-location_history_timeline_2017}{}
@@
 \url{//ianlondon.github.io/blog/encoding-cyclical-features-24hour-time/}
+
+\hypertarget{ref-R-bindrcpp}{}
+Müller, K. (2017). \emph{Bindrcpp: An 'rcpp' interface to active
+bindings}. Retrieved from
+\url{https://CRAN.R-project.org/package=bindrcpp}
+
+\hypertarget{ref-R-RColorBrewer}{}
+Neuwirth, E. (2014). \emph{RColorBrewer: ColorBrewer palettes}.
+Retrieved from \url{https://CRAN.R-project.org/package=RColorBrewer}
 
@@
 
+\hypertarget{ref-R-sp}{}
+Pebesma, E. J., \& Bivand, R. S. (2005). Classes and methods for spatial
+data in R. \emph{R News}, \emph{5}(2), 9--13. Retrieved from
+\url{https://CRAN.R-project.org/doc/Rnews/}
+
 \hypertarget{ref-preisler_modeling_2004}{}
@@
 
+\hypertarget{ref-sobrado_my_2018}{}
+Sobrado, B. (2018). \emph{My thesis on missing data in GPS
+measurements.} Retrieved from \url{https://github.com/sobradob/thesis}
+
 \hypertarget{ref-_st-dbscan:_2007}{}
@@
 doi:\href{https://doi.org/10.1016/j.datak.2006.01.013}{10.1016/j.datak.2006.01.013}
+
+\hypertarget{ref-tatsiramos_geographic_2009}{}
+Tatsiramos, K. (2009). Geographic labour mobility and unemployment
+insurance in europe. \emph{Journal of Population Economics},
+\emph{22}(2), 267--283.
+doi:\href{https://doi.org/10.1007/s00148-008-0194-7}{10.1007/s00148-008-0194-7}
+
+\hypertarget{ref-R-padr}{}
+Thoen, E. (2017). \emph{Padr: Quickly get datetime data ready for
+analysis}. Retrieved from \url{https://CRAN.R-project.org/package=padr}
+
+\hypertarget{ref-R-tibbletime}{}
+Vaughan, D., \& Dancho, M. (2018). \emph{Tibbletime: Time aware
+tibbles}. Retrieved from
+\url{https://CRAN.R-project.org/package=tibbletime}
 
@@
 
+\hypertarget{ref-R-scales}{}
+Wickham, H. (2017). \emph{Scales: Scale functions for visualization}.
+Retrieved from \url{https://CRAN.R-project.org/package=scales}
+
+\hypertarget{ref-R-tidyr}{}
+Wickham, H., \& Henry, L. (2018). \emph{Tidyr: Easily tidy data with
+'spread()' and 'gather()' functions}. Retrieved from
+\url{https://CRAN.R-project.org/package=tidyr}
+
+\hypertarget{ref-R-dbplyr}{}
+Wickham, H., \& Ruiz, E. (2018). \emph{Dbplyr: A 'dplyr' back end for
+databases}. Retrieved from
+\url{https://CRAN.R-project.org/package=dbplyr}
+
+\hypertarget{ref-R-dplyr}{}
+Wickham, H., Francois, R., Henry, L., \& Müller, K. (2017). \emph{Dplyr:
+A grammar of data manipulation}. Retrieved from
+\url{https://CRAN.R-project.org/package=dplyr}
+
+\hypertarget{ref-R-readr}{}
+Wickham, H., Hester, J., \& Francois, R. (2017). \emph{Readr: Read
+rectangular text data}. Retrieved from
+\url{https://CRAN.R-project.org/package=readr}
+
 \hypertarget{ref-wolf_impact_2003}{}
@@
 
+\hypertarget{ref-wu_spatial-temporal-semantic_2017}{}
+Wu, F., Fu, K., Wang, Y., Xiao, Z., \& Fu, X. (2017). A
+spatial-temporal-semantic neural network algorithm for location
+prediction on moving objects. \emph{Algorithms}, \emph{10}(2), 37.
+doi:\href{https://doi.org/10.3390/a10020037}{10.3390/a10020037}
+
 \hypertarget{ref-R-knitr}{}
@@
 
+\hypertarget{ref-R-kableExtra}{}
+Zhu, H. (2018). \emph{KableExtra: Construct complex table with 'kable'
+and pipe syntax}. Retrieved from
+\url{https://CRAN.R-project.org/package=kableExtra}
+
 \endgroup

### 2018-05-03 Thesis Report/thesisReport.Rmd
restructure and organise

--- 2017-12-01 thesisReport.Rmd
@@
 accuracy.
-$G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ...
+$G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ...
 < t_{n+1}$.
-For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window.
+For simplicity, let each entry in the discrete index set $t$ represent a 5 minute window.
 The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
 If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+
+When several data sets are available from individuals living in overlapping areas we can construct a $t \times i$ matrix $M$ where the entry $M(t,i)$ contains $G(t)$ for the individual $i$.
 

### 2018-05-03 PalmiusImplementation/results.R

@@ -1,0 +1,145 @@
+# Full Palmius Script Results
+
+#set directory to currently open script
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# read data in
+data<- readRDS("../ownalgo/finalBinnedDataNL.rds")
+
+# load auxiliary functions
+source("../../thesis/scripts/auxFuns.R")
+
+
+# read in all functions necessary
+file.sources <- list.files(c("functions"), 
+                           pattern="*.R$", full.names=TRUE, 
+                           ignore.case=TRUE)
+sapply(file.sources,source,.GlobalEnv)
+
+#select the month of march
+dataExample <- select(data,time,lat,lon,accuracy,nextMeas) %>%
+  arrange(time) %>% 
+  mutate( nextLon = lead(lon),
+          nextLat = lead(lat),
+          index = 1:length(lat),
+          nextTimeSec = nextMeas
+  ) %>% 
+  as_tbl_time(time) %>% 
+  filter_time( ~ '2017-03')
+
+# filter the data based on Palmius' filtering script
+filtered<- filteringPalmius(data = dataExample)
+
+# downsample the data based on Palmius' method
+downsampledPalmius <-  downSamplingPalmius(data = filtered)
+
+## calculate deviance
+
+deviancePalmius <- left_join(dataExample %>% thicken("5 min"),
+                  downsampledPalmius,
+          by =c("time_5_min" = "time_5min")) %>% 
+  select(lon,lat,lonF,latF,accuracy, time_5_min)
+
+deviancePalmius$dist <- raster::pointDistance(as.matrix(cbind(test$lon,test$lat),ncol = 2),
+                      as.matrix(cbind(test$lonF,test$latF),ncol = 2),
+                      lonlat = TRUE)
+
+# extract deviance measures
+summary(deviancePalmius$dist)
+
+# Calculate percentage withhin accuracy
+deviancePalmius %>%  summarise(mean(accuracy >= dist)) 
+
+# by bin
+deviancePalmius %>% group_by(time_5_min) %>% summarise(distMean = mean(dist),
+                                            distSD   = sd(dist)) %>% summarise(mean(distMean))
+
+# Create a plot to investigate results
+library(ggplot2)
+library(ggthemes)
+ggplot(deviancePalmius, aes(x=accuracy, y=dist)) +
+  geom_point()+theme_tufte()+
[…]

### 2018-05-03 PalmiusImplementation/scripts/implementation copy.R

@@ -1,0 +1,63 @@
+## Implementing measures from the paper
+#http://ieeexplore.ieee.org/document/7676335/
+
+# read data
+data<- readRDS("C:\\Users\\user1\\Desktop\\Life\\Professional\\Statistics Nederland\\gps\\boaz\\Takeout\\Location History\\myGPSRDS.rds")
+
+#or just a day?
+
+day <- readRDS("C:\\Users\\user1\\Desktop\\Life\\Academic\\Utrecht\\year2\\thesis\\kalmanFilters\\implementation\\feb17.rds")
+
+# part 1: Filtering
+
+## based on speed. How was speed calculated?
+
+# distance
+
+distances<- apply(day, 1, FUN = function(row) {
+  raster::pointDistance(c(as.numeric(as.character(row["lon"])),
+                  as.numeric(as.character(row["lat"]))),
+                c(as.numeric(as.character(row["lonNext"])), as.numeric(as.character(row["latNext"]))),
+                lonlat = T) # Parameter 'lonlat' has to be TRUE!
+})
+
+# output is in meters, convert to kilometers
+# does not match original, perhaps due to lat lon mixup in data pre processing file 
+
+distkm<- distances/1000
+time.hr<- as.numeric(day$timeNextMeasure, units="hours")
+
+speed<- distkm/time.hr
+
+day$speed <- speed
+
+#inspect
+
+library(leaflet)
+library(sp)
+
+speedingFilter<- which(day$speed>100)
+speeding<- day[speedingFilter,]
+notSpeeding <- day[-speedingFilter,]
+lns <- Line(cbind(day$lon,day$lat))
+
+#addCircles(lng = ~lon, lat = ~lat, weight = 1,
+#           radius = ~accuracy,fillOpacity = 0.02,color = "#DF2935")%>%
+  
+
+leaflet(day)%>%
+  addTiles() %>%
+#  addPolylines(data = lns,color = "#DF2935")%>%
+  addProviderTiles(providers$CartoDB.Positron) %>%
+  addCircles(data=speeding,lng = ~lon, lat = ~lat, weight = 1,
+             radius = ~accuracy,fillOpacity = 0.02,color = "#5F53AD")%>%
+  addPolylines(data = lns2,color = "#5F53AD")
+
+# basically it gets stuck there. Remove those locations?
+#stopped here
+
+pointsToRemove<- day[which(day$speed>100),c("lon","lat")]
+
[…]

### 2018-05-03 implementingBipolar/aggregateResults.R

@@ -1,138 +1,0 @@
-# results: aggregation time spent at home
-
-leaflet(allPoints) %>% 
-  addTiles() %>% 
-  addCircles(lng = ~lon, lat = ~lat, label = ~as.character(clust))
-
-home <- c(5.114051,52.10415)
-
-test <- fin %>% 
-  as_tbl_time(index = time) %>%
-  filter_time(~ "2017-03") %>% 
-  select(time, clust, lon, lat) %>% 
-  thicken("5 min", "thick") %>% 
-  left_join(., allPoints, by = "clust", suffix = c(".measured",".clust")) %>% 
-  mutate( distHClust = raster::pointDistance(p1 = home, p2 = as.matrix(cbind(lon.clust,lat.clust)), lonlat = T))
-
-library(ggplot2)
-plot(test %>%as_tbl_time(time) %>% 
-       filter_time(~"2017-03-07") %>% pull(distH))
-
-test %>%as_tbl_time(time) %>% 
-  filter_time(~"2017-03-07") %>%
-  filter(distH<400) %>% 
-leaflet() %>% 
-  addTiles() %>% 
-  addCircles(lng = ~lon.clust, lat = ~lat.clust, label = ~as.character(clust))
-
-#clusts 49, 348 and 336
-
-test2 <- test %>% group_by(thick) %>% 
-  summarise(lon  = mean(lon.measured),
-            lat  = mean(lat.measured),
-            clust = Mode(clust),
-            home1 = clust %in% c(49,348,336),
-            home2 = raster::pointDistance(p1 = home, p2 = as.matrix(cbind(lon,lat)), lonlat = T))
-
-test2 %>%  group_by(home1) %>% summarise(m = min(home2), mean = mean(home2), max = max(home2))  
-
-test2 %>%  filter ( home1 == T) %>% 
-  leaflet() %>% 
-  addTiles() %>% 
-  addCircles(lng = ~lon, lat = ~lat)
-
-
-test2 %>% pad() %>% summarise(atHomeProp = sum(home1, na.rm = T)/nrow(.),
-                              missingProp = sum(is.na(home1))/nrow(.))
-
-# predicted values
-t <- test2 %>%  pad()
-t[which(t$thick %in% rems),"clust"] <- as.integer(model %>% predict_classes(pred))
-
-# model prediction
-t %>% mutate( home1 = clust %in% c(49,348,336)) %>% summarise(atHomeProp = sum(home1)/nrow(.))
-
[…]

### 2018-05-03 implementingBipolar/fullPalmiusResults.R

@@ -1,147 +1,0 @@
-# Full Palmius Script Results
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-data<- readRDS("../ownalgo/finalBinnedDataNL.rds")
-
-# load auxiliary functions
-source("../../thesis/scripts/auxFuns.R")
-
-file.sources <- list.files(c("functions"), 
-                           pattern="*.R$", full.names=TRUE, 
-                           ignore.case=TRUE)
-sapply(file.sources,source,.GlobalEnv)
-
-#select the month of march
-data <-  fin
-dataExample <- select(data,time,lat,lon,accuracy,nextMeas) %>%
-  arrange(time) %>% 
-  mutate( nextLon = lead(lon),
-          nextLat = lead(lat),
-          index = 1:length(lat),
-          nextTimeSec = nextMeas
-  ) %>% 
-  as_tbl_time(time) %>% 
-  filter_time( ~ '2017-03')
-
-filtered<- filteringPalmius(data = dataExample) #no unique locations removed
-
-downsampled <-  downSamplingPalmius(data = filtered) #change name of time column, lat and lon
-
-## calculate deviance
-
-test <- left_join(dataExample %>% thicken("5 min"),
-          downsampled,
-          by =c("time_5_min" = "time_5min")) %>% 
-  select(lon,lat,lonF,latF,accuracy, time_5_min)
-
-test$dist <- raster::pointDistance(as.matrix(cbind(test$lon,test$lat),ncol = 2),
-                      as.matrix(cbind(test$lonF,test$latF),ncol = 2),
-                      lonlat = TRUE)
-
-summary(test$dist)
-
-test %>%  summarise(mean(accuracy >= dist)) #each measurement
-# by bin
-test %>% group_by(time_5_min) %>% summarise(distMean = mean(dist),
-                                            distSD   = sd(dist)) %>% summarise(mean(distMean))
-library(ggplot2)
-library(ggthemes)
-ggplot(test, aes(x=accuracy, y=dist)) +
-  geom_point()+theme_tufte()+
-  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0,
-              na.rm = FALSE, show.legend = NA)
-
-
-#there is something wierd happening with these outliers
-test %>% 
-  filter(dist>100000) %>% 
-  leaflet() %>% 
[…]

### 2018-05-03 ownAlgo/imgCreation.R

@@ -1,3 +1,0 @@
-# image creation
-install.packages("magick")
-library(rmagick)

### 2018-05-05 Correspondence/Paths & clusters/Clusters.Rmd
finishing touches

---
title: "Finding Endpoints"
author: "Boaz Sobrado"
date: "3/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(leaflet)
test2 <- readRDS("clusters.rds")
data2 <- readRDS("data2.rds")
```

## Clusters

Endpoints are defined as locations where an individual spends at least an X amount of time within a radius r. It is assumed that an individual will often (but not always) visit an endpoint several times. An endpoint can be home, work, a bus or train station, a bar etc.

Palmius extracts endpoints by looking at the speed between measurements. This requires downsampling and does not take into account data drift and the fact that people tend to visit the same locations several times.Barnett does so by looking at the time that an individual spends within a radius. Then, once all endpoints are extracted endpoints close to each other are merged into a single endpoint.

There are abundant other endpoint extraction methods in the literature. Such as  [Fu et al.](http://www.mdpi.com/2220-9964/5/10/166) and  [Chen et al](http://adsabs.harvard.edu/abs/2014ISPAr.XL2...55C).

I will not dwell on those but rather use a rather simple model for now. In the future, it might be worth using other methods.

First, let us take the data and extract only the bits within the Netherlands, then we'll take the pauses. Pauses are defined as a measurement where the next measurement and the previous measurement is within X amount of seconds (so it is not missing) as well as

```{r dataExtract, eval=F}
# not actually run in the RMD 
home<- c(5.113919,52.10421)
test <- data %>% mutate( distanceHome = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                                              home,
                                                              longlat = T),
                         timestampMs = as.numeric(time))%>%
  filter(distanceHome < 300000) %>% 
  select(time,lon,lat,accuracy, timestampMs,distanceHome) %>%
  arrange(timestampMs) %>%

rm(data)

timeLim <- 600 # time limit between measurements in s
distLim <- 100 # distance limit in m 
minPause <- 60 # minimum pause length in s

test2 <- test %>% filter(accuracy <25) %>% # get high accuracy measurements only 
  mutate( prevLon = lag(lon),
                   prevLat = lag(lat),
                   nextLon = lead(lon),
                   nextLat = lead(lat),
                   nextDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                                    matrix(c(nextLon,nextLat),ncol = 2),
                                                    longlat = T),
                   prevDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                                    matrix(c(prevLon,prevLat),ncol = 2),
                                                    longlat = T),
                   nextMeas = as.numeric(timestampMs-lead(timestampMs)),
                   prevMeas = as.numeric(lag(timestampMs)-timestampMs),
                   isPause = case_when(prevMeas <= timeLim &nextMeas <= timeLim & nextDist <= distLim~1,
                                       TRUE ~0)
) %>%
  group_by(run = {run = rle(isPause); rep(seq_along(run$lengths), run$lengths)}) %>%
  summarize(isPause = mean(isPause),
            lon = mean(lon),
            lat = mean(lat),
            t0  = min(timestampMs),
            t1  = max(timestampMs),
            duration = t1-t0)

#perhaps take weighted mean in the future to reduce error

```

Here we get the continuous pauses grouped together, as well as the length of each pause in seconds. See below for points filtered where the duration is more than 10 minutes.

```{r mapres}
test2 %>% filter( isPause == 1 & duration > 600) %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat,
             radius = 30, fillOpacity = 0.02,color = "#18206F",label = ~as.character(duration))
```

However, many pauses are next to each other. In order to segment them I use the following code:

```{r segment}
toClust <- test2 %>% filter( isPause == 1 & duration > 600) %>%
  select(lon,lat) %>% as.matrix()

library(sp)
library(rgdal)
library(geosphere)

# convert data to a SpatialPointsDataFrame object
xy <- SpatialPointsDataFrame(
  toClust, data.frame(ID=seq(1:nrow(toClust))),
  proj4string=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))

# use the distm function to generate a geodesic distance matrix in meters
mdist <- distm(xy)

# cluster all points using a hierarchical clustering approach
hc <- hclust(as.dist(mdist), method="complete")

# define the distance threshold, in this case 200 m
d=200

# define clusters based on a tree "height" cutoff "d" and add them to the SpDataFrame
xy$clust <- cutree(hc, h=d)

xy <- cbind(xy@data,xy@coords)

xy %>% as.data.frame() %>% 
  group_by(clust) %>% 
  summarise(lon = mean(lon),
            lat = mean(lat),
            n   = n()) %>% 
  leaflet() %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat,
             radius = 50, fillOpacity = 0.02,color = "#18206F",label = ~as.character(n))

clustMap <- xy %>% as.data.frame() %>% 
  group_by(clust) %>% 
  summarise(lon = mean(lon),
            lat = mean(lat),
            n   = n())

```

As far as I can tell this does a pretty good job. The next steps are the following:
1. Assign each measured point to the nearest cluster
2. Check whether the nearest point is within the measured accuracy.
3. Check whether the measured point is within a "pause".

As a result we will be able to extract the nearest travelled paths.

```{r cs, eval = F}
data <- test
distance <-  raster::pointDistance(data[,2:3], clustMap[,2:3], longlat = T)
clust<- apply(distance,MARGIN = 1, FUN = which.min)
data$distClustMin<- apply(distance,MARGIN = 1, FUN = min)
data$clust <- clust

data2<- data %>%
  arrange(timestampMs) %>% mutate(
  withinAcc = case_when(accuracy >= distClustMin ~1,
                        TRUE ~ 0),
  prevLon = lag(lon),
  prevLat = lag(lat),
  nextLon = lead(lon),
  nextLat = lead(lat),
  nextDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                   matrix(c(nextLon,nextLat),ncol = 2),
                                   longlat = T),
  prevDist = raster::pointDistance(matrix(c(lon,lat),ncol = 2),
                                   matrix(c(prevLon,prevLat),ncol = 2),
                                   longlat = T),
  nextMeas = as.numeric(timestampMs-lead(timestampMs)),
  prevMeas = as.numeric(lag(timestampMs)-timestampMs),
  isPause = case_when(prevMeas <= timeLim &nextMeas <= timeLim & nextDist <= distLim~1,
                      TRUE ~0),
  pauseClust = case_when(isPause == 1 & withinAcc == 1~ as.numeric(clust),
                         TRUE ~NA_real_),
  nextPauseClust = pauseClust,
  prevPausClust = pauseClust
) %>% 
  fill(nextPauseClust,.direction ="up") %>%
  fill(prevPausClust,.direction ="down")
```

Then we can map out the most common paths between two endpoints.

```{r pathBetweenCluster}
checkPath <- function( data2 = data, start = 14, end = 3){
  data2 %>%
    filter(nextPauseClust == end & prevPausClust == start) %>% 
    leaflet() %>% 
    addTiles() %>% 
    addCircles(lng = ~lon, lat = ~lat, fillOpacity = 0.02,color = "#18206F",label = ~as.character(clust)) %>% 
    addCircles(data = filter(clustMap, clust ==end),lng = ~lon, lat = ~lat, color = "red", radius = 100) %>% 
    addCircles(data = filter(clustMap, clust ==start),lng = ~lon, lat = ~lat, color = "green", radius = 100)
}

checkPath(data2, start = 144, end = 3)
checkPath(data2, start = 48, end = 36)
checkPath(data2, start = 55, end = 139)
```

Here you can see a cross section of paths and issues. Some are just sparse, some have a lot of noise, and some have other issues (e.g. underground signal loss).

### 2018-05-05 Thesis Manuscript/apaManu.Rmd
finishing touches

--- 2018-05-03 manuscript/apaManu.Rmd
@@
 The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data.
-Nonetheless, smartphone collected personal mobility data has remained underutilised in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it.
+Nonetheless, smartphone collected personal mobility data has remained underused in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it.
 Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain.
@@
 The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data.
-In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurments and impute missing data values.
+In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurements and impute missing data values.
 By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals.

### 2018-05-05 PPMI/mapBuilding.R

@@ -81,10 +81,11 @@ fin<- rbind(pause, notpause) %>% arrange(timestampMs)
 rm(pause,notpause,routePoints,pathClust,test3,clusterMap,toClust,test2)
 Sys.time()
 
+# save data frames if needed
 saveRDS(fin,file = "finalBinnedDataNL.rds")
 saveRDS(allPoints, file = "finalAllClusts2.rds")
 
-# load 
+# load if needed.
 
 fin <- readRDS("finalBinnedDataNL.rds")
 allPoints <- readRDS("finalAllClusts2.rds")

### 2018-05-08 Thesis Manuscript/manuscript.Rmd
final commit

--- 2018-05-05 Thesis Manuscript/apaManu.Rmd
@@
 
-  Submitted as a research report conforming to APA manuscript guidelines (6th edition).
+  Submitted as a thesis manuscript conforming to APA manuscript guidelines (6th edition).
 
 abstract: |
-  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health.
+  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and health.
 The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data.
@@
 The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.
-This results in a high proportion of missing data and significant noise.
+This results in a high proportion of missing data.
 In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs.
@@
 output            : papaja::apa6_pdf
+keep_tex: true
 
@@
 Why is human mobility important?
+Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010].
 Human mobility measures are quantified metrics of how people move about in their environment.
-Human mobility affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010].
 The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@goldberg_international_2006].
@@
 The extensive use of measures like these across different domains in social science strongly suggests that mobility metrics are linked to important real world outcomes.
-Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017].
-Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough.
-When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space.
-To fully understand behaviour we must understand how behaviour can vary across environments.
+Behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017].
+To fully understand behaviour not only must we understand how behaviour can vary within and between individuals, but also across environments.
 
-Despite the importance of mobility measures, the majority of these metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries.
+Despite the importance of mobility measures, the majority of  metrics are obtained through the use of questionnaires (such as the aforementioned personality questionnaire) or specifically through pen-and-paper travel diaries.
 As a thought experiment we encourage the reader to try to remember where exactly he or she was five weeks ago on Saturday at midday.
-Most people find this sort of task very difficult (if not impossible), which illustrates that questionnaires and travel diaries are have well-known methodological flaws.
-This method of collecting data is difficult and relies on accurate self-reporting.
-Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals due to human forgetfulness.
-This makes the data collection expensive and limits the duration of data collection in practice.
-In addition, the frequent reporting duties of the participant may bias the participants behaviour.
+Most people find this sort of task difficult (if not impossible), which illustrates the main methodological flaws of questionnaires and travel diaries.
+Travel diaries are burdensome to collect because participants must be explicitly asked to write down on their movement patterns at frequent intervals (lest they forget them).
+The burden to the participants makes data collection expensive, which in turn limits the duration of data collection in practice.
+In addition, the frequent reporting duties of the participant may bias the participant's behaviour.
 The limits of human cognition also limit the accuracy of self-reported measures.
-There is evidence that participants are systematically biased when self-reporting mobility measures.
+There is evidence that statistics derived from self-reported mobility measures are systematically biased.
 For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017].
@@
 Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017].
-Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.
+Researchers now have the ability to easily access the objective movement records of millions of individuals spanning several years at little cost and without a significant burden to the participants.
 
@@
 We call location logs resulting from these custom-made apps *custom logs*.
-The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor with which they are made.
-The disadvantage with this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation.
+The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor used to make the measurement.
+The disadvantage of this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation.
 Participants may dislike tracking apps because they view them as more intrusive and these apps regularly drain the battery of the device [@harari_using_2016].
-Moreover, researchers have distribute this application among research participants and convince them not to turn it off.
 
-We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*) .
-The advantages are clear: repositories such as Google's Location History contains information on millions of users spanning years [@location_history_timeline_2017], participants can share the data by the click of a button and there can be no behavioural changes due to participation in the study as the participant share past data.
+We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*).
+The advantages are clear: repositories such as Google's Location History contain years of information on millions of users [@location_history_timeline_2017] and participants can share the data by clicking of a button.
 The disadvantage is that researchers have no control over the tracking parameters, often resulting in logs with sparse and inaccurate measurements.
-Hence, two important challenges are dealing with missing data and measurement noise.
+Hence, two important challenges in using secondary logs are how to deal with missing data and measurement noise.
 
@@
 Technical reasons include signal loss, battery failure and device failure.
-Behavioural reasons include leaving the phone at home or switching the device off.
+Behavioural reasons include switching the device or its location services off.
 As a result, secondary logs often contain wide temporal gaps with no measurements.
@@
 @harari_using_2016;@jankowska_framework_2015].
-In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparsity it is also evident.
+Figure \@ref(fig:longMeasurementsPerDay) shows a secondary log spanning years where despite its long duration the sparsity of measurements is also evident.
 
@@
 Importantly, spatiotemporal data measurements are auto-correlated in both time and space.
-This means that best practices with other types of data, such as mean imputation, are unsuitable.
+This means that common practices with other types of data, such as mean imputation, are unsuitable.
 For example, imagine an individual who splits almost all her time between work and home.
@@
 
-Another methodological problem is related to the noise in the measurements that are collected.
+Another methodological problem is related to the noise in the measurements.
 The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors.
@@
 Suppose we wish to calculate an individual's movement in a day.
-A simple approach would be to calculate the sum of the distance between each measurement.
+A simple approach would be to calculate the sum of the distances between each measurement.
 But if there is noise, the coordinates will vary even though the individual is not moving.
 If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all!
-This issue is also visualised in Figure \@ref(fig:palmiusVsMe1).
 The problem is further complicated because missing data and noisy measurements are related.
@@
 In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs.
-We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simulated as missing and their imputed  counterparts.
-Finally we will calculate time spent at home as a function of the imputation method.
+PPMI creates a discrete personalised map based on the individuals mobility history.
+Then it assigns each measurement to a discrete location based on the individuals entire mobility history.
+Finally, it uses a classification method to impute missing periods.
 
@@
 This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons_false-positive_2011].
-The vagueness is possibly also due to the fact that most researchers are unfamiliar with possible solutions.
-Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g.
-@palmius_detecting_2017).
-The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016.
-We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.
+Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations [e.g.
+@palmius_detecting_2017].
+The only principled approach (that we know of) aiming to solve the issue of missing data in this context (i.e.
+location logs as they relate to human mobility) is that of @barnett_inferring_2016.
+We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing other spatiotemporal methods.
 
@@
 Ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008].
-These models can be quite complex.
 @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements.
@@
 Because SSMs are based on the Markov property, they cannot incorporate this information.
-In other words, the estimated location $G(t)$ at time-point $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$.
+In other words, the estimated location at a time-point $t$ is often based only upon measurements at timepoints $t$ and ${t-1}$.
 Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models.
 These solutions are often computationally intractable or unfeasible [@sadilek_far_2016].
-Moreover, these models often assume time and space invariance (location is not a direct function of time or space).
+Moreover, these models often assume time and space invariance (i.e.
+location is not a direct function of time or space).
 These mathematical assumptions are violated in the case of human movement patterns.
-For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.
+For this reason we do not consider existing SSMs to be useful for imputing missing data in smartphone mobility logs.
 
@@
 
-On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @@wu_spatial-temporal-semantic_2017].
-Importantly, none of them worked with secondary logs.
+On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu_spatial-temporal-semantic_2017].
+Importantly, none of them worked with secondary logs, but rather with custom logs that have higher sampling frequency.
 Nonetheless we will detail what they did as informative examples.
-@palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$.
+@palmius_detecting_2017 deal with the measurement inaccuracy in custom logs by removing from the data set all unique low-accuracy data points with a speed greater than 100 km/h.
 Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter.
@@
 
->"If the standard deviation of [$D$] in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".
+>"If the standard deviation of in both latitude and longitude within a 1 h epoch was less than 0.01 km, then all samples within the hour were set to the mean value of the recorded data, otherwise a 5 min median filter window was applied to the recorded latitude and longitude in the epoch".
 
-Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm.
+Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ if this occurred after 9pm.
 In cases where the previous conditions are not met no values are imputed.
@@
 @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes.
-In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses.
+@barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses.
 Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move.
@@
 
-Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidean plane.
+Subsequently let us simplify by assuming that a persons location is on a two-dimensional Euclidean plane.
 Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively.
@@
 < t_{n+1}$.
+If $D_t = \emptyset$ it is considered *missing*.
 The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle.
-If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.
+The reader should keep in mind that a high accuracy value $a_t$ for a measurement $D_t$ means that the measurement is inaccurate (this can be rather confusing).
 
-When several data sets are available from individuals living in overlapping areas we can construct a $t \times i$ matrix $M$ where the entry $M(t,i)$ contains $G(t)$ for the individual $i$.
-
-## Personalised Map Matching Imputation
+## Personalised Map Matched Imputation
 
@@
 $$t_a = t_i < ... < t_{i+n}$$
-Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$:
-$$t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j$$
+Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$ and $t_b$ is the first pause after the end of $t_a$:
+$$t_b = t_j < ... < t_{j+m}|t_{i+n}< t_j$$
 Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
@@
 = G(t_{an})$.
-Moreover,it is possible for $G(t_a) = G(t_b)$ such that $t_a \ne t_b$.
+Moreover,it is possible for two consecutive pauses $t_a$ and $t_b$ to share a location $G(t_a) = G(t_b)$ such that $t_a \ne t_b$.
 For example, if the individual leaves home for a run and returns home without stopping anywhere else.
@@
 
-## Personalised Map Matching Imputation algorithm
+## Personalised Map Matched Imputation algorithm
 
@@
 
-1. *Map building*: Extract from measurements $D$ all pause location bins and path location bins to create a personalised map.
-2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
+1. *Map building*: Extract from measurements $D$ all pause locations and path locations to create a personalised map consisting of discrete bins.
+2. *Binning*: Assign each measurement $D$ to a unique discrete location. 
 3. *Imputing*: Use a classification method to predict missing measurements based on all the available information.
@@
 
-Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences.
+Following @wu_spatial-temporal-semantic_2017's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete locations, which we will call location bins.
 There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007].
-We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
+We will focus on methods which explicitly deal with mobility patterns extracted from smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].
 Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity.
@@
 4.
-Let the measurements of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$.
-These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.
+Let the measurements of a possible pause which fit the aforementioned criteria be $X = D_{t,t+1,...,t+n}$.
+Let $\bar{x}$ be the mean coordinates of $X$, let $x_{\text{dist}} \in X$ be the furthest away points from $\bar{x}$ and $\bar{a}$ be the the mean accuracy of $X$.
+Then $X$ is a pause if and only if the distances between $\bar{x}$ and $x_{\text{dist}}$ are $\leq 2 \times \bar{a}$.
 
-This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.
+The resulting set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location.
+Each pause location is a bin.
 
@@
 2.
-Create a distance matrix for all remaining measurements $D_t \in F$ and hierarchically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.
+Create a distance matrix for all remaining measurements $D_t \in F$ and hierarchically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single flight bin.
 
@@
 This means that if a pause bin and a path bin are within less than $d$ meters of each other, the path bin is removed.
-The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster.
+The reasoning for this is that the threshold for not being in a pause bin should be higher, as individuals spend the majority of time at a pause bin.
 The end result is a discretised map which contains pause and flight bins based on the entire log history of the individual.
@@
 PMMI uses a similar logic, but without using any external geographic database.
-The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses a map from outside the persons location logs, we use the total location history of the individual to create a personalised map.
+The key modification in PMMI is that whilst @wu_spatial-temporal-semantic_2017 uses an external map, we use the total location history of the individual to create a personalised map.
 This map is subsequently used to bin measurements.
@@
 
-Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map.
+Thus, all measurements $D$ are then assigned to a discrete bin on the personal map.
 This includes previous measurements which were discarded from the map building exercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$.
-In this implementation we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.
+In this implementation we used a simple assigning function, whereby the measurements where assigned to the nearest bin.
 
@@
 In this implementation we trained an artificial neural network (ANN) to do so.
-For more information on the precise architecture of the artificial neural network please consult the appendix.
+The ANN has a simple architecture: it consists of an input layer, a hidden layer and an output layer.
+The input layer and the hidden layer employ a rectified linear unit activation function and the output layer uses a softmax activation function.
+
 The input variables to the ANN are:
@@
 3. The time of the day encoded as a cyclical two-dimensional feature.
-4. The day of the month as a binary class matrix.
+4. The day of the week as a binary class matrix.
 5. The month of the year as a binary class matrix.
@@
 For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london_encoding_2016].
-This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.
+This is necessary for the model to be able to understand that one second past midnight and one second before midnight are in fact two seconds away from each other.
 Moreover we scaled the non-binary values to occupy a range between 0 and 1 in order to ensure convergence.
 
-For a missing time-point at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster.
+For a missing time-point at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location bin.
 That is, for each missing time-point the model returns a vector of probability estimates (with one estimate per bin) associated with where the individual is.
@@
 About 54% of the data is missing for the entire duration of the log.
-This may be misleading as there are several long periods with no measurements whatsoever.
+There are several long periods with no measurements whatsoever, as well as periods with far less missing data.
 For days which were not entirely missing, approximately 22% of all five minute segments were missing.
-The structure of missingness of a day with measurements is shown in Figure \@ref(fig:measurementsPerDay).
+The structure of missingness of a non-missing day is shown in Figure \@ref(fig:measurementsPerDay).
 As you can see, there are several long periods over the course of the log for which there are no measurements.
@@
 
-@palmius_detecting_2017's algorithm was implemented in R based on the original MATLAB code and pseudocode kindly provided by the author.
-@barnett_inferring_2016's method was slightly adapted in R to fit the the data structure from the original R code provided by the author.
+@palmius_detecting_2017's algorithm was implemented in R based on the original MATLAB code and pseudocode kindly provided by the authors.
+@barnett_inferring_2016's method was slightly adapted in R to fit the data structure from the original R code provided by the authors.
 When executing their models we used the same parameters as the authors did.
@@
 
-All the code is available on a public repository [@sobrado_my_2018].
+All the code is available on a public repository[^1] .
+
+[^1]: [Github ](https://github.com/sobradob/thesis) [@sobrado_my_2018]
 
@@
 1. Evaluating the performance of the map building and assigning functions.
-2. Comparing the performance of PMMI compared to alternative methods [[@palmius_detecting_2017; @barnett_inferring_2016]] and a baseline model using cross-validation.
-3. Comparing the performance of PPMI compared to the aformentioned methods using objective ground-truth data (public transportation time-stamps locations).
+2. Comparing the performance of PMMI compared to alternative methods [@palmius_detecting_2017; @barnett_inferring_2016] and a baseline model using cross-validation.
+3. Comparing the performance of PPMI compared to the aformentioned methods using objective ground-truth data (public transportation time-stamped locations).
 
@@
 All methods smoothen noise differently in an attempt to extract true location $G(t)$ from measurements $D_t$.
-In the subsequent cross-validation step we will remove random smoothened blocks and compare the imputed value to the smoothened value.
-For this reason we want to ensure that the smoothened method actually does reflect the true location $G(t)$.
+In the subsequent cross-validation step we will remove random smoothened 5 minute periods and compare the imputed value to the smoothened value.
+For this reason we want to ensure that the smoothened value actually does reflect an approximation to the true location $G(t)$.
 Otherwise we run the risk of over-estimating the accuracy of a model in the sense that we will compute the extent to which an imputation method can correctly impute a smoothened value which has little to do with the true location $G(t)$.
@@
 In order to evaluate the map building and binning we will first visually evaluate the paths and pause locations.
-A visual evaluation of paths superimposed on is an established way to heuristically check their accuracy [e.g.
+A visual evaluation of paths superimposed on a map is an established way to heuristically check their accuracy [e.g.
 @brunsdon_path_2007].
@@
 
-1. A positive relationship between the deviation distance and the accuracy of each measurement.
+1. A positive relationship between the deviation distance and the accuracy $a$ (in meters) of each measurement.
 2. Roughly 67% of the deviation distances $\delta_{\text{dev}}$ are within accuracy $a$ of each measurement.
@@
 
-In addition, other measures of interest for PMMI are *accuracy* (in what percentage of the cases was the appropriate cluster predicted), the *confidence* and the *distance expectation*.
-The confidence is the probability with which the model predicts the most likely cluster.
+In addition, other measures of interest for PMMI are *accuracy* (in what percentage of the cases was the appropriate bin predicted), the *confidence* and the *distance expectation*.
+The confidence is the probability with which the model predicts the most likely bin.
 For instance, if the model predicts the missing bin to be bin A with a probability of 0.9 then we can say this is a high confidence prediction.
-Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given clusters and the distances between the clusters to the true cluster.
+Similarly, the distance expectation is the cross-product of the estimated probabilities that the individual is at any of all given bins and the distances between the bins to the true bin.
 For example, if the true location of an individual is bin A (bin A is 1000 meters away from bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at bin B, then the distance expectation would be 100 meters.
@@
 We will compare the performance of PMMI, @palmius_detecting_2017, @barnett_inferring_2016 and the naive baseline model with regards to imputing real missing measurments.
-To do will use information from the Dutch public transportation card.
+To do so we will use information from the Dutch public transportation card.
 The Dutch public transportation service provides users with time-stamped location data of when and where they board, change lines or leave public transportation.
@@
 We used the following parameters to extract pauses: an accuracy limit $a_{\text{Pause lim}}$ of 250 meters, a time limit $t_{\text{Pause lim}}$ of 300 seconds, a distance limit of $d_{\text{Pause lim}}$ 50 meters and a minimum pause duration limit $\delta_{\text{Pause lim}}$ of 100 seconds.
-Moreover, to extract path clusters we used the parameters: $a_{\text{Path Lim}} = 150  \text{meters}$ and $d_{\text{Path Lim}} =300  \text{meters}$.
+Moreover, to extract path bins we used the parameters: $a_{\text{Path Lim}} = 150  \text{meters}$ and $d_{\text{Path Lim}} =300  \text{meters}$.
 
@@
 1. Plausible extraction of pause locations. For example, a pause in a field with no path leading to it is less plausible than a pause at a train station, airport, or office building.
-2. A suitable extraction of paths. For instance, are ensuring that individual paths are complete and not merged unneccessarily in the case of closely parallel paths.
+2. A suitable extraction of paths. Ensuring that individual paths are logically consistent (e.g. do not go over rivers or other unplausable behaviour).
 3.
-Less than a fifth of extracted clusters are assigned only one measurement.
-Unique clusters are unavoidable as people sometimes visit a location briefly and only once, nonetheless they pose problems for the classification algorithms.
+Less than a fifth of extracted bins are assigned only one measurement.
+Unique bins are unavoidable as people sometimes visit a location briefly and only once, nonetheless they pose problems for the classification algorithms.
 
 When selecting parameters there is a trade off is between bias and precision.
-This is because an increase in precision (in the form of a higher resolution of locations location) comes at the expense of precision (assigning measurements to bins becomes more difficult).
+This is because an increase in precision (in the form of a higher resolution of bin locations) comes at the expense of bias (assigning measurements to bins becomes more difficult).
 For instance, by increasing the clustering distance parameter $d_{\text{Pause lim}}$ we can extract more valid pause locations at the risk of falsely categorising certain measurements to the wrong bin.
@@
 
-Map building results in a personalised map with pause and path clusters.
+Map building results in a personalised map with pause and path bins.
 An excerpt can be seen in Figure \@ref(fig:mapClusterMap).
-It is important to remind the reader that PMMI is map agnostic and uses no information from the map.
-Therefore, the close overlap with features on the map, such as pause bins at relevant buildings and transportation clusters, as well as the flight bins following roads and railway lines indicate a high degree of precision in personal map building.
+It is important to remind the reader that PMMI is map agnostic and uses no external information.
+Therefore, the close overlap with features on the map, such as pause bins at relevant buildings and transportation locations, as well as the flight bins following roads and railway lines indicate a high degree of precision in personal map building.
 As expected, PMMI's path extraction yields greater accuracy for frequently occurring paths.
@@
 
-```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
+```{r, mapClusterMap,echo = FALSE,fig.cap="Excerpt of the personalised binned map of an individual. Red points are pause locations, blue points are path locations.",out.width = '100%'}
 #fix width
@@
 
-For the entire period examined period the we find a deviance of 40 meters and a median deviance of 15 meters.
-Around 69% of the deviance values are within their corresponding accuracy value, which is close to the theoretical 67% value that is expected.
-Approximately 9% of values were not taken into account when creating the bins,given that their the accuracy $a$ exceeded $a_{\text{Path Lim}} = 300$.
+For the entire period examined period the we find a mean deviance $\delta_{\text{dev}}$ of 40 meters and a median deviance of 15 meters.
+Around 69% of the deviance values are within their corresponding accuracy value, which is close to the expected theoretical 67%.
+Approximately 9% of values were not taken into account when creating the bins, given that their the accuracy $a$ exceeded $a_{\text{Path Lim}} = 300$.
 
 In comparison, @palmius_detecting_2017's method has a median deviance $\delta_{\text{dev}}$ of 3 meters, with a mean of 115 due to high deviance outliers.
-On the other hand, @barnett_inferring_2016's method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
-@barnett_inferring_2016's deviance is necessarily higher than @palmius_detecting_2017', as they down-sample temporally (like @palmius_detecting_2017) and subsequently aggregate into pauses and linear flights.
+On the other hand, the @barnett_inferring_2016 method has mean deviance $\delta_{\text{dev}}$ of 343 meters and a median deviance of 8 meters.
+@barnett_inferring_2016's deviance is necessarily higher than @palmius_detecting_2017, as they down-sample temporally [like @palmius_detecting_2017] and subsequently aggregate into pauses and linear flights.
 
-The key difference between temporal and spatial down-sampling is shown in Figure \@ref(fig:palmiusVme1).
+The key difference between temporal and spatial down-sampling is shown in Figure \@ref(fig:palmiusVsMe1).
 Temporal down-sampling is much more sensitive to noise in sparsely measured periods because it averages out values within five minute periods.
-Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay) ), which leads to a noise in the down-sampled values.
+Often there are only a few noisy measurements in those periods (see Figure \@ref(fig:longMeasurementsPerDay)  which shows that a large proportion of 5 minute periods have less than 3 measurements), which leads to a noise in the down-sampled values.
 Unsurprisingly, there is a positive relationship between deviance and the amount of measurements in each down-sampled interval.
@@
 The first factor we must take into account is the coverage of each imputation model.
-The results show us that while PPMI and the naive baseline model impute all missing values, the other two models have difficulties doing so (Table 1).
+The results show us that while PPMI and the baseline models impute all missing values, the other two models have difficulties doing so (Table 1).
 In particular, the models of @palmius_detecting_2017 and @barnett_inferring_2016's fail to impute an increasing amount of missing values with an increase in the duration of the missing period.
-@palmius_detecting_2017's method in particular failed to impute a single missing time period.
+@palmius_detecting_2017's method in particular failed to impute a single missing time period when the duration of the removed periods was one day.
 The longer a log, the higher the probability that long time periods will be missing.
@@
 With respect to the distance metrics, Table 1 shows the results of the distance metrics for each method.
-We can see that PPMI predicted the location perfectly for over half of the removed values in all three time categories.
+We can see that PPMI predicted the location perfectly for at least about half of the removed values in all three time categories.
 To be precise PPMI's prediction accuracy was 88%, 73% and 47% for the 5 minute, 1 hour and 1 day periods respectively.
@@
 The explanation is that the alternative models failed to impute more difficult time periods.
+This increases the magnitude of errors of the PPMI method.
 
-On the other hand, the confidence score for each time period were heavily skewed left, withh a median of 1.
+On the other hand, the confidence score for each time period were heavily skewed left, with a median of 1.
 Accordingly, the distance expectation (the crossproduct of the model probabilities of an estimate with the distances of the true value to the estimates) values were quite similar to the distance scores, albeit approximately 5% higher.
@@
 In addition, the comparison to the performance of the @barnett_inferring_2016 and @palmius_detecting_2017 models is somewhat unfair, as they were created for custom logs, not secondary logs.
-Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone GPS logs.
+Nonetheless, the comparison remains valid as they are the closest we found to a missing data imputation methods in smartphone location logs.
 
@@
 A known-drawback of single imputation is that it takes an imputed value and treats it as observed.
-Simple rule-based methods such as @palmius_detecting_2017' are essentially algorithmic single imputation methods.
+Simple rule-based methods such as @palmius_detecting_2017's are essentially algorithmic single imputation methods.
 With PPMI it is possible to model uncertainty using the predicted probabilities of each estimate.
@@
 
-While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with with a close degree of spatial proximity, neither of them do the same for non-pause locations.
+While both @barnett_inferring_2016 and @palmius_detecting_2017 use historical data to smoothen pause locations by clustering pause locations with a close degree of spatial proximity, neither of them do the same for non-pause locations.
 This may be feasible with high frequency, regularly sampled short duration logs but creates noise with secondary logs.
@@
 For instance, although the mean sampling frequency during train travels on the Amsterdam-Utrecht line is low (about 0.01 Hz) the personalised map manages to recreate the train line almost perfectly, despite being completely map agnostic.
-Map agnisticism makes the model more flexible as it can be used for areas where no good geographical databases are available.
+Map agnosticism makes the model more flexible as it can be used for areas where no good geographical databases are available.
 
@@
 Most importantly, the evaluation methods are imperfect.
-The golden standard would be to use at least one highly accurate professional grade GPS device with high sampling frequency to compare our data to.
+The golden standard would be to use at least one highly accurate professional grade GPS device with high sampling frequency to validate results.
 Until that is available, the use of public transport data and cross-validation is just a substitute.
@@
 Improved methods would do well to do so.
-As for paths, a drawback of the current method is that the density of the clusters is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
+As for paths, a drawback of the current method is that the density of the bins is a function of the clustering parameter $d$, the distance between the observed points and their sampling density.
 It does not take into account the length of the path as well as the average sampling frequency of the path.
@@
 It does not take into account any contextual information that can be gleaned from the entire movement history of the information, such as what path they are on.
-For instance, assume that it is known that an individual is travelling from point A to point B along path AB, and there is an inaccurate measurement closest to a cluster which belongs to path AC.
-By only taking distance into account, the measurement can get assigned to the wrong cluster on path AC.
+For instance, assume that it is known that an individual is travelling from point A to point B along path AB, and there is an inaccurate measurement closest to a bin which belongs to path AC.
+By only taking distance into account, the measurement can get assigned to the wrong bin on path AC.
 An improvement would be use a Bayesian method, whereby assignment is a function of both the measurement and a model of the individuals movement history.
@@
 Objective location logs are vastly superior to alternatives, such as questionnaires and travel diares, which rely on accurate self-reporting.
-Social science researchers must take advantage of regulatory changes in with regard to data portability and put the vast wealth of data collected by commercial entities to scientific use.
+Social science researchers must take advantage of regulatory changes with regard to data portability and put the vast wealth of data collected by commercial entities to scientific use.
 

### 2018-05-08 Thesis Manuscript/manuscript.tex
final commit

--- 2018-05-03 Thesis Manuscript/apaManu.tex
@@
 
-    Submitted as a research report conforming to APA manuscript guidelines
+    Submitted as a thesis manuscript conforming to APA manuscript guidelines
     (6th edition).
@@
 associated with a vast range of behavioural traits and outcomes, such as
-socioeconomic status, personality and mental health. The widespread
-adoption of location-sensor equipped smartphones has generated a wealth
-of objective personal mobility data. Nonetheless, smartphone collected
-personal mobility data has remained underutilised in behavioural
-research, partly due to the practical difficulties associated with
-obtaining the data and partly because of the methodological complexity
-associated with analysing it. Recent changes in European regulation have
-made it easier for researchers to obtain this data, but the
-methodological difficulties remain. The difficulty lies in that
-smartphone location data is irregularly sampled, sparse and often
-inaccurate. This results in a high proportion of missing data and
-significant noise. In this paper we present a method called Personal Map
-Matched Imputation (PPMI) to deal with missing data and noise in
-smartphone location logs. The main innovation of PPMI is that it creates
-a personalised spatial map for each individual based on all the
-available data. In doing so PPMI leverages the regularity of human
-mobility in order to smoothen noisy measurments and impute missing data
+socioeconomic status, personality and health. The widespread adoption of
+location-sensor equipped smartphones has generated a wealth of objective
+personal mobility data. Nonetheless, smartphone collected personal
+mobility data has remained underused in behavioural research, partly due
+to the practical difficulties associated with obtaining the data and
+partly because of the methodological complexity associated with
+analysing it. Recent changes in European regulation have made it easier
+for researchers to obtain this data, but the methodological difficulties
+remain. The difficulty lies in that smartphone location data is
+irregularly sampled, sparse and often inaccurate. This results in a high
+proportion of missing data. In this paper we present a method called
+Personal Map Matched Imputation (PPMI) to deal with missing data and
+noise in smartphone location logs. The main innovation of PPMI is that
+it creates a personalised spatial map for each individual based on all
+the available data. In doing so PPMI leverages the regularity of human
+mobility in order to smoothen noisy measurements and impute missing data
 values. By simulating missing periods in real data we find that a simple
@@
 
-Why is human mobility important? Human mobility measures are quantified
-metrics of how people move about in their environment. Human mobility
-affects a wide range of outcomes, such as health, income and social
-capital (Goodchild \& Janelle, 2010). The most widely administered
+Why is human mobility important? Human mobility affects a wide range of
+outcomes, such as health, income and social capital (Goodchild \&
+Janelle, 2010). Human mobility measures are quantified metrics of how
+people move about in their environment. The most widely administered
 personality questionnaires ask individuals to what extent they agree
@@
 across different domains in social science strongly suggests that
-mobility metrics are linked to important real world outcomes. Perhaps it
-is unsurprising that behavioural researchers have found that mobility
-measures can be used to predict academic performance (R. Wang, Harari,
-Hao, Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz,
-\& Odoms-Young, 2009) or even the onset of a depressive episode in
-bipolar depression patients (Palmius et al., 2017). Indeed, there is an
-argument to be made that perhaps psychologists have not been studying
-mobility enough. When studying behavioural differences within
-individuals behavioural scientists have often neglected the fact that
-individuals vary not only over time but also over space. To fully
-understand behaviour we must understand how behaviour can vary across
-environments.
-
-Despite the importance of mobility measures, the majority of these
-metrics are obtained through the use of questionnaires (such as the
-aforementioned personality questionnaire) or specifically through
-pen-and-paper travel diaries. As a thought experiment we encourage the
-reader to try to remember where exactly he or she was five weeks ago on
-Saturday at midday. Most people find this sort of task very difficult
-(if not impossible), which illustrates that questionnaires and travel
-diaries are have well-known methodological flaws. This method of
-collecting data is difficult and relies on accurate self-reporting.
-Travel diaries are burdensome to collect because participants must be
-explicitly asked to write down on their movement patterns at frequent
-intervals due to human forgetfulness. This makes the data collection
-expensive and limits the duration of data collection in practice. In
-addition, the frequent reporting duties of the participant may bias the
-participants behaviour. The limits of human cognition also limit the
-accuracy of self-reported measures. There is evidence that participants
-are systematically biased when self-reporting mobility measures. For
-instance, participants under-report the frequency of short trips (Wolf,
-Oliveira, \& Thompson, 2003) and underestimate the duration of regular
-commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017). These
-obstacles can only be overcome by using objective data on human
+mobility metrics are linked to important real world outcomes.
+Behavioural researchers have found that mobility measures can be used to
+predict academic performance (R. Wang, Harari, Hao, Zhou, \& Campbell,
+2015), the incidence of obesity (Zenk, Schulz, \& Odoms-Young, 2009) or
+even the onset of a depressive episode in bipolar depression patients
+(Palmius et al., 2017). To fully understand behaviour not only must we
+understand how behaviour can vary within and between individuals, but
+also across environments.
+
+Despite the importance of mobility measures, the majority of metrics are
+obtained through the use of questionnaires (such as the aforementioned
+personality questionnaire) or specifically through pen-and-paper travel
+diaries. As a thought experiment we encourage the reader to try to
+remember where exactly he or she was five weeks ago on Saturday at
+midday. Most people find this sort of task difficult (if not
+impossible), which illustrates the main methodological flaws of
+questionnaires and travel diaries. Travel diaries are burdensome to
+collect because participants must be explicitly asked to write down on
+their movement patterns at frequent intervals (lest they forget them).
+The burden to the participants makes data collection expensive, which in
+turn limits the duration of data collection in practice. In addition,
+the frequent reporting duties of the participant may bias the
+participant's behaviour. The limits of human cognition also limit the
+accuracy of self-reported measures. There is evidence that statistics
+derived from self-reported mobility measures are systematically biased.
+For instance, participants under-report the frequency of short trips
+(Wolf, Oliveira, \& Thompson, 2003) and underestimate the duration of
+regular commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
+These obstacles can only be overcome by using objective data on human
 mobility.
@@
 Google Location History contains movement information on millions of
-users, often spanning years (Location History, 2017). Moreover, recent
+users, often spanning years (\emph{Timeline}, 2017). Moreover, recent
 changes in EU regulations with regard to consumer data-portability
@@
 this information with researchers at no cost to the either the
-participant or the researchers (Commission, 2017). Taken together, this
-means that researchers now have at their disposal the ability to easily
-access the objective human mobility data of millions of individuals
-spanning several years at little cost and without a significant burden
-of the participants.
+participant or the researchers (European Commission, 2017). Researchers
+now have the ability to easily access the objective movement records of
+millions of individuals spanning several years at little cost and
+without a significant burden to the participants.
 
@@
 given in meters such that it represents the radius of a 67\% confidence
-circle (Location History, 2017). In other words, the true location of a
+circle (\emph{Timeline}, 2017). In other words, the true location of a
 device should be within the radius \(a\) of the measurement 67\% of the
@@
 of custom logs is that the researchers can adjust tracking parameters,
-such as the the frequency of measurements and the sensor with which they
-are made. The disadvantage with this approach is that researchers have
-to develop or adapt a custom-made tracking application (which is not
-easy given hundreds of different types of smartphone models), distribute
-it among research participants and enforce participation. Participants
-may dislike tracking apps because they view them as more intrusive and
-these apps regularly drain the battery of the device (G. M. Harari et
-al., 2016). Moreover, researchers have distribute this application among
-research participants and convince them not to turn it off.
+such as the the frequency of measurements and the sensor used to make
+the measurement. The disadvantage of this approach is that researchers
+have to develop or adapt a custom-made tracking application (which is
+not easy given hundreds of different types of smartphone models),
+distribute it among research participants and enforce participation.
+Participants may dislike tracking apps because they view them as more
+intrusive and these apps regularly drain the battery of the device
+(Harari et al., 2016).
 
 We focus on another solution, which is to take advantage of existing
-smartphone location logs (\emph{secondary logs}) . The advantages are
-clear: repositories such as Google's Location History contains
-information on millions of users spanning years (Location History,
-2017), participants can share the data by the click of a button and
-there can be no behavioural changes due to participation in the study as
-the participant share past data. The disadvantage is that researchers
-have no control over the tracking parameters, often resulting in logs
-with sparse and inaccurate measurements. Hence, two important challenges
-are dealing with missing data and measurement noise.
+smartphone location logs (\emph{secondary logs}). The advantages are
+clear: repositories such as Google's Location History contain years of
+information on millions of users (\emph{Timeline}, 2017) and
+participants can share the data by clicking of a button. The
+disadvantage is that researchers have no control over the tracking
+parameters, often resulting in logs with sparse and inaccurate
+measurements. Hence, two important challenges in using secondary logs
+are how to deal with missing data and measurement noise.
 
@@
 reasons. Technical reasons include signal loss, battery failure and
-device failure. Behavioural reasons include leaving the phone at home or
-switching the device off. As a result, secondary logs often contain wide
+device failure. Behavioural reasons include switching the device or its
+location services off. As a result, secondary logs often contain wide
 temporal gaps with no measurements. For instance, several research
@@
 50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
-Other researchers report similar trends in different fields (e.g. G. M.
-Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015). In Figure
-\ref{fig:longMeasurementsPerDay} shows that despite the long duration of
-the log the sparsity it is also evident.
+Other researchers report similar trends in different fields (e.g. Harari
+et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015). Figure
+\ref{fig:longMeasurementsPerDay} shows a secondary log spanning years
+where despite its long duration the sparsity of measurements is also
+evident.
 
@@
 (Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
-are auto-correlated in both time and space. This means that best
+are auto-correlated in both time and space. This means that common
 practices with other types of data, such as mean imputation, are
@@
 Another methodological problem is related to the noise in the
-measurements that are collected. The accuracy of smartphone location
-measurements is substantially lower than that of professional GPS
-location trackers because smartphones often use less accurate sensors.
-In professional GPS trackers less than 80\% of measurements fall within
-10 meters of the true location. GPS measures are most inaccurate in
-dense urban locations and indoors (S. Duncan et al., 2013; Schipperijn
-et al., 2014). Unfortunately for researchers, this is where people in
-the developed world spend most of their time. Figure
-\ref{fig:accuracyPlot} shows how accuracy can vary as a function of user
-behaviour, time and location. Most notably, low accuracy is often (but
-not always) associated with movement (see Figure
-\ref{fig:accuracyPlot2}).
+measurements. The accuracy of smartphone location measurements is
+substantially lower than that of professional GPS location trackers
+because smartphones often use less accurate sensors. In professional GPS
+trackers less than 80\% of measurements fall within 10 meters of the
+true location. GPS measures are most inaccurate in dense urban locations
+and indoors (S. Duncan et al., 2013; Schipperijn et al., 2014).
+Unfortunately for researchers, this is where people in the developed
+world spend most of their time. Figure \ref{fig:accuracyPlot} shows how
+accuracy can vary as a function of user behaviour, time and location.
+Most notably, low accuracy is often (but not always) associated with
+movement (see Figure \ref{fig:accuracyPlot2}).
 
@@
 for. Suppose we wish to calculate an individual's movement in a day. A
-simple approach would be to calculate the sum of the distance between
+simple approach would be to calculate the sum of the distances between
 each measurement. But if there is noise, the coordinates will vary even
@@
 and noisy, we will calculate a lot of movement, even if the individual
-did not move at all! This issue is also visualised in Figure
-\ref{fig:palmiusVsMe1}. The problem is further complicated because
-missing data and noisy measurements are related. Methods used by
-researchers to reduce noise, such as throwing out inaccurate
-measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
-the missing data problem.
+did not move at all! The problem is further complicated because missing
+data and noisy measurements are related. Methods used by researchers to
+reduce noise, such as throwing out inaccurate measurements (e.g. Palmius
+et al., 2017), can exacerbate the severity of the missing data problem.
 
 In this paper we will propose PPMI as a method for dealing with missing
-data and measurement error in secondary location logs. We will compare
-PPMI to similar solutions in the literature by evaluating the distance
-between points which were simulated as missing and their imputed
-counterparts. Finally we will calculate time spent at home as a function
-of the imputation method.
+data and measurement error in secondary location logs. PPMI creates a
+discrete personalised map based on the individuals mobility history.
+Then it assigns each measurement to a discrete location based on the
+individuals entire mobility history. Finally, it uses a classification
+method to impute missing periods.
 
@@
 worrisome as it invites solutions which contain significant researcher
-degrees of freedom (Simmons, Nelson, \& Simonsohn, 2011). The vagueness
-is possibly also due to the fact that most researchers are unfamiliar
-with possible solutions. Most researchers simply down-sample temporally
-and remove missing observations or use some sort of rule-based common
-sense imputations (e.g. Palmius et al. (2017)). The only principled
-approach that we know of that aims to solve the issue of missing data in
-location logs as they relate to human mobility is that of Barnett and
-Onnela (2016). We will explore the methods of Barnett and Onnela (2016)
-and Palmius et al. (2017) in detail subsequently, after introducing
-exploring other spatiotemporal methods.
+degrees of freedom (Simmons, Nelson, \& Simonsohn, 2011). Most
+researchers simply down-sample temporally and remove missing
+observations or use some sort of rule-based common sense imputations
+(e.g. Palmius et al., 2017). The only principled approach (that we know
+of) aiming to solve the issue of missing data in this context
+(i.e.~location logs as they relate to human mobility) is that of Barnett
+and Onnela (2016). We will explore the methods of Barnett and Onnela
+(2016) and Palmius et al. (2017) in detail subsequently, after
+introducing other spatiotemporal methods.
 
@@
 how animals interact with their environment (T. A. Patterson, Thomas,
-Wilcox, Ovaskainen, \& Matthiopoulos, 2008). These models can be quite
-complex. Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
-processes to characterise the effect of roads, food patches and streams
-on cyclical elk movements. The most well studied SSM is the Kalman
-filter, which is the optimal algorithm for inferring linear Gaussian
-systems. The extended Kalman filter is the de facto standard for GPS
-navigation (Z. Chen \& Brown, 2013). The advantage of state space models
-is that they are flexible, deal with measurement inaccuracy, include
-information from different sources and can be used in real time.
+Wilcox, Ovaskainen, \& Matthiopoulos, 2008). Preisler, Ager, Johnson,
+and Kie (2004) uses Markovian movement processes to characterise the
+effect of roads, food patches and streams on cyclical elk movements. The
+most well studied SSM is the Kalman filter, which is the optimal
+algorithm for inferring linear Gaussian systems. The extended Kalman
+filter is the de facto standard for GPS navigation (Z. Chen \& Brown,
+2013). The advantage of state space models is that they are flexible,
+deal with measurement inaccuracy, include information from different
+sources and can be used in real time.
 
@@
 property, they cannot incorporate this information. In other words, the
-estimated location \(G(t)\) at time-point \(t\) is often based only upon
-measurements \(D_t\), \(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\).
-Hierarchical structuring and conditioning on a larger context have been
-suggested as ways to add periodicity to Markovian models. These
-solutions are often computationally intractable or unfeasible (Sadilek
-\& Krumm, 2016). Moreover, these models often assume time and space
-invariance (location is not a direct function of time or space). These
+estimated location at a time-point \(t\) is often based only upon
+measurements at timepoints \(t\) and \({t-1}\). Hierarchical structuring
+and conditioning on a larger context have been suggested as ways to add
+periodicity to Markovian models. These solutions are often
+computationally intractable or unfeasible (Sadilek \& Krumm, 2016).
+Moreover, these models often assume time and space invariance
+(i.e.~location is not a direct function of time or space). These
 mathematical assumptions are violated in the case of human movement
 patterns. For this reason we do not consider existing SSMs to be useful
-for imputing missing data in this case.
+for imputing missing data in smartphone mobility logs.
 
@@
 missing data from human mobility patterns (Barnett \& Onnela, 2016;
-Palmius et al., 2017 ; @ Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
-none of them worked with secondary logs. Nonetheless we will detail what
+Palmius et al., 2017 ; Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
+none of them worked with secondary logs, but rather with custom logs
+that have higher sampling frequency. Nonetheless we will detail what
 they did as informative examples. Palmius et al. (2017) deal with the
-measurement inaccuracy of \(D\) in custom logs by removing from the data
-set all unique low-accuracy \(a\) data points that had
-\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
-sample the data to a sample rate of 12 per hour using a median filter.
-Moreover, Palmius et al. (2017) explain:
+measurement inaccuracy in custom logs by removing from the data set all
+unique low-accuracy data points with a speed greater than 100 km/h.
+Subsequently the researchers down sample the data to a sample rate of 12
+per hour using a median filter. Moreover, Palmius et al. (2017) explain:
 
 \begin{quote}
-\enquote{If the standard deviation of {[}\(D\){]} in both latitude and
-longitude within a 1 h epoch was less than 0.01 km, then all samples
-within the hour were set to the mean value of the recorded data,
-otherwise a 5 min median filter window was applied to the recorded
-latitude and longitude in the epoch}.
+\enquote{If the standard deviation of in both latitude and longitude
+within a 1 h epoch was less than 0.01 km, then all samples within the
+hour were set to the mean value of the recorded data, otherwise a 5 min
+median filter window was applied to the recorded latitude and longitude
+in the epoch}.
 \end{quote}
@@
 section and the missing section had a length of \(\leq 2h\) or
-\(\leq 12h\) after 9pm. In cases where the previous conditions are not
-met no values are imputed.
+\(\leq 12h\) if this occurred after 9pm. In cases where the previous
+conditions are not met no values are imputed.
 
@@
 custom logs where location is measured for 2 minutes and subsequently
-not measured for 10 minutes. In the words of the authors, Barnett and
-Onnela (2016) handle missing data by first converting data to mobility
-traces, which are defined as a sequence of flights and pauses. Flights
-are segments of linear movements and pauses corresponding to periods of
-time where a person does not move. Subsequently, the authors impute
-missing data by:
+not measured for 10 minutes. Barnett and Onnela (2016) handle missing
+data by first converting data to mobility traces, which are defined as a
+sequence of flights and pauses. Flights are segments of linear movements
+and pauses corresponding to periods of time where a person does not
+move. Subsequently, the authors impute missing data by:
 
@@
 
-Subsequently let us simplify by assuming that a persons location is on
+Subsequently let us simplify by assuming that a persons location is on a
 two-dimensional Euclidean plane. Let a person's true location on this
@@
 recorded data. Furthermore, \(G(t)\), \(D\) and \(a\) are indexed by
-time labeled by the countable set \(t = t_1 < ... < t_{n+1}\). The
-measure of accuracy \(a_t\) is given in meters such that it represents
-the radius of a 67\% confidence circle. If \(D_t = \emptyset\) it is
-considered \emph{missing} and it is not missing otherwise.
-
-When several data sets are available from individuals living in
-overlapping areas we can construct a \(t \times i\) matrix \(M\) where
-the entry \(M(t,i)\) contains \(G(t)\) for the individual \(i\).
-
-\subsection{Personalised Map Matching
-Imputation}\label{personalised-map-matching-imputation}
+time labeled by the countable set \(t = t_1 < ... < t_{n+1}\). If
+\(D_t = \emptyset\) it is considered \emph{missing}. The measure of
+accuracy \(a_t\) is given in meters such that it represents the radius
+of a 67\% confidence circle. The reader should keep in mind that a high
+accuracy value \(a_t\) for a measurement \(D_t\) means that the
+measurement is inaccurate (this can be rather confusing).
+
+\subsection{Personalised Map Matched
+Imputation}\label{personalised-map-matched-imputation}
 
@@
 \(t_b\) be a pause of length \(m\) such that there is no temporal
-overlap between \(t_a\) and \(t_b\):
-\[t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j\] Then it follows that between
-the two pauses there must be a flight indexed by \(t_x\) of length
-\(j-i+n\). \[t_x =  t_{i+n} < ... < t_{j} |t_x \in F\] We define pause
-locations \(G(t_a), G(t_b) | t_a, t_b \in P\) as locations where an
-individual spends an extended amount of time in the same space
-(e.g.~school, home, work, train station, barber shop, bar, gym).
-Importantly, our model assumes period and cyclic human movement such
-that there are many pauses \(t_{a1},t_{a2},...,t_{an}\) such that
+overlap between \(t_a\) and \(t_b\) and \(t_b\) is the first pause after
+the end of \(t_a\): \[t_b = t_j < ... < t_{j+m}|t_{i+n}< t_j\] Then it
+follows that between the two pauses there must be a flight indexed by
+\(t_x\) of length \(j-i+n\). \[t_x =  t_{i+n} < ... < t_{j} |t_x \in F\]
+We define pause locations \(G(t_a), G(t_b) | t_a, t_b \in P\) as
+locations where an individual spends an extended amount of time in the
+same space (e.g.~school, home, work, train station, barber shop, bar,
+gym). Importantly, our model assumes period and cyclic human movement
+such that there are many pauses \(t_{a1},t_{a2},...,t_{an}\) such that
 \(G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})\). Moreover,it is possible for
+two consecutive pauses \(t_a\) and \(t_b\) to share a location
 \(G(t_a) = G(t_b)\) such that \(t_a \ne t_b\). For example, if the
@@
 
-\subsection{Personalised Map Matching Imputation
-algorithm}\label{personalised-map-matching-imputation-algorithm}
+\subsection{Personalised Map Matched Imputation
+algorithm}\label{personalised-map-matched-imputation-algorithm}
 
@@
   \emph{Map building}: Extract from measurements \(D\) all pause
-  location bins and path location bins to create a personalised map.
+  locations and path locations to create a personalised map consisting
+  of discrete bins.
 \item
   \emph{Binning}: Assign each measurement \(D\) to a unique discrete
-  location bin.
+  location.
 \item
@@
 extraction algorithm our aim is to transform pause and path locations
-into machine learning friendly discrete location sequences. There are
-multiple ways of extracting such measurement clusters in the literature,
-such as Spatio-Temporal Density-Based Spatial Clustering of Applications
-with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
-(``ST-DBSCAN,'' 2007). We will focus on two methods which explicitly
-with mobility patterns from unevenly sampled smartphone logs (Barnett \&
-Onnela, 2016; Palmius et al., 2017 ). Both of these methods pre-process
-the data and subsequently use two steps to extract pause locations:
-first they extract pauses and their corresponding locations, then they
-cluster pause locations based on spatial proximity. This implementation
-of PMMI uses a stricter version of Barnett and Onnela (2016)'s approach
-to extract pauses.
+into machine learning friendly discrete locations, which we will call
+location bins. There are multiple ways of extracting such measurement
+clusters in the literature, such as Spatio-Temporal Density-Based
+Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence
+oriented clustering (SOC) (Birant \& Kut, 2007). We will focus on
+methods which explicitly deal with mobility patterns extracted from
+smartphone logs (Barnett \& Onnela, 2016; Palmius et al., 2017 ). Both
+of these methods pre-process the data and subsequently use two steps to
+extract pause locations: first they extract pauses and their
+corresponding locations, then they cluster pause locations based on
+spatial proximity. This implementation of PMMI uses a stricter version
+of Barnett and Onnela (2016)'s approach to extract pauses.
 
@@
   Let the measurements of a possible pause which fit the aforementioned
-  criteria be \(D_{t,t+1,...,t+n}\). These points are only a pause if
-  the distance between the mean coordinates of \(D_{t,t+1,...,t+n}\) and
-  the furthest away points of \(D_{t,t+1,...,t+n}\) is within 2 times
-  the mean accuracy \(a\) of \(D_{t,t+1,...,t+n}\).
+  criteria be \(X = D_{t,t+1,...,t+n}\). Let \(\bar{x}\) be the mean
+  coordinates of \(X\), let \(x_{\text{dist}} \in X\) be the furthest
+  away points from \(\bar{x}\) and \(\bar{a}\) be the the mean accuracy
+  of \(X\). Then \(X\) is a pause if and only if the distances between
+  \(\bar{x}\) and \(x_{\text{dist}}\) are \(\leq 2 \times \bar{a}\).
 \end{enumerate}
 
-This set of points were then hierarchically clustered using a distance
-matrix, such that all points within \(d\) meters of each other were
-clustered into a pause location. Each pause location is a bin.
+The resulting set of points were then hierarchically clustered using a
+distance matrix, such that all points within \(d\) meters of each other
+were clustered into a pause location. Each pause location is a bin.
 
@@
   \(d_{\text{Path Lim}}\) meters of each other are clustered into a
-  single pause point.
+  single flight bin.
 \end{enumerate}
@@
 The reasoning for this is that the threshold for not being in a pause
-cluster should be higher, as individuals spend the majority of time at a
-pause cluster. The end result is a discretised map which contains pause
-and flight bins based on the entire log history of the individual.
+bin should be higher, as individuals spend the majority of time at a
+pause bin. The end result is a discretised map which contains pause and
+flight bins based on the entire log history of the individual.
 
@@
 database. The key modification in PMMI is that whilst Wu et al. (2017)
-uses a map from outside the persons location logs, we use the total
-location history of the individual to create a personalised map. This
-map is subsequently used to bin measurements. This is feasible for two
-reasons: humans tend to have repetitive movement habits and secondary
-logs tend to be long. To put it in simpler terms, we consider each
-measurement at \(D_x\) as a sample of \(Path_{ab}\), and by aggregating
-many measurements we can use them to map out \(Path_{ab}\).
-
-Thus, all measurements \(D\) were are then assigned to a discrete bin on
-the personal map. This includes previous measurements which were
-discarded from the map building exercise due to an accuracy \(a\) value
-which exceeded \(a_{\text{P lim}}\) or \(a_{\text{F lim}}\). In this
+uses an external map, we use the total location history of the
+individual to create a personalised map. This map is subsequently used
+to bin measurements. This is feasible for two reasons: humans tend to
+have repetitive movement habits and secondary logs tend to be long. To
+put it in simpler terms, we consider each measurement at \(D_x\) as a
+sample of \(Path_{ab}\), and by aggregating many measurements we can use
+them to map out \(Path_{ab}\).
+
+Thus, all measurements \(D\) are then assigned to a discrete bin on the
+personal map. This includes previous measurements which were discarded
+from the map building exercise due to an accuracy \(a\) value which
+exceeded \(a_{\text{P lim}}\) or \(a_{\text{F lim}}\). In this
 implementation we used a simple assigning function, whereby the
-measurements where assigned to the bin nearest to the measurement.
+measurements where assigned to the nearest bin.
 
@@
 missing value. In this implementation we trained an artificial neural
-network (ANN) to do so. For more information on the precise architecture
-of the artificial neural network please consult the appendix. The input
-variables to the ANN are:
+network (ANN) to do so. The ANN has a simple architecture: it consists
+of an input layer, a hidden layer and an output layer. The input layer
+and the hidden layer employ a rectified linear unit activation function
+and the output layer uses a softmax activation function.
+
+The input variables to the ANN are:
 
@@
 \item
-  The day of the month as a binary class matrix.
+  The day of the week as a binary class matrix.
 \item
@@
 transforms of the amount of seconds that have elapsed after midnight
-(London, 2016). This is necessary so that the model can understand that
-one second past midnight and one second before midnight are in fact two
-seconds away from each other. Moreover we scaled the non-binary values
-to occupy a range between 0 and 1 in order to ensure convergence.
+(London, 2016). This is necessary for the model to be able to understand
+that one second past midnight and one second before midnight are in fact
+two seconds away from each other. Moreover we scaled the non-binary
+values to occupy a range between 0 and 1 in order to ensure convergence.
 
@@
 model is a set of probability estimates associated with every location
-cluster. That is, for each missing time-point the model returns a vector
-of probability estimates (with one estimate per bin) associated with
-where the individual is.
+bin. That is, for each missing time-point the model returns a vector of
+probability estimates (with one estimate per bin) associated with where
+the individual is.
 
@@
 single individual. About 54\% of the data is missing for the entire
-duration of the log. This may be misleading as there are several long
-periods with no measurements whatsoever. For days which were not
-entirely missing, approximately 22\% of all five minute segments were
-missing. The structure of missingness of a day with measurements is
-shown in Figure \ref{fig:measurementsPerDay}. As you can see, there are
-several long periods over the course of the log for which there are no
-measurements. The median sampling frequency per day for non-missing days
-is around 0.006 Hz.
+duration of the log. There are several long periods with no measurements
+whatsoever, as well as periods with far less missing data. For days
+which were not entirely missing, approximately 22\% of all five minute
+segments were missing. The structure of missingness of a non-missing day
+is shown in Figure \ref{fig:measurementsPerDay}. As you can see, there
+are several long periods over the course of the log for which there are
+no measurements. The median sampling frequency per day for non-missing
+days is around 0.006 Hz.
 
@@
 Palmius et al. (2017)'s algorithm was implemented in R based on the
-original MATLAB code and pseudocode kindly provided by the author.
+original MATLAB code and pseudocode kindly provided by the authors.
 Barnett and Onnela (2016)'s method was slightly adapted in R to fit the
-the data structure from the original R code provided by the author. When
+data structure from the original R code provided by the authors. When
 executing their models we used the same parameters as the authors did.
@@
 
-All the code is available on a public repository (Sobrado, 2018).
+All the code is available on a public repository\footnote{\href{https://github.com/sobradob/thesis}{Github}
+  (Sobrado, 2018)} .
 
@@
   Comparing the performance of PMMI compared to alternative methods
-  {[}(Barnett \& Onnela, 2016; Palmius et al., 2017){]} and a baseline
-  model using cross-validation.
+  (Barnett \& Onnela, 2016; Palmius et al., 2017) and a baseline model
+  using cross-validation.
 \item
@@
   methods using objective ground-truth data (public transportation
-  time-stamps locations).
+  time-stamped locations).
 \end{enumerate}
@@
 \(G(t)\) from measurements \(D_t\). In the subsequent cross-validation
-step we will remove random smoothened blocks and compare the imputed
-value to the smoothened value. For this reason we want to ensure that
-the smoothened method actually does reflect the true location \(G(t)\).
-Otherwise we run the risk of over-estimating the accuracy of a model in
-the sense that we will compute the extent to which an imputation method
-can correctly impute a smoothened value which has little to do with the
-true location \(G(t)\).
+step we will remove random smoothened 5 minute periods and compare the
+imputed value to the smoothened value. For this reason we want to ensure
+that the smoothened value actually does reflect an approximation to the
+true location \(G(t)\). Otherwise we run the risk of over-estimating the
+accuracy of a model in the sense that we will compute the extent to
+which an imputation method can correctly impute a smoothened value which
+has little to do with the true location \(G(t)\).
 
@@
 evaluate the paths and pause locations. A visual evaluation of paths
-superimposed on is an established way to heuristically check their
+superimposed on a map is an established way to heuristically check their
 accuracy (e.g. Brunsdon, 2007). Then, let the average distance between
@@
   A positive relationship between the deviation distance and the
-  accuracy of each measurement.
+  accuracy \(a\) (in meters) of each measurement.
 \item
@@
 In addition, other measures of interest for PMMI are \emph{accuracy} (in
-what percentage of the cases was the appropriate cluster predicted), the
+what percentage of the cases was the appropriate bin predicted), the
 \emph{confidence} and the \emph{distance expectation}. The confidence is
-the probability with which the model predicts the most likely cluster.
-For instance, if the model predicts the missing bin to be bin A with a
+the probability with which the model predicts the most likely bin. For
+instance, if the model predicts the missing bin to be bin A with a
 probability of 0.9 then we can say this is a high confidence prediction.
 Similarly, the distance expectation is the cross-product of the
-estimated probabilities that the individual is at any of all given
-clusters and the distances between the clusters to the true cluster. For
-example, if the true location of an individual is bin A (bin A is 1000
-meters away from bin B) and the model assigns a probability of 0.9 at
-bin A and 0.1 at bin B, then the distance expectation would be 100
-meters.
+estimated probabilities that the individual is at any of all given bins
+and the distances between the bins to the true bin. For example, if the
+true location of an individual is bin A (bin A is 1000 meters away from
+bin B) and the model assigns a probability of 0.9 at bin A and 0.1 at
+bin B, then the distance expectation would be 100 meters.
 
@@
 and Onnela (2016) and the naive baseline model with regards to imputing
-real missing measurments. To do will use information from the Dutch
-public transportation card. The Dutch public transportation service
-provides users with time-stamped location data of when and where they
-board, change lines or leave public transportation. In this paper we
-used 97 such events in a single month.
+real missing measurments. To do so we will use information from the
+Dutch public transportation card. The Dutch public transportation
+service provides users with time-stamped location data of when and where
+they board, change lines or leave public transportation. In this paper
+we used 97 such events in a single month.
 
@@
 \(\delta_{\text{Pause lim}}\) of 100 seconds. Moreover, to extract path
-clusters we used the parameters:
-\(a_{\text{Path Lim}} = 150 \text{meters}\) and
-\(d_{\text{Path Lim}} =300 \text{meters}\).
+bins we used the parameters: \(a_{\text{Path Lim}} = 150 \text{meters}\)
+and \(d_{\text{Path Lim}} =300 \text{meters}\).
 
@@
 \item
-  A suitable extraction of paths. For instance, are ensuring that
-  individual paths are complete and not merged unneccessarily in the
-  case of closely parallel paths.
-\item
-  Less than a fifth of extracted clusters are assigned only one
-  measurement. Unique clusters are unavoidable as people sometimes visit
-  a location briefly and only once, nonetheless they pose problems for
-  the classification algorithms.
+  A suitable extraction of paths. Ensuring that individual paths are
+  logically consistent (e.g.~do not go over rivers or other unplausable
+  behaviour).
+\item
+  Less than a fifth of extracted bins are assigned only one measurement.
+  Unique bins are unavoidable as people sometimes visit a location
+  briefly and only once, nonetheless they pose problems for the
+  classification algorithms.
 \end{enumerate}
@@
 precision. This is because an increase in precision (in the form of a
-higher resolution of locations location) comes at the expense of
-precision (assigning measurements to bins becomes more difficult). For
-instance, by increasing the clustering distance parameter
-\(d_{\text{Pause lim}}\) we can extract more valid pause locations at
-the risk of falsely categorising certain measurements to the wrong bin.
-This is illustrated in Figure \ref{fig:pauseLocUithof1}.
+higher resolution of bin locations) comes at the expense of bias
+(assigning measurements to bins becomes more difficult). For instance,
+by increasing the clustering distance parameter \(d_{\text{Pause lim}}\)
+we can extract more valid pause locations at the risk of falsely
+categorising certain measurements to the wrong bin. This is illustrated
+in Figure \ref{fig:pauseLocUithof1}.
 
@@
 
-Map building results in a personalised map with pause and path clusters.
-An excerpt can be seen in Figure \ref{fig:mapClusterMap}. It is
-important to remind the reader that PMMI is map agnostic and uses no
-information from the map. Therefore, the close overlap with features on
-the map, such as pause bins at relevant buildings and transportation
-clusters, as well as the flight bins following roads and railway lines
-indicate a high degree of precision in personal map building. As
-expected, PMMI's path extraction yields greater accuracy for frequently
-occurring paths. For example, the frequently travelled Amsterdam-Utrecht
-railway line has been extracted almost perfectly, while the less
-frequently travelled Utrecht-Enschede line is far sparser.
+Map building results in a personalised map with pause and path bins. An
+excerpt can be seen in Figure \ref{fig:mapClusterMap}. It is important
+to remind the reader that PMMI is map agnostic and uses no external
+information. Therefore, the close overlap with features on the map, such
+as pause bins at relevant buildings and transportation locations, as
+well as the flight bins following roads and railway lines indicate a
+high degree of precision in personal map building. As expected, PMMI's
+path extraction yields greater accuracy for frequently occurring paths.
+For example, the frequently travelled Amsterdam-Utrecht railway line has
+been extracted almost perfectly, while the less frequently travelled
+Utrecht-Enschede line is far sparser.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{img/mapClusterMap} \caption{Excerpt of the cluster map of an individual. Red points are pause locations, blue points are path locations.}\label{fig:mapClusterMap}
+\includegraphics[width=1\linewidth]{img/mapClusterMap} \caption{Excerpt of the personalised binned map of an individual.
+Red points are pause locations, blue points are path locations.}\label{fig:mapClusterMap}
 \end{figure}
 
-For the entire period examined period the we find a deviance of 40
-meters and a median deviance of 15 meters. Around 69\% of the deviance
-values are within their corresponding accuracy value, which is close to
-the theoretical 67\% value that is expected. Approximately 9\% of values
-were not taken into account when creating the bins,given that their the
-accuracy \(a\) exceeded \(a_{\text{Path Lim}} = 300\).
-
-In comparison, Palmius et al. (2017)\enquote{s method has a median
-deviance \(\delta_{\text{dev}}\) of 3 meters, with a mean of 115 due to
-high deviance outliers. On the other hand, Barnett and Onnela (2016)'s
+For the entire period examined period the we find a mean deviance
+\(\delta_{\text{dev}}\) of 40 meters and a median deviance of 15 meters.
+Around 69\% of the deviance values are within their corresponding
+accuracy value, which is close to the expected theoretical 67\%.
+Approximately 9\% of values were not taken into account when creating
+the bins, given that their the accuracy \(a\) exceeded
+\(a_{\text{Path Lim}} = 300\).
+
+In comparison, Palmius et al. (2017)'s method has a median deviance
+\(\delta_{\text{dev}}\) of 3 meters, with a mean of 115 due to high
+deviance outliers. On the other hand, the Barnett and Onnela (2016)
 method has mean deviance \(\delta_{\text{dev}}\) of 343 meters and a
 median deviance of 8 meters. Barnett and Onnela (2016)'s deviance is
-necessarily higher than Palmius et al. (2017)}, as they down-sample
-temporally (like Palmius et al. (2017)) and subsequently aggregate into
+necessarily higher than Palmius et al. (2017), as they down-sample
+temporally (like Palmius et al., 2017) and subsequently aggregate into
 pauses and linear flights.
@@
 The key difference between temporal and spatial down-sampling is shown
-in Figure \ref{fig:palmiusVme1}. Temporal down-sampling is much more
+in Figure \ref{fig:palmiusVsMe1}. Temporal down-sampling is much more
 sensitive to noise in sparsely measured periods because it averages out
@@
 measurements in those periods (see Figure
-\ref{fig:longMeasurementsPerDay} ), which leads to a noise in the
-down-sampled values. Unsurprisingly, there is a positive relationship
-between deviance and the amount of measurements in each down-sampled
-interval. The fact that over 90\% of deviance values are within accuracy
-(substantially higher than the expected theoretical 67\%) confirms that
-temporal down-sampling is not sufficiently filtering out the noise.
+\ref{fig:longMeasurementsPerDay} which shows that a large proportion of
+5 minute periods have less than 3 measurements), which leads to a noise
+in the down-sampled values. Unsurprisingly, there is a positive
+relationship between deviance and the amount of measurements in each
+down-sampled interval. The fact that over 90\% of deviance values are
+within accuracy (substantially higher than the expected theoretical
+67\%) confirms that temporal down-sampling is not sufficiently filtering
+out the noise.
 
@@
 The first factor we must take into account is the coverage of each
-imputation model. The results show us that while PPMI and the naive
-baseline model impute all missing values, the other two models have
-difficulties doing so (Table 1). In particular, the models of Palmius et
-al. (2017) and Barnett and Onnela (2016)'s fail to impute an increasing
-amount of missing values with an increase in the duration of the missing
-period. Palmius et al. (2017)'s method in particular failed to impute a
-single missing time period. The longer a log, the higher the probability
-that long time periods will be missing. Secondary logs are typically
-long. Hence missing data imputation methods designed to work with
-secondary logs must be capable of imputing longer time periods.
+imputation model. The results show us that while PPMI and the baseline
+models impute all missing values, the other two models have difficulties
+doing so (Table 1). In particular, the models of Palmius et al. (2017)
+and Barnett and Onnela (2016)'s fail to impute an increasing amount of
+missing values with an increase in the duration of the missing period.
+Palmius et al. (2017)'s method in particular failed to impute a single
+missing time period when the duration of the removed periods was one
+day. The longer a log, the higher the probability that long time periods
+will be missing. Secondary logs are typically long. Hence missing data
+imputation methods designed to work with secondary logs must be capable
+of imputing longer time periods.
 
@@
 distance metrics for each method. We can see that PPMI predicted the
-location perfectly for over half of the removed values in all three time
-categories. To be precise PPMI's prediction accuracy was 88\%, 73\% and
-47\% for the 5 minute, 1 hour and 1 day periods respectively. As a
-comparison, the baseline naive model's prediction accuracy was 87\%,
-68\% and 22\%, and the home baseline model's were 46\%, 52\% and 40\%.
-Although the home baseline model and PPMI achieved similar distance
-metrics in the 1 day condition, this does not mean that PPMI defaulted
-to simply predicting the individual is home at all times (only 40\% of
-predictions were home). PPMI outperformed the alternative models when
-taking into account the median scores, but not the mean. The explanation
-is that the alternative models failed to impute more difficult time
-periods.
+location perfectly for at least about half of the removed values in all
+three time categories. To be precise PPMI's prediction accuracy was
+88\%, 73\% and 47\% for the 5 minute, 1 hour and 1 day periods
+respectively. As a comparison, the baseline naive model's prediction
+accuracy was 87\%, 68\% and 22\%, and the home baseline model's were
+46\%, 52\% and 40\%. Although the home baseline model and PPMI achieved
+similar distance metrics in the 1 day condition, this does not mean that
+PPMI defaulted to simply predicting the individual is home at all times
+(only 40\% of predictions were home). PPMI outperformed the alternative
+models when taking into account the median scores, but not the mean. The
+explanation is that the alternative models failed to impute more
+difficult time periods. This increases the magnitude of errors of the
+PPMI method.
 
 On the other hand, the confidence score for each time period were
-heavily skewed left, withh a median of 1. Accordingly, the distance
+heavily skewed left, with a median of 1. Accordingly, the distance
 expectation (the crossproduct of the model probabilities of an estimate
@@
 The amount of time spent at home has been found to be a reliable
-predictor of extroversion (G. M. Harari et al., 2016) and the onset of
+predictor of extroversion (Harari et al., 2016) and the onset of
 depressive episodes in bipolar patients (Palmius et al., 2017). The 12\%
@@
 valid as they are the closest we found to a missing data imputation
-methods in smartphone GPS logs.
+methods in smartphone location logs.
 
@@
 of single imputation is that it takes an imputed value and treats it as
-observed. Simple rule-based methods such as Palmius et al. (2017)' are
+observed. Simple rule-based methods such as Palmius et al. (2017)'s are
 essentially algorithmic single imputation methods. With PPMI it is
@@
 historical data to smoothen pause locations by clustering pause
-locations with with a close degree of spatial proximity, neither of them
-do the same for non-pause locations. This may be feasible with high
+locations with a close degree of spatial proximity, neither of them do
+the same for non-pause locations. This may be feasible with high
 frequency, regularly sampled short duration logs but creates noise with
@@
 line almost perfectly, despite being completely map agnostic. Map
-agnisticism makes the model more flexible as it can be used for areas
+agnosticism makes the model more flexible as it can be used for areas
 where no good geographical databases are available.
@@
 would be to use at least one highly accurate professional grade GPS
-device with high sampling frequency to compare our data to. Until that
-is available, the use of public transport data and cross-validation is
-just a substitute.
+device with high sampling frequency to validate results. Until that is
+available, the use of public transport data and cross-validation is just
+a substitute.
 
@@
 would do well to do so. As for paths, a drawback of the current method
-is that the density of the clusters is a function of the clustering
+is that the density of the bins is a function of the clustering
 parameter \(d\), the distance between the observed points and their
@@
 that an individual is travelling from point A to point B along path AB,
-and there is an inaccurate measurement closest to a cluster which
-belongs to path AC. By only taking distance into account, the
-measurement can get assigned to the wrong cluster on path AC. An
-improvement would be use a Bayesian method, whereby assignment is a
-function of both the measurement and a model of the individuals movement
-history. In terms of state-space models the state equation would
-represent a probabilistic representation of where the individual could
-be at that given time based on the individuals entire movement history.
-The space side of the model would be a measurement equation representing
-the measurement and the uncertainty surrounding it in the form of \(a\).
-This would be a similar implementation to that used in Liao et al.
-(2007), but rather taking the personalised map instead of information
-about road networks.
+and there is an inaccurate measurement closest to a bin which belongs to
+path AC. By only taking distance into account, the measurement can get
+assigned to the wrong bin on path AC. An improvement would be use a
+Bayesian method, whereby assignment is a function of both the
+measurement and a model of the individuals movement history. In terms of
+state-space models the state equation would represent a probabilistic
+representation of where the individual could be at that given time based
+on the individuals entire movement history. The space side of the model
+would be a measurement equation representing the measurement and the
+uncertainty surrounding it in the form of \(a\). This would be a similar
+implementation to that used in Liao et al. (2007), but rather taking the
+personalised map instead of information about road networks.
 
@@
 accurate self-reporting. Social science researchers must take advantage
-of regulatory changes in with regard to data portability and put the
-vast wealth of data collected by commercial entities to scientific use.
+of regulatory changes with regard to data portability and put the vast
+wealth of data collected by commercial entities to scientific use.
 
@@
 Retrieved from \url{http://arxiv.org/abs/1606.06328}
+
+\hypertarget{ref-_st-dbscan:_2007}{}
+Birant, D., \& Kut, A. (2007). ST-dbscan: An algorithm for clustering
+spatial--temporal data. \emph{Data \& Knowledge Engineering},
+\emph{60}(1), 208--221.
 
@@
 
-\hypertarget{ref-commission_protecting_2017}{}
-Commission, E. (2017). \emph{Protecting your data: Your rights -
-european commission}. Retrieved from
-\url{http://ec.europa.eu/justice/data-protection/individuals/rights/index_en.htm}
-
 \hypertarget{ref-delclos-alio_keeping_2017}{}
@@
 
+\hypertarget{ref-commission_protecting_2017}{}
+European Commission. (2017). \emph{Protecting your data: Your rights -
+european commission}. Retrieved from
+\url{http://ec.europa.eu/justice/data-protection/individuals/rights/index_en.htm}
+
 \hypertarget{ref-feng_cutoff:_2014}{}
@@
 \hypertarget{ref-harari_using_2016}{}
-Harari, G. M., Lane, N. D., Wang, R., Crosier, B. S., Campbell, A. T.,
-\& Gosling, S. D. (2016). Using smartphones to collect behavioral data
-in psychological science: Opportunities, practical considerations, and
+Harari, G., Lane, N. D., Wang, R., Crosier, B. S., Campbell, A. T., \&
+Gosling, S. D. (2016). Using smartphones to collect behavioral data in
+psychological science: Opportunities, practical considerations, and
 challenges. \emph{Perspectives on Psychological Science}, \emph{11}(6),
@@
 
-\hypertarget{ref-location_history_timeline_2017}{}
-Location History, G. (2017). \emph{Timeline}. Retrieved from
-\url{https://www.google.com/maps/timeline?pb}
-
 \hypertarget{ref-london_encoding_2016}{}
@@
 
-\hypertarget{ref-_st-dbscan:_2007}{}
-ST-DBSCAN: An algorithm for clustering spatial--temporal data. (2007).
-\emph{Data \& Knowledge Engineering}, \emph{60}(1), 208--221.
-doi:\href{https://doi.org/10.1016/j.datak.2006.01.013}{10.1016/j.datak.2006.01.013}
-
 \hypertarget{ref-tatsiramos_geographic_2009}{}
@@
 analysis}. Retrieved from \url{https://CRAN.R-project.org/package=padr}
+
+\hypertarget{ref-location_history_timeline_2017}{}
+\emph{Timeline}. (2017). Retrieved from
+\url{https://www.google.com/maps/timeline?pb}
 
